---
title: "Compile Final Database and Query"
author: "Madeline Berger"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))



```

# Final Databse and Query

This markdown contains code to assemble and clean the final compiled database. It also contains the development code for the effluent query.


## Assemble Tables Into Final Dataabse

Up until this point, I've worked on each dataset separately due to small differences in each islands' most recent dataset. 
Here I will combine them so we have one database with 5 tables:

-Table 1: Authoritative TMK List, Spatial Information, Analysis Flags
-Table 2: Bedrooms per year
-Table 3a: Permits applied per year
-Table 3b: Cummulative OSDS status per year
-Table 4: Non-residential OSDS

```{r}


# table1s

oahu_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv"))
maui_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"))
kauai_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table1_tmk_registry_all.csv"))

hi_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table1_tmk_registry_ALL_draft.csv")) %>% select(-class_match_flag, -pitt_assignment,-min_year,-max_year,-scenario) %>% 
  rename(permit_story = unique_story,bed_gapfill_flag = bedroom_data_source) %>% 
  mutate(year_built_combined = 9999)

# missing key column names for Hi

final_table1_kmo <- rbind(oahu_1,
                                 maui_1) %>% 
  rbind(kauai_1) %>% 
  select(-starts_with("use"), -n_buildings)
  
final_table1_allislands <-  rbind(hi_1,final_table1_kmo)

# table 2
oahu_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2a_bedrooms_sum.csv")) %>% 
  mutate(`2021` = NA,
         `2022` = NA,
         `2023` = NA)

# NOTE - we don't have most recent data for oahu
maui_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2a_bedrooms_sum.csv")) %>% 
  select(-`2009`)

kauai_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table2a_bedrooms_sum.csv")) %>%
  select(-`2009`,-`2024`,-`2025`)

hi_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table2a_bedrooms_sum.csv")) %>% 
  rename(tmk = full_taxid)

final_table2_allislands <- rbind(oahu_2,
                                 maui_2,
                                 kauai_2,
                                 hi_2)

write_csv(final_table2_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands.csv"))

# table 3s

oahu_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_all.csv")) %>% 
  rename(tmk = full_taxid)
names(oahu_3a)[-1] <- paste0("osds_class_",names(oahu_3a)[-1])




maui_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

kauai_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

hi_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3_osdscl_ALL_draft.csv")) %>% 
  select(-osds_cl_pre_2009)
names(hi_3a)[-1] <- gsub("osds_cl_", "osds_class_", names(hi_3a)[-1])

# removed years not present in oahu - 2009 and pre2009 
final_table3a_allislands <- rbind(oahu_3a,
                                  maui_3a,
                                  kauai_3a,
                                  hi_3a)
write_csv(final_table3a_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))

# 3b - need to change
oahu_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3b_allyears_draft.csv")) 
names(oahu_3b)[-1] <- gsub("osds_class_", "osds_present_", names(oahu_3b)[-1])
names(oahu_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(oahu_3b)[-1])

maui_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3b_allyears_draft.csv"))
names(maui_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(maui_3b)[-1])
names(maui_3b)[-1] <- gsub("osds_class_", "osds_present_", names(maui_3b)[-1])

kauai_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3b_allyears_draft.csv"))
names(kauai_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(kauai_3b)[-1])
names(kauai_3b)[-1] <- gsub("osds_class_", "osds_present_", names(kauai_3b)[-1])

hi_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3b_allyears_draft.csv"))
names(hi_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(hi_3b)[-1])

final_table3b_allislands <- rbind(oahu_3b,
                                  maui_3b,
                                  kauai_3b,
                                  hi_3b)

write_csv(final_table3b_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))

# table 4s - this one has already been combined 



```

### Check to ensure Table 1 TMKS are accounted for. 

```{r}




```

### 