---
title: "AFS Analysis and Visualization"
output: html_document
date: "2024-08-28"
---

# Summary

For the talk at the American Fisheries Society, I am exploring how the driver layers can be used to examine fisheries and pollution overlap. This is a sandbox area to explore ways to move the data around and analyze it that might be interesting. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
source(here("workflow/setup.R"))


```

# Idea 1: Create product for input to bivisualziations ArcPro (not working yet)

```{r}

# Attempt 1: this didnt work
all_rec_fish <- rast(file.path(data_dir,"hi_otp_all_fishing_rec.tif"))
all_comm_fish <- rast(file.path(data_dir,"hi_otp_all_fishing_com.tif"))

# convert to polygons using terra - this takes for ever
poly_rec_fish <- as.polygons(all_rec_fish, round = FALSE)
poly_comm_fish <- as.polygons(all_comm_fish)

# convert to sf
sf_rec_fish <- st_as_sf(poly_rec_fish)


# try again, this time converting to a table: 

rec_df <- as.data.frame(all_rec_fish, xy = TRUE, cells = TRUE, na.rm = FALSE)
comm_df <- as.data.frame(all_comm_fish, xy = TRUE, cells = TRUE)

# join tables - this isnt working because the tables are not the same

joined <- left_join(rec_df,comm_df, by = c("x","y"))

# check if coordinates match


```

The above did not work because the rasters are not aligned. Try some code to realign, although not sure it will work


```{r}

# try to just align: 

comm_aligned <- terra::resample(all_rec_fish,all_comm_fish)

comm_aligned_df <- as.data.frame(comm_aligned, xy = TRUE, cells = TRUE, na.rm = FALSE)

all.equal(rec_df[, c("x","y")], comm_aligned_df[, c("x","y")])

all.equal(ext(all_rec_fish), ext(comm_aligned))
all.equal(res(all_rec_fish), res(comm_aligned))
all.equal(origin(all_rec_fish), origin(comm_aligned))

# define chunk

chunk <- 500

# write a function that aligns each chunk

realign_chunk <- function(raster1,raster2,row_start, nrows){
  # Extract chunk
  r1_chunk <- crop(raster1, ext(raster1, row_start, row_start + nrows - 1, 1, ncol(raster1)))
  r2_chunk <- crop(raster2, ext(raster2, row_start, row_start + nrows - 1, 1, ncol(raster2)))
  
  # Align the chunk from raster2 to raster1
  r2_chunk_aligned <- resample(r2_chunk, r1_chunk)
  
  # Return the processed chunk
  return(list(r1_chunk = r1_chunk, r2_chunk_aligned = r2_chunk_aligned))

}

# Process the fishing raster in chunks



```


# Idea 2: Calculate percent overlap between "hotspots" for different drivers 

```{r}

# read in funtions for calculations

extremes = function(r,q){
        qua = quantile(r[],q,na.rm=TRUE) # get the value that demarcates the top theshold
        Q = r > qua # filter your raster for only values above that 
        Q[Q==0]=NA # set all other values to NA
        r*Q # multiply original raster by Q to zero out any values you dont want
        }

# Apply to non commercial

all_rec_fish_5 <- extremes(noncomm_boat_spear, q = 0.95)
all_rec_fish_10 <- extremes(noncomm_boat_spear, q = 0.90) # these are not very different

# using the 10% threshold, calculate the pixels and then area (area might not be necessary)

total_pixels <- global(all_rec_fish_10, fun = "notNA")
pixel_area <- 100*100
top_fishing_area_m = total_pixels[1,1] * pixel_area
top_fishing_area_km = top_fishing_area_m / 1e+6

# also do it by island - ie where are the top 10 most fished areas by island?






```


# Idea 3: Look at overlap between pollution and MMAs

```{r}

# read in MMAs 

mmas <- st_read(file.path(data_dir,"Marine_Managed_Areas_(DAR)/Marine_Managed_Areas_(DAR).shp"))

mapview(mmas)

# filter out for focus ones: fisheries replenishment & reserves

mmas_preservation_focus <- mmas %>% 
  filter(mma_design %in% c("Marine Life Conservation District","Fish Replenishment Area","Fish Replenishment Area / Netting Restricted Area"))

mapview(mmas_preservation_focus)
  
# throw in FMAS if interesting

  

```

We can to assess overlap:
1. Intersecting the top 10 pollution polygon with the fmas
2. Some kind of zonal stats - get total amount of fishing and divide that by the total (I think this would only work for )
