---
title: "AFS Analysis and Visualization"
output: html_document
date: "2024-08-28"
---

# Summary

For the talk at the American Fisheries Society, I am exploring how the driver layers can be used to examine fisheries and pollution overlap. This is a sandbox area to explore ways to move the data around and analyze it that might be interesting. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
source(here("workflow/setup.R"))


```

# Idea 1: Create product for input to bivisualziations ArcPro

```{r}

# Attempt 1: this didnt work
all_rec_fish <- rast(file.path(data_dir,"hi_otp_all_fishing_rec.tif"))
all_comm_fish <- rast(file.path(data_dir,"hi_otp_all_fishing_com.tif"))

# convert to polygons using terra - this takes for ever
poly_rec_fish <- as.polygons(all_rec_fish, round = FALSE)
poly_comm_fish <- as.polygons(all_comm_fish)

# convert to sf
sf_rec_fish <- st_as_sf(poly_rec_fish)


# try again, this time converting to a table: 

rec_df <- as.data.frame(all_rec_fish, xy = TRUE, cells = TRUE)
comm_df <- as.data.frame(all_comm_fish, xy = TRUE, cells = TRUE)

# join tables - this isnt working because the tables are not the same

joined <- left_join(rec_df,comm_df, by = c("x","y"))


```


# Idea: Calculate percent overlap between "hotspots" for different drivers 



