---
title: "Cartography - misc."
author: "Madeline Berger"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(tidyverse)
library(terra)
library(tmap)
library(mapview)

library(here)

source(here("workflow/setup.R"))
```

## Misc Cartography

This markdown is a place to make maps while I figure out a better long term solution on my mac

## Grey 3D basemap 

This code wrangles the DEM we have from Joey into small, island-size pieces that are more easily used for cartographic purposes. I've saved each so we don't have to keep reading in the large file. this code does not need to be run again, unless we want to adjust how the og DEM is aggregated at some point (ie if in large figs it looks blurry, etc).

Inputs: 
- MHI coastline / land mass shapefile
- DEM?
- TIGER roads

```{r}

# can we get if from drop box?


coastlines <- read_sf(here("data/Coastlines_Watersheds/shoreline_CorrectedInlets_2022-23.shp")) #nad83utm4

# gdb

basemap_gdb <- file.path(data_dir,"base-map-layers.gdb")
st_layers(basemap_gdb)

 ##################################### Roads #################################
roads <- st_read(basemap_gdb, layer = "Roads_layer")

roads_01 <- roads %>% 
  filter(COUNTYFP == "001") %>% 
  filter(Access_Cat == "Normal Roads") %>% 
  st_transform(crs(dem))

roads_kauai <- roads %>%
  filter(COUNTYFP == "007") # just molokai

roads_maui <- roads %>% 
  filter(COUNTYFP == "009")

roads_oahu <- roads %>% 
  filter(COUNTYFP == "003")

#mapview(roads_01)

############################### Prep DEM ############################################

# read in island extents for cropping the DEM

oahu_ext <- st_read(here("workflow/extents/oahu_extent.shp"))
maui_ext <- st_read(here("workflow/extents/maui_extent.shp"))
kauai_ext <- st_read(here("workflow/extents/kaui_extent.shp"))
hi_ext <- st_read(here("workflow/extents/hi_extent.shp"))

# read in DEM 

dem <- terra::rast(file.path(data_dir,"MHI_DEM_Corrected_hill.tif"))

#rm(list = setdiff(ls(), "dem")) <- code for cleaning environment 
# crop and aggregate for hi island only

dem_hi <- crop(dem,hi_ext)

dem_smaller <- aggregate(dem_hi, fact = 8, fun = mean)

dem_smaller[dem_smaller == 0] <- NA

plot(dem_smaller)

# repeat for other islands

dem_oahu <- crop(dem, oahu_ext)
dem_oahu[dem_oahu < 0] <- NA # this doesnt work for oahu for some reason
dem_oahu_sm <- aggregate(dem_oahu, fact = 8, fun = mean)
dem_oahu_sm[dem_oahu_sm == 0] <- NA
dem_oahu_sm[dem_oahu_sm > 250] <- NA

plot(dem_oahu_sm)# some weird artifacts?

# try cropping to coastline
oahu_coast <- coastlines %>%  filter(Island == "Oahu") %>% st_transform(crs = crs(dem_oahu_sm))

dem_oahu_sm_cl <- crop(dem_oahu_sm, oahu_coast)
dem_oahu_sm_msk <- mask(dem_oahu_sm_cl, oahu_coast)
plot(dem_oahu_sm_msk)

dem_mauinui <- crop(dem, maui_ext)
dem_mauinui_sm <- aggregate(dem_mauinui, fact = 8, fun = mean)
dem_mauinui_sm[dem_mauinui_sm == 0] <- NA

plot(dem_mauinui_sm)

dem_kauai <- crop(dem, kauai_ext)
dem_kauai[dem_kauai < 1] <- NA # this gets rid of ocean cells before aggregating
dem_kauai_sm <- aggregate(dem_kauai, fact = 10, fun = mean, na.rm = TRUE)
dem_kauai_sm[dem_kauai_sm == 0] <- NA
dem_kauai_sm[dem_kauai_sm > 250] <- NA

plot(dem_kauai_sm)

# save
writeRaster(dem_oahu_sm_msk, file.path(data_dir,"dem_oahu_smaller.tif"), overwrite = TRUE)
writeRaster(dem_mauinui_sm, file.path(data_dir,"dem_mauinui_smaller.tif"), overwrite = TRUE)
writeRaster(dem_kauai_sm, file.path(data_dir,"dem_kauai_smaller.tif"), overwrite = TRUE)


################# Basemaps for HI island ##########################################

hi_coast <- coastlines %>%
  filter(Island == "Hawaii") %>% 
  st_transform(crs = crs(dem))

coastlines_only <- coastlines %>% 
  filter(Island == "Hawaii") %>% 
  ggplot()+
  geom_sf(color = "grey")+
  theme_minimal()

coastlines_only





# convert to data frame for ggplot

dem_df <- terra::as.data.frame(dem_smaller, xy = TRUE,na.rm = TRUE)

# 
 colnames(dem_df)[3] <- "elevation"

with_dem <- ggplot()+
  geom_sf(data = hi_coast, fill = "white", color = "#696969", size = 0.3)+
  geom_raster(data = dem_df, aes(x = x, y = y, fill = elevation), alpha = 0.7)+
  scale_fill_gradientn(colours=c("#454545", "#e9e9e9"))+
  guides(fill = "none")+
  coord_sf()+
  theme_bw()

with_dem

#library(elevatr)

#elevation <- get_elev_raster(hi_coast, z = 14, clip = "bbox") # still too huge

```

## Maui case study maps 

This map was initially created for the HIMB scholarship symposium as a simple illustration for a slide. Will likely be useful in the future as well. 

```{r}

# read in shorelines and watersheds and moku

shoreline <- st_read(here("data/Coastlines_Watersheds/shoreline_CorrectedInlets_2022-23.shp")) %>% 
  filter(Island == "Maui")

watersheds <- st_read(here("data/Coastlines_Watersheds/WBD_HU12_n83utm4_2023fixed.shp"))

moku <- st_read(here("data/Moku_New/Moku_NEW.shp"))

mapview(moku)


# get maui watersheds only

maui_watersheds <- watersheds %>%
  st_filter(shoreline, .predicates = st_intersects)

mapview(maui_watersheds)

# make some point for kihei and honolua bay

maui_poi <- data.frame(
  names = c("Kihei","Honolua Bay"),
  lon = c(-156.4542,-156.6385),
  lat = c(20.7399,21.0139)
)

maui_poi_sf <- st_as_sf(maui_poi, coords = c("lon","lat"), crs = 4326)

mapview(maui_poi_sf)

library(RColorBrewer)
# add column to map colors to

maui_watersheds$color_map <- ifelse(maui_watersheds$Name %in% c("Honokohau Stream","Kulanihakoi Gulch"),"wheat","cadetblue")


# create tmap

maui_map <- tm_shape(shoreline)+
  tm_fill(fill = "darkslategrey")+
  # tm_borders(lwd = 2,
  #           col = "black")+
  tm_shape(maui_watersheds)+
  tm_fill(col = "color_map",
          #palette = c("seagreen","darkgrey"),
          style = "cat")+
  tm_shape(maui_poi_sf)+
  tm_symbols(col = "black",
             size = 0.17)
  # tm_text("names",
  #         size = 1,
  #         just = "left")+
  # tm_layout(
  #   frame = FALSE
  # )
  # tm_borders(lwd = 0.2,
  #            col = "black") 



maui_map

## 

tmap_save(maui_map, filename = "maui_map_himbsymp.png", bg = "transparent", dpi = 300)

```

## Template figures?

