---
title: "OSDS Nearshore Impact Model"
author: "Madeline Berger"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))


# read in osds points 

osds_gdb <- here("data/osds/FINAL_OSDS_PRODUCTS.gdb")
st_layers(osds_gdb)
oahu_2023 <- st_read(osds_gdb, layer = "Oahu_OSDS_2023")


```

# Modelling OSDS Impact to Nearshore

This markdown walks through the steps of estimating the relative impact of OSDS on the nearshore of the Main Hawaiian Islands. 

Input data needed: 
- Point locations of all functional OSDS and the year they were permitted. We are using year permitted as a proxy for year built although it may be that some are not built right away.
- Ocean mask
- Euclidean distance mask 


## Step 1: Aggregate effluent and convert to 100 x 100 m raster

**Goal:**
**Assumptions:**

```{r}


# create raster template and rasterize with terra

r_om <- rast(ocean_mask)

osds_raster <- rasterize(vect(oahu_2023), r_om)

# 



```


## Step 2: Apply focal statistics to estimate "sphere of impact" into nearshore

**Goal:**
**Assumptions:**


## Step 3: Use ocean mask to limit area of impact to marine space & add decay function




