---
title: "OSDS Nearshore Impact Model"
author: "Madeline Berger"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)


source(here("workflow/setup.R"))


# read in osds points and explore

osds_gdb <- here("data/osds/FINAL_OSDS_PRODUCTS.gdb")
st_layers(osds_gdb)
oahu_2023 <- st_read(osds_gdb, layer = "Oahu_OSDS_2023")

plot(oahu_2023)




```

# Modelling OSDS Impact to Nearshore

This markdown walks through the steps of estimating the relative impact of OSDS on the nearshore of the Main Hawaiian Islands. 

Input data needed: 
- Point locations of all functional OSDS and the year they were permitted. We are using year permitted as a proxy for year built although it may be that some are not built right away.
- Ocean mask
- Euclidean distance mask 

## Step 0: Clean and explore OSDS Data

```{r}



hist(oahu_2023$Effluent)
hist(oahu_2023$Nitrogen_Flux[!is.na(oahu_2023$Nitrogen_Flux)])

max(oahu_2023$Effluent)

nas <- oahu_2023_nad83 %>% 
  filter(!is.na(Effluent)) # 129

max(oahu_2023$Effluent, na.rm = T)
min(oahu_2023$Effluent, na.rm = T)
mean(oahu_2023$Effluent, na.rm = T)


median(oahu_2023$Effluent, na.rm = T)

nas <- oahu_2023_nad83 %>% 
  filter(is.na(Nitrogen_Flux)) 


nozeros <- oahu_2023_nad83 %>% 
  filter(Effluent > 0 ) 


mapview(nozeros, zcol = "Effluent")


# ok so effluent has a crazy range. Need to:
# remove injection wells and anything after 2019
# multiply yearly effluent by the years in serivce, which = 2019 - year 
# multiply N by effluent


osds_2015_2019 <- oahu_2023 %>% 
  filter(Year < 2020) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(years_active_2015_2019 = (2020 - Year) - (2015 - Year)) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5))


# questions - does Effluent scale with N flux?

plot(osds_2015_2019$Effluent, osds_2015_2019$Nitrogen_Flux)


```

## Step 1: Aggregate effluent and convert to 100 x 100 m raster

**Goal:**
**Assumptions:**

```{r}

# read in ocean mask and use extent of oahu points to crop
ocean_mask <- terra::rast(file.path(data_dir,"OceanMask_LBSP_post2018.tif")) #

osds_2015_2019_nad83 <- st_transform(osds_2015_2019, crs = crs(ocean_mask))
extent <- ext(oahu_2023_nad83)
extent_vec <- as.vector(extent)

# expand extent to fit entire ocean mask
add_vec <- c(-14283,14300,-20892,20000)
new_extent <- extent_vec + add_vec
expanded <- ext(new_extent)

# crop ocean mask by extent
ocean_mask_oahu <- terra::crop(ocean_mask, expanded, mask = T)
plot(ocean_mask_oahu)

# create raster template and rasterize points with terra

r_om <- rast(ocean_mask_oahu)
osds_raster <- rasterize(vect(osds_2015_2019_nad83), r_om, field = "Nflux_kgpday_5yr_avg", fun = "sum", na.rm = T)
plot(osds_raster)


```


## Step 2: Apply focal statistics to estimate "sphere of impact" into nearshore

**Goal:**
- Estimate how OSDS impacts the nearshore

**Assumptions:**
- Ocean areas closest to pixels with higher effluent, from aggregated OSDS point data, will experience the highest impact from the leaching / runoff 

```{r}
  
# create a matrix of the raster values, and then replace any non zero value with 1
  fw_mat <- focalMat(osds_raster,2500, type = "circle") #2100 does not cover all indicator maps
  fw_mat[fw_mat > 0 ] <- 1
  
# apply focal statistics
  focalStat <- terra::focal(osds_raster,w = fw_mat, fun = "sum", na.rm = T) # na.rm key to making it expand out

plot(focalStat)


```



## Step 3: Use ocean mask to limit area of impact to marine space & add decay function

**Assumptions:**
- Intensity of impact lessens with distance from shore, due to chemical decay of N or P and / or wave or currents
- Decay occurs as a linear function of distance from shore 

```{r}

# extract focal statistics by ocean mask

ocean_mask_oahu[ocean_mask_oahu != 10] <- NA # I had to change this because it reads in the new om weird

focalStat_nearshore <- mask(focalStat, ocean_mask_oahu)

plot(focalStat_nearshore)

# get custon decay function

source(here("workflow/offshoreDecay.R"))

# read in Euclidean distance, check the crs and crop by the custom extent

euc_distance <- rast(file.path(data_dir,"DistFromShore_Euc_2023.tif"))
euc_distance_oahu <- terra::crop(euc_distance, expanded, mask = T)
plot(euc_distance_oahu)

# apply to nearshore raster
  
final_osds_2019_2015 <- offshore_decay(focalStat_nearshore,euc_distance_oahu)
# save final raster
  
plot(final_osds_2019_2015)

final_osds_2019_2015[final_osds_2019_2015 == 0] <- NA

plot(final_osds_2019_2015)

writeRaster(
  final_osds_2019_2015,
  filename = here("outputs/Oahu_OSDS_2015_2019_draft.tif"),
  overwrite = T
)

```


