---
title: "OSDS Nearshore Impact Model"
author: "Madeline Berger"
date: "2025-03-29"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)


source(here("workflow/setup.R"))


# read in osds points and explore

osds_gdb <- here("data/osds/FINAL_OSDS_PRODUCTS.gdb")
st_layers(osds_gdb)
oahu_2023 <- st_read(osds_gdb, layer = "Oahu_OSDS_2023")

mapview(oahu_2023)



```

# Modelling OSDS Impact to Nearshore

This markdown walks through the steps of estimating the relative impact
of OSDS on the nearshore of the Main Hawaiian Islands.

Input data needed:

-   Point locations of all functional OSDS and the year they were
    permitted. We are using year permitted as a proxy for year built
    although it may be that some are not built right away

-   Ocean mask

-   Euclidean distance mask

## Step 0: Clean and explore OSDS Data

### Oahu Dataset
```{r}


# check valid geometries 

oahu_2023 <- oahu_2023 %>% 
  mutate(valid = st_is_valid(.),
         empty = st_is_empty(.))

oahu_empty <- oahu_2023 %>% 
  filter(empty == TRUE) # note these are empty to Julia


oahu_2023_cl <- oahu_2023 %>% 
  filter(empty == FALSE)

# look at distributions

hist(oahu_2023_cl$Effluent)
hist(oahu_2023_cl$Nitrogen_Flux[!is.na(oahu_2023$Nitrogen_Flux)])

max(oahu_2023_cl$Effluent, na.rm = T)

nas_effluent <- oahu_2023_cl %>% 
  filter(is.na(Effluent)) # 5 once we take out the non valid geometries

nas_N <- oahu_2023_cl %>%
  filter(is.na(Nitrogen_Flux)) # 5 more

# check on stats
max(oahu_2023_cl$Effluent, na.rm = T)
min(oahu_2023_cl$Effluent, na.rm = T) # this is zero
mean(oahu_2023_cl$Effluent, na.rm = T) #533
median(oahu_2023$Effluent, na.rm = T)

nozeros <- oahu_2023_cl %>% 
  filter(Effluent > 0 ) # removed another 10 observations, just to note

injection_wells <- oahu_2023_cl %>% 
  filter(PermitType == "Injection Well") 

mapview(injection_wells)

layout(mat = matrix(c(1,2),2,1, byrow=TRUE),  height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(log(oahu_2023$Effluent) , horizontal=TRUE , ylim=c(4,10), xaxt="n" , col=rgb(0.8,0.8,0,0.5) , frame=F)
par(mar=c(4, 3.1, 1.1, 2.1))
hist(log(oahu_2023$Effluent) , breaks=40 , col=rgb(0.2,0.8,0.5,0.5) , border=F , main="" , xlab="Log of Daily Effluent for Oahu Locations All Years", xlim=c(4,10))


#mapview(nozeros, zcol = "Effluent")


# ok so effluent has a crazy range. Need to:
# remove injection wells and anything after 2019
# multiply yearly effluent by the years in serivce, which = 2019 - year 
# multiply N by effluent


osds_2015_2019 <- oahu_2023_cl %>% 
  filter(Year < 2020) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(years_active_2015_2019 = (2020 - Year) - (2015 - Year)) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2025_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2025_2019 / (365*5))

# recheck stats after year and type filters

hist(osds_2015_2019$Effluent, breaks = 30)
hist(log(osds_2015_2019$Effluent)) # can see distribution better this way 
# nicer code
# Layout to split the screen
layout(mat = matrix(c(1,2),2,1, byrow=TRUE),  height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(log(osds_2015_2019$Effluent) , horizontal=TRUE , ylim=c(4,10), xaxt="n" , col=rgb(0.8,0.8,0,0.5) , frame=F)
par(mar=c(4, 3.1, 1.1, 2.1))
hist(log(osds_2015_2019$Effluent) , breaks=40 , col=rgb(0.2,0.8,0.5,0.5) , border=F , main="" , xlab="Log of Daily Effluent for Oahu Locations, 2015 - 2019", xlim=c(4,10))


max(osds_2015_2019$Effluent, na.rm = T) #14150
min(osds_2015_2019$Effluent, na.rm = T) #39
mean(osds_2015_2019$Effluent, na.rm = T) # 582
median(osds_2015_2019$Effluent, na.rm = T) #400

# questions - does Effluent scale with N or P flux? Plot, and see read me for documentation

plot(osds_2015_2019$Effluent, osds_2015_2019$Nitrogen_Flux)
plot(osds_2015_2019$Effluent, osds_2015_2019$Phosphorous_Flux)

# create versions of the points for other scenarios: with injection wells, and for previous time period with and without injection wells. Repeat summary statistics for older data


osds_2015_2019_IW <- oahu_2023_cl %>% 
  filter(Year < 2020 | is.na(Year)) %>% # this grabs the iws
  mutate(Year = ifelse(is.na(Year),2009,Year)) %>%  # so we can count iws
  #filter(PermitType != "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(years_active_2015_2019 = (2020 - Year) - (2015 - Year)) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2025_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2025_2019 / (365*5)) %>% 
  st_transform(., crs = crs(ocean_mask_oahu))


max(osds_2015_2019_IW$Effluent, na.rm = T) #539941.2
min(osds_2015_2019_IW$Effluent, na.rm = T) #39
mean(osds_2015_2019_IW$Effluent, na.rm = T) # 1681
median(osds_2015_2019_IW$Effluent, na.rm = T) #420
    
nastest <- osds_2015_2019_IW[is.na(osds_2015_2019_IW$Year),]# all nas are injection wells so this filter works

osds_2010_2014 <- oahu_2023_cl %>% 
  filter(Year < 2015) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(years_active_2010_2014 = (2015 - Year) - (2010 - Year)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask_oahu))# this accounts for OSDS that were active for only part of the period from 2010 - 2014


max(osds_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #14150
min(osds_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #60
mean(osds_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) # 499
median(osds_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #420


osds_2010_2014_IW <- oahu_2023_cl %>% 
  filter(Year < 2015 | is.na(Year)) %>% # this grabs the iws
  mutate(Year = ifelse(is.na(Year),2009,Year)) %>% # this just allows us to account IW for all POI
  #filter(PermitType != "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(years_active_2010_2014 = (2015 - Year) - (2010 - Year)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>% 
  st_transform(., crs = crs(ocean_mask_oahu))

max(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #548841.2
min(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #60
mean(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #2716.339
median(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #420

# copying some cool code just for future reference

# library(magrittr)
# 
# osds_2010_2014_IW <- oahu_2023_cl %>%
#   subset(Year < 2015 | is.na(Year)) %>%
#   within({
#     Year <- as.numeric(Year)
#     years_active_2010_2014 <- (2015 - Year) - (2010 - Year)
#     total_N_toenv2010_2014 <- years_active_2010_2014 * 365 * Nitrogen_Flux
#     total_Eff_toenv2010_2014 <- years_active_2010_2014 * 365 * Effluent
#     Nflux_kgpday_5yr_avg <- total_N_toenv2010_2014 / (365 * 5)
#     Effluent_galpday_5yr_avg <- total_Eff_toenv2010_2014 / (365 * 5)
#   })


```

### Maui Nui Dataset 



## Step 1: Aggregate effluent and convert to 100 x 100 m raster

**Goal:**

**Assumptions:**

```{r}

# read in ocean mask and use extent of oahu and maui points to crop
ocean_mask <- terra::rast(file.path(data_dir,"OceanMask_LBSP_post2018.tif")) #

osds_2015_2019_nad83 <- st_transform(osds_2015_2019, crs = crs(ocean_mask))

extent <- ext(oahu_2023_nad83)
extent_vec <- as.vector(extent)

# expand extent to fit entire ocean mask
add_vec <- c(-14283,14300,-20892,20000)
new_extent <- extent_vec + add_vec
expanded <- ext(new_extent)

ocean_mask_oahu <- terra::crop(ocean_mask, expanded, mask = T)
  plot(ocean_mask_oahu)


# rasterize function, to apply to multiple point subsets

rasterizePoints <- function(points,mask){
  
  # create raster template and rasterize points with terra
  r_om <- rast(mask)
  #rasterize points, using sum to aggregate effluent 
  osds_raster <- rasterize(vect(points), r_om, field = "Effluent_galpday_5yr_avg", fun = "sum", na.rm = T)
  
  return(osds_raster)
  
}

# apply to points for different scenarios 

oahu_2015_2019_eff <- rasterizePoints(osds_2015_2019_nad83,ocean_mask_oahu)
oahu_2015_2019_IWeff <- rasterizePoints(osds_2015_2019_IW,ocean_mask_oahu)
oahu_2010_2014_eff <- rasterizePoints(osds_2010_2014,ocean_mask_oahu)
oahu_2010_2014_IWeff <- rasterizePoints(osds_2010_2014_IW,ocean_mask_oahu)

plot(oahu_2015_2019_eff)
plot(oahu_2010_2014_eff)
plot(oahu_2010_2014_IWeff)


```

## Step 2: Apply focal statistics to estimate "sphere of impact" into nearshore

**Goal:**

-   Estimate how OSDS impacts the nearshore

**Assumptions:**

-   Ocean areas closest to pixels with higher effluent, from aggregated
    OSDS point data, will experience the highest impact from the
    leaching / runoff

```{r}

# prep and funciton for Oahu 

ocean_mask_oahu[ocean_mask_oahu != 10] <- NA  

applyFocalStats_extractByMask <- function(raster,mask){
  
  # create a matrix of the raster values, and then replace any non zero value with 1
  fw_mat <- focalMat(raster,2500, type = "circle") #2100 does not cover all indicator maps
  fw_mat[fw_mat > 0 ] <- 1
  
# apply focal statistics
  focalStat <- terra::focal(raster,w = fw_mat, fun = "sum", na.rm = T) # na.rm key to making it expand out
 
  # Extract by ocean mask 
  focalStat_nearshore <- mask(focalStat, mask)
  
  return(focalStat_nearshore)

}

# apply to scenario

focStat_2015_2019 <- applyFocalStats_extractByMask(oahu_2015_2019_eff,ocean_mask_oahu)
plot(focStat_2015_2019)
focStat_2015_2019_IW <- applyFocalStats_extractByMask(oahu_2015_2019_IWeff,ocean_mask_oahu)
plot(focStat_2015_2019_IW)
focStat_2010_2014 <- applyFocalStats_extractByMask(oahu_2010_2014_eff,ocean_mask_oahu)
plot(focStat_2010_2014)
focStat_2010_2014_IW <- applyFocalStats_extractByMask(oahu_2010_2014_IWeff,ocean_mask_oahu)
plot(focStat_2010_2014_IW)


```

## Step 3: Use ocean mask to limit area of impact to marine space & add decay function

**Assumptions:**

-   Intensity of impact lessens with distance from shore, due to
    chemical decay of N or P and / or wave or currents

-   Decay occurs as a linear function of distance from shore

```{r}


# get custom decay function

source(here("workflow/offshoreDecay.R")) # identical to all other decay for LBSP

# read in Euclidean distance, check the crs and crop by the custom extent

euc_distance <- rast(file.path(data_dir,"DistFromShore_Euc_2023.tif"))
euc_distance_oahu <- terra::crop(euc_distance, expanded, mask = T)
plot(euc_distance_oahu)

# apply to nearshore rasters
  
final_osds_2015_2019 <- offshore_decay(focStat_2015_2019,euc_distance_oahu)
final_osds_2015_2019[final_osds_2015_2019 == 0] <- NA
plot(final_osds_2015_2019)

final_osds_2015_2019_IW <- offshore_decay(focStat_2015_2019_IW,euc_distance_oahu)
final_osds_2015_2019_IW[final_osds_2015_2019_IW == 0] <- NA
plot(final_osds_2015_2019_IW)

final_osds_2010_2014 <- offshore_decay(focStat_2010_2014,euc_distance_oahu)
final_osds_2010_2014[final_osds_2010_2014 == 0] <- NA
plot(final_osds_2010_2014)

final_osds_2010_2014_IW <- offshore_decay(focStat_2010_2014_IW,euc_distance_oahu)
final_osds_2010_2014_IW[final_osds_2010_2014_IW == 0] <- NA
plot(final_osds_2010_2014_IW)

# save rasters to output then add to Dropbox

writeRaster(
  final_osds_2015_2019,
  filename = here("outputs/Oahu_OSDS_2015_2019_eff.tif"),
  overwrite = T
)

writeRaster(
  final_osds_2015_2019_IW,
  filename = here("outputs/Oahu_OSDS_2015_2019_IWeff.tif"),
  overwrite = T
)

writeRaster(
  final_osds_2010_2014,
  filename = here("outputs/Oahu_OSDS_2010_2014_eff.tif"),
  overwrite = T
)


writeRaster(
  final_osds_2010_2014_IW,
  filename = here("outputs/Oahu_OSDS_2010_2014_IWeff.tif"),
  overwrite = T
)


```
