---
title: "OSDS Nearshore Impact Model"
author: "Madeline Berger"
date: "2025-03-29"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = FALSE
                      )

library(here)
source(here("workflow/setup.R"))

# file path for osds on dropbox 

osds_dir <- file.path(driver_dir,"4_OSDS_N_Runoff/OSDS_2025_Julia")


# read in osds points and explore

osds_gdb <- here("data/osds/FINAL_OSDS_PRODUCTS.gdb")
st_layers(osds_gdb)
oahu_2023 <- st_read(osds_gdb, layer = "Oahu_OSDS_2023")

#oahu_2023 <- st_read(here("data/osds/oahu_2023_raw.shp"))

#mapview(oahu_2023)

# read in ocean mask and use extent of oahu and maui points to crop
ocean_mask <- terra::rast(file.path(data_dir,"OceanMask_LBSP_post2018.tif"))


# Save point data so we can zip up gdbs and not have them clog up my computer

# st_write(
#   oahu_2023,
#   dsn = here("data/osds/oahu_2023_raw.shp")
# )


```

# Modelling OSDS Impact to Nearshore

This markdown walks through the steps and code for estimating the
relative impact of OSDS on the nearshore of the Main Hawaiian Islands.

Important note: No ability to assess difference any upgrades if original
osds points were replaced by upgraded points FOR ALL YEARS they exist.
Our only comparison would be to Joey's layer. Would be cool to also make
sure any reductions we're seeing might be influenced by upgrades.

Input data needed:

-   Point locations of all functional OSDS and the year they were
    permitted. We are using year permitted as a proxy for year built
    although it may be that some are not built or become functional
    right away

-   Ocean mask

-   Euclidean distance mask

## Step 0: Clean and explore OSDS Data

### Oahu Dataset

```{r}


# check valid geometries 

oahu_2023 <- oahu_2023 %>% 
  mutate(valid = st_is_valid(.),
         empty = st_is_empty(.))

oahu_empty <- oahu_2023 %>% 
  filter(empty == TRUE) # note these are empty to Julia

# clean out empty geometries and remove characters from year
oahu_2023_cl <- oahu_2023 %>% 
  filter(empty == FALSE) %>% 
  mutate(Year = ifelse(Year == "pre 2009", "2008", Year))
  

mapview(oahu_2023_cl)

# look at distributions

hist(oahu_2023_cl$Effluent)
hist(oahu_2023_cl$Nitrogen_Flux[!is.na(oahu_2023$Nitrogen_Flux)])

max(oahu_2023_cl$Effluent, na.rm = T)

nas_effluent <- oahu_2023_cl %>% 
  filter(is.na(Effluent)) # 5 once we take out the non valid geometries

nas_N <- oahu_2023_cl %>%
  filter(is.na(Nitrogen_Flux)) # 5 more, same ones as above

# check on stats
max(oahu_2023_cl$Effluent, na.rm = T)
min(oahu_2023_cl$Effluent, na.rm = T) # this is zero
mean(oahu_2023_cl$Effluent, na.rm = T) #533
median(oahu_2023$Effluent, na.rm = T)

zeros <- oahu_2023_cl %>% 
  filter(Effluent == 0 ) # removed another 10 observations, just to note

oahu_nas <- oahu_2023_cl %>% 
  filter(is.na(Effluent)) # 5


# mapview(zeros, col.regions = "blue")+
#   mapview(oahu_nas, col.regions = "green")

injection_wells <- oahu_2023_cl %>% 
  filter(PermitType == "Injection Well") 

#mapview(injection_wells)

layout(mat = matrix(c(1,2),2,1, byrow=TRUE),  height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(log(oahu_2023$Effluent) , horizontal=TRUE , ylim=c(4,10), xaxt="n" , col=rgb(0.8,0.8,0,0.5) , frame=F)
par(mar=c(4, 3.1, 1.1, 2.1))
hist(log(oahu_2023$Effluent) , breaks=40 , col=rgb(0.2,0.8,0.5,0.5) , border=F , main="" , xlab="Log of Daily Effluent for Oahu Locations All Years", xlim=c(4,10))


#mapview(nozeros, zcol = "Effluent")


# ok so effluent has a crazy range. Need to:
# remove injection wells and anything after 2019
# multiply yearly effluent by the years in serivce, which = 2019 - year 
# multiply N by effluent



oahu_2015_2019 <- oahu_2023_cl %>% 
  filter(Year < 2020) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(year_calc = as.numeric(Year)) %>% 
  mutate(years_active_2015_2019 = case_when(
    year_calc >= 2015 ~ 2020 - year_calc,
    year_calc < 2015 ~ (2020 - year_calc) - (2015 - year_calc),
    TRUE ~ NA
  )) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2025_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2025_2019 / (365*5))

#mapview(osds_2015_2019) # where did all the pearl harbor ones go?


oahu_post2020 <- oahu_2023_cl %>% 
  filter(Year > 2020)

# recheck stats after year and type filters to make sure they changed

#hist(osds_2015_2019$Effluent, breaks = 30)
#hist(log(osds_2015_2019$Effluent)) # can see distribution better this way 
# nicer code
# Layout to split the screen
layout(mat = matrix(c(1,2),2,1, byrow=TRUE),  height = c(1,8))
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(log(oahu_2015_2019$Effluent) , horizontal=TRUE , ylim=c(4,10), xaxt="n" , col=rgb(0.8,0.8,0,0.5) , frame=F)
par(mar=c(4, 3.1, 1.1, 2.1))
hist(log(oahu_2015_2019$Effluent) , breaks=40 , col=rgb(0.2,0.8,0.5,0.5) , border=F , main="" , xlab="Log of Daily Effluent for Oahu Locations, 2015 - 2019", xlim=c(4,10))


max(oahu_2015_2019$Effluent, na.rm = T) #14150
min(oahu_2015_2019$Effluent, na.rm = T) #39
mean(oahu_2015_2019$Effluent, na.rm = T) # 441
median(oahu_2015_2019$Effluent, na.rm = T) #420

# questions - does Effluent scale with N or P flux? Plot, and see read me for documentation
# par(mfrow = c(1,2))
# plot(oahu_2015_2019$Effluent, oahu_2015_2019$Nitrogen_Flux)
# plot(oahu_2015_2019$Effluent, oahu_2015_2019$Phosphorous_Flux)

# create versions of the points for other scenarios: with injection wells, and for previous time period with and without injection wells. Repeat summary statistics for older data


oahu_IW <- oahu_2023_cl %>% 
  filter(Year < 2020 | is.na(Year)) %>% # this grabs the iws
  mutate(Year = ifelse(is.na(Year),2009,Year)) %>%  # so we can count iws
  filter(PermitType == "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  # mutate(years_active_2015_2019 = (2020 - Year) - (2015 - Year)) %>%
  # mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>%
  # mutate(total_Eff_toenv2015_2019 = years_active_2015_2019 * 365 * Effluent) %>%
  # mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = Effluent) %>%  
  st_transform(., crs = crs(ocean_mask))


max(oahu_IW$Effluent, na.rm = T) #539941.2
min(oahu_IW$Effluent, na.rm = T) 
mean(oahu_IW$Effluent, na.rm = T) # 28208.1
median(oahu_IW$Effluent, na.rm = T) #3488.114
    
nastest <- oahu_IW[is.na(oahu_IW$Year),]# all nas for year are injection wells so this filter works

oahu_2010_2014 <- oahu_2023_cl %>% 
  filter(Year < 2015) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(year_calc = as.numeric(Year)) %>% 
  mutate(years_active_2010_2014 = case_when(
    year_calc > 2010 ~ (2015 - year_calc),
    year_calc <= 2010 ~ (2015 - year_calc) - (2010 - year_calc),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))


max(oahu_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #14150
min(oahu_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #60
mean(oahu_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #436 - oahu has more
median(oahu_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #420


# osds_2010_2014_IW <- oahu_2023_cl %>% 
#   filter(Year < 2015 | is.na(Year)) %>% # this grabs the iws
#   mutate(Year = ifelse(is.na(Year),2009,Year)) %>% # this just allows us to account IW for all POI
#   filter(PermitType == "Injection Well") %>% 
#   mutate(Year = as.numeric(Year)) %>% 
#   mutate(years_active_2010_2014 = (2015 - Year) - (2010 - Year)) %>% 
#   mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
#   mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
#   mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
#   mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>% 
#   st_transform(., crs = crs(ocean_mask))
# 
# plot(osds_2010_2014_IW)
# 
# max(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #539941.2
# min(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #
# mean(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #2716.339
# median(osds_2010_2014_IW$Effluent_galpday_5yr_avg, na.rm = T) #420


#mapview(osds_2015_2019)

# copying some cool code just for future reference

# library(magrittr)
# 
# osds_2010_2014_IW <- oahu_2023_cl %>%
#   subset(Year < 2015 | is.na(Year)) %>%
#   within({
#     Year <- as.numeric(Year)
#     years_active_2010_2014 <- (2015 - Year) - (2010 - Year)
#     total_N_toenv2010_2014 <- years_active_2010_2014 * 365 * Nitrogen_Flux
#     total_Eff_toenv2010_2014 <- years_active_2010_2014 * 365 * Effluent
#     Nflux_kgpday_5yr_avg <- total_N_toenv2010_2014 / (365 * 5)
#     Effluent_galpday_5yr_avg <- total_Eff_toenv2010_2014 / (365 * 5)
#   })


```

### Maui Nui Dataset

```{r}

maui_gdb <- here("data/osds/Final_MauiNui_OSDS_Products.gdb")
st_layers(maui_gdb)
mauinui_2023 <- st_read(maui_gdb, layer = "MauiNui_OSDS_2023")

plot(mauinui_2023)

# 
# st_write(
#   mauinui_2023,
#   dsn = here("data/osds/mauinui_2023_raw.shp")
# )
# check valid geoms 

mauinui_2023 <- mauinui_2023 %>% 
  mutate(valid = st_is_valid(.),
         empty = st_is_empty(.))

maui_empty <- mauinui_2023 %>% 
  filter(empty == TRUE) # none are empty, nice


# check stats

min(mauinui_2023$Effluent, na.rm = T) # 0 , and check NA
max(mauinui_2023$Effluent, na.rm = T) # 4393710
mean(mauinui_2023$Effluent, na.rm = T) #1192.007
median(mauinui_2023$Effluent, na.rm = T) #315 

# nas?

mauinui_nas <- mauinui_2023 %>% 
  filter(is.na(Effluent)) # 0 

mauinui_zeros <- mauinui_2023 %>% 
  filter(Effluent == 0) # 2 obesrvations

# more checks 

length(mauinui_2023$Effluent)
sum(is.infinite(mauinui_2023$Effluent))
sum(is.nan(mauinui_2023$Effluent))


# summarize all potential weirdness

mauinui_df <- st_drop_geometry(mauinui_2023)

summary_check <- function(df) {
  data.frame(
    Column = names(df),
    NA_count = sapply(df, function(x) sum(is.na(x))),
    Inf_count = sapply(df, function(x) sum(is.infinite(x))),
    NaN_count = sapply(df, function(x) sum(is.nan(x)))
  )
}

mauinui_check <- summary_check(mauinui_df) 

# all of the NAs in effluent are new. what percentage of them are NA?

mauinui_new_osds <- mauinui_2023 %>% 
  filter(comments == "new OSDS for 2023 update") # 638

mapview(mauinui_nas, col.regions = "green")+
  mapview(mauinui_zeros, col.regions = "blue")

# how many injection wells?

mauinui_iw <- mauinui_2023 %>% 
  filter(PermitType == "Injection Well")
  
mapview(mauinui_iw)

# clean version

maui_2023_cl <- mauinui_2023 %>%
  filter(!is.na(Effluent))

# look at lanai points

mapview(maui_2023_cl, zcol = "PermitType")


# subset points by injection wells versus no injection wells for each 5 year interval

mauinui_2010_2014 <- maui_2023_cl %>% 
  filter(Year < 2015) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(year_calc = as.numeric(Year)) %>% 
  mutate(years_active_2010_2014 = case_when(
    year_calc > 2010 ~ (2015 - year_calc),
    year_calc <= 2010 ~ (2015 - year_calc) - (2010 - year_calc),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(mauinui_2010_2014$Effluent) #682.72
max(mauinui_2010_2014$Effluent) #27000

mauinui_IW <- maui_2023_cl %>% 
  #filter(Year < 2015) %>% 
  mutate(Year = ifelse(Year == "pre2009",2009,Year)) %>% # this just allows us to account IW for all POI
  filter(PermitType == "Injection Well") %>% 
  mutate(Year = as.numeric(Year)) %>%
  mutate(years_active_2010_2014 = (2015 - Year) - (2010 - Year)) %>%
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>%
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>%
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014) %>%
  st_transform(., crs = crs(ocean_mask))

mean(mauinui_IW$Effluent) #6590.429
max(mauinui_IW$Effluent) #18004
  
mapview(mauinui_IW)

mauinui_2015_2019 <- maui_2023_cl %>% 
  filter(Year < 2020) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(year_calc = as.numeric(Year)) %>% 
  mutate(years_active_2015_2019 = case_when(
    year_calc >= 2015 ~ 2020 - year_calc,
    year_calc < 2015 ~ (2020 - year_calc) - (2015 - year_calc),
    TRUE ~ NA
  )) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2015_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2015_2019 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(mauinui_2015_2019$Effluent) # 607
max(mauinui_2015_2019$Effluent) #27000



# mauinui_2015_2019_IW <- maui_2023_cl %>% 
#   filter(Year < 2020) %>% 
#   mutate(Year = ifelse(is.na(Year),2009,Year)) %>% # this just allows us to account IW for all POI
#   filter(PermitType == "Injection Well") %>% 
#   mutate(Year = as.numeric(Year)) %>% 
#   mutate(years_active_2010_2014 = (2020 - Year) - (2015 - Year)) %>% 
#   mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
#   mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
#   mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
#   mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
#   st_transform(., crs = crs(ocean_mask))
# 
# mean(mauinui_2015_2019_IW$Effluent) #6590.429
# max(mauinui_2015_2019_IW$Effluent) #18004
```

### Kauai Dataset

```{r}

kauai_gdb <- file.path(data_dir, "osds/Kauai_OSDS_Derivative.gdb")
st_layers(kauai_gdb)
kauai_2023 <- st_read(kauai_gdb, layer = "Kauai_OSDS_2023")

mapview(kauai_2023)

# fix columns tht might be messing with save

kauai_2023 <- rename(kauai_2023, c1 = class_i, c2 = class_ii, c3 = class_iii, c4 = class_iv, c5 = class_V)

kauai_df <- st_drop_geometry(kauai_2023)

kauai_qaqc <- kauai_2023 %>% 
  mutate(valid = st_is_valid(.),
         empty = st_is_empty(.))
# save raw kauai point data so I can remove gdb from device
# st_write(
#   kauai_2023,
#   dsn = here("data/osds/kauai_2023_raw.shp"),
#   delete_dsn = T
# )

# qa qc

min(kauai_2023$Effluent, na.rm = T) #105
max(kauai_2023$Effluent, na.rm = T) #539228
mean(kauai_2023$Effluent, na.rm = T) #602
median(kauai_2023$Effluent, na.rm = T) #312

kauai_check <- summary_check(kauai_df) # some nas

# map any points with addresses but missing data 
kauai_nas <- kauai_2023 %>% 
  filter(is.na(Effluent)) # 39

kauai_zeros <- kauai_2023 %>% 
  filter(Effluent == 0) # 0

mapview(kauai_nas, col.regions = "green")

# check out injection wells
kauai_iw <- kauai_2023 %>%filter(PermitType == "Injection Well")

mapview(kauai_iw) #20 


kauai_cl <- kauai_2023 %>%  filter(!is.na(Effluent)) %>% 
  mutate(year_calc = ifelse(Year == "pre2009",2009,Year)) %>% 
  mutate(year_calc = as.numeric(year_calc))
## subset points 2010 - 2024, 2015 - 2019 and IW

kauai_2010_2014 <- kauai_cl %>% 
  filter(year_calc < 2015) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(years_active_2010_2014 = case_when(
    year_calc > 2010 ~ (2015 - year_calc),
    year_calc <= 2010 ~ (2015 - year_calc) - (2010 - year_calc),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(kauai_2010_2014$Effluent) # 485.3942
max(kauai_2010_2014$Effluent) #10080

kauai_2015_2019 <- kauai_cl %>% 
  filter(year_calc < 2020) %>% 
  filter(PermitType != "Injection Well") %>% 
  mutate(years_active_2015_2019 = case_when(
    year_calc >= 2015 ~ 2020 - year_calc,
    year_calc < 2015 ~ (2020 - year_calc) - (2015 - year_calc),
    TRUE ~ NA
  )) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2015_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2015_2019 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(kauai_2015_2019$Effluent) # 483.4839
max(kauai_2015_2019$Effluent) # 10815


### repeat for injection wells by themselves

kauai_IW <- kauai_cl %>%
  filter(year_calc < 2015) %>% 
  filter(PermitType == "Injection Well") %>%
  mutate(years_active_2010_2014 = case_when(
    year_calc > 2010 ~ (2015 - year_calc),
    year_calc <= 2010 ~ (2015 - year_calc) - (2010 - year_calc),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

# mean(kauai_2010_2014_IW$Effluent) # 87436
# max(kauai_2010_2014_IW$Effluent) # 539228

# kauai_2015_2019_IW <- kauai_cl %>%
#   filter(year_calc < 2020) %>% 
#   filter(PermitType == "Injection Well") %>% 
#   mutate(years_active_2015_2019 = (2020 - year_calc) - (2015 - year_calc)) %>% 
#   mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
#   mutate(total_Eff_toenv2015_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
#   mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
#   mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2015_2019 / (365*5)) %>%
#   st_transform(., crs = crs(ocean_mask))


# mean(kauai_2015_2019_IW$Effluent) # 87436
# max(kauai_2015_2019_IW$Effluent) # 539228 - nothing has changed, no new injection wells


```

### Hawaii Dataset

-   Note the Hawaii dataset has a "shut down" column for osds no longer
    active from Kilauaea eruption removals

```{r}

hawaii_gdb <- file.path(data_dir, "osds/Hawaii_OSDS_Derivative.gdb")
st_layers(hawaii_gdb)
hawaii_2023 <- st_read(hawaii_gdb, layer = "Hawaii_OSDS_2023")

# save raw version

# st_write(
#   hawaii_2023,
#   dsn = here("data/osds/hawaii_2023_raw.shp"),
#   delete_dsn = T
# )

#mapview(hawaii_2023)

## run checks

hi_df <- st_drop_geometry(hawaii_2023)

hi_qaqc <- hawaii_2023 %>% 
  mutate(valid = st_is_valid(.),
         empty = st_is_empty(.)) # all good on geometries

hi_check <- summary_check(hi_df) # 493 na

hi_zeros <- hawaii_2023 %>% 
  filter(EFFLUENT == 0) # 399

hi_nas <- hawaii_2023 %>% 
  filter(is.na(EFFLUENT)) # 493 

# mapview(hi_zeros, col.regions = "blue")+
#   mapview(hi_nas, col.regions = "green")

# no injection wells

# mess around with eff date a bit - extract year and compare to "Year" columns

class(hawaii_2023$EFF_DATE) # date object

hi_dates_cl <- hawaii_2023 %>%
  mutate(year_extract = year(EFF_DATE)) %>% 
  mutate(year_match = ifelse(year_extract == Year,"Yes","No")) %>% 
  rename(Effluent = EFFLUENT)


# summarize

hi_dates_check <- hi_dates_cl %>% 
  group_by(year_match, Year) %>% 
  summarize(
    count_no = length(year_match == "No")
  )
  
# USE THIS ONE replace "pre2009" with year from eff date

hi_2023_cl <- hi_dates_cl %>% 
  rename(year_char = Year) %>% 
  rename(Year = year_extract)


# also look at shutdown year. should not contain characters

hi_shutdown <- hi_2023_cl %>% 
  filter(!is.na(Shutdown))

#mapview(hi_shutdown, zcol = "Shutdown")


# calculate years active for these separately


hi_2023_cl <- hi_2023_cl %>% 
  mutate(year_shutdown = case_when(
    Shutdown == "post1992" ~ "2018",
    Shutdown == "not determined " ~ NA,
    TRUE ~ Shutdown
  )) %>% 
  mutate(year_shutdown = as.numeric(year_shutdown)) %>% 
  mutate(years_active = year_shutdown - Year) %>% 
  mutate(years_active = ifelse(years_active < 0, 0, years_active)) # this doesnt matter for first time step



hi_2010_2014 <- hi_2023_cl %>% 
  filter(Year < 2015) %>% 
  filter(is.na(InjectionWell_PermitNo)) %>% # for some reason these were treated differently in this dataset
  mutate(years_active_2010_2014 = case_when(
    Year < 2010 ~ (2015 - Year) - (2010 - Year),
    Year >= 2010 ~ (2015 - Year),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(hi_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #302
max(hi_2010_2014$Effluent_galpday_5yr_avg, na.rm = T) #36960

hi_2015_2019 <- hi_2023_cl %>% 
  filter(Year < 2020) %>% 
  filter(is.na(InjectionWell_PermitNo)) %>% 
  mutate(years_active_2020 = case_when(
    is.na(Shutdown) ~ (2020 - Year),
    !is.na(year_shutdown) ~ (2018 - Year), # to  account for IW shutdown
    TRUE ~ NA),
    years_active_2015 = (2015 - Year),
    years_active_2015_2019 = case_when(
      Year >= 2015 ~ years_active_2020,
      Year < 2015 ~ years_active_2020 - years_active_2015,
      TRUE ~ NA
    )
  ) %>% 
  mutate(total_N_toenv2015_2019 = years_active_2015_2019 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2015_2019 = years_active_2015_2019 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2015_2019 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2015_2019 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))

mean(hi_2015_2019$Effluent_galpday_5yr_avg, na.rm = T) #306.32
max(hi_2015_2019$Effluent_galpday_5yr_avg, na.rm = T) #36960

# injection wells only until 2011

hi_IW <- hi_2023_cl %>% 
  filter(!is.na(InjectionWell_PermitNo)) %>%
  mutate(years_active_2010_2014 = case_when(
    Year < 2010 ~ (2015 - Year) - (2010 - Year),
    Year >= 2010 ~ (2015 - Year),
    TRUE ~ NA)) %>% 
  mutate(total_N_toenv2010_2014 = years_active_2010_2014 * 365 * Nitrogen_Flux) %>% 
  mutate(total_Eff_toenv2010_2014 = years_active_2010_2014 * 365 * Effluent) %>% 
  mutate(Nflux_kgpday_5yr_avg = total_N_toenv2010_2014 / (365 *5)) %>%
  mutate(Effluent_galpday_5yr_avg = total_Eff_toenv2010_2014 / (365*5)) %>%
  st_transform(., crs = crs(ocean_mask))
  
mapview(hi_IW)

```

## Step 1: Aggregate effluent and convert to 100 x 100 m raster

**Goal:**

\- Sum up effluent from points within each 100 x 100 grid cell to get
total osds effluent for that pixel

\- Separate OSDS from Injection wells

**Assumptions:**

\- The impact of OSDS is cummulative

**Notes**

I first create custom extents for each island. I'm keeping the code here
in case they need to be adjusted, but I also saved shapefile versions of
each in the `workflow/extents` folder for use in other scripts.

```{r}

# read in ocean mask and use extent of oahu and maui points to crop
ocean_mask <- terra::rast(file.path(data_dir,"OceanMask_LBSP_post2018.tif")) #

oahu_2015_2019_nad83 <- st_transform(oahu_2015_2019, crs = crs(ocean_mask))

maui_osds_nad83 <- st_transform(mauinui_2023, crs = crs(ocean_mask))

kauai_osds_nad83 <- st_transform(kauai_2023, crs = crs(ocean_mask))

hi_osds_nad83 <- st_transform(hawaii_2023, crs = crs(ocean_mask))

# create oahu specific ocean mask and expand extent to fit entire ocean mask. covert to polygon for saving. 
extent <- ext(oahu_2015_2019_nad83)
extent_vec <- as.vector(extent)

add_vec <- c(-14283,14300,-20892,20000)
new_extent <- extent_vec + add_vec
oahu_expanded <- ext(new_extent)

oahu_poly <- as.polygons(oahu_expanded)
crs(oahu_poly) <- crs(oahu_2015_2019_nad83)
oahu_poly <- st_as_sf(oahu_poly)

# here you you can use the SpatExtent since crop is a terra function. 
ocean_mask_oahu <- terra::crop(ocean_mask, oahu_expanded, mask = T)
  plot(ocean_mask_oahu)
  
# repeat for maui

mauinui_ext <- ext(maui_osds_nad83)
extent_vec <- as.vector(mauinui_ext)

add_vec <- c(-20000,20000,-30000,20000)
new_maui_ext <- extent_vec + add_vec
maui_expanded <- ext(new_maui_ext)

maui_poly <- as.polygons(maui_expanded)
crs(maui_poly) <- crs(oahu_2015_2019_nad83)
maui_poly <- st_as_sf(maui_poly)

ocean_mask_maui <- terra::crop(ocean_mask, maui_expanded, mask = T)
plot(ocean_mask_maui)

# repeat for Kauai

kauai_ext <- ext(kauai_osds_nad83)
kauai_vec <- as.vector(kauai_ext)
add_vec <- c(-15000,15000,-10892,10000)
new_kauai_ext <- kauai_vec + add_vec
kauai_expanded <- ext(new_kauai_ext)

kauai_poly <- as.polygons(kauai_expanded)
crs(kauai_poly) <- crs(oahu_2015_2019_nad83)
kauai_poly <- st_as_sf(kauai_poly)

ocean_mask_kauai <- terra::crop(ocean_mask, kauai_expanded, mask = T)
plot(ocean_mask_kauai)

# repeat for Hawaii

hi_ext <- ext(hi_osds_nad83)
hi_vect <- as.vector(hi_ext)
add_vec <- c(-15000, 15000, -20000, 15000)
new_hi_ext <- hi_vect + add_vec
hi_expanded <- ext(new_hi_ext)

hi_poly <- as.polygons(hi_expanded)
crs(hi_poly) <- crs(oahu_2015_2019_nad83)
hi_poly <- st_as_sf(hi_poly)

ocean_mask_hi <- terra::crop(ocean_mask,new_hi_ext, mask = T)
plot(ocean_mask_hi)


# Save extents


# st_write(oahu_poly, dsn = here("workflow/extents_rds/oahu_extent.shp"))
# st_write(maui_poly, dsn = here("workflow/extents_rds/maui_extent.shp"))
#st_write(kauai_poly, dsn = here("workflow/extents_rds/kauai_extent.shp"))
# st_write(hi_poly, dsn = here("workflow/extents_rds/hi_extent.shp"))


# to read these back in use v <- vect(readRDS("test.rds"))

#################### rasterize function #######################################

rasterizePoints <- function(points,mask){
  
  #points <- mauinui_IW
  #mask <- ocean_mask_maui
  # create raster template and rasterize points with terra
  r_om <- rast(mask)
  #values(r_om) <- 1
  #rasterize points, using sum to aggregate effluent 
  osds_raster <- rasterize(vect(points), r_om, field = "Effluent_galpday_5yr_avg", fun = "sum", na.rm = T)
  
  #plot(r_om, col = "blue")
  #plot(ocean_mask_maui, add = T)
  
  return(osds_raster)
  
}

# apply to points for different scenarios 

oahu_2015_2019_eff <- rasterizePoints(oahu_2015_2019_nad83,ocean_mask_oahu)
oahu_IWeff <- rasterizePoints(oahu_IW,ocean_mask_oahu)
oahu_2010_2014_eff <- rasterizePoints(oahu_2010_2014,ocean_mask_oahu)
#oahu_2010_2014_IWeff <- rasterizePoints(osds_2010_2014_IW,ocean_mask_oahu) # the same


par(mfrow = c(2,2))
plot(oahu_2015_2019_eff)
plot(oahu_IWeff)
plot(oahu_2010_2014_eff)

minmax(oahu_IWeff)

# repeat with maui

maui_2010_2014_eff <- rasterizePoints(mauinui_2010_2014,ocean_mask_maui)
maui_IWeff <- rasterizePoints(mauinui_IW,ocean_mask_maui) # this one is coming out pretty strange
maui_2015_2019_eff <- rasterizePoints(mauinui_2015_2019, ocean_mask_maui)
#maui_2015_2019_IWeff <- rasterizePoints(mauinui_2015_2019_IW, ocean_mask_maui)

par(mfrow = c(2,2))
plot(maui_2010_2014_eff)
plot(maui_IWeff)
plot(maui_2015_2019_eff)

res(maui_2010_2014_eff)
res(oahu_2010_2014_eff)


# kauai
kauai_2010_2014_eff <- rasterizePoints(kauai_2010_2014, ocean_mask_kauai)
kauai_IWeff <- rasterizePoints(kauai_IW, ocean_mask_kauai)
kauai_2015_2019_eff <- rasterizePoints(kauai_2015_2019, ocean_mask_kauai)

plot(kauai_IWeff)
plot(kauai_2010_2014_eff)
plot(kauai_2015_2019_eff)


# hi

hi_2015_2019_eff <- rasterizePoints(hi_2015_2019, ocean_mask_hi)
hi_2010_2014_eff <- rasterizePoints(hi_2010_2014, ocean_mask_hi)
hi_IW_eff <- rasterizePoints(hi_IW, ocean_mask_hi)

par(mfrow = c(1,3))
plot(hi_2015_2019_eff)
plot(hi_2010_2014_eff)
plot(hi_IW_eff)
```

## Step 2: Apply focal statistics to estimate "sphere of impact" into nearshore

**Goal:**

-   Estimate how OSDS impacts the nearshore

**Assumptions:**

-   Ocean areas closest to pixels with higher effluent, from aggregated
    OSDS point data, will experience the highest impact from the
    leaching / runoff

```{r}

# prep and funciton for Oahu 

ocean_mask_oahu[ocean_mask_oahu != 10] <- NA  
ocean_mask_maui[ocean_mask_maui != 10] <- NA
ocean_mask_kauai[ocean_mask_kauai != 10] <- NA
ocean_mask_hi[ocean_mask_hi != 10] <- NA

#plot(ocean_mask_oahu)

applyFocalStats_extractByMask <- function(raster,mask){
  
  #raster <- oahu_2015_2019_eff
  #mask <- ocean_mask_oahu
  
  # create a matrix of the raster values, and then replace any non zero value with 1
  fw_mat <- focalMat(raster,1500, type = "circle") #1500 is what Joey used in his osds model builder. ask why
  fw_mat[fw_mat > 0 ] <- 1
  
  # apply focal statistics
  focalStat <- terra::focal(raster, w = fw_mat, fun = "sum", na.rm = T) # na.rm key to making it expand out
 
  # Extract by ocean mask 
  focalStat_nearshore <- mask(focalStat, mask)
  
  plot(focalStat_nearshore)
  
  # Normalize for km2
  focalStat_nearshore_7km <- ifel(mask == 10,focalStat_nearshore/7, NA) # joey about this also. I think I get it but not sure
  
  plot(focalStat_nearshore_7km)
  
  return(focalStat_nearshore)

}

# apply to oahu scenarios

focStat_2015_2019 <- applyFocalStats_extractByMask(oahu_2015_2019_eff,ocean_mask_oahu)
focStat_IW <- applyFocalStats_extractByMask(oahu_IWeff,ocean_mask_oahu)
focStat_2010_2014 <- applyFocalStats_extractByMask(oahu_2010_2014_eff,ocean_mask_oahu)
#focStat_2010_2014_IW <- applyFocalStats_extractByMask(oahu_2010_2014_IWeff,ocean_mask_oahu)


par(mfrow = c(2,2))
plot(focStat_2015_2019)
plot(focStat_2010_2014)
plot(focStat_IW)

# maui scenarios

maui_focStat_2010_2014 <- applyFocalStats_extractByMask(maui_2010_2014_eff, ocean_mask_maui)
maui_focStat_IW <- applyFocalStats_extractByMask(maui_IWeff, ocean_mask_maui)
maui_focStat_2015_2019 <- applyFocalStats_extractByMask(maui_2015_2019_eff, ocean_mask_maui)
#maui_focStat_2015_2019IW <- applyFocalStats_extractByMask(maui_2015_2019_IWeff, ocean_mask_maui)


par(mfrow = c(2,2))
plot(maui_focStat_2010_2014)
plot(maui_focStat_2015_2019)
plot(maui_focStat_IW)

minmax(maui_focStat_2010_2014)
minmax(maui_focStat_2015_2019)


# kauai scenarios
kauai_focStat_2010_2014 <- applyFocalStats_extractByMask(kauai_2010_2014_eff, ocean_mask_kauai)
kauai_focStat_2015_2019 <- applyFocalStats_extractByMask(kauai_2015_2019_eff, ocean_mask_kauai)
kauai_focStat_IW <- applyFocalStats_extractByMask(kauai_IWeff, ocean_mask_kauai)


# hi scenarios
hi_focStat_2010_2014 <- applyFocalStats_extractByMask(hi_2010_2014_eff, ocean_mask_hi)
hi_focStat_2015_2019 <- applyFocalStats_extractByMask(hi_2015_2019_eff,ocean_mask_hi)
hi_focStat_IW <- applyFocalStats_extractByMask(hi_IW_eff, ocean_mask_hi)

```

## Step 3: Merge all rasters for statewide dataset

```{r}

# create a collection of rasters for each time step version

all_focStat_2010_2014 <- merge(sprc(
  focStat_2010_2014,
  maui_focStat_2010_2014,
  kauai_focStat_2010_2014,
  hi_focStat_2010_2014
  ))


plot(all_focStat_2010_2014)


all_focStat_2015_2019 <- merge(sprc(
  focStat_2015_2019,
  maui_focStat_2015_2019,
  kauai_focStat_2015_2019,
  hi_focStat_2015_2019
  ))

plot(all_focStat_2015_2019) # where is lanai?

all_focStat_IW <- merge(sprc(
  focStat_IW,
  maui_focStat_IW,
  kauai_focStat_IW,
  hi_focStat_IW
  ))


plot(all_focStat_IW)

```

## Step 4: Use ocean mask to limit area of impact to marine space & add decay function

**Assumptions:**

-   Intensity of impact lessens with distance from shore, due to
    chemical decay of N or P and / or wave or currents

-   Decay occurs as an exponential function of distance from shore
    (Gaussian Decay)

```{r}


# get custom decay function

source(here("workflow/offshoreDecay.R")) # identical to all other decay for LBSP

# read in Euclidean distance, check the crs and crop by the custom extent

euc_distance <- rast(file.path(data_dir,"DistFromShore_Euc_2023.tif"))
#euc_distance_oahu <- terra::crop(euc_distance, ocean_mask_oahu, mask = T)
#euc_distance_maui <- terra::crop(euc_distance, ocean_mask_maui, mask = T)
#euc_distance_kauai <- terra::crop(euc_distance, ocean_mask_kauai, mask = T)

#plot(euc_distance_oahu)
#plot(euc_distance_maui)
#plot(euc_distance_kauai)

# apply to nearshore rasters - Oahu
  
final_osds_2015_2019 <- offshore_decay(all_focStat_2015_2019,euc_distance)
final_osds_2015_2019[final_osds_2015_2019 == 0] <- NA
minmax(final_osds_2015_2019)


final_IW <- offshore_decay(all_focStat_IW,euc_distance)
final_IW[final_IW == 0] <- NA


final_osds_2010_2014 <- offshore_decay(all_focStat_2010_2014,euc_distance)
final_osds_2010_2014[final_osds_2010_2014 == 0] <- NA
minmax(final_osds_2010_2014)

#####################
par(mfrow = c(2,2))
plot(final_osds_2015_2019)
plot(final_IW)
plot(final_osds_2010_2014)


minmax(final_osds_2010_2014)
minmax(final_osds_2015_2019)



```

## Save Rasters

```{r}

writeRaster(
  final_osds_2015_2019,
  filename = here("outputs/AllIsl_OSDS_2015_2019_eff.tif"),
  overwrite = T
)

writeRaster(
  final_osds_2010_2014,
  filename = here("outputs/AllIsl_OSDS_2010_2014_eff.tif"),
  overwrite = T
)

writeRaster(
  final_IW,
  filename = here("outputs/AllIsl_IW_eff.tif"),
  overwrite = T
)




```


## Make some maps

```{r}
# read in basemaps
source(here("workflow/3D_basemaps.R"))

# read in final rasters from output folder

inject_wells <- rast(here("outputs/AllIsl_IW_eff.tif"))

iw_df <- terra::as.data.frame(inject_wells, xy = TRUE, na.rm = TRUE)
colnames(iw_df)[3] <- "effluent_gal_day"

eff_2015_2019 <- rast(here("outputs/AllIsl_OSDS_2015_2019_eff.tif"))

eff_df <- terra::as.data.frame(eff_2015_2019, xy = TRUE,na.rm = TRUE)

colnames(eff_df)[3] <- "effluent_gal_day"

plot(inject_wells)
plot(eff_2015_2019)

library(ggnewscale)

x_limits <- range(hi_dem_df$x, na.rm = TRUE)
y_limits <- range(hi_dem_df$y, na.rm = TRUE)


hi_test <- hi_with_dem + 
  new_scale_fill() + # necessary to maintain basemap
  geom_raster(data = eff_df, aes(x = x, y = y, fill = effluent_gal_day), alpha = 0.7)+
  scale_fill_gradientn(colours=c("white", "olivedrab"),
                       trans = "sqrt")+
  coord_sf(xlim = x_limits, ylim = y_limits, expand = FALSE)

hi_test

# reset limits for oahu
x_limits <- range(oahu_dem_df$x, na.rm = TRUE)
y_limits <- range(oahu_dem_df$y, na.rm = TRUE)

oahu_eff <- oahu_with_dem + 
  new_scale_fill() + # necessary to maintain basemap
  geom_raster(data = eff_df, aes(x = x, y = y, fill = effluent_gal_day))+
  scale_fill_gradientn(colours=c("white", "#5C4033"),
                       trans = "sqrt")+
  coord_sf(xlim = x_limits, ylim = y_limits, expand = FALSE)

oahu_eff

oahu_IW <- oahu_with_dem + 
  new_scale_fill() + # necessary to maintain basemap
  geom_raster(data = iw_df, aes(x = x, y = y, fill = effluent_gal_day))+
  scale_fill_gradientn(colours=c("white", "#6A6E09"),
                       trans = "log")+
  coord_sf(xlim = x_limits, ylim = y_limits, expand = FALSE)

oahu_IW



```

