---
title: "OSDS_change"
author: "Madeline Berger"
date: "2025-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



library(here)

source(here("workflow/setup.R"))


```

## Comparing past point data

This markdown contains code that explores and compares past versions of the OSDS point data with the 2023 version. The reason is to see if I can get an idea of spatial patterns of upgrades. We have two previous versions to work with:

- 2017 OSDS data for Hawaii Island only, used in the West Hawaii paper. Unpublished
- Data that Joey used for OTP. 

Why this is important / useful:

-   the new dataset replaced data for a point with its most up to date information. So, for example, if a point was a cesspool from 2003

-   2011, but then was converted to a septic system in 2011, this dataset will have that point classified as that septic system for its entire existence.

-   This means that if we wanted to compare past estimates of effluent to what we find today, we actually can't really do that accurately with just the 2023 dataset.

-   Since it seems like very few points are ever removed (ie we can't confirm complete shutdown or removal, unless the house was covered by lava), these upgrades to a better treatment class are the only way we would be able to "detect" a reduction in OSDS impact over time. Otherwise, we should just expect later time periods to always show increasing impacts.

-   Because as time goes on, more points are added (as more permits are acquired) but if none are taken away, than across the board we'll just see increases, even if small.

### 2023 to 2017


```{r}

# read in prior version form 2017

hi_2017_raw <- read_sf(here("data/osds/Hawaii_2017/Haw_Island_OSDS_2017.shp"))

# what happens if we run distinct?

hi_2017_dist <- distinct(hi_2017_raw) # 300 removed that were exactly the same

hi_2023_raw <- read_sf(here("data/osds/hawaii_2023_raw.shp"))

hi_2023_dist <- distinct(hi_2023_raw) # none removed, but weirdly the same raw length as 2017


# so there are 300 unique points in 2023 that are not unique in 2017. 

```

**Check distinct points based on different columns**

Which data rows are completely distinct?

- All rows contain unique combinations of data in both datasets

#### Which geometries are distinct?

```{r}

hi_2017_dist_geoms <- hi_2017_raw %>% distinct(geometry, .keep_all = T) # removed 138 points

hi_2017_dup_geoms <- setdiff(hi_2017_raw, hi_2017_dist_geoms) # look at duplicates

hi_2017_dups2 <- hi_2017_raw %>% filter(geometry %in% hi_2017_dup_geoms$geometry)

#mapview(hi_2017_dist_geoms) + mapview(hi_2017_dup_geoms, col.regions = "red")

plot(st_geometry(hi_2017_dist_geoms), col = "blue", pch = 16, main = "Unique 2017 Points")
plot(st_geometry(hi_2017_dup_geoms), col = "red", pch = 16, add = T)


####### 2023

hi_2023_dist_geoms <- hi_2023_raw %>%  distinct(geometry, .keep_all = T) # removed 5
hi_2023_dup_geoms <- setdiff(hi_2023_raw, hi_2023_dist_geoms)

mapview(hi_2023_dup_geoms)

# look at them all together

hi_2023_doubles <- hi_2023_raw %>%  filter(geometry %in% hi_2023_dup_geoms$geometry)


```

From this investigation, it seems like many of the duplicate geometries are also duplicate TMKS that are labeled as cesspool upgrades.

#### Which TMKs are distinct?

Thourough QA QC of all the duplicate TMKs.

```{r}

# are the non-distinct TMKS upgrades, multiple properties on the same TMK, or duplicates

#hi_2017_dist_tmks <- hi_2017_raw %>% distinct(TMK, .keep_all = T)
library(data.table)
dt <- as.data.table(hi_2017_raw)

tmk_counts <- dt %>%
  group_by(TMK) %>%
  tally() %>%
  filter(n > 1) #2089

# exemine duplicates togther more closely

hi_2017_dup_tmks <- hi_2017_raw %>% filter(TMK %in% tmk_counts$TMK) #4178

mapview(hi_2017_dup_tmks)

# isolate possible cesspool replacements

hi_2017_replacements <- hi_2017_dup_tmks %>% 
  filter(COMMENTS == "Possible Cesspool Replacement") #3143
  #filter(TYPE == "Septic")
  

hi_replacement_tmk <-  hi_2017_replacements %>%  pull(TMK) %>%
  unique()

hi_2017_non_upgrade_dups <- hi_2017_dup_tmks %>%
  filter(!(TMK %in% hi_replacement_tmk)) # this is zero

# a few others checks: 

## Which TMKS are exactly the same

exact_copes <- hi_2017_dup_tmks %>%
  group_by(across(everything())) %>%
  tally() %>%
  filter(n > 1) # none

## Are the comments on all the duplicate tmks the same?
comments <- hi_2017_dup_tmks %>%
  group_by(TMK) %>%
  summarize(all_comments = paste(unique(COMMENTS), collapse = ", ")) %>%
  filter(grepl(",", all_comments))

# are there any with more than 2 entries?

tmk_counts %>% filter(n > 2) # no

```

Wrangle confirmed cesspool upgrades (as determined by notes)

```{r}


hi_2017_cess2sep <- hi_2017_raw %>% filter(TMK %in% hi_2017_replacements$TMK) 

# hi_2017_cess2sep2 <- hi_2017_raw %>% filter(TMK %in% hi_2017_dup_tmks$TMK) # identical to above so we are good

# group and pivot to examine differences

hi_2017_summary <- hi_2017_cess2sep %>% 
  as.data.frame() %>% 
  st_drop_geometry() %>% 
  select(TMK, OSDS_TYPE,starts_with("CLASS"),EFF_DATE, DATE_BASIS,EFFLUENT, N_FLUX) %>% 
  pivot_longer(cols = starts_with("CLASS"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity > 0)

length(unique(hi_2017_summary$TMK)) #2029 if not

mapview(hi_2017_cess2sep, zcol = "OSDS_TYPE")


```


From this, we can see that many of the TMKS in the `hi_2017_summary` dataset are likely upgrades. It is important that we use the "cesspool" version prior to the upgrade. Weirdly, some are also still cesspools even after upgrade. 

Are these possible upgrades in the 2023 dataset? Filter 2023 by TMKs in `hi_2017_summary` to find out.

```{r}

hi_2023_poss_upgrades <- hi_2023_raw %>% filter(TMK %in% hi_2017_summary$TMK) # 2085

# which two are missing? map it
missing_from_2023 <- setdiff(hi_2017_summary$TMK,hi_2023_poss_upgrades$TMK)

missing_from_2023geom <- hi_2017_raw %>% filter(TMK %in% missing_from_2023)
mapview(missing_from_2023geom, label = "EFF_DATE")

# are any duplicates in the 2023 dataset?

dt_2023 <- as.data.table(hi_2023_raw)


tmk_counts_2023 <-dt_2023 %>% 
  group_by(TMK) %>%
  tally() # no repeat tmks

#  lets look at the tmks that are marked as duplicates in 2017 in 2023 - are they all "upgraded"?

hi_2023_poss_upgr_types <- hi_2023_poss_upgrades %>% 
  group_by(PrmtTyp,DspslTy,osds_cl) %>% 
  summarize(
    total_count = sum(OSDS_QT)
  )

# from this table we can see most of them are still cesspools. 

hi_2023_still_cesspools <- hi_2023_poss_upgrades %>%
  # select(TMK, OSDS_TYPE = PrmtTyp,starts_with("CLASS"),EFF_DAT,EFFLUENT = EFFLUEN, N_FLUX = Ntrgn_F) %>% 
  # pivot_longer(cols = starts_with("CLASS"), names_to = "class", values_to = "quantity") %>% 
  # filter(quantity > 0)


```

These TMK are in 2017 but not in 2023:
- 375032019
- 383007015
- 384005033
- 392011037

Findings from this we see that all but 3 of TMKS with possible cesspool upgrades are still present in the 2023 dataset. None are duplicated. By 2023, none are classified as cesspools.

However, some still are "class 5", which is the same treatment level as cesspools. This appears to be due to the fact that they TMK has multiple and in that instance per Julia's methods, are assumed to be Class V. 


#### New Septic: What are the differences between old and new TMKs with possible upgrades, just using 2017 data?

Separate out newer versions to compare with 2023. Do they match?
Also, calculate the effluent difference if we were to use the older version.

```{r}

# assess changes from the 2017 data

#hi_2017_summary <- filter(hi_2017_summary, str_detect(OSDS_TYPE,"Septic"))

class(hi_2017_summary$EFF_DATE)

older_2017 <- hi_2017_summary %>% 
  group_by(TMK) %>% 
  slice_min(EFF_DATE, n = 1, with_ties = FALSE) %>% 
  mutate(instance = "older") %>% 
  ungroup()

newer_2017 <- hi_2017_summary %>% 
  group_by(TMK) %>% 
  slice_max(EFF_DATE, n = 1, with_ties = FALSE) %>% 
  mutate(instance = "newer") %>% 
  ungroup()


# which classes have changed, and how much has quantity and effluent changed?

hi_2017_upgrade_summ <- bind_rows(newer_2017,older_2017)

hi_2017_upgrade_wide <- hi_2017_upgrade_summ %>% 
  select(TMK,EFF_DATE,instance,class,quantity,EFFLUENT,N_FLUX) %>% 
  pivot_wider(id_cols = TMK,
              names_from = instance,
              values_from = c(EFF_DATE,class,quantity,EFFLUENT,N_FLUX),
              names_sep = "_") %>% 
  mutate(
    class_changed = class_older != class_newer,
    effluent_diff = EFFLUENT_older - EFFLUENT_newer, # if this is NEGATIVE than the new effluent is more??
    qty_diff = quantity_older - quantity_newer,
    nflux_diff = N_FLUX_older - N_FLUX_newer
  )


# make mini df for join

instance_join <- hi_2017_upgrade_summ %>% 
  select(TMK, EFF_DATE, instance)


```

Note - these may have additional changed geometry. But right now we are just focused on the TMKS

##### Visualize Findings

First we need to re-join the data to geometries.

```{r}

# figure out if the geoms are similar enough to just grab whichever, or if it matters

hi_2017_upgrade_geoms <- hi_2017_raw %>% 
  filter(TMK %in% hi_2017_upgrade_wide$TMK) %>%
  select(TMK, EFF_DATE, geometry) %>% 
  distinct() %>% 
  left_join(., instance_join, by = c("TMK","EFF_DATE"))

upgrade_nosf <- hi_2017_upgrade_geoms %>% 
  mutate(geom = geometry) %>% 
  st_drop_geometry()

geom_compare <- upgrade_nosf %>% 
  select(TMK, instance, geom) %>% 
  pivot_wider(
    id_cols = TMK,
    names_from = instance,
    values_from = geom,
    names_prefix = "geometry_"
  )


geoms_wide_sf <- geom_compare %>% 
  st_as_sf()

geoms_wide_sf$geom_sameish <- mapply(
  function(g1, g2) st_distance(g1, g2) <= 300,
  geoms_wide_sf$geometry_older,
  geoms_wide_sf$geometry_newer
)
# ok great they are all the same. note this awesome function

# slice first geom for each duplicate tmk so we have one geom per TMK 

hi_upgrade_one_geom <- hi_2017_upgrade_geoms %>% 
  group_by(TMK) %>% 
  slice(1) %>% 
  ungroup()

# join to the data frame with our "change metrics" so we can plot and look for spatial patters

hi_2017_upgrade_changes <- hi_upgrade_one_geom %>% 
  select(TMK,geometry) %>% 
  right_join(.,hi_2017_upgrade_wide, by = "TMK")


```


Create some plots

```{r}


# read in basemap script to load basic 3D grey basemaps for each island

source(here("workflow/3D_basemaps.R"))

# create some columns for each color and size mapping

hi_2017_upchange_plot <- hi_2017_upgrade_changes %>% 
  mutate(
    color_eff = ifelse(effluent_diff >= 0,"dodgerblue","red"), #red meaning effluent has increased over time
    magnitude_eff = abs(effluent_diff),
    direction_qty = ifelse(effluent_diff > 0,"Decreased or No Change","Increased"),
    magnitude_qty = abs(qty_diff),
    direction_n = ifelse(nflux_diff > 0, "Decreased or No Change","Increased"),
    magnitude_n = abs(nflux_diff)
  )


# distributions of the variables to plot

hist(hi_2017_upchange_plot$effluent_diff)

hist(hi_2017_upchange_plot$qty_diff)


# ggplot(data = hi_2017_upchange_plot)+
#   #geom_sf(data = land, fill = "grey95", color = "grey80")+
#   geom_sf(aes(size = magnitude, color = color), alpha = 0.6) + 
#   scale_color_identity()+
#   theme_minimal()+
#   labs(title = "Possible Cesspool Upgrades",
#        size = "Magnitude of Effluent Change")
# 
# 
# 
# 
# ggplot(data = hi_2017_upchange_plot)+
#   #geom_sf(data = land, fill = "grey95", color = "grey80")+
#   geom_sf(aes(size = abs(effluent_diff), color = effluent_diff), alpha = 0.6) + 
#   scale_color_gradient2(low = "dodgerblue", high = "red", midpoint = 0)+
#   scale_size_continuous(range = c(2,10))+
#   theme_minimal()+
#   labs(title = "Possible Cesspool Upgrades",
#        size = "Magnitude of Effluent Change")



# what about quantity of OSDS at that tmk?

effluent_change_final <- hi_with_dem +
  geom_sf(data = hi_2017_upchange_plot,aes(size = magnitude_eff, color = color_eff), alpha = 0.6) + 
  scale_color_identity()+
  #theme_minimal()+
  labs(title = "Possible Cesspool Upgrades on Hawaii",
       size = "Magnitude of Effluent Change",
       #color = "Direction of Magnitude Change"
       )


effluent_change_final

ggsave(plot = last_plot(),
       filename = here("figs/osds_2017_upgradesEffChange.png"),
       width = 5.5,
       height = 5, 
       units = "in",
       dpi = 300)

# quantity change

quantity_change <- hi_with_dem + 
  geom_sf(data = hi_2017_upchange_plot,aes(size = magnitude_qty, color = direction_qty), alpha = 0.6)+
  scale_color_manual(
    values = c(
      "Increased" = "coral",
      "Decreased or No Change" = "cadetblue"
    )
  )+
  scale_size_continuous(range = c(1, 5), name = "Magnitude of change in Qty")+
  #theme_minimal()+
  labs(title = "Possible Cesspool Upgrades on Hawaii",
       size = "Change in OSDS Number",
       #color = "Direction of Magnitude Change"
       )
  

quantity_change

ggsave(plot = last_plot(),
       filename = here("figs/osds_2017_upgradesQtyChange.png"),
       width = 5.5,
       height = 5, 
       units = "in",
       dpi = 300)


# nflux change

nflux_change <- hi_with_dem + 
  geom_sf(data = hi_2017_upchange_plot,aes(size = magnitude_n, color = direction_n), alpha = 0.6)+
  scale_color_manual(
    values = c(
      "Increased" = "brown",
      "Decreased or No Change" = "aquamarine"
    )
  )+
  scale_size_continuous(range = c(1, 5), name = "Magnitude of change in N Flux")+
  labs(title = "Possible Cesspool Upgrades on Hawaii",
       size = "Change in N Flux",
       color = "Direction of Change"
       )
  

nflux_change

ggsave(plot = last_plot(),
       filename = here("figs/osds_2017_upgradesNChange.png"),
       width = 5.5,
       height = 5, 
       units = "in",
       dpi = 300)


```

The mystery here to me is why are we seeing an increase in effluent following a decrease in class assignment? Is it because we changed the calculation? But, I didn't do that here.

Second question to explore: Does N scale directly with effluent? Probably not but double check metadata. Then map N as well. 

#### Are the newer versions of the upgraded cesspools in 2017 showing the same data for 2023?
Ie, is the effluent, N, and qty the same? Hypothesis: no because calculation might be different.

```{r}

# grab just newer versions of the 2017 cesspool upgrades
newer_2017 <- hi_2017_upgrade_wide %>% 
  select(TMK, ends_with("newer")) %>% 
  rename_with(~str_replace(., "newer","2017"))

# create wide version of 2023

hi_2023_upgrades_wide <- hi_2023_poss_upgrades %>%
  select(TMK, OSDS_TYPE = PrmtTyp,class = osds_cl,EFF_DATE = EFF_DAT,EFFLUENT = EFFLUEN, N_FLUX = Ntrgn_F, quantity = OSDS_QT) %>% 
  #pivot_longer(cols = starts_with("CLASS"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity > 0) %>% 
  rename_with(~paste0(.x,"_2023"), -TMK,-geometry) #2370 why duplicates? because how we have more than 1 class


# join together and re-calculate the impact change metrics

hi_2023_2017_compare <- left_join(hi_2023_upgrades_wide, newer_2017, by = "TMK") %>% 
  mutate(
    class_2017 = case_when(
      class_2017 == "CLASS_I" ~ 1,
      class_2017 == "CLASS_II" ~ 2,
      class_2017 == "CLASS_III" ~ 3,
      class_2017 == "CLASS_IV" ~ 4,
      class_2017 == "CLASS_V" ~ 5,
      TRUE ~ NA
    )
  ) %>% 
  mutate(
    date_changed = EFF_DATE_2017 != EFF_DATE_2023,
    class_changed = class_2017 != class_2023,
    effluent_diff = EFFLUENT_2017 - EFFLUENT_2023, # if this is NEGATIVE than the new effluent is more??
    qty_diff = quantity_2017 - quantity_2023,
    nflux_diff = N_FLUX_2017 - N_FLUX_2023,
    all_match = !date_changed & !class_changed & effluent_diff == 0 & qty_diff == 0 & nflux_diff == 0
  )



# which have changed class
hi_2017_class_diff <- hi_2023_2017_compare %>% 
  filter(class_changed == TRUE) # so some of these have different classes, same date but the same N flux? # n = 214

# which have changed DATE
hi_2017_date_diff <- hi_2023_2017_compare %>%
  filter(date_changed == TRUE) # n = 171
  

summary_counts <- hi_2023_2017_compare %>%
  distinct(TMK, .keep_all = TRUE) %>% 
  summarize(
    total = n(),
    exact_matches = sum(all_match, na.rm = TRUE),
    differing = total - exact_matches # ok so 582 are different while the rest match what is in our dataset. What is driving these differences? hunch = osds assigned as multiple, maybe
  )



```

Ok SO - we see there are key differences between 2017 and 2023. Does this mean that these have been upgraded again? is the information just different? What is driving these changes. 

I think we can create a strong framework for what to consider changes. For example: if class and date are changing together, is that a new permit and should be considered an upgrade. If tha class changed but the date didnt...what? if date changed but class didnt....what is this.


Next steps: 

- Compare to 2010 datasets? This may not be a good exercise if the calculation is different. Wait until post meeting
- Possibly create new yearly versions

