---
title: "Oahu Timeseries Streamlined"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))

```


### Step 1: Clean Up Dwell Data
About dwelling data: 
 - obtained from each county, and maintained by each county differently 
 - records are created for tax assessment purposes, and therefore usually have two key time columns: year built and year assessed.
 - if the tmk has a second card number, that usually indicates an auxiliary building or additional structure built on the property.

**Data Cleaning steps:**
- parse data into correct columns (for HI, this was very sketchy)
- look for exact duplicates of data with ALL COLUMNS - note which columns are differentiating them
  - note: you canʻt do this for HI or Oahu because the parsing is so sketch
- select columns of interest
- look for duplicates again
  - if we find duplicates, need to figure out which column is differentiating dwellings. this is very important. we also need to add a unique ID to each row. 
  
  Parse Oahu Raw files, join all years

This can all be found in the OG driver repository on github. Here I'll start from the clean CSV I wrote to generate products for QA QC report 2

This report should tell us:

- total number of TMKS assessed each year for Oahu
- how many TMKS are single owner versus multiple owner per year of data
- how many TMKS have more than one card (i.e. more than one building assessed on the parcel) per year
- Summary of bedrooms and bathrooms per year?

Note: go back to parsing data and check the card column. 

Resources: 

https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/

```{r dwelldata}

library(data.table)

oahu_dwell_path <- file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_2015_2020.csv")

oahu_dwells_cleanedup <- fread(file = oahu_dwell_path)

oahu_dwells_cleanedup <- oahu_dwells_cleanedup %>% 
  mutate(full_taxid = substr(paste0(jurisid,taxid),1,9))

length(unique(oahu_dwells_cleanedup$full_taxid)) #how many? 14,8527


```


### Step 2: Bring in Permit Data (IWS, Dept. of Health)

```{r permit data}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) #1741 rows, # 1412 tax ids

tmks_permits <- unique(iws_oahu_permits$full_taxid)

# execute join to see which tmks have dwell data

iws_oahu_dwell_join <- inner_join(iws_oahu_permits, oahu_dwells_cleanedup, by = "full_taxid") %>% 
  distinct() #28215 rows and 1166 tmks

iws_oahu_dwell_join_norepeats <- iws_oahu_permits %>% 
  filter(full_taxid %in% oahu_dwells_cleanedup$full_taxid) #1355 rows, 1666 tmks yay!!

# now filter out the ones that do not have dwelling data and save

iws_oahu_NOdwell <- iws_oahu_permits %>% 
  filter(full_taxid %!in% iws_oahu_dwell_join$full_taxid) # 386. perfect

write_csv(iws_oahu_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/oahu_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**:
- There were 1741 permits issued for HI island over the past 10 years
- 1355 have dwelling information, and this represents 6011 tmks
- 1105 do not have dwelling information

### Step 2b: Clean and Pivot New Permits

Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 

```{r}
library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by adding a year column, assigning osds class, and trying to standardize the address

master_clean <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

master_clean2 <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) # Arranging helps you see them side-by-side

view(duplicates_to_check)


length(unique(master_clean$tmk)) # 1412, so we still have all tmks represented after distinct

# N rows = 1683 after distinct, 1741 before. Lost 58
# N tmks = 1412

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #


### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(full_taxid) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(full_taxid %in% oahu_dwells_cleanedup$full_taxid),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- oahu_dwells_cleanedup %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    total_beds = sum(bedrooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)

table2b_bedrooms <- oahu_dwells_cleanedup %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bedrooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)


#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(full_taxid, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(full_taxid, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)

# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2a_bedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2b_bedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_allcsv"))

```




### Step 3: Compare and Integrate Past Datasets