---
title: "Oahu Timeseries Streamlined"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))

```

### Step 0: Create authorative TMK lists

```{r}

library(janitor)
# read in new permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  clean_names()

iws_oahu_permits_tmk <- iws_oahu_permits %>% pull(tmk)

# read in old otp

otp_oahu_raw <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/osds_oah.shp")) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk_9)) %>% 
  st_drop_geometry()

otp_oahu_tmk <- otp_oahu_raw %>%
  pull(tmk)

# new ones:

oahu_unchanged <- otp_oahu_raw %>% 
  filter(tmk %!in% iws_oahu_permits_tmk) #12519

unchanged_tmks <- oahu_unchanged %>% pull(tmk) %>% unique() #12461

oahu_updated <- otp_oahu_raw %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_oahu_permits_tmk) #1165

updated_tmks <- oahu_updated %>% pull(tmk) %>% unique() # 1164

oahu_new <- iws_oahu_permits %>% 
  filter(tmk %!in% otp_oahu_tmk) # 314

new_tmks <- oahu_new %>% pull(tmk) %>% unique() #
# create authoritative tmk list - add ups!

master_tmk_list <- c(iws_oahu_permits_tmk,otp_oahu_tmk)
master_tmk_list <- unique(master_tmk_list) # 13873



```

### Step 1: Clean Up Dwell Data

About dwelling data: - obtained from each county, and maintained by each county differently - records are created for tax assessment purposes, and therefore usually have two key time columns: year built and year assessed. - if the tmk has a second card number, that usually indicates an auxiliary building or additional structure built on the property.

**Data Cleaning steps:** - parse data into correct columns (for HI, this was very sketchy) - look for exact duplicates of data with ALL COLUMNS - note which columns are differentiating them - note: you canʻt do this for HI or Oahu because the parsing is so sketch - select columns of interest - look for duplicates again - if we find duplicates, need to figure out which column is differentiating dwellings. this is very important. we also need to add a unique ID to each row.

Parse Oahu Raw files, join all years

Part of this can all be found in the OG driver repository on github.

The earlier years I got later (2010 - 2014) I'll do here

```{r}

# read in packge
library(readr)

# define col specs
col_specs_old <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92, 112, 132, 152, 154,
    179, 186, 193, 199, 201, 202, 206, 211, 216, 218, 238, 239,
    259, 279, 299, 319, 343, 355, 357, 361, 364, 384, 404, 434,
    438, 443, 447, 448, 470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 18, 23, 43, 44, 47, 71, 91, 111, 131, 151, 153, 173,
    180, 187, 194, 200, 201, 201, 207, 212, 217, 237, 238, 258,
    278, 298, 318, 342, 354, 356, 360, 363, 383, 403, 433, 437,
    442, 446, 448, 469, 478, 489, 494, 502, 509, 515 
  )
)

# try new ones

col_specs_new <- fwf_positions(
  start = c(
    1, 15, 20, 24, 44, 49, 52, 72, 92, 112, 132, 152, 154, 
    179, 186, 193, 199, 201, 202, 206, 211, 217, 218, 238, 
    239, 259, 279, 299, 319, 343, 349, 356, 358, 365, 373, 
    375, 379, 382, 402, 422, 443, 448, 452, 453, 457, 479, 
    488, 499, 504, 512, 519
  ),
  end = c(
    14, 19, 23, 43, 48, 51, 71, 91, 111, 131, 151, 153, 178, 
    185, 192, 198, 200, 201, 205, 210, 216, 217, 237, 238, 
    258, 278, 298, 318, 342, 348, 355, 357, 364, 372, 374, 
    378, 381, 401, 421, 442, 447, 451, 452, 456, 478, 487, 
    498, 503, 511, 518, 525 # Adjust final end based on actual row length
  ),
  col_names = c(
    "tax_map_key", "parcel_year", "card_no", "building_no", "indicator", 
    "story_height", "roof_design", "roof_structure", "roof_material", 
    "ext_wall", "framing", "style_occupancy", "occupancy", "full_baths", 
    "half_baths", "add_fixtures", "total_fixtures", "ac", "attic_rating", 
    "total_rooms", "bedrooms", "family_rooms", "foundation", "basement", 
    "floor_construction", "flooring", "int_wall_struct", "int_wall_mat", 
    "ceiling", "rec_room_size", "misc_desc_1", "misc_qty_1", "misc_desc_2", 
    "misc_qty_2", "condo_level", "condo_type", "condo_view", "condo_parking", 
    "condo_style", "condo_common", "year_built", "eff_year_built", 
    "phys_cond", "bldg_code", "bldg_grade", "bldg_shape", "cost_design_pct", 
    "econ_cond_factor", "pct_complete", "ground_floor_area", "living_area"
  )
)

# read in each year and check how it worked out

oahu_2014 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2014-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2013 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2013-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2012 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2012-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))



oahu_2011 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2011-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2010 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2010-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))

```

Make sure columns we need are clean:

```{r}


check_parsing_quality <- function(df, year_label = "Unknown Year") {
  # 1. Calculate lengths
  results <- df %>%
    mutate(
      len_year = nchar(as.character(year_built)),
      is_messy = len_year != 4
    )
  
  # 2. Count totals
  total_rows <- nrow(results)
  messy_rows <- results %>% filter(is_messy)
  n_messy    <- nrow(messy_rows)
  pct_messy  <- (n_messy / total_rows) * 100
  
  # 3. Print a pretty summary to the console
  cat("\n==========================================\n")
  cat("PARSING REPORT FOR:", year_label, "\n")
  cat("Total Rows:    ", total_rows, "\n")
  cat("Clean Rows:    ", total_rows - n_messy, "\n")
  cat("Messy Rows:    ", n_messy, "\n")
  cat("Error Rate:    ", round(pct_messy, 2), "%\n")
  
  if(n_messy > 0) {
    cat("\nCommon 'Messy' Lengths detected:\n")
    print(table(messy_rows$len_year))
    
    cat("\nExample of messy value: '", head(messy_rows$year_built, 1), "'\n")
  }
  cat("==========================================\n")
  
  # Return the messy data invisibly so you can assign it if needed
  return(invisible(messy_rows))
}

oahu_2014_messy <- check_parsing_quality(oahu_2014, "2014 Data")
oahu_2013_messy <- check_parsing_quality(oahu_2013, "2013 Data")
oahu_2012_messy <- check_parsing_quality(oahu_2012, "2012 Data")
oahu_2011_messy <- check_parsing_quality(oahu_2011, "2011 Data")
oahu_2010_messy <- check_parsing_quality(oahu_2010, "2012 Data")

# check bedroom

check_bedroom_quality <- function(df, year_label = "Dataset") {
  
  # 1. Clean and standardize the bedroom column
  # We convert to character and trim to catch "00", "0", or "   0"
  results <- df %>%
    mutate(
      bd_char = str_trim(as.character(bedrooms)),
      status = case_when(
        is.na(bedrooms) | bd_char == "" | bd_char == "NA" ~ "Missing/NA",
        bd_char %in% c("0", "00", "000") ~ "Zero (00)",
        TRUE ~ "Valid Count"
      )
    )
  
  # 2. Generate summary stats
  summary_table <- table(results$status, useNA = "always")
  total_rows <- nrow(df)
  n_na   <- sum(results$status == "Missing/NA")
  n_zero <- sum(results$status == "Zero (00)")
  
  # 3. Print the report
  cat("\n==========================================\n")
  cat("BEDROOM QUALITY REPORT:", year_label, "\n")
  cat("Total Records: ", total_rows, "\n")
  cat("------------------------------------------\n")
  cat("NA / Missing:  ", n_na, " (", round(n_na/total_rows*100, 1), "%)\n")
  cat("Zero (00):     ", n_zero, " (", round(n_zero/total_rows*100, 1), "%)\n")
  cat("Valid (> 0):   ", sum(results$status == "Valid Count"), "\n")
  cat("==========================================\n")
  
  # Return the summarized counts as a data frame
  return(as.data.frame(summary_table))
}

check_bedroom_quality(oahu_2014, "2014")

table(oahu_2014$indicator, is.na(oahu_2014$bedrooms))

# lets just join and filer if possible

oahu_2010_2014_cl <- rbind(
  oahu_2014,
  oahu_2013,
  oahu_2012,
  oahu_2011,
  oahu_2010
) %>% 
  select(tax_map_key,
         tax_year = parcel_year,
         card_no,
         total_rooms,
         bedrooms,
         year_built,
         eff_year_built)

oahu_cl_tmk <- oahu_2010_2014_cl %>% 
  mutate(tmk = substr(tax_map_key,1,9))


iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date))

oahu_dwell_osds_2010_2014 <- oahu_cl_tmk %>% 
  filter(tmk %in% master_tmk_list) # 9759


write_csv(oahu_dwell_osds_2010_2014,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_2010_2014.csv"))

```

This report should tell us:

-   total number of TMKS assessed each year for Oahu
-   how many TMKS are single owner versus multiple owner per year of data
-   how many TMKS have more than one card (i.e. more than one building assessed on the parcel) per year
-   Summary of bedrooms and bathrooms per year?

Note: go back to parsing data and check the card column.

Resources:

<https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/>

```{r dwelldata}

library(data.table)

oahu_dwell_path_2015_2020 <- file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_2015_2020.csv")


oahu_dwell_path_2010_2014 <- file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_2010_2014.csv")

oahu_dwells_new <- fread(file = oahu_dwell_path_2015_2020) %>% 
  mutate(tmk = substr(paste0(jurisid,taxid),1,9)) %>% 
  filter(tmk %in% master_tmk_list)

oahu_dwells_old <- fread(file = oahu_dwell_path_2010_2014) %>% 
  select(-tax_map_key, -eff_year_built)

oahu_dwells_all <- oahu_dwells_new %>% 
  select(tmk,tax_year,card_no = card, total_rooms, bedrooms,year_built) %>% 
  rbind(oahu_dwells_old)

write_csv(oahu_dwells_all,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_alltmks.csv"))


length(unique(oahu_dwells_all$tmk)) #how many? 11163

length(unique(master_tmk_list)) #13873 <- so some of these are commercial or not in dwell

```

### Step 2: Bring in Permit Data (IWS, Dept. of Health)

```{r permit data}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk,"1")) %>%
  rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk))

tmks_permits <- unique(iws_oahu_permits$tmk) #1412

# execute join to see which tmks have dwell data

iws_oahu_dwell_join <- inner_join(iws_oahu_permits, oahu_dwells_all, by = "tmk") %>% 
  distinct() 

iws_oahu_dwell_join_norepeats <- iws_oahu_permits %>% 
  filter(tmk %in% oahu_dwells_all$tmk) #1369, 1178

# now filter out the ones that do not have dwelling data and save

iws_oahu_NOdwell <- iws_oahu_permits %>% 
  filter(tmk %!in% iws_oahu_dwell_join$tmk) # 372, 234 

write_csv(iws_oahu_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/oahu_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**: - There were 1741 permits issued for HI island over the past 10 years - 1355 have dwelling information, and this represents 6011 tmks - 1105 do not have dwelling information

### Step 2b: Clean and Pivot New Permits

Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. are as follows: 1. Assign class, using a custom function 2. Create master list (table 1) 3. Create tables 2 and 3 based on master list.

```{r}

library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by adding a year column, assigning osds class, and trying to standardize the address

master_clean <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$tmk,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

master_clean2 <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$tmk,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) # Arranging helps you see them side-by-side

view(duplicates_to_check)


length(unique(master_clean$tmk)) # 1412, so we still have all tmks represented after distinct

# N rows = 1683 after distinct, 1741 before. Lost 58
# N tmks = 1412

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #


### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(tmk) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(tmk %in% oahu_dwells_all$tmk),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- oahu_dwells_all %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    total_beds = sum(bedrooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE)

table2b_bedrooms <- oahu_dwells_all %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bedrooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE)

#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(tmk, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(tmk, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)

# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2a_bedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2b_bedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_allcsv"))

```

### Step 3: Compare and Integrate Past Datasets

#### Unchanged

```{r}

# first read in the dwelling data for all kauai

oahu_dwell_alltmks <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/oahu_dwell_clean_alltmks.csv")) %>% 
  mutate(tmk = as.character(tmk))


# check on duplicates in unchanged
oahu_dupes <- oahu_unchanged %>% 
  # Count occurrences per TMK and Class
  group_by(tmk, osds_class) %>% 
  filter(n() > 1) %>% 
  arrange(tmk) %>% 
  select(-fid_1) %>% 
  distinct() #92

oahu_xytmk_dupes <- oahu_unchanged %>% 
  select(-fid_1, -x,-y) %>% 
  distinct() %>% 
  filter(ttl_osds != 0) %>% 
  # Count occurrences per TMK and X and Y
  group_by(tmk) %>%
  filter(n_distinct(x,y) > 1) %>% 
  ungroup() %>% 
  arrange(tmk,x) %>% 
  select(tmk, x, y, everything())

# the only thing that is different about these entires are the lat lon. which could mean multiple buildings, but I am suspicous that there are two buildings each with 10 cesspools, so am leaning towards simply removing the duplciates. check dwell data just to be sure: 

oahu_unch_dupes_dwell <- oahu_dwells_all %>% 
  filter(tmk %in% oahu_dupes$tmk) %>% 
  select(tmk, year_built, card_no, total_rooms,bedrooms) %>% 
  distinct()

# one more way of checking for duplicates

oahu_hidden_dupes <- oahu_unchanged %>%
  # Group by everything EXCEPT the things you want to drop
  group_by(across(-c(fid_1, x, y))) %>%
  # Filter for rows that have "twins"
  filter(n() > 1) %>%
  arrange(tmk)

# encode classes

oahu_unchanged_cl <- oahu_unchanged %>% 
  select(tmk, osds_class, ttl_osds, type, x,y)

oahu_unchanged_multiples <- oahu_unchanged_cl %>% 
  filter(osds_class == "Multiple")

# encode classes
oahu_unchanged_encoded <- oahu_unchanged_cl %>% 
  #filter(if_any(starts_with("class"), ~ .x != 0 & !is.na(.x))) %>% # remove 147 rows
  #select(-x,-y) %>% 
  #distinct() %>%  # remove 6 more rows
  # pivot_longer(cols = starts_with("class"),
  #              names_to = "class",
  #              values_to = "quantity") %>% 
  # filter(quantity != 0) %>% 
  # filter(quantity != -9999) %>% 
  rename(ttl_osds_2010 = ttl_osds) %>%
  mutate(osds_class_otp = case_when(
    osds_class %in% "Class IV" ~ 8, # cesspool 
    osds_class %in% "Class III" ~ 4,
    osds_class %in% "Class II" ~ 2,
    osds_class %in% "Class I" ~ 1,
    osds_class %in% "Multiple"  ~ 9, #multiple, onc is a cesspool.
    TRUE ~ NA_real_
  )) %>%
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "osds_class", 
    values_from = "osds_class_otp",
    values_fn = max # sum or max? thinking max because I think these are multi buildings. 
  ) %>% 
  rename(Class_mult = Multiple) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("Class")), na.rm = TRUE))

# check we didnt loose any tmks - we didn, but some duplicated dropped out

n_distinct(oahu_unchanged_encoded$tmk, na.rm = TRUE) == n_distinct(oahu_unchanged$tmk, na.rm = TRUE)
setdiff(kauai_unchanged_encoded$tmk,kauai_unchanged$tmk)
  

# 2. join to dwells

oahu_unchgd_encoded_dwell <- left_join(oahu_unchanged_encoded, oahu_dwell_alltmks, by = "tmk") %>% 
  distinct()

# 3. Create the three tables 

oahu_card_dwell_summary <-  oahu_unchgd_encoded_dwell %>% 
  group_by(tmk) %>% 
  summarize(
    n_buildings = n_distinct(card_no),# remember this can also be units. 
    in_dwelldat = any(!is.na(tax_year))
  ) %>% 
  ungroup() 

oahu_otp_unchanged_table1 <- oahu_unchgd_encoded_dwell %>%
  mutate(n_permits = case_when(
    encoded_sum %in% c(1,2,4,8) ~ 1,
    encoded_sum %in% c(3,5,6,9,10,12) ~ 2,
    encoded_sum %in% c(7,11,13,14) ~ 3,
    encoded_sum == 15 ~ 4,
    TRUE ~ 99999),
    n_years = 1,
    min_year = 2010,
    max_year = 2010,
    n_addresses = "unknown",
    status_since_otp = "unchanged since otp"
  ) %>%
  left_join(oahu_card_dwell_summary) %>% 
  mutate(
      scenario = case_when(
      n_permits == 1 & n_years == 1 & n_buildings == 1 ~ "Simple",
      n_buildings > 1 ~ "Multi-building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Unknown"
    )) %>%
    select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    scenario,# might not need this
    status_since_otp
  )
  
# ----------- Table 2 ------------------- # 

# no commercial here - removed at the end of this code chunk

oahu_otp_unchanged_table2 <- oahu_unchgd_encoded_dwell %>%
  mutate(bed_rooms = as.numeric(bedrooms)) %>% 
  group_by(tmk, tax_year) %>% 
  summarize(total_beds = sum(bed_rooms, na.rm = TRUE),
            .groups = "drop") %>% 
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  select(-`NA`) %>% 
  filter(!if_all(2:last_col(), is.na)) %>% # remove commercial
  rename_with(~ paste0("bedrooms_", .), matches("^[0-9]"))
  


# --------- Table 3 ---------------- #

oahu_otp_unchanged_table3 <- oahu_unchgd_encoded_dwell %>% 
  mutate(year = 2010) %>% 
  select(tmk,year, encoded_sum) %>% 
  distinct() %>% 
  pivot_wider(id_cols = tmk,names_from = year, values_from = encoded_sum) %>% 
  rename_with(~ paste0("osds_cl_", .), matches("^[0-9]"))
  
  
# ---------- Table 4 -------------- #

# these should include all marked as commercial in the data + any we dont have dwell data for
oahu_otp_unchanged_commxy <- oahu_unchanged %>% 
  select(tmk, x,y)

oahu_otp_unchanged_table4_googleplaces <- oahu_otp_unchanged_table1 %>% 
  filter(in_dwelldat == FALSE) %>% 
  left_join(oahu_unchanged, by = "tmk") %>% 
  select(tmk, x, y, in_dwelldat)

# for oahu we'll just go with this


# ------------- Final checks and save ----------------- #

write_csv(oahu_otp_unchanged_table1,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table2,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2_classes_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table3,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_classes_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table4_googleplaces,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_commercial_UNCHANGED_draft.csv"))



```

#### Updated

```{r updated}

# ------------ gap fill bed rooms --------------- #

table_2_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2a_bedrooms_sum.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename_with(., ~paste0("newbeds",.x), .cols= !tmk)

# oahu doesn't have any info other than xy, and osds class
oahu_updated_forjoin <- oahu_updated %>% 
  select(tmk,x,y) 
 # backfill with pittcode

# are all the updated ones in table 2? if they arenʻt are all those commercial?

table_2_updated <- table_2_new %>% 
  filter(tmk %in% oahu_updated_forjoin$tmk) # missing alott - check to see if those are commercial, or they are new?

updates_notInTable2 <- setdiff(oahu_updated_forjoin$tmk, table_2_updated$tmk) # 114 

# these should go in table 4
table4_withUpdates <- oahu_updated_forjoin %>% 
  filter(tmk %in% updates_notInTable2)
  

# since we don't have old data, we can skip the gapfilling. but, to match the other data, we need to create the bed trend, gapfill flag and years built





# --------- gap fill class ------------ # 

# read in table 3 class 

table_3_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_allcsv")) %>% 
  rename_with(., ~paste0("newclass",.x), .cols= !tmk) %>% 
  mutate(tmk = as.character(tmk))

# join to maui updated, with specific columsn accepted. also do encoding

oahu_updated_withClass_enc <- oahu_updated %>% 
  #filter(if_any(starts_with("class"), ~ .x != 0 & !is.na(.x))) %>% # remove 147 rows
  #select(-x,-y) %>% 
  #distinct() %>%  # remove 6 more rows
  # pivot_longer(cols = starts_with("class"),
  #              names_to = "class",
  #              values_to = "quantity") %>% 
  # filter(quantity != 0) %>% 
  # filter(quantity != -9999) %>% 
  rename(ttl_osds_2010 = ttl_osds) %>%
  mutate(osds_class_otp = case_when(
    osds_class %in% "Class IV" ~ 8, # cesspool 
    osds_class %in% "Class III" ~ 4,
    osds_class %in% "Class II" ~ 2,
    osds_class %in% "Class I" ~ 1,
    osds_class %in% "Multiple"  ~ 9, #multiple, onc is a cesspool.
    TRUE ~ NA_real_
  )) %>%
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "osds_class", 
    values_from = "osds_class_otp",
    values_fn = max # sum or max? thinking max because I think these are multi buildings. 
  ) %>% 
  rename(Class_mult = Multiple) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("Class")), na.rm = TRUE))

oahu_updated_withClassEnc_New <- oahu_updated_withClass_enc %>%
  left_join(table_3_new, by = "tmk")

# get cols to run match on
class_cols <- oahu_updated_withClassEnc_New %>% select(starts_with("new")) %>% names()

table3_updated_classmatch <- oahu_updated_withClassEnc_New %>% 
  mutate(
   # CREATE MATCH INDEX: 
    # This finds the index (1, 2, 3...) of the first column that matches OTP bedrooms
    # We use matrix comparison (df[cols] == df$otp) to keep it vectorized
    match_index = {
      match_matrix <- pick(all_of(class_cols)) == encoded_sum
      # replace NAs with FALSE so max.col doesn't trip
      match_matrix[is.na(match_matrix)] <- FALSE 
      
      # If a row has at least one match, get the index. Otherwise, NA.
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    
    # CREATE MATCH NAME: Convert that index (e.g., 1) into the column name (e.g., "new_2010")
    match_name = class_cols[match_index],
    # COLUMN 1: The Match Year Flag
    # Priority: If it matches, show which one. Otherwise, it's unmatched.
    # match_status = case_when(
    #   !is.na(match_index) ~ paste0("Matches with ", match_name),
    #   TRUE                ~ "Unmatched"
    #),

    # COLUMN 3: Data Presence (The "All Zeros/NAs" check)
    # Checks if ALL columns starting with 'NEW' are either 0 or NA
    data_presence = if_else(
      if_all(starts_with("new"), ~ .x == 0 | is.na(.x)),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )) %>% 
  unite("class_history", starts_with("new"), sep = "|", remove = FALSE)

# separate df to create a flag. you need the iws table 1 registry for this one

table_1_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv")) %>% 
  mutate(tmk = as.character(tmk))

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  dplyr::select(tmk, encoded_sum, min_year, max_year, n_addresses, class_history,
                starts_with("new")) %>% 
  # change this logic: first define siutation in "new", ie one or two upgrades etc. then compare that to OSDS. 
  # 
  mutate(
    legacy_val = as.numeric(encoded_sum),
    v2009 = as.numeric(newclass2009),
    v2010 = as.numeric(newclass2010),
    
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010) ~ "no data to compare",
      TRUE ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = case_when(
    scenario_match_flag == "matched 2009 or 2010" ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE ~ NA_real_
  )) %>% 
  select(tmk, newpre_2009, newclass2009, newclass2010,
         starts_with("newclass201"),starts_with("newclass202"), everything()) %>% 
  # Use rowwise to build the unique story without the split/paste overhead
  rowwise() %>%
  mutate(unique_story = {
    vals <- c_across(starts_with("new"))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>%
  ungroup() %>%
  mutate(
    upgrade_scenario = case_when(
      # High-priority special overrides
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1" ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2" ~ "septic and aerobic - replacement of aerobic",
      
      unique_story == "3|4" ~ "septic combo to aerobic",
      
      unique_story == "1|4" ~ "single septic to aerobic",
      
      unique_story == "7|4" ~ "complex septic system to aerobic",
      
      unique_story == "8|5" ~ "cesspool to septic / aerobic hybrid",
      
      unique_story == "5|1" ~ "septic and aerobic to septic",
      
      unique_story == "7|3" ~ "2 septics and aerobic to just 2 septics - aerobic removal",
      
      unique_story == "7|1|4" ~ "2 septics and aerobic + one septic repair + one aerobic repair",
      
      unique_story == "10|9" ~ "1 septic + cesspool to better septic and cesspool",
      
      unique_story == "1|5" ~ "1 septic with a new aerobic installed",
      
      unique_story == "8|9" ~ "1 cesspool + one other type installed",
      
      
      # Standard Logic
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1234]$") ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$") ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$") ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1234]$") ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$") ~ "multiple osds with new cesspool application",
      
      # Errors and Laterals
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$") ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$") ~ "lateral change (clean to clean)",
      
      TRUE ~ "other scenario"
    )
  )


# split into data for table 3 and a flag for table 1
table_3_new_updated <- update_logic_refined_cl %>% 
  select(tmk, starts_with("new"))

class_flag_forTable1 <- update_logic_refined_cl %>%
    select(tmk,class_match_flag = scenario_match_flag,unique_story,upgrade_scenario)


```

#### Complete Tables 1 - 3a

```{r complete-tables}

# read in table 1 new and unchanged versions, and add gapfill flags

table_1_new <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv")) %>% 
  rename(in_dwelldat = is_dwelling) %>% 
  mutate(status_since_otp = "New Permit",
         n_addresses = as.character(n_addresses),
         tmk = as.character(tmk)) %>% 
  #select(-full_taxid) %>% 
  distinct()

table_1_unchanged <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_UNCHANGED.csv")) %>% 
  mutate(scenario = "No new permits") %>% 
  mutate(tmk = as.character(tmk)) %>% 
  distinct()

# are there duplicates? no

table_1_unchangeddups <- table_1_unchanged %>% filter(tmk %in% table_1_new$tmk)

table_1_complete <- bind_rows(table_1_new, table_1_unchanged)

# ------------ Table 2 --------------------- #

# this one should have all the tmks - commercial tmks 
table_2_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2_classes_UNCHANGED.csv")) %>% 
  mutate(tmk = as.character(tmk))

table_2_updatedforbind <- table_2_updated %>% 
  rename_with(~ str_replace(.x, "newbeds", "bedrooms_"), 
              starts_with("newbeds"))

table_2_new_only <- table_2_new %>% 
  filter(tmk %!in% table_2_updatedforbind$tmk)

table_2_complete <- bind_rows(table_2_unchanged, table_2_updatedforbind, table_2_new_only) %>% 
  distinct()#2000 ish commerical?

year_built <- oahu_dwell_alltmks %>% 
  filter(tmk %in% table_2_complete$tmk) %>% 
  select(tmk, year_built) %>%
  filter(!is.na(year_built)) %>% 
  distinct() %>% 
  group_by(tmk) %>%
  summarize(
    # str_flatten to join them with a comma
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    #count how many years were found
    count_buildyears = n()
  ) %>%
  ungroup() %>% 
  mutate(tmk = as.character(tmk))


table_2_flags <- table_2_complete %>% 
  mutate(
    bed_trend = case_when(
      bedrooms_2010 < bedrooms_2020  ~ "Expansion since OTP",
      bedrooms_2010 > bedrooms_2020  ~ "Shrink - double check",
      bedrooms_2010 == bedrooms_2020 ~ "Perfect Match",
      TRUE                          ~ NA_character_
    ),
  ) %>% 
  left_join(year_built, by = "tmk") %>% 
  select(tmk, bed_trend, year_built_combined, count_buildyears)

table_1_complete_bedflags <- table_1_complete %>% 
  mutate(gapfill_flag = "not gapfilled - no earlier info") %>% 
  left_join(table_2_flags, by = "tmk")

# -------------- Table 3 --------------------- #
  
table_3_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_classes_UNCHANGED.csv")) %>% 
  distinct() %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename(newclass2010 = osds_cl_2010)

table_3_updatedforbind <- table_3_new_updated %>% 
  mutate(tmk = as.character(tmk))

table_3_newonly <- table_3_new %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  distinct()

table_3_complete <- bind_rows(table_3_new_updated, table_3_unchanged, table_3_newonly) %>%   rename_with(~ str_replace(.x, "newclass", "osds_class_"), 
              starts_with("newclass")) 

table_1_withClassflags <- table_1_complete_bedflags %>% 
  left_join(class_flag_forTable1, by = "tmk")


# ------------ Table 4 --------------------- # 

table_4_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_commercial_UNCHANGED_draft.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  select(tmk, x,y)

table_4_googleplaces <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table4_googleplaces_UNCHANGED_draft.csv")) %>% 
  mutate(
    tmk = as.character(tmk),
    commercial = "unknown",
    pitt_code = -9999) %>% 
  select(tmk, commercial,pitt_code)
  
table_4_withUpdates_forbind <- table4_withUpdates %>% 
  select(tmk,x,y)
  
table_4_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_commercial.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk)
  

table_4_complete <- bind_rows(table_4_unchanged,
                              table_4_new,
                              table_4_withUpdates_forbind)



# ----------- Clean up Table 1 ---------------- #

table_1_final <- table_1_withClassflags %>% 
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete$tmk, TRUE,FALSE)
  ) %>% 
  mutate(bed_trend = "not able to discern bed trend") %>% 
  select(
    tmk,
    starts_with("n_"),
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag = gapfill_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4 = use_table4
  )

# ----------- Tmk checks ----------------- #

# tmk checks
# All TMKs in table 1 accounted for table 3? 

all(table_1_final$tmk %in% table_3_complete$tmk)

# All TMKS in table 2 in table 1?
all(table_2_complete$tmk %in% table_1_final$tmk)

# Authoritative list all in table 1?

all(master_tmk_list %in% table_1_final$tmk) # only missing 1, and its 0

shouldbe_intable1 <- setdiff(master_tmk_list, table_1_final$tmk)

# Difference in table 2 and table 1 == table 4?

shouldbein_table_4 <- setdiff(table_1_final$tmk,table_2_complete$tmk) #2710

shouldbe_intable2 <- setdiff(shouldbein_table_4,table_4_complete$tmk) # 0

head(shouldbe_intable2)

all(shouldbe_intable2 %in% table_2_complete$tmk) #TRUE

table_4_complete_cl <- table_4_complete %>% 
  filter(tmk %!in% shouldbe_intable2)


# update use table 4 flag
table_1_final_final <- table_1_final %>%
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete_cl$tmk, TRUE,FALSE)
  )

# Save all! 


write_csv(table_1_final_final,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_ALL.csv"))

write_csv(table_2_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2_bedrooms_ALL.csv"))

write_csv(table_3_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3a_permitsApplied_ALL.csv"))

write_csv(table_4_complete_cl,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_percommercial_draft.csv"))


```

## Step 4: Create table 3b

The additive encoding is what is known as a binary bit mask. Each class represents a power of 2, so every unique sum corresponds to exactly one combination of classes.

We can think of table 3a as an "event" table, since it records the permit application event. However, we want to create a "status" table, which records the actual OSDS present on the property in each year, BASED on the events we know happened. This will be table 3b

We can make multiple versions of this, using different math. In the first, I'll assume that: 1. If there was only one OSDS on the property, the new application is replacing it 2. If there were multiple OSDS on the property, the new application is replacing the grossest, ie highest class. This correspond to the largest power of 2 within the encoded value

This translates to the following math: 1. Identify the largest power of 2 inside the current value. 2. Subtract that power of 2. 3. Add the new application value.

```{r}

# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% 
  rename(osds_cl_2008 = newpre_2009) %>% # note, this is a stand in for "pre 2009"
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(Applied,0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3b_allyears_draft.csv"))



```

## Step 5: Get Data for Non-Dwells, Google places

```{r}


# read in non-dwell data

oahu_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_percommercial_draft.csv"))


coordinates_fortbl4 <- oahu_table4 %>% 
  select(tmk, x,y) %>% 
  filter(!is.na(x) & !is.na(y)) %>% 
  st_as_sf(coords = c("x","y"), crs = 32604)

coordinates_latlon <- st_transform(coordinates_fortbl4, crs = 4326)

mapview(coordinates_latlon)

api_ready_coords <- coordinates_latlon %>%
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1],
    google_coords = paste(lat, lon, sep = ",")
  ) %>%
  st_drop_geometry()

# get addresses - if there are coordinates, then use these rather than address because it will be more precise.

addresses_fortbl4 <- iws_oahu_permits %>% 
  filter(tmk %in% oahu_table4$tmk) %>% 
  filter(tmk %!in% api_ready_coords$tmk) %>% 
  select(tmk, starts_with("project")) %>% 
  mutate(street_house = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
         # Create the full address with a conditional check for the zip code
    address_full = if_else(is.na(project_zip_code) | project_zip_code == "NA",
                           paste(street_house, project_city, sep = ", "),
                           paste(street_house, project_city, project_zip_code, sep = ", ")),
    
    # Final step: convert everything to lower case
    address_full = str_to_lower(address_full)
    )


api_ready_address <- addresses_fortbl4 %>%
  select(tmk, address_full)


## Alternative information to search with: TMK parcel centroids

table1_withspatial <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"))

tmk_for_api <- table1_withspatial %>% 
  st_transform(crs = 4326) %>% 
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1]
  ) %>%
  st_drop_geometry() %>%
  select(tmk, lat, lon) %>% 
  filter(tmk %in% addresses_fortbl4$tmk)


```


Activate Google Places API

```{r}

# See if we have the key

# Load the key from the local file
source(here("config.txt"))

# Set it as an environment variable so packages can find it
Sys.setenv(GOOGLE_PLACES_KEY = GOOGLE_PLACES_KEY)

# Verify (should print your key)
print(Sys.getenv("GOOGLE_PLACES_KEY"))


# Verify it immediately
if(Sys.getenv("GOOGLE_PLACES_KEY") == "") {
  stop("API Key is missing!")
} else {
  print("Success! Key is loaded.")
}

key <- Sys.getenv("GOOGLE_PLACES_KEY")

# Load packages
library(googleway)
library(dplyr)


# --- PART A: SEARCH BY ADDRESS ---
# try one from address dataframe
addr_example <- "66-1009 kaukona hua road, waialua"

res_addr <- google_find_place(
  input = addr_example,
  key = key,
  inputtype = "textquery",
  fields = c("formatted_address", "name", "type")
)

addr_types <- res_addr$candidates$types[[1]]
print(paste("Address is a:", paste(addr_types, collapse = ", ")))

# set up saftey limit

# 1. Initialize the counter at the very top
api_count <- 0
safety_limit <- 200

# 2. Logic for Address Search + Coordinate Extraction
get_detailed_address_info <- function(addr_string) {
  # Safety count
  # Access the counter from the global environment
  api_count <<- api_count + 1 
  
  # THE TOLL BOOTH: Check the limit BEFORE calling Google
  if (api_count > safety_limit) {
    stop(paste("Safety limit of", safety_limit, "reached. Stopping to save your bank account!"))
  }
  
  # We ask for 'geometry' so we get the lat/lon even if the type is boring
  res <- google_find_place(
    input = addr_string, 
    key = key,
    inputtype = "textquery", 
    fields = c("name", "type", "geometry")
  )
  
  if (length(res$candidates) > 0) {
    top_cand <- res$candidates
    main_type <- top_cand$types[[1]][1]
    coords <- top_cand$geometry$location
    
    # Check if we need to dig deeper
    needs_sweep <- main_type %in% boring_types
    
    return(list(
      name = top_cand$name[1], 
      type = main_type, 
      lat = coords$lat[1], 
      lon = coords$lng[1], 
      sweep = needs_sweep
    ))
  }
  return(list(name = NA, type = NA, lat = NA, lon = NA, sweep = FALSE))
}

# 3. The Radius Sweep (Same as before, using the extracted coordinates)
radius_sweep <- function(lat, lon) {
  if (is.na(lat)) return(NA)
  
  # Saftey stop for bank account
  api_count <- api_count + 1
  if (api_count > safety_limit){
    stop("Saftey limit reached during radius sweep")
  }
  
  res <- google_places(location = c(lat, lon), radius = 100, key = key)
  
  if (nrow(res$results) > 0) {
    # Look for the first thing that ISN'T boring
    interesting_finds <- res$results %>%
      unnest(cols = c(types)) %>%
      filter(!types %in% boring_types)
    
    if (nrow(interesting_finds) > 0) {
      return(paste0("Part of: ", interesting_finds$name[1], " (", interesting_finds$types[1], ")"))
    }
  }
  return("No larger landmark found")
}

# 4. Execute on address df

results_addresses <- addresses_fortbl4 %>%
  head(5) %>% # Test batch
  rowwise() %>%
  mutate(
    # Step 1: Initial Address Lookup
    init = list(get_detailed_address_info(address_full)),
    found_name = init$name,
    found_type = init$type,
    temp_lat = init$lat,
    temp_lon = init$lon,
    needs_sweep = init$sweep
  )

results_sweep <- results_addresses %>% 
  mutate(
    # Step 2: If address was 'premise', look around that coordinate
    final_insight = if_else(needs_sweep, 
                            radius_sweep(temp_lat, temp_lon), 
                            found_type)
  ) %>%
  select(-init) # Clean up the list column


# --------------- PART B: Search by Coordinates ----------------------
# Workflow: 1) if it has a specific place, get it. if it doesn't flag for radius search

boring_types <- c("premise", "establishment", "point_of_interest", "political", "floor", "locality")

# 2. Logic for Step 1: Specific Check
get_initial_info_coords <- function(lat, lon) {
  # Safety count
  api_count <<- api_count + 1 
  
  # THE TOLL BOOTH: Check the limit BEFORE calling Google
  if (api_count > safety_limit) {
    stop(paste("Safety limit of", safety_limit, "reached. Stopping to save your bank account!"))
  }
  # Perform the search
  res <- google_places(location = c(lat, lon), radius = 10, key = key)
  
  # Check if 'results' exists and has rows safely
  if (!is.null(res$results) && length(res$results) > 0 && nrow(res$results) > 0) {
    top_place <- res$results[1, ]
    
    # Safely extract the first type
    main_type <- if (!is.null(top_place$types[[1]])) top_place$types[[1]][1] else "unknown"
    
    # Check if we need a radius sweep (boring_types defined earlier)
    needs_sweep <- main_type %in% boring_types
    
    return(list(name = top_place$name, type = main_type, sweep = needs_sweep))
  }
  
  # Fallback if nothing is found at all
  return(list(name = "No result", type = "unknown", sweep = TRUE))
}
# 3. Logic for radius sweep

radius_sweep_coords <- function(lat, lon) {
  # We look up to 100 meters away for a "landmark" type
  res <- google_places(location = c(lat, lon), radius = 100, key = key)
  
  if (nrow(res$results) > 0) {
    # Flatten all types found in the area
    all_nearby <- res$results %>%
      select(name, types) %>%
      tidyr::unnest(types) %>%
      filter(!types %in% boring_types) # Keep only the interesting ones
    
    if (nrow(all_nearby) > 0) {
      # Return the name of the most prominent interesting thing
      return(paste0("Part of: ", all_nearby$name[1], " (", all_nearby$types[1], ")"))
    }
  }
  return("No landmark found")
}

# --- 4. EXECUTION ---
# This assumes your df has 'lat' and 'lon'

safe_sweep <- possibly(radius_sweep_coords, otherwise = "Sweep Failed")

processed_data <- api_ready_coords %>%
  head(5) %>% # Test with 5 first
  rowwise() %>%
  mutate(
    init = list(get_initial_info_coords(lat, lon)),
    place_name = init$name,
    primary_type = init$type,
    needs_sweep = init$sweep
  ) %>%
  mutate(
    final_category = if_else(needs_sweep, 
                             safe_sweep(lat, lon), 
                             primary_type)
  )

```

