---
title: "OSDS Points Explore"
author: "Madeline Berger"
date: "2025-03-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}


library(sf)
library(terra)
library(here)
library(janitor)
library(tidyverse)

library(gt)
library("webshot")
#install.packages("webshot")
webshot::install_phantomjs()  # Needed for saving as PNG

source(here("workflow/setup.R"))

```

# Summary

This markdown is to explore and re-form the data products that Julia
sent over. The goal is a database that we can query for each year,
capturing difference in wastewater pollution that might result in
upgrades, new builds etc.

From our team meeting in May 2025:

Objective is to isolate / determine what is driving any changes we are
seeing between time-stamps in the coastal impact rasters (answer for the
“why”)

Reasons one might expect there is a difference:

-   TMK / point no longer exists (house demolished)

-   TMK / point is a new build (house built or replaced with something
    larger)

-   Cesspool was upgraded to septic or higher OSDS was eliminated and
    hooked up to central system

Other reasons we know there might be a difference:

-   We changed how we quantify impact

![Drawing from Brainstorm with Mary, May
2025](images/whiteboard_drawing.png){width="376"}


# Step 1: Determine starting point


## Examine Raw Data

Julia shared the OSDS permits as well as the Dwelling data, but she did not share the TMK data. So I'll have to get that myself. I also had to manually re-save the data she sent because some of it was in xlsx format, and some of it was in txt formats. I saved all to csvs before reading into R.

Issues to look out for:
- different TMK styles?
- dates: where are the pre-2009 data points?



```{r}

# read in raw data
permit_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))
spec(permit_data)

dwell_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))

# after looking at the mess of the data here is some code to try and read it in nicely

# Load necessary package
library(readr)

# Read in the tab-delimited data for Hawaii
hi_df <- read_delim(here("data/osds/shared-data/Dwellings/Hawaii_DWELDAT2023.TXT"),
                    delim = "\t",
                    col_names = FALSE, na = c("", "NA"),
                    col_types = cols(
                      X1 = col_character()
                    ))

# Assign useful names to selected columns based best guess. Its not confirmed though and the raw data would be better. 

hi_df_selected <- hi_df %>%
  transmute(
    tax_map_key = X1,
    tax_year = X3,
    dwelling_id = X4,
    style = X13,
    full_baths = X20,
    half_baths = X21,
    bedrooms = X22,
    total_rooms = X19,# this might not be right
    year_built = X23,
    occupancy = X25,
    #sqft = X30
  )

# View the result
head(hi_df_selected)


```

Conclusions so far: this is going to require more data from Julia because what she sent does not have enough information for me to feel confident about the column names. 

It would be really helpful to have year built as well as type of building / occupancy to help us figure out those "grey areas". so does the original data that matches the metadata exactly still exist somewhere? 

## Examine Intermediate Products

```{r}

int_gdb_path <- here("data/osds/shared-data/Intermediate_Products.gdb")

int_layers <- st_layers(int_gdb_path)

hi_int <- st_read(file.path(int_gdb_path), layer= "HAW_2023_update_backup_06_06_24")

hi_int_clean <- hi_int %>% 
  

```





Old code

```{r}
gdb_path <- here("data/osds/FINAL_OSDS_Products.gdb")

layers <- st_layers(gdb_path)

oahu_osds_23 <- st_read(file.path(gdb_path), layer = "Oahu_OSDS_2023") %>% 
  clean_names() %>% 
  mutate(class_desc = case_when(
    osds_class == 1 ~ "1: Septic - Tank, Bed, or Trench",
    osds_class == 2 ~ "2: Septic - Seepage Pit (or unknown)",
    osds_class == 3 ~ "3: Compost Toilet, Grey Water System, or ATU Pit",
    osds_class == 4 ~ "4: Cesspool",
    osds_class == 5 ~ "5: Multiple types",
    TRUE ~ "unknown"
  ))


total_pre2008 <- sum(oahu_osds_23$osds_qty[oahu_osds_23$year == "pre 2009"], na.rm = TRUE) #13175


oahu_osds_2009_2023 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year != "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()

total <- sum(oahu_osds_2009_2023$osds_qty)

oahu_osds_agg <- oahu_osds_2009_2023 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(year, class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  )
  
oahu_osds_table <- oahu_osds_agg %>%
  pivot_wider(id_cols = "class_desc", names_from = "year", values_from = "total") %>% 
  adorn_totals(where = c("row","col"))


table_gt <- gt(oahu_osds_table)
gtsave(table_gt, "table_output.png")

# number by year, oahu
year_bar <- ggplot(oahu_osds_agg, aes(x = year, y = total, group = class_desc))+
  geom_bar(stat = "identity", position = "stack",aes(fill = class_desc))+
  theme_bw()

year_bar



# older 

oahu_osds_pre2009 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year == "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()


oahu_osds_old_agg <- oahu_osds_pre2009 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  ) %>% 
  adorn_totals(where = "row")
  
old_table_gt <- gt(oahu_osds_old_agg)

gtsave(old_table_gt, "old_table_output.png")
```
