---
title: "OSDS Points Explore"
author: "Madeline Berger"
date: "2025-03-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}


library(sf)
library(terra)
library(here)
library(readxl)
library(janitor)
library(tidyverse)


library(gt)
#library("webshot")
#install.packages("webshot")
#webshot::install_phantomjs()  # Needed for saving as PNG

source(here("workflow/setup.R"))

```

# Summary

This markdown is to explore and re-form the data products that Julia
sent over. The goal is a database that we can query for each year,
capturing difference in wastewater pollution that might result in
upgrades, new builds etc.

From our team meeting in May 2025:

Objective is to isolate / determine what is driving any changes we are
seeing between time-stamps in the coastal impact rasters (answer for the
“why”)

Reasons one might expect there is a difference:

-   TMK / point no longer exists (house demolished)

-   TMK / point is a new build (house built or replaced with something
    larger)

-   Cesspool was upgraded to septic or higher OSDS was eliminated and
    hooked up to central system

Other reasons we know there might be a difference:

-   We changed how we quantify impact

![Drawing from Brainstorm with Mary, May
2025](images/whiteboard_drawing.png){width="376"}

# Step 1: Data Prep and Cleaning

## Clean Raw Dwell Data

Julia shared the OSDS permits as well as the Dwelling data, but she did
not share the TMK data. So I'll have to get that myself. I also had to
manually re-save the data she sent because some of it was in xlsx
format, and some of it was in txt formats. I saved all to csvs before
reading into R.

Issues to look out for: - different TMK styles? - dates: where are the
pre-2009 data points?

```{r}


# the hawaii dwell data is tab delimited but I think that is a result of some kind of transformation that occurrred. 

# read in raw data
permit_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))
spec(permit_data)

# after looking at the mess of the data here is some code to try and read it in nicely

# Load necessary package
library(readr)



# oahu data has the fixed width preserved, which I counted by hand and confirmed at least the first bunch columns match. The last few are also easy to tell which columns so below I assign them manually.

# hi data was a mess, but the most promising outcome was to try and replace the tabs with spaces (8 seemed like the best bet), and then read it in using the same column specs from the metatdata


# Define widths and optionally names
col_specs <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92, 112, 132, 152, 154,
    179, 186, 193, 199, 201, 202, 206, 211, 216, 218, 238, 239,
    259, 279, 299, 319, 343, 355, 357, 361, 364, 384, 404, 434,
    438, 443, 444, 448, 470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 18, 23, 43, 44, 47, 71, 91, 111, 131, 151, 153, 173,
    180, 187, 194, 200, 201, 201, 207, 212, 217, 237, 238, 258,
    278, 298, 318, 342, 354, 356, 360, 363, 383, 403, 433, 437,
    442, 443, 447, 469, 478, 489, 494, 502, 509, 515  # <-- adjust last few by hand
  )
)



# Read the fixed-width file
oahu_raw_dwell <- read_fwf(
  here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"),
  col_positions = col_specs,   # or actual names from metadata
  na = c("", "NA"),
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))

# remame columns based on visual examination


oahu_dwell_rename <- oahu_raw_dwell %>% 
  select(jurisid = X1,
         taxid = X2,
         condo_dwell = X6,
         full_baths = X15,
         half_baths = X16,
         total_rooms = X21,
         bedrooms = X22,
         year_built = X42) %>% 
  mutate(year_built = substr(year_built,1,4))

# fix three that were off

oahu_fix <- oahu_raw_dwell %>% 
  filter(X2 %in% c("920060090000","130110480001","130110480003")) %>% 
  select(jurisid = X1,
         taxid = X2,
         condo_dwell = X6,
         full_baths = X15,
         half_baths = X16,
         total_rooms = X21,
         bedrooms = X22,
         year_built = X42) %>% 
  mutate(year_built = substr(year_built,5,8)
  )


# re join to full data set and save

oahu_dwell_clean <- oahu_dwell_rename %>% 
  filter(taxid %!in% c("920060090000","130110480001","130110480003")) %>% 
  rbind(oahu_fix)


st_write(oahu_dwell_clean,
         here("data/osds/generated-data/oahu_dwelldat_clean.csv"))



############ hawaii data, expanded version #####################
# note to create this I ran the following code in my terminal : expand -t 8 Hawaii_DWELDAT2023.TXT > hi_expanded.txt


# hi_df <- read_fwf(
#   file = here("data/osds/shared-data/Dwellings/hi_expanded.txt"),
#   col_positions = col_specs,  # from your metadata start/end bytes
#   col_types = cols(.default = col_character())
# )
# 
# head(hi_df,5)


# 1) Read as TAB-delimited (keep everything as text)
df_raw <- read_delim(here("data/osds/shared-data/Dwellings/Hawaii_DWELDAT2023.TXT"),
                    delim = "\t",
                    col_names = FALSE, 
                    na = c("", "NA"),
                    col_types = cols(.default = col_character())
) %>% mutate(row_id = row_number(), .before = 1)

# ok this is better
hi_since_2020 <- df_raw %>% 
  mutate(year_built = as.numeric(X23)) %>% 
  filter(X23 > 2019)

# which are nas?

hi_since_2020na <- hi_since_2020 %>% filter(is.na(year_built)) # we dont want anyway

# filter out the nas

hi_since_2020 <- hi_since_2020 %>% 
  filter(!is.na(year_built))

issues_summ <- hi_since_2020 %>%
  summarise(
    n_na_X5 = sum(is.na(X5)), 
    unique_X5 = list(unique(na.omit(X5))),
    n_na_X12 = sum(is.na(X12)),
    unique_X12 = list(unique(na.omit(X12)))
  )


suspect_rows <- hi_since_2020 %>%
  mutate(
    X5_has_space  = grepl(" ", X5),
    X12_has_space = grepl(" ", X12)
  ) %>%
  filter(X5_has_space == T | X12_has_space ==T)
  
  
suspect_row_ids <- suspect_rows %>% pull(row_id)

# now read those in raw to see what they should be. see if you can work with this one
raw_lines <- read_lines(here("data/osds/shared-data/Dwellings/hi_expanded.txt"))[suspect_row_ids]
raw_lines

# rename columsn to our best guess so we can get on with our lives

hi_since_2020_cl <- hi_since_2020 %>% 
  select(
    row_id,
    taxid = X1,
    full_baths = X15,
    half_baths = X16,
    total_rooms = X17,# this was x21 for oahu
    bedrooms = X18, # this was x22 for oahu
    year_built = X23
  )


st_write(hi_since_2020_cl,
         here("data/osds/generated-data/hi_2023_dwell_clean.csv"))

```

Sending these to Julia to make sure they line up. 

Ok at this point I am just going to pick ones to go with:

x16 = halfbath
x15 = full bath
x17 = total rooms
x18 = total beddroms


Look at islands that are not separated like this:

```{r}


maui_dwell_raw <- read_xlsx(here("data/osds/shared-data/Dwellings/Maui_Dwelling_2015_2022.xlsx"))

maui_dwell_cl <- maui_dwell_raw %>% 
  select(
    parid = PARID, # need to add 2 to this number
    year_built = `Year Built`,
    full_baths = `Full Baths`,
    half_baths = `Half Baths`,
    total_rooms = `Total Rooms`,
    bedrooms = `Bed Rooms`
  ) %>% 
  mutate(taxid = paste0(2,parid)) %>% 
  mutate(across(3:6, ~as.numeric(.x))) %>% 
  select(-parid) %>% 
  mutate(island = "Maui")


kauai_dwell <- read_xlsx(here("data/osds/shared-data/Dwellings/Kauai_BedBathCount_2014_2020_081423.xlsx"))

kauai_dwell_cl <- kauai_dwell %>% 
  select(
    parid = PARID, # need to add 2 to this number
    full_baths = `Full Baths`,
    half_baths = `Half Baths`,
    bedrooms = `Bed Rooms`
  ) %>% 
  mutate(taxid = paste0(4,parid)) %>% 
  mutate(across(2:4, ~as.numeric(.x))) %>% 
  mutate(total_rooms = NA,
         year_built = NA,
         island = "Kauai") %>% 
  select(-parid)

oahu_dwell_clean <- read_csv(here("data/osds/generated-data/oahu_dwelldat_clean.csv")) %>% 
  mutate(taxid = paste0(jurisid,taxid)) %>% 
  mutate(across(4:7, ~as.numeric(.x))) %>% 
  select(-jurisid, -condo_dwell) %>% 
  mutate(island = "Oahu")


# merge with Oahu because at this point, all actions applied to each of these datasets will be the same. Hi remain separate 
# so we can look at WH work. 

oahu_maui_kaui_dwell_join <- rbind(oahu_dwell_clean, maui_dwell_cl, kauai_dwell_cl)


st_write(oahu_maui_kaui_dwell_join,
         here("data/osds/generated-data/oahu_maui_kaui_dwell_merged.csv"))
```

Questions / thoughts from Maui and Kauai Raw: 
Maui looks pretty good
It would be really helpful to have year built as well as type of building / occupancy to help us
figure out those "grey areas". so does the original data that matches
the metadata exactly still exist somewhere? Or did this come this way? 
Especially for Kauai, because we dont have total rooms OR year built. 



# Step 2: Join Clean Datasets to most recent TMK points and record differences

```{r}

# read in tmk points from HI state GIS

# join to hawaii


# 





```























## Examine Intermediate Products


### Hawaii

```{r}

# hawaii - only one with just one file
int_gdb_path <- here("data/osds/shared-data/Intermediate_Products.gdb")

int_layers <- st_layers(int_gdb_path)

hi_int <- st_read(file.path(int_gdb_path), layer= "HAW_2023_update_backup_06_06_24")

# what are the three datasets? 

hi_pre2010 <- hi_int %>% filter(DATA_SET == "2010&Prior") #53530
hi_post2010 <- hi_int %>% filter(DATA_SET == "Post 2010") #6093
hi_na <- hi_int %>% filter(is.na(DATA_SET)) #4069 


# how many duplicate TMKS? do all of these have different dates?

hi_dup_tmks <- hi_int %>% group_by(TMK) %>% 
  filter(n()>1) 

# remove all information related to OSDS except date, bed and bath and reshape
# clean up dates so we just have a year column for each one

hi_dup_tmks_cl <- hi_dup_tmks %>% 
  select(TMK,OSDS_TYPE,OSDS_QTY,COMMERCIAL,B_ROOMS,BATHS,EFF_DATE,DATE_BASIS,DATA_SET) %>% 
  mutate(date_clean = as_date(EFF_DATE)) %>% 
  mutate(year = year(date_clean)) %>% 
  st_drop_geometry()


# osds type by year

hi_dup_tmks_osds_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = OSDS_TYPE)


# osds qty by year

hi_dup_tmks_qty_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = OSDS_QTY)


# bedrooms by year

hi_dup_tmks_bedrooms_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = B_ROOMS)

# bathrooms by year

hi_dup_tmks_bathrooms_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = BATHS)


```

Overall Hawaii seems pretty "clean" I guess. My main questions are:
- Why would a TMK have -999 for bedrooms and an actual number for baths? I think this might be from a weird join prior. But might be good to check if that's coming from the dwell data


### Maui 
```{r}

st_layers(int_gdb_path)

# these look exactly the same so not sure, maybe one just has Julia's update columns and the other does not?
maui_iws1 <- st_read(file.path(int_gdb_path), layer= "MauiNui_2023_IWS_update") #2024

maui_iws2 <- st_read(file.path(int_gdb_path), layer= "Mau_IWS_31_01_2010_to_03_15_2023") #2388

# do these have the same data? what are the extra 364?

maui_in1_butnot2 <- setdiff(maui_iws1$tmk, maui_iws2$tmk) 




```

### Kauai

```{r}

kauai_iws1 <- st_read(file.path(int_gdb_path), layer= "KAU_IWS_2023_update")
 
kauai_iws2 <- st_read(file.path(int_gdb_path), layer= "On_site_Sewage_Disposal_Systems_Kauai_nutrient_update") #



```


### Oahu

```{r}


oahu_points_with_dwell <- st_read(file.path(int_gdb_path), layer= "Oahu_address_Points_Including_Dwellings_IWS") #11
 
oahu_iws2 <- st_read(file.path(int_gdb_path), layer= "OAH_OSDS_2019_V2_merge_missing_points_10_01_24")



```

So these are confusing to me. 




Old code

```{r}
gdb_path <- here("data/osds/FINAL_OSDS_Products.gdb")

layers <- st_layers(gdb_path)

oahu_osds_23 <- st_read(file.path(gdb_path), layer = "Oahu_OSDS_2023") %>% 
  clean_names() %>% 
  mutate(class_desc = case_when(
    osds_class == 1 ~ "1: Septic - Tank, Bed, or Trench",
    osds_class == 2 ~ "2: Septic - Seepage Pit (or unknown)",
    osds_class == 3 ~ "3: Compost Toilet, Grey Water System, or ATU Pit",
    osds_class == 4 ~ "4: Cesspool",
    osds_class == 5 ~ "5: Multiple types",
    TRUE ~ "unknown"
  ))


total_pre2008 <- sum(oahu_osds_23$osds_qty[oahu_osds_23$year == "pre 2009"], na.rm = TRUE) #13175


oahu_osds_2009_2023 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year != "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()

total <- sum(oahu_osds_2009_2023$osds_qty)

oahu_osds_agg <- oahu_osds_2009_2023 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(year, class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  )
  
oahu_osds_table <- oahu_osds_agg %>%
  pivot_wider(id_cols = "class_desc", names_from = "year", values_from = "total") %>% 
  adorn_totals(where = c("row","col"))


table_gt <- gt(oahu_osds_table)
gtsave(table_gt, "table_output.png")

# number by year, oahu
year_bar <- ggplot(oahu_osds_agg, aes(x = year, y = total, group = class_desc))+
  geom_bar(stat = "identity", position = "stack",aes(fill = class_desc))+
  theme_bw()

year_bar



# older 

oahu_osds_pre2009 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year == "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()


oahu_osds_old_agg <- oahu_osds_pre2009 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  ) %>% 
  adorn_totals(where = "row")
  
old_table_gt <- gt(oahu_osds_old_agg)

gtsave(old_table_gt, "old_table_output.png")
```
