---
title: "OSDS Points Explore"
author: "Madeline Berger"
date: "2025-03-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}


library(sf)
library(terra)
library(here)
library(readxl)
library(janitor)
library(tidyverse)


library(gt)
#library("webshot")
#install.packages("webshot")
#webshot::install_phantomjs()  # Needed for saving as PNG

source(here("workflow/setup.R"))

```

# Summary

This markdown is to explore and re-form the data products that Julia
sent over. The goal is a database that we can query for each year,
capturing difference in wastewater pollution that might result in
upgrades, new builds etc.

From our team meeting in May 2025:

Objective is to isolate / determine what is driving any changes we are
seeing between time-stamps in the coastal impact rasters (answer for the
“why”)

Reasons one might expect there is a difference:

-   TMK / point no longer exists (house demolished)

-   TMK / point is a new build (house built or replaced with something
    larger)

-   Cesspool was upgraded to septic or higher OSDS was eliminated and
    hooked up to central system

Other reasons we know there might be a difference:

-   We changed how we quantify impact

![Drawing from Brainstorm with Mary, May
2025](images/whiteboard_drawing.png){width="376"}

# Step 1: Determine starting point

## Examine Raw Data

Julia shared the OSDS permits as well as the Dwelling data, but she did
not share the TMK data. So I'll have to get that myself. I also had to
manually re-save the data she sent because some of it was in xlsx
format, and some of it was in txt formats. I saved all to csvs before
reading into R.

Issues to look out for: - different TMK styles? - dates: where are the
pre-2009 data points?

```{r}

# read in raw data
permit_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))
spec(permit_data)

dwell_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))

# after looking at the mess of the data here is some code to try and read it in nicely

# Load necessary package
library(readr)

# Read in the tab-delimited data for Hawaii
hi_df <- read_delim(here("data/osds/shared-data/Dwellings/Hawaii_DWELDAT2023.TXT"),
                    delim = "\t",
                    col_names = FALSE, na = c("", "NA"),
                    col_types = cols(
                      X1 = col_character()
                    ))

# Assign useful names to selected columns based best guess. Its not confirmed though and the raw data would be better. 

hi_df_selected <- hi_df %>%
  transmute(
    tax_map_key = X1,
    tax_year = X3,
    dwelling_id = X4,
    style = X13,
    full_baths = X20,
    half_baths = X21,
    bedrooms = X22,
    total_rooms = X19,# this might not be right
    year_built = X23,
    occupancy = X25,
    #sqft = X30 # this was definitely not right
  )

# View the result
head(hi_df_selected)

# oahu data used fixed width

library(readr)

# Define widths and optionally names
widths <- c(1, 1, 4, 4, 20, 1, 4, 20, 20, 20, 20, 20, 2, 20, 2, 2, 2, 2, 1, 1,
            2, 2, 2, 20, 1, 20, 20, 20, 20, 20, 6, 2, 4, 3, 20, 20, 30, 4, 4, 1,
            4, 20, 6, 3, 3, 6, 6, 6)

names <- c("JUR", "PARID", "TAXYR", "CARD", "USER20", "VALMETH", "STORIES", "USER6", "USER7", "USER8",
           "USER9", "USER10", "STYLE", "USER11", "EXTWALL", "ROOFMAT", "INTWALL", "FLOORTYP", "HEAT", "AIRC",
           "RMTOT", "BEDROOM", "FIXBATH", "FIXHALF", "FPLACE", "USER12", "USER13", "USER14", "USER15", "USER16",
           "USER17", "USER18", "YRBUILT", "EFFYR", "USER19", "USER21", "USER22", "CONDITION", "GRADE", "OCCUPANCY",
           "VALMETH2", "USER23", "SQFT_LIV", "SQFT_BSMT", "SQFT_ADJ", "USER24", "USER25", "USER26")

# Read the fixed-width file?
df <- read_fwf(here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"), fwf_widths(widths, col_names = names), col_types = cols(.default = "c"))

# This didnt work
#df_2 <- read_delim(here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"), delim = "\t", col_names = FALSE)

# This didnt work
# df <- read_delim(here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"),
#                  delim = " ",         # space delimiter
#                  col_names = FALSE,
#                  trim_ws = TRUE,
#                  skip_empty_rows = TRUE,
#                  progress = FALSE)


lines <- read_lines(here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"))
split_data <- str_split_fixed(lines, pattern = " {2,}", n = 48)

df <- as.data.frame(split_data, stringsAsFactors = FALSE)

colnames(df) <- c(
  "JUR", "PARID", "TAXYR", "CARD", "USER20", "VALMETH", "STORIES", "USER6", "USER7", "USER8",
  "USER9", "USER10", "STYLE", "USER11", "EXTWALL", "ROOFMAT", "INTWALL", "FLOORTYP", "HEAT", "AIRC",
  "RMTOT", "BEDROOM", "FIXBATH", "FIXHALF", "FPLACE", "USER12", "USER13", "USER14", "USER15", "USER16",
  "USER17", "USER18", "YRBUILT", "EFFYR", "USER19", "USER21", "USER22", "CONDITION", "GRADE", "OCCUPANCY",
  "VALMETH2", "USER23", "SQFT_LIV", "SQFT_BSMT", "SQFT_ADJ", "USER24", "USER25", "USER26"
)

```

Conclusions so far: this is going to require more data from Julia
because what she sent does not have enough information for me to feel
confident about the column names.

Look at islands that are not separated like this:

```{r}


maui_dwell_raw <- read_xlsx(here("data/osds/shared-data/Dwellings/Maui_Dwelling_2015_2022.xlsx"))


kauai_dwell <- read_xlsx(here("data/osds/shared-data/Dwellings/Kauai_BedBathCount_2014_2020_081423.xlsx"))



```

Questions thoughts from Maui and Kauai Raw: It would be really helpful
to have year built as well as type of building / occupancy to help us
figure out those "grey areas". so does the original data that matches
the metadata exactly still exist somewhere? A bit confused by the TMKs
for Maui, seem to all start with different numbers yet are the same length as kauai?

## Examine Intermediate Products


### Hawaii

```{r}

# hawaii - only one with just one file
int_gdb_path <- here("data/osds/shared-data/Intermediate_Products.gdb")

int_layers <- st_layers(int_gdb_path)

hi_int <- st_read(file.path(int_gdb_path), layer= "HAW_2023_update_backup_06_06_24")

# what are the three datasets? 

hi_pre2010 <- hi_int %>% filter(DATA_SET == "2010&Prior") #53530
hi_post2010 <- hi_int %>% filter(DATA_SET == "Post 2010") #6093
hi_na <- hi_int %>% filter(is.na(DATA_SET)) #4069 


# how many duplicate TMKS? do all of these have different dates?

hi_dup_tmks <- hi_int %>% group_by(TMK) %>% 
  filter(n()>1) 

# remove all information related to OSDS except date, bed and bath and reshape
# clean up dates so we just have a year column for each one

hi_dup_tmks_cl <- hi_dup_tmks %>% 
  select(TMK,OSDS_TYPE,OSDS_QTY,COMMERCIAL,B_ROOMS,BATHS,EFF_DATE,DATE_BASIS,DATA_SET) %>% 
  mutate(date_clean = as_date(EFF_DATE)) %>% 
  mutate(year = year(date_clean)) %>% 
  st_drop_geometry()


# osds type by year

hi_dup_tmks_osds_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = OSDS_TYPE)


# osds qty by year

hi_dup_tmks_qty_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = OSDS_QTY)


# bedrooms by year

hi_dup_tmks_bedrooms_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = B_ROOMS)

# bathrooms by year

hi_dup_tmks_bathrooms_wd <- hi_dup_tmks_cl %>% 
  arrange(year) %>% 
  pivot_wider(id_cols = c(TMK,COMMERCIAL), names_from = year, values_from = BATHS)


```

Overall Hawaii seems pretty "clean" I guess. My main questions are:
- Why would a TMK have -999 for bedrooms and an actual number for baths? I think this might be from a weird join prior. But might be good to check if that's coming from the dwell data


### Maui 
```{r}

st_layers(int_gdb_path)

# these look exactly the same so not sure, maybe one just has Julia's update columns and the other does not?
maui_iws1 <- st_read(file.path(int_gdb_path), layer= "MauiNui_2023_IWS_update") #2024

maui_iws2 <- st_read(file.path(int_gdb_path), layer= "Mau_IWS_31_01_2010_to_03_15_2023") #2388

# do these have the same data? what are the extra 364?

maui_in1_butnot2 <- setdiff(maui_iws1$tmk, maui_iws2$tmk) 




```

### Kauai

```{r}

kauai_iws1 <- st_read(file.path(int_gdb_path), layer= "KAU_IWS_2023_update")
 
kauai_iws2 <- st_read(file.path(int_gdb_path), layer= "On_site_Sewage_Disposal_Systems_Kauai_nutrient_update") #



```


### Oahu

```{r}


oahu_points_with_dwell <- st_read(file.path(int_gdb_path), layer= "Oahu_address_Points_Including_Dwellings_IWS") #11
 
oahu_iws2 <- st_read(file.path(int_gdb_path), layer= "OAH_OSDS_2019_V2_merge_missing_points_10_01_24")



```

So these are confusing to me. 




Old code

```{r}
gdb_path <- here("data/osds/FINAL_OSDS_Products.gdb")

layers <- st_layers(gdb_path)

oahu_osds_23 <- st_read(file.path(gdb_path), layer = "Oahu_OSDS_2023") %>% 
  clean_names() %>% 
  mutate(class_desc = case_when(
    osds_class == 1 ~ "1: Septic - Tank, Bed, or Trench",
    osds_class == 2 ~ "2: Septic - Seepage Pit (or unknown)",
    osds_class == 3 ~ "3: Compost Toilet, Grey Water System, or ATU Pit",
    osds_class == 4 ~ "4: Cesspool",
    osds_class == 5 ~ "5: Multiple types",
    TRUE ~ "unknown"
  ))


total_pre2008 <- sum(oahu_osds_23$osds_qty[oahu_osds_23$year == "pre 2009"], na.rm = TRUE) #13175


oahu_osds_2009_2023 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year != "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()

total <- sum(oahu_osds_2009_2023$osds_qty)

oahu_osds_agg <- oahu_osds_2009_2023 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(year, class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  )
  
oahu_osds_table <- oahu_osds_agg %>%
  pivot_wider(id_cols = "class_desc", names_from = "year", values_from = "total") %>% 
  adorn_totals(where = c("row","col"))


table_gt <- gt(oahu_osds_table)
gtsave(table_gt, "table_output.png")

# number by year, oahu
year_bar <- ggplot(oahu_osds_agg, aes(x = year, y = total, group = class_desc))+
  geom_bar(stat = "identity", position = "stack",aes(fill = class_desc))+
  theme_bw()

year_bar



# older 

oahu_osds_pre2009 <- oahu_osds_23 %>% 
  #clean_names() %>% 
  filter(year == "pre 2009") %>% 
  mutate(osds_class = as.integer(osds_class)) %>% 
  #mutate(year = ifelse(year == "pre 2009",2008,year)) %>% 
  mutate(year = as.Date(year, format = "%Y")) %>% 
  mutate(year = year(year)) %>% 
  rowid_to_column()


oahu_osds_old_agg <- oahu_osds_pre2009 %>% 
  as.data.frame() %>% 
  select(-starts_with("point")) %>% 
  group_by(class_desc) %>% 
  summarize(
    total = sum(osds_qty)
  ) %>% 
  adorn_totals(where = "row")
  
old_table_gt <- gt(oahu_osds_old_agg)

gtsave(old_table_gt, "old_table_output.png")
```
