---
title: "OSDS Points Explore"
author: "Madeline Berger"
date: "2025-03-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}


library(sf)
library(terra)
library(here)
library(readxl)
library(janitor)
library(tidyverse)


#library(gt)
#library("webshot")
#install.packages("webshot")
#webshot::install_phantomjs()  # Needed for saving as PNG

source(here("workflow/setup.R"))

```

# Summary

This markdown is to explore and re-form the data products that Julia
sent over. The goal is a database that we can query for each year,
capturing difference in wastewater pollution that might result in
upgrades, new builds etc.

From our team meeting in May 2025:

Objective is to isolate / determine what is driving any changes we are
seeing between time-stamps in the coastal impact rasters (answer for the
“why”)

Reasons one might expect there is a difference:

-   TMK / point no longer exists (house demolished)

-   TMK / point is a new build (house built or replaced with something
    larger)

-   Cesspool was upgraded to septic or higher OSDS was eliminated and
    hooked up to central system

Other reasons we know there might be a difference:

-   We changed how we quantify impact

![Drawing from Brainstorm with Mary, May
2025](images/whiteboard_drawing.png){width="376"}

Each of the phases below corresponds to the workflow chart in the Hawaii Driver Data Google Drive. To reduce environment clutter, phases stand alone, i.e. you can start at a phase and all dependencies should be loaded in the contained code chunks. 

# Phase 1: Data Prep and Cleaning

## Clean Raw Dwell Data

Julia shared the OSDS permits as well as the Dwelling data, but she did
not share the TMK data. So I'll have to get that myself. I also had to
manually re-save the data she sent because some of it was in xlsx
format, and some of it was in txt formats. I saved all to csvs before
reading into R.

Data Cleaning steps:
- parse data into correct columns (for HI, this was very sketchy)
- look for exact duplicates of data with ALL COLUMNS - note which columns are differentiating them
  - note: you canʻt do this for HI or Oahu because the parsing is so sketch
- select columns of interest
- look for duplicates again
  - if we find duplicates, need to figure out which column is differentiting dwellings. this is very important. we also need to add a unique ID to each row. 

### Parse Oahu and Hawaii
```{r}


# the hawaii dwell data is tab delimited but I think that is a result of some kind of transformation that occurrred. 

# read in raw data
permit_data_raw <- read_csv(here("data/osds/shared-data/2009_Present_IWSList_Permits.csv"))
spec(permit_data_raw)

# after looking at the mess of the data here is some code to try and read it in nicely

# Load necessary package
library(readr)



# oahu data has the fixed width preserved, which I counted by hand and confirmed at least the first bunch columns match. The last few are also easy to tell which columns so below I assign them manually.

# hi data was a mess, but the most promising outcome was to try and replace the tabs with spaces (8 seemed like the best bet), and then read it in using the same column specs from the metatdata


# Define widths and optionally names
col_specs <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92, 112, 132, 152, 154,
    179, 186, 193, 199, 201, 202, 206, 211, 216, 218, 238, 239,
    259, 279, 299, 319, 343, 355, 357, 361, 364, 384, 404, 434,
    438, 443, 444, 448, 470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 18, 23, 43, 44, 47, 71, 91, 111, 131, 151, 153, 173,
    180, 187, 194, 200, 201, 201, 207, 212, 217, 237, 238, 258,
    278, 298, 318, 342, 354, 356, 360, 363, 383, 403, 433, 437,
    442, 443, 447, 469, 478, 489, 494, 502, 509, 515  # <-- adjust last few by hand
  )
)



# Read the fixed-width file
oahu_raw_dwell <- read_fwf(
  here("data/osds/shared-data/Dwellings/Oahu_Dweldat2020.txt"),
  col_positions = col_specs,   # or actual names from metadata
  na = c("", "NA"),
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))

# remame columns based on visual examination

oahu_dwell_rename <- oahu_raw_dwell %>% 
  select(jurisid = X1,
         taxid = X2,
         tax_year = X3,
         card = X4,
         condo_dwell = X6,
         full_baths = X15,
         half_baths = X16,
         total_rooms = X21,
         bedrooms = X22,
         year_built = X42) %>% 
  mutate(year_built = substr(year_built,1,4))

# fix three that were off

oahu_fix <- oahu_raw_dwell %>% 
  filter(X2 %in% c("920060090000","130110480001","130110480003")) %>% 
  select(jurisid = X1,
         taxid = X2,
         tax_year = X3,
         card = X4,
         condo_dwell = X6,
         full_baths = X15,
         half_baths = X16,
         total_rooms = X21,
         bedrooms = X22,
         year_built = X42) %>% 
  mutate(year_built = substr(year_built,5,8)
  )


# re join to full data set and save

oahu_dwell_clean <- oahu_dwell_rename %>% 
  filter(taxid %!in% c("920060090000","130110480001","130110480003")) %>% 
  rbind(oahu_fix)


write_csv(oahu_dwell_clean,
         here("data/osds/generated-data/oahu_dwelldat_clean.csv"),
         append = FALSE)



############ hawaii data, expanded version #####################
# note to create this I ran the following code in my terminal : expand -t 8 Hawaii_DWELDAT2023.TXT > hi_expanded.txt


# hi_df <- read_fwf(
#   file = here("data/osds/shared-data/Dwellings/hi_expanded.txt"),
#   col_positions = col_specs,  # from your metadata start/end bytes
#   col_types = cols(.default = col_character())
# )
# 
# head(hi_df,5)


# 1) Read as TAB-delimited (keep everything as text)
df_raw <- read_delim(here("data/osds/shared-data/Dwellings/Hawaii_DWELDAT2023.TXT"),
                    delim = "\t",
                    col_names = FALSE, 
                    na = c("", "NA"),
                    col_types = cols(.default = col_character())
) %>% mutate(row_id = row_number(), .before = 1)

# ok this is better
hi_since_2020 <- df_raw %>% 
  mutate(year_built = as.numeric(X23)) %>% 
  filter(X23 > 2019)

# which are nas?

hi_since_2020na <- hi_since_2020 %>% filter(is.na(year_built)) # we dont want anyway

# filter out the nas

hi_since_2020 <- hi_since_2020 %>% 
  filter(!is.na(year_built))

issues_summ <- hi_since_2020 %>%
  summarise(
    n_na_X5 = sum(is.na(X5)), 
    unique_X5 = list(unique(na.omit(X5))),
    n_na_X12 = sum(is.na(X12)),
    unique_X12 = list(unique(na.omit(X12)))
  )


suspect_rows <- hi_since_2020 %>%
  mutate(
    X5_has_space  = grepl(" ", X5),
    X12_has_space = grepl(" ", X12)
  ) %>%
  filter(X5_has_space == T | X12_has_space ==T)
  
  
suspect_row_ids <- suspect_rows %>% pull(row_id)

# now read those in raw to see what they should be. see if you can work with this one
raw_lines <- read_lines(here("data/osds/shared-data/Dwellings/hi_expanded.txt"))[suspect_row_ids]
raw_lines

# rename columsn to our best guess so we can get on with our lives

hi_since_2020_cl <- hi_since_2020 %>% 
  select(
    row_id,
    taxid = X1,
    tax_year = X3,
    card = X4,
    full_baths = X15,
    half_baths = X16,
    total_rooms = X17,# this was x21 for oahu
    bedrooms = X18, # this was x22 for oahu
    year_built = X23
  )


write_csv(hi_since_2020_cl,
         here("data/osds/generated-data/hi_2023_dwell_selectedcols.csv"))

```

Sending these to Julia to make sure they line up. 

Ok at this point I am just going to pick ones to go with:

x16 = halfbath
x15 = full bath
x17 = total rooms
x18 = total beddroms


### Clean Maui and Kauai

Look at county data that are not separated like the HI and Oahu data.


**Maui Data Notes**
In this data, the columns that identify either a unique dwelling OR A CHANGE to a in dwelling are as follows:
- tax id (parid)
- year built
- CARD number - Property-record-card docs note that multi-card parcels are used when a parcel has several primary buildings; the card is “# of #” for that parcel. These could be ADUs or other outbuildings. 

These are variable columns
- number of bedrooms
- number of bathrooms
- half baths
- total rooms
- year built

Volatile columns:
- Tax year
- first column id (#)

```{r}

# read in maui data
maui_dwell_raw <- read_xlsx(here("data/osds/shared-data/Dwellings/Maui_Dwelling_2015_2022.xlsx")) %>% 
  clean_names()

colnames(maui_dwell_raw)

# drop any fully duplicated rows (note - there are none due to id col)

maui_df0 <- maui_dwell_raw %>% select(-number) %>% distinct()

# set columns 

id_cols <- c("year_built","parid","card")
tax_year_col <- "tax_year"
total_rooms <- "total_rooms"
beds_col <- "bed_rooms"
baths_col <- "full_baths"


# add a flag for cards that are clearly residential. any without bed and bath information might be non res buildings

# df1 <- maui_df0 %>%
#   mutate(
#     is_res_card = ( !!sym(beds_col) > 0 | !!sym(baths_col) > 0 ), "res")

df1 <- maui_df0

# get one row per unique parcel id and card number, ignoring tax year

unique_dwellings <- df1 %>%
  filter(is_res_card) %>%
  distinct(across(all_of(id_cols)), .keep_all = TRUE)

# list cols that might vary and this matters because of the way we calculate OSDS

var_cols <- c("total_rooms","half_baths","full_baths","bed_rooms")


# 2) Order by year and flag changes in attributes versus the previous year 

df_check <- df1 %>% 
  filter(is_res_card) %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) 

df_changes <- df1 %>% 
  filter(is_res_card) %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) %>% 
  mutate(
    across(
      all_of(var_cols),
      list(
        changed = ~. != lag(.),
        increased = ~. > lag(.),
        decreased = ~. < lag(.)
      ),
      .names = ".{col}_{.fn}"
    ),
    change_any = rowSums(across(ends_with("_changed"), ~ .x %in% TRUE), na.rm = TRUE) > 0,
    change_category = dplyr::case_when(
      rowSums(across(ends_with("_increased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "remodel_or_expansion",
      rowSums(across(ends_with("_decreased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "demolition_or_correction",
      change_any ~ "attribute_changed",
      TRUE ~ NA_character_
    )
  ) %>% 
  ungroup()



# which houses have multiple years, id cols and differnt cards? this is likely multiple dwellings on a parcel

same_year_diff_cards <- df1 %>%
  filter(is_res_card) %>%
  group_by(parid = .data[["parid"]], Year = .data[[tax_year_col]]) %>%
  summarise(n_cards = n_distinct(.data[["card"]]),
            beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(n_cards > 1 & (beds_var > 1 | baths_var > 1))

# which houses have the same year, the same card, but conflicting bed and bath values?

conflicts_same_card_same_year <- df1 %>%
  filter(is_res_card) %>%
  group_by(across(all_of(c(id_cols, tax_year_col)))) %>%
  summarise(beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(beds_var > 1 | baths_var > 1) # yay this is zero!

######### create final df of events and summary df for QAQC report

# keep baseline row per dwelling (earliest year) + every change row
final_events <- df_changes %>%
  group_by(across(all_of(id_cols))) %>%
  arrange(tax_year, .by_group = TRUE) %>%
  mutate(is_baseline = row_number() == 1) %>% # which is the og observation
  filter(is_baseline | change_any) %>% # keep the baseline and any we flagged as a change
  # label the event
  mutate(event_type = case_when(
    is_baseline & !change_any ~ "baseline",
    change_category == "remodel_or_expansion" ~ "remodel_or_expansion",
    change_category == "demolition_or_correction" ~ "demolition_or_correction",
    change_any ~ "attribute_changed",
    TRUE ~ "baseline"
  )) %>%
  # list WHICH fields changed on each change row (blank for baseline)
  rowwise() %>%
  mutate(changed_fields = {
    flags <- c_across(ends_with("_changed"))
    cols <- names(flags)[flags]
    cols <- str_remove(cols, "_changed$")
    paste(cols, collapse = ", ")
  }) %>%
  ungroup()

# create a summary df and save both

final_events_view <- final_events %>%
  select(all_of(id_cols), tax_year, event_type, all_of(var_cols))


write_csv(final_events,
          here("data/osds/generated-data/final_maui_dwelling_withflags.csv"))

write_csv(final_events_view,
          here("data/osds/generated-data/final_maui_dwelling_clean.csv"))

# so now we can grab the unique ids from these (tmks) and use that for the spatial data.





```

Questions / thoughts from Maui and Kauai Raw: 
Maui looks pretty good - they key columns ended up being card, tmk year along with the parid, number of bed and bath etc. This has allowed me to flag entries that are duplicate properties, but have undergone a change that matters to osds.

But a quick google search revealed this is way low. I think we are missing a lot of condo data

Likely missing stock:

Condos and apartment complexes often get represented in tax data as one improvement record (one (parid, card)) for the entire building, without per-unit bedroom counts.

In those cases, the bed_rooms field might be blank, zero, or just reflect a model unit, not the sum across all units.

Some tax datasets also exclude certain categories like timeshares, vacation rentals, or “common element” parcels from bedroom counts.

**Maui QA QC**

```{r}

total_bedrooms <- sum(unique_dwellings$bed_rooms, na.rm = T)

total_na_bedrooms <- sum(is.na(unique_dwellings$bed_rooms)) #72 are NA

total_0_bedrooms <- sum(unique_dwellings$bed_rooms == 0, na.rm = T) # 15 are 0

# which are NA or zero? do they all have the same condo flag?

unique_dwellings_na0 <- unique_dwellings %>% 
  filter(bed_rooms == 0 | is.na(bed_rooms))


# look at the raw data for these guys

maui_condos <- maui_dwell_raw %>% 
  filter(parid %in% unique_dwellings_na0$parid)

# how many unique id columns do we have in this data?

unique_combos <- maui_df0 %>%
  distinct(parid, card, year_built)

head(unique_combos, 20)

# counts by year build

coutns_by_yearbuilt <- maui_df0 %>%
  distinct(parid, card, year_built) %>%
  count(year_built, sort = TRUE)

```

Ok mystery solved I think - this is only data from 2010. so I can only track changes since then. That's ok. 

**Kauai Data Notes** 

From the Maui experience, it would be really helpful to have year built, card number AND tax year to sort out changes to a dwelling. But since we have card number let's see if we can figure it out: 

```{r}

kauai_dwell_raw <- read_xlsx(here("data/osds/shared-data/Dwellings/Kauai_BedBathCount_2014_2020_081423.xlsx")) %>% clean_names()

colnames(kauai_dwell_raw)

# remove any total duplicates

kauai_df0 <- kauai_dwell_raw %>% select(-number) %>% distinct() # the same nrow as raw, good

# create columns 

# set columns 

id_cols <- c("parid","card")
total_rooms <- "total_rooms"
beds_col <- "bed_rooms"
baths_col <- "full_baths"

# flag for residential buildings only

df1 <- kauai_df0 %>%
  mutate(
    is_res_card = ( !!sym(beds_col) > 0 | !!sym(baths_col) > 0 ), "res") # all residential

# flag and filter for buildings with toilets
df1 <- kauai_df0 %>%
  mutate(
    no_toilet_flag = !(coalesce(full_baths, 0) > 0 | coalesce(half_baths, 0) > 0)
  )

df2 <- df1 %>% filter(no_toilet_flag == FALSE) # 136 obs dont have a toilet, rest are good



# get one row per unique parcel id and card number, ignoring tax year

unique_dwellings <- df2 %>%
  #filter(is_res_card) %>%
  distinct(across(all_of(id_cols)), .keep_all = TRUE) # fewer. this means there are duplicate parids that indicate changes rather than duplicates

# list cols that might vary and this matters because of the way we calculate OSDS

var_cols <- c("half_baths","full_baths","bed_rooms")

df_check <- df2 %>% 
  #filter(is_res_card) %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(parid, .by_group = TRUE) 

df_changes <- df2 %>% 
  #filter(is_res_card) %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(parid, .by_group = TRUE) %>% 
  mutate(
    across(
      all_of(var_cols),
      list(
        changed = ~. != lag(.),
        increased = ~. > lag(.),
        decreased = ~. < lag(.)
      ),
      .names = ".{col}_{.fn}"
    ),
    change_any = rowSums(across(ends_with("_changed"), ~ .x %in% TRUE), na.rm = TRUE) > 0,
    change_category = dplyr::case_when(
      rowSums(across(ends_with("_increased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "remodel_or_expansion",
      rowSums(across(ends_with("_decreased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "demolition_or_correction",
      change_any ~ "attribute_changed",
      TRUE ~ NA_character_
    )
  ) %>% 
  ungroup()

# interesting - so there are no changes recorded for kauai. without tax year or year built, we can't really determine that


final_events_view <- df_changes %>%
  select(all_of(id_cols),all_of(var_cols), change_any) %>% 
  mutate(year_built = NA,
         tax_year = NA,
         island = "Kauai") %>% 
  distinct()


write_csv(final_events_view,
          here("data/osds/generated-data/final_oahu_dwelling_clean.csv"))



```

Because we dont have tax year or year built, and because none of the rows here flagged as a possible change, I can only assume that this dwelling data represents a snap shot in time (ie 2023). It would be better if we could re-request the data with all tax assessments. 

### Clean Oahu and Hawaii

After figuring out a good workflow with the Maui data (the best dataset, I'd say), apply to the Oahu and Hawaii parsed data.

**Oahu Data Notes**

I went back and made sure I had all the key columns in the parsing code: parid, taxyear, card, year built, and the beds and baths. then I repeated the Maui code.

The card column doesn't look quite right, some are really high, but it could still help us ID duplicates...maybe. I'll try!


```{r}

# read in parsed data

oahu_raw <- read_csv(here("data/osds/generated-data/oahu_dwelldat_clean.csv"))

# check for any fully duplicated rows

oahu_df0 <- oahu_raw %>% distinct() %>% 
  mutate(across(6:9, ~as.numeric(.x))) %>% 
  rename(bed_rooms = bedrooms)# none dropped, good!

# set columns 

id_cols <- c("year_built","taxid","card")
tax_year_col <- "tax_year"
total_rooms <- "total_rooms"
beds_col <- "bed_rooms"
baths_col <- "full_baths"

# set cols and find unique entries

unique_dwellings <- oahu_df0 %>%
  distinct(across(all_of(id_cols)), .keep_all = TRUE) #all unique, like kauai

# define columns that vary
var_cols <- c("total_rooms","half_baths","full_baths","bed_rooms")

# Group by id cols, order by year, and flag any differences

df_check <- oahu_df0 %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) 

df_changes <- oahu_df0 %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) %>% 
  mutate(
    across(
      all_of(var_cols),
      list(
        changed = ~. != lag(.),
        increased = ~. > lag(.),
        decreased = ~. < lag(.)
      ),
      .names = ".{col}_{.fn}"
    ),
    change_any = rowSums(across(ends_with("_changed"), ~ .x %in% TRUE), na.rm = TRUE) > 0,
    change_category = dplyr::case_when(
      rowSums(across(ends_with("_increased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "remodel_or_expansion",
      rowSums(across(ends_with("_decreased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "demolition_or_correction",
      change_any ~ "attribute_changed",
      TRUE ~ NA_character_
    )
  ) %>% 
  ungroup()

# this didnt change anything. 

# but this hints taht this dataset has the condos:

same_year_diff_cards <- oahu_df0 %>%
  group_by(taxid = .data[["taxid"]], Year = .data[[tax_year_col]]) %>%
  summarise(n_cards = n_distinct(.data[["card"]]),
            beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(n_cards > 1 & (beds_var > 1 | baths_var > 1)) # 7452, can check condo against this. 

# which houses have the same year, the same card, but conflicting bed and bath values?

conflicts_same_card_same_year <- oahu_df0 %>%
  group_by(across(all_of(c(id_cols, tax_year_col)))) %>%
  summarise(beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(beds_var > 1 | baths_var > 1) # yay zero!


final_events_view <- df_changes %>%
  select(all_of(id_cols),all_of(var_cols), change_any) %>% 
  mutate(island = "Hawaii") %>% 
  distinct()


write_csv(same_year_diff_cards,
          here("data/osds/generated-data/oahu_condos_likley.csv"))


write_csv(final_events_view,
          here("data/osds/generated-data/final_oahu_dwelling_clean.csv"))


```

**Hawaii data notes**

Lastly feed Hawaii data through the process - still working with 2020

```{r}


hi_parsed <- read_csv(here("data/osds/generated-data/hi_2023_dwell_selectedcols.csv"))

# check for any fully duplicated rows

hi_df0 <- hi_parsed %>% distinct() %>% 
  select(-row_id) %>% 
  mutate(across(4:8, ~as.numeric(.x))) %>% 
  rename(bed_rooms = bedrooms)# none dropped, good!

# set columns 

id_cols <- c("year_built","taxid","card")
tax_year_col <- "tax_year"
total_rooms <- "total_rooms"
beds_col <- "bed_rooms"
baths_col <- "full_baths"

# find unique dwellings - with dwellings defined as buildings

unique_dwellings <- hi_df0 %>%
  distinct(across(all_of(id_cols)), .keep_all = TRUE) # none dropped


# define columns that vary
var_cols <- c("total_rooms","half_baths","full_baths","bed_rooms")

# Group by id cols, order by year, and flag any differences

df_check <- hi_df0 %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) 

df_changes <- hi_df0 %>% 
  group_by(across(all_of(id_cols))) %>% 
  arrange(.data[[tax_year_col]], .by_group = TRUE) %>% 
  mutate(
    across(
      all_of(var_cols),
      list(
        changed = ~. != lag(.),
        increased = ~. > lag(.),
        decreased = ~. < lag(.)
      ),
      .names = ".{col}_{.fn}"
    ),
    change_any = rowSums(across(ends_with("_changed"), ~ .x %in% TRUE), na.rm = TRUE) > 0,
    change_category = dplyr::case_when(
      rowSums(across(ends_with("_increased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "remodel_or_expansion",
      rowSums(across(ends_with("_decreased"), ~ .x %in% TRUE), na.rm = TRUE) > 0 ~
        "demolition_or_correction",
      change_any ~ "attribute_changed",
      TRUE ~ NA_character_
    )
  ) %>% 
  ungroup()

# which houses have mulitple cards and buildings potentially?

same_year_diff_cards <- hi_df0 %>%
  group_by(taxid = .data[["taxid"]], Year = .data[[tax_year_col]]) %>%
  summarise(n_cards = n_distinct(.data[["card"]]),
            beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(n_cards > 1 & (beds_var > 1 | baths_var > 1)) # 13 can check condo # since 2020 against this. 

# which houses have the same year, the same card, but conflicting bed and bath values?

conflicts_same_card_same_year <- hi_df0 %>%
  group_by(across(all_of(c(id_cols, tax_year_col)))) %>%
  summarise(beds_var = n_distinct(.data[[beds_col]], na.rm = TRUE),
            baths_var = n_distinct(.data[[baths_col]], na.rm = TRUE),
            .groups = "drop") %>%
  filter(beds_var > 1 | baths_var > 1) # yay zero!


final_events_view <- df_changes %>%
  select(all_of(id_cols),all_of(var_cols), change_any) %>% 
  mutate(island = "Hawaii") %>% 
  distinct()

write_csv(final_events_view,
          here("data/osds/generated-data/final_hawaii_dwelling_clean.csv"))

write_csv(same_year_diff_cards,
          here("data/osds/generated-data/hi_condos_likley.csv"))

```



# Phase 2: Join dwell data to most recent HI statewide parcels, permit data, and record differences

First Iʻll work on spatializing the dwell data. Then Iʻll work on joining in the permit data. 

Important: the TMK field for the parcels is only 9 digits. This is because the last 4 digits in the dwell data, as per the metadata, are codes for the horizontal property regime. This is needed for units that share a parcel of land but may have many different owners, like a condo where each unit is individually owned. If the last 4 digits are 0000, then usually that means the parcel is a single owner. 

I will split the dwell TMKS into a TMK column and an HPR column, so we preserve that info, then use the shortened 9 digit tmk to join to parcel centroids. 

```{r clean-parcels}

# read in tmk points from HI state GIS, select columsn we need

library(sf)

tmk_parcels <- st_read(here("data/osds/raw-data-downloaded/tmk_state_parcels.shp/tmk_state.shp")) # note lots of warnings about non closures

tmk_hi <- filter(tmk_parcels,county == "Hawaii")

mapview(head(tmk_hi)) # just check out a few

# valid?

tmk_valid <- tmk_parcels %>% mutate(valid_yn = st_is_valid(geometry))

not_valid <- tmk_valid[tmk_valid$valid_yn == FALSE, ] ## 444 observations, or 0.1 percent

tmk_valid <- tmk_valid[tmk_valid$valid_yn == TRUE, ] # filter out the non valid ones, weʻll come back to those

# get centroids of valid geoms only, save

tmk_parcel_centroid <- st_centroid(tmk_valid)

st_write(tmk_parcel_centroid,
         here("data/osds/generated-data/tmk_centroids.gpkg"),
         layer = "tmk_centroids") # using geopacakge because .shp cant deal with tmk field

```


Since some parcels are not valid. weʻll check the OSDS permit data against these as well as the dwell data. if they match any, figure out how to include. if not, then yay we can just move on

Note - this will be a one to many join, in that the same centroid will be added to mulitple rows. THAT IS OK. the HPR will help us determine if that is a mistake or because there are mulitple individually owned dwellings on the property. 

Also note - for HI, we only grabbed dwell data for units built since 2020 for now. Re-do once we have better data. 

```{r}

options(scipen = 999)

# read in clean dwell data and create a new column 

oahu_maui_kauai_dwell <- read_csv(here("data/osds/generated-data/oahu_maui_kaui_dwell_merged.csv"))

oahu_maui_kauai_dwell_forjoin <- oahu_maui_kauai_dwell %>% 
  mutate(tmk = substr(taxid,1,9),
         hpr = substr(taxid,10,13))

hi_dwell <- read_csv(here("data/osds/generated-data/hi_2023_dwell_clean.csv")) #2487

hi_dwell_forjoin <- hi_dwell %>% 
  mutate(tmk = substr(taxid,1,9),
         hpr = substr(taxid,10,13))

# read in tmk centroids, filter out HI for clarity 

tmk_centroids <- st_read(here("data/osds/generated-data/tmk_centroids.gpkg"),
                         layer = "tmk_centroids")

tmk_oahu_maui_kauai <- tmk_centroids %>% filter(county != "Hawaii") %>% rename(tmk = TMK_txt)

tmk_hi <- tmk_centroids %>% filter(county == "Hawaii") %>% rename(tmk = TMK_txt)

# join to dwell data - left join

hi_dwell_sf <- left_join(hi_dwell_forjoin,tmk_hi, by = "tmk") 


oahu_maui_kaui_dwell_sf <- left_join(oahu_maui_kauai_dwell_forjoin,tmk_oahu_maui_kauai, by = "tmk") 

# after confirm no major join issues, clean up columns and save

hi_dwell_sf_clean <- hi_dwell_sf %>% 
  select(2:9,zone,qpub_link,valid_yn)

oahu_maui_kauai_dwell_sf_clean <- oahu_maui_kaui_dwell_sf %>% select(1:9,zone,qpub_link,valid_yn)

# generate data first QAQC report - which tmks do not have dwells? we can look at and summarize zone ids (digit 3) to see how these look. And, do any dwells have no tmks?

hi_dwell_nosf <- hi_dwell_sf %>% 
  mutate(geom_yn = st_is_empty(geom)) %>% 
  filter(geom_yn == TRUE) # yay 0 observations

oahu_etc_dwell_nosf <- oahu_maui_kaui_dwell_sf %>%
  mutate(geom_yn = st_is_empty(geom)) %>% 
  filter(geom_yn == TRUE) # ok there are 1037 that are missing a spatial point. does this mean they no longer exist? save for QA QC report


# tmk no dwell summary data

tmk_not_indwell <- tmk_centroids %>% 
  filter(TMK_txt %!in% oahu_maui_kauai_dwell_sf_clean$tmk) %>% 
  filter(TMK_txt %!in% hi_dwell_sf_clean$tmk)

length(unique(tmk_not_indwell$TMK_txt)) #212868
nrow(tmk_not_indwell) #212934 <- some tmks are repeated

total_tmkwdwell <- length(unique(oahu_maui_kauai_dwell_sf_clean$tmk)) + length(unique(oahu_maui_kauai_dwell_sf_clean$tmk)) #337296

total_tmk_check <- total_tmkwdwell + nrow(tmk_not_indwell)

nrow(tmk_centroids)
total_tmk_check


```

Ok so two things came out of this:
1) there are ~1000 dwellings in the Oahu, Maui and Kauai dataset that did not have a matching parcel TMK. 

2) there are duplicate TMKS in the parcel dataset. I can tell because when I joined them to the oahu maui kauai dwell data, the overall row count got larger. Look more closely at these below. 


```{r}
# add a column that ids whether the data within is duplicated
tmk_oahu_maui_kauai <- tmk_centroids %>% filter(county != "Hawaii") %>% rename(tmk = TMK_txt) %>% 
  mutate(is_duplicate_tmk = duplicated(.[,"tmk"]))

tmk_oahu_maui_kauai_dups <- filter(tmk_oahu_maui_kauai,is_duplicate_tmk == TRUE) # weird

# if its not duplicate tmks, what is it?

# do a set dif on the dwell data before and after the tmk join to examine the extra ~200 rows

oahu_etc_setdiff <- setdiff(oahu_maui_kauai_dwell_sf_clean$tmk, oahu_maui_kauai_dwell_forjoin$tmk)

oahu_etc_setdiff <- setdiff(oahu_maui_kauai_dwell_sf_clean, oahu_maui_kauai_dwell_forjoin$tmk)

# since the tmks arenʻt the issue, add a row id and see what is "exploding" the numbers out by 200

# what happens if we remove EXACT duplicates?

oahu_maui_kauai_dwell_sf_clean_nodups <- distinct(oahu_maui_kauai_dwell_sf_clean)

```

Ah ok so there are a lot of duplicates in the RAW dwell data. The key thing we need to decide is if we want to just move ahead and treat these as duplicates, or if we want to think about if their duplication means something. For example, I want to be careful with kauai because we have less information for those dwellings.  

















