---
title: "Oahu Timeseries Streamlined"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))


# set up out paths for each step or chunk - change upstream 


```

### Step 0: Create Authoratative TMK lists - RUN FIRST

```{r}

library(janitor)
# read in new permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"),
                            col_types = cols(tmk = col_character()))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  clean_names()

iws_oahu_permits_tmk <- iws_oahu_permits %>% pull(tmk)

# read in old otp

otp_oahu_raw <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/osds_oah.shp")) %>% 
  clean_names() %>% 
  mutate(tmk = sprintf("%09.0f", as.numeric(tmk_9))) %>% 
  st_drop_geometry()

otp_oahu_tmk <- otp_oahu_raw %>%
  pull(tmk)

# new ones:

oahu_unchanged <- otp_oahu_raw %>% 
  filter(tmk %!in% iws_oahu_permits_tmk) #12519

unchanged_tmks <- oahu_unchanged %>% pull(tmk) %>% unique() #12461

oahu_updated <- otp_oahu_raw %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_oahu_permits_tmk) #1165

updated_tmks <- oahu_updated %>% pull(tmk) %>% unique() # 1164

oahu_new <- iws_oahu_permits %>% 
  filter(tmk %!in% otp_oahu_tmk) # 314

new_tmks <- oahu_new %>% pull(tmk) %>% unique() #


# create authoritative tmk list - add ups!

master_tmk_list <- c(iws_oahu_permits_tmk,otp_oahu_tmk)
master_tmk_list <- unique(master_tmk_list) # 13873



```

### Step 1: Clean Up Dwell Data

About dwelling data: - obtained from each county, and maintained by each
county differently - records are created for tax assessment purposes,
and therefore usually have two key time columns: year built and year
assessed. - if the tmk has a second card number, that usually indicates
an auxiliary building or additional structure built on the property.

**Dwell Data Cleaning steps:** - parse data into correct columns (for
HI, this was very sketchy) - look for exact duplicates of data with ALL
COLUMNS - note which columns are differentiating them - note: you canʻt
do this for HI or Oahu because the parsing is so sketch - select columns
of interest - look for duplicates again - if we find duplicates, need to
figure out which column is differentiating dwellings. this is very
important. we also need to add a unique ID to each row.

Parse Oahu Raw files, join all years

Part of this can all be found in the OG driver repository on github.

The earlier years I got later (2010 - 2014) and I'll do the cleaning
here

```{r}

# read in packge
library(readr)

# define col specs
col_specs_old <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92, 112, 132, 152, 154,
    179, 186, 193, 199, 201, 202, 206, 211, 216, 218, 238, 239,
    259, 279, 299, 319, 343, 355, 357, 361, 364, 384, 404, 434,
    438, 443, 447, 448, 470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 18, 23, 43, 44, 47, 71, 91, 111, 131, 151, 153, 173,
    180, 187, 194, 200, 201, 201, 207, 212, 217, 237, 238, 258,
    278, 298, 318, 342, 354, 356, 360, 363, 383, 403, 433, 437,
    442, 446, 448, 469, 478, 489, 494, 502, 509, 515 
  )
)

# try new ones

col_specs_new <- fwf_positions(
  start = c(
    1, 15, 20, 24, 44, 49, 52, 72, 92, 112, 132, 152, 154, 
    179, 186, 193, 199, 201, 202, 206, 211, 217, 218, 238, 
    239, 259, 279, 299, 319, 343, 349, 356, 358, 365, 373, 
    375, 379, 382, 402, 422, 443, 448, 452, 453, 457, 479, 
    488, 499, 504, 512, 519
  ),
  end = c(
    14, 19, 23, 43, 48, 51, 71, 91, 111, 131, 151, 153, 178, 
    185, 192, 198, 200, 201, 205, 210, 216, 217, 237, 238, 
    258, 278, 298, 318, 342, 348, 355, 357, 364, 372, 374, 
    378, 381, 401, 421, 442, 447, 451, 452, 456, 478, 487, 
    498, 503, 511, 518, 525 # Adjust final end based on actual row length
  ),
  col_names = c(
    "tax_map_key", "parcel_year", "card_no", "building_no", "indicator", 
    "story_height", "roof_design", "roof_structure", "roof_material", 
    "ext_wall", "framing", "style_occupancy", "occupancy", "full_baths", 
    "half_baths", "add_fixtures", "total_fixtures", "ac", "attic_rating", 
    "total_rooms", "bedrooms", "family_rooms", "foundation", "basement", 
    "floor_construction", "flooring", "int_wall_struct", "int_wall_mat", 
    "ceiling", "rec_room_size", "misc_desc_1", "misc_qty_1", "misc_desc_2", 
    "misc_qty_2", "condo_level", "condo_type", "condo_view", "condo_parking", 
    "condo_style", "condo_common", "year_built", "eff_year_built", 
    "phys_cond", "bldg_code", "bldg_grade", "bldg_shape", "cost_design_pct", 
    "econ_cond_factor", "pct_complete", "ground_floor_area", "living_area"
  )
)

# read in each year and check how it worked out

oahu_2014 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2014-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2013 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2013-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2012 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2012-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))



oahu_2011 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2011-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))


oahu_2010 <- read_fwf(
  file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Oahu_2010 to 2014-Dweldat_04142026/2010-Dweldat.txt"),
  col_positions = col_specs_new,
  col_types = cols(.default = col_character())
) %>% mutate(across(everything(), ~trimws(.)))

```

Make sure columns we need are clean:

```{r}


check_parsing_quality <- function(df, year_label = "Unknown Year") {
  # 1. Calculate lengths
  results <- df %>%
    mutate(
      len_year = nchar(as.character(year_built)),
      is_messy = len_year != 4
    )
  
  # 2. Count totals
  total_rows <- nrow(results)
  messy_rows <- results %>% filter(is_messy)
  n_messy    <- nrow(messy_rows)
  pct_messy  <- (n_messy / total_rows) * 100
  
  # 3. Print a pretty summary to the console
  cat("\n==========================================\n")
  cat("PARSING REPORT FOR:", year_label, "\n")
  cat("Total Rows:    ", total_rows, "\n")
  cat("Clean Rows:    ", total_rows - n_messy, "\n")
  cat("Messy Rows:    ", n_messy, "\n")
  cat("Error Rate:    ", round(pct_messy, 2), "%\n")
  
  if(n_messy > 0) {
    cat("\nCommon 'Messy' Lengths detected:\n")
    print(table(messy_rows$len_year))
    
    cat("\nExample of messy value: '", head(messy_rows$year_built, 1), "'\n")
  }
  cat("==========================================\n")
  
  # Return the messy data invisibly so you can assign it if needed
  return(invisible(messy_rows))
}

oahu_2014_messy <- check_parsing_quality(oahu_2014, "2014 Data")
oahu_2013_messy <- check_parsing_quality(oahu_2013, "2013 Data")
oahu_2012_messy <- check_parsing_quality(oahu_2012, "2012 Data")
oahu_2011_messy <- check_parsing_quality(oahu_2011, "2011 Data")
oahu_2010_messy <- check_parsing_quality(oahu_2010, "2012 Data")

# check bedroom

check_bedroom_quality <- function(df, year_label = "Dataset") {
  
  # 1. Clean and standardize the bedroom column
  # We convert to character and trim to catch "00", "0", or "   0"
  results <- df %>%
    mutate(
      bd_char = str_trim(as.character(bedrooms)),
      status = case_when(
        is.na(bedrooms) | bd_char == "" | bd_char == "NA" ~ "Missing/NA",
        bd_char %in% c("0", "00", "000") ~ "Zero (00)",
        TRUE ~ "Valid Count"
      )
    )
  
  # 2. Generate summary stats
  summary_table <- table(results$status, useNA = "always")
  total_rows <- nrow(df)
  n_na   <- sum(results$status == "Missing/NA")
  n_zero <- sum(results$status == "Zero (00)")
  
  # 3. Print the report
  cat("\n==========================================\n")
  cat("BEDROOM QUALITY REPORT:", year_label, "\n")
  cat("Total Records: ", total_rows, "\n")
  cat("------------------------------------------\n")
  cat("NA / Missing:  ", n_na, " (", round(n_na/total_rows*100, 1), "%)\n")
  cat("Zero (00):     ", n_zero, " (", round(n_zero/total_rows*100, 1), "%)\n")
  cat("Valid (> 0):   ", sum(results$status == "Valid Count"), "\n")
  cat("==========================================\n")
  
  # Return the summarized counts as a data frame
  return(as.data.frame(summary_table))
}

check_bedroom_quality(oahu_2014, "2014")

table(oahu_2014$indicator, is.na(oahu_2014$bedrooms))

# lets just join and filer if possible

oahu_2010_2014_cl <- rbind(
  oahu_2014,
  oahu_2013,
  oahu_2012,
  oahu_2011,
  oahu_2010
) %>% 
  select(tax_map_key,
         tax_year = parcel_year,
         card_no,
         total_rooms,
         bedrooms,
         year_built,
         eff_year_built)

oahu_cl_tmk <- oahu_2010_2014_cl %>% 
  mutate(tmk = substr(tax_map_key,1,9))


iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date))

oahu_dwell_osds_2010_2014 <- oahu_cl_tmk %>% 
  filter(tmk %in% master_tmk_list) # 9759


write_csv(oahu_dwell_osds_2010_2014,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/oahu_dwell_clean_2010_2014.csv"))

```

This report should tell us:

-   total number of TMKS assessed each year for Oahu
-   how many TMKS are single owner versus multiple owner per year of
    data
-   how many TMKS have more than one card (i.e. more than one building
    assessed on the parcel) per year
-   Summary of bedrooms and bathrooms per year?

Note: go back to parsing data and check the card column.

Resources:

<https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/>

```{r dwelldata}

library(data.table)

oahu_dwell_path_2015_2020 <- file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/oahu_dwell_clean_2015_2020.csv") # where was this created?


oahu_dwell_path_2010_2014 <- file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/oahu_dwell_clean_2010_2014.csv")

oahu_dwells_new <- fread(file = oahu_dwell_path_2015_2020) %>% 
  mutate(tmk = substr(paste0(jurisid,taxid),1,9)) %>% 
  filter(tmk %in% master_tmk_list)

oahu_dwells_old <- fread(file = oahu_dwell_path_2010_2014) %>% 
  select(-tax_map_key, -eff_year_built)

oahu_dwells_all <- oahu_dwells_new %>% 
  select(tmk,tax_year,card_no = card, total_rooms, bedrooms,year_built) %>% 
  rbind(oahu_dwells_old)

write_csv(oahu_dwells_all,
         file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/oahu_dwell_clean_allyears.csv"))


length(unique(oahu_dwells_all$tmk)) #how many? 11163

length(unique(master_tmk_list)) #13873 <- so some of these are commercial or not in dwell

# important: filter dwell data by authortative tmk to create "dwell" and "nondwell lists"

oahu_dwells_osds <- oahu_dwells_all %>% 
  filter(tmk %in% master_tmk_list) %>% 
  pull(tmk) %>% 
  unique() #11163 - this includes legacy 

```

### Step 2: Bring in Permit Data (IWS, Dept. of Health)

This code chunk separates permits into ones that match our dwelling
database and ones that don't.

```{r permit-data}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk,"1")) %>%
  rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk)) # N = 1741

tmks_permits <- unique(iws_oahu_permits$tmk) #1412

# execute join to see which tmks have dwell data

iws_oahu_dwell_join <- inner_join(iws_oahu_permits, oahu_dwells_all, by = "tmk") %>% 
  distinct() 

# filter using oahu dwells
iws_oahu_dwell_join_norepeats <- iws_oahu_permits %>% 
  filter(tmk %in% oahu_dwells_all$tmk) #1369, 1178 unique tmks

write_csv(iws_oahu_dwell_join_norepeats,
         file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_NewIWSpermits_with_dwellinginfo.csv"))


# now filter out the ones that do not have dwelling data and save

iws_oahu_NOdwell <- iws_oahu_permits %>% 
  filter(tmk %!in% iws_oahu_dwell_join$tmk) # 372, 234 unique tmks 

write_csv(iws_oahu_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_NewIWSpermits_nodwellinginfo.csv"))


```

**Results so Far**: - There were 1741 permits issued for HI island over
the past 10 years - 1369 have dwelling information, and 372 do not. The
second number written next to N are the number of unique tmks.

### Step 2b: Clean and Pivot New Permits

Instead of doing a bunch of stuff that identifies easy, tricky and hard
scenarios, I'm condensing everything here into one big clean. are as
follows: 1. Assign class, using a custom function 2. Create master list
(table 1) 3. Create tables 2 and 3 based on master list.

##### Wastewater Class Encoding Reference Table

We only have classes 1 - 4 in this data. From the metadata, this
represents:

-   CLASS_I: Class I OSDS are those systems where the effluent is
    subjected to soil treatment
-   CLASS_II: Class II OSDS are septic systems discharging to a seepage
    pit.\
-   CLASS_III: Class III OSDS are aerobic systems discharging to a
    seepage pit.
-   CLASS_IV: Class IV OSDS are cesspools.\
-   EFFLUENT: The total effluent flux in million gallons per day
    discharged within the TMK parcel.

| Detected Classes     | Encoded Sum | Description                          |
|:---------------------|:-----------:|:-------------------------------------|
| **Class 1**          |      1      | Cleanest Septic Type                 |
| **Class 2**          |      2      | Secondary Septic Type                |
| **Class 1 + 2**      |      3      | **Combination of Class 1 + Class 2** |
| **Class 3**          |      4      | Aerobic System                       |
| **Class 1 + 3**      |      5      | Class 1 Septic + Aerobic             |
| **Class 2 + 3**      |      6      | Class 2 Septic + Aerobic             |
| **Class 1 + 2 + 3**  |      7      | Classes 1, 2, and 3                  |
| **Class 4**          |      8      | Cesspool                             |
| **Class 1 + 4**      |      9      | Class 1 + Class 4                    |
| **Class 2 + 4**      |     10      | Class 2 + Class 4                    |
| **Class 1 + 2 + 4**  |     11      | Classes 1, 2, and 4                  |
| **Class 3 + 4**      |     12      | Aerobic + Class 4                    |
| **Class 1 + 3 + 4**  |     13      | Classes 1, 3, and 4                  |
| **Class 2 + 3 + 4**  |     14      | Classes 2, 3, and 4                  |
| **All (1, 2, 3, 4)** |     15      | Full suite of permits                |

```{r new-tables}

library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by adding a year column, assigning osds class, and trying to standardize the address

master_clean <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$tmk,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

master_clean2 <- iws_oahu_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_oahu_NOdwell$tmk,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) # arranging helps see them side-by-side

view(duplicates_to_check)


length(unique(master_clean$tmk)) # 1412, so we still have all tmks represented after distinct

# N rows = 1683 after distinct, 1741 before. Lost 58
# N tmks = 1412

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #


### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(tmk) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(tmk %in% oahu_dwells_all$tmk),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- oahu_dwells_all %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    total_beds = sum(bedrooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE)

table2b_bedrooms <- oahu_dwells_all %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bedrooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bedrooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE)

#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(tmk, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(tmk, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)

# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table2a_bedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table2b_newbedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table3_newclass_all.csv"))

```

### Step 3: Compare and Integrate Past Datasets

#### Records Unchanged Since OTP

```{r unchanged}

#### Unchanged

# 1. Read in dwelling data
oahu_dwell_alltmks <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/oahu_dwell_clean_allyears.csv")
) %>% 
  mutate(tmk = as.character(tmk))

# 2. Duplicate checks (Exploratory)
oahu_dupes <- oahu_unchanged %>% 
  group_by(tmk, osds_class) %>% 
  filter(n() > 1) %>% 
  arrange(tmk) %>% 
  select(-any_of("fid_1")) %>% 
  distinct()

oahu_xytmk_dupes <- oahu_unchanged %>% 
  select(-any_of("fid_1")) %>% 
  distinct() %>% 
  filter(ttl_osds != 0) %>% 
  group_by(tmk) %>% 
  filter(n_distinct(x, y) > 1) %>% 
  ungroup() %>% 
  arrange(tmk)

# Look at duplicates from the dwell data
oahu_unch_dupes_dwell <- oahu_dwell_alltmks %>% 
  filter(tmk %in% oahu_dupes$tmk) %>% 
  select(tmk, year_built, card_no, total_rooms, bedrooms) %>% 
  distinct()

oahu_unchanged %>% 
  group_by(tmk) %>% 
  summarize(n_counts = n_distinct(ttl_osds)) %>% 
  filter(n_counts > 1)

# 3. Encode classes
oahu_unchanged_cl <- oahu_unchanged %>% 
  select(tmk, osds_class, ttl_osds, type, x, y)

oahu_unchanged_encoded <- oahu_unchanged_cl %>% 
  rename(ttl_osds_2010 = ttl_osds) %>% 
  mutate(osds_class_otp = case_when(
    osds_class == "Class IV"  ~ 8, # cesspool 
    osds_class == "Class III" ~ 4,
    osds_class == "Class II"  ~ 2,
    osds_class == "Class I"   ~ 1,
    osds_class == "Multiple"  ~ 9, # multiple, at least one is a cesspool
    TRUE ~ NA_real_
  )) %>% 
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "osds_class", 
    values_from = "osds_class_otp",
    values_fn = max 
  ) %>% 
  rename(Class_mult = any_of("Multiple")) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("Class")), na.rm = TRUE))

# Check TMK counts
stopifnot(n_distinct(oahu_unchanged_encoded$tmk) == n_distinct(oahu_unchanged$tmk))

# 4. Join to dwellings
oahu_unchgd_encoded_dwell <- left_join(oahu_unchanged_encoded, oahu_dwell_alltmks, by = "tmk") %>% 
  distinct()

# --- Table 1: Registry Summary ---
oahu_card_dwell_summary <- oahu_unchgd_encoded_dwell %>% 
  group_by(tmk) %>% 
  summarize(
    n_buildings = n_distinct(card_no, na.rm = TRUE),
    in_dwelldat = any(!is.na(tax_year)),
    .groups = "drop"
  )

oahu_otp_unchanged_table1 <- oahu_unchgd_encoded_dwell %>% 
  distinct(tmk, .keep_all = TRUE) %>% # Ensures strict 1 row per TMK
  mutate(
    n_permits = case_when(
      encoded_sum %in% c(1, 2, 4, 8) ~ 1,
      encoded_sum %in% c(3, 5, 6, 9, 10, 12) ~ 2,
      encoded_sum %in% c(7, 11, 13, 14) ~ 3,
      encoded_sum == 15 ~ 4,
      TRUE ~ 99999
    ),
    n_years = 1,
    min_year = 2010,
    max_year = 2010,
    n_addresses = "unknown",
    status_since_otp = "unchanged since otp"
  ) %>% 
  left_join(oahu_card_dwell_summary, by = "tmk") %>% 
  mutate(
    scenario = case_when(
      n_permits == 1 & n_years == 1 & n_buildings == 1 ~ "Simple",
      n_buildings > 1 ~ "Multi-building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Unknown"
    )
  ) %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    scenario,
    status_since_otp,
    n_buldings
  )

# --- Table 2: Bedrooms Time Series ---
oahu_otp_unchanged_table2 <- oahu_unchgd_encoded_dwell %>% 
  filter(!is.na(tax_year)) %>% 
  mutate(bed_rooms = as.numeric(bedrooms)) %>% 
  group_by(tmk, tax_year) %>% 
  summarize(total_beds = sum(bed_rooms, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  rename_with(~ paste0("bedrooms_", .), matches("^[0-9]")) %>% 
  view()

# --- Table 3: OSDS Encoded Class History ---
oahu_otp_unchanged_table3 <- oahu_unchgd_encoded_dwell %>% 
  mutate(year = 2010) %>% 
  select(tmk, year, encoded_sum) %>% 
  distinct() %>% 
  pivot_wider(id_cols = tmk, names_from = year, values_from = encoded_sum) %>% 
  rename_with(~ paste0("osds_class_", .), matches("^[0-9]")) # Aligned to osds_class_2010

# --- Table 4: Non-Dwelling Coordinates ---
# Deduplicate spatial coordinates before joining
oahu_unchanged_xy <- oahu_unchanged %>% 
  group_by(tmk) %>% 
  summarize(x = mean(x, na.rm = TRUE), y = mean(y, na.rm = TRUE), .groups = "drop")

oahu_otp_unchanged_table4_googleplaces <- oahu_otp_unchanged_table1 %>% 
  filter(in_dwelldat == FALSE) %>% 
  left_join(oahu_unchanged_xy, by = "tmk") %>% 
  select(tmk, x, y, in_dwelldat)

# --- Save CSV Outputs ---
write_csv(oahu_otp_unchanged_table1,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table1_tmk_registry_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table2,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table2_beds_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table3,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table3_classes_UNCHANGED.csv"))

write_csv(oahu_otp_unchanged_table4_googleplaces,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table4_commercial_UNCHANGED_draft.csv"))
```

#### Records Updated Since OTP

```{r updated}

# 1. Load missing prerequisite objects
# Make sure oahu_updated is loaded from intermediate output if not in memory
if (!exists("oahu_updated")) {
  oahu_updated <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_updated_draft.csv")) %>% 
    mutate(tmk = as.character(tmk))
}

# 2. Bed count gap-fill checks
table_2_new <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table2a_bedrooms_sum.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename_with(~ paste0("newbeds", .x), .cols = !tmk)

oahu_updated_forjoin <- oahu_updated %>% 
  select(tmk, x, y) %>% 
  distinct(tmk, .keep_all = TRUE)

table_2_updated <- table_2_new %>% 
  filter(tmk %in% oahu_updated_forjoin$tmk)

updates_notInTable2 <- setdiff(oahu_updated_forjoin$tmk, table_2_updated$tmk)

# Route non-dwelling / commercial updates to Table 4
table4_withUpdates <- oahu_updated_forjoin %>% 
  filter(tmk %in% updates_notInTable2)

# 3. Class Gap-filling
table_3_new <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table3_newclass_all.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename_with(~ paste0("newclass", .x), .cols = !tmk)

oahu_updated_withClass_enc <- oahu_updated %>% 
  rename(ttl_osds_2010 = ttl_osds) %>%
  mutate(osds_class_otp = case_when(
    osds_class == "Class IV"  ~ 8, # cesspool 
    osds_class == "Class III" ~ 4,
    osds_class == "Class II"  ~ 2,
    osds_class == "Class I"   ~ 1,
    osds_class == "Multiple"  ~ 9, # multiple, at least one is a cesspool
    TRUE ~ NA_real_
  )) %>%
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "osds_class", 
    values_from = "osds_class_otp",
    values_fn = max 
  ) %>% 
  rename(Class_mult = any_of("Multiple")) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("Class")), na.rm = TRUE))

oahu_updated_withClassEnc_New <- oahu_updated_withClass_enc %>%
  left_join(table_3_new, by = "tmk")

class_cols <- oahu_updated_withClassEnc_New %>% select(starts_with("newclass")) %>% names()

table3_updated_classmatch <- oahu_updated_withClassEnc_New %>% 
  mutate(
    match_index = {
      match_matrix <- pick(all_of(class_cols)) == encoded_sum
      match_matrix[is.na(match_matrix)] <- FALSE 
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    match_name = class_cols[match_index],
    data_presence = if_else(
      if_all(starts_with("newclass"), ~ .x == 0 | is.na(.x)),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )
  ) %>% 
  unite("class_history", starts_with("newclass"), sep = "|", remove = FALSE)

table_1_new <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table1_tmk_registry_all.csv")) %>% 
  mutate(tmk = as.character(tmk))

has_2009 <- "newclass2009" %in% names(table3_updated_classmatch) || "newclass2009" %in% names(table_1_new)
has_2010 <- "newclass2010" %in% names(table3_updated_classmatch) || "newclass2010" %in% names(table_1_new)

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  dplyr::select(tmk, encoded_sum, min_year, max_year, any_of("n_addresses"), class_history, starts_with("newclass")) %>% 
  mutate(
    legacy_val = as.numeric(encoded_sum),
# Safely pull the columns if they exist, otherwise assign NA
    v2009 = if (has_2009) as.numeric(.data[["newclass2009"]]) else NA_real_,
    v2010 = if (has_2010) as.numeric(.data[["newclass2010"]]) else NA_real_,
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010) ~ "no data to compare",
      TRUE ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = case_when(
    scenario_match_flag == "matched 2009 or 2010" ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE ~ NA_real_
  )) %>% 
  select(tmk, newpre_2009, starts_with("newclass"), everything()) %>% 
  rowwise() %>%
  mutate(unique_story = {
    # Strictly isolate class value columns to build history string
    vals <- c_across(c(newpre_2009, starts_with("newclass")))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>%
  ungroup() %>%
  mutate(
    upgrade_scenario = case_when(
      # 1. Exact-Match Overrides
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1"  ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2"    ~ "septic and aerobic - replacement of aerobic",
      unique_story == "3|4"    ~ "septic combo to aerobic",
      unique_story == "1|4"    ~ "single septic to aerobic",
      unique_story == "7|4"    ~ "complex septic system to aerobic",
      unique_story == "8|5"    ~ "cesspool to septic / aerobic hybrid",
      unique_story == "5|1"    ~ "septic and aerobic to septic",
      unique_story == "7|3"    ~ "2 septics and aerobic to just 2 septics - aerobic removal",
      unique_story == "7|1|4"  ~ "2 septics and aerobic + one septic repair + one aerobic repair",
      unique_story == "10|9"   ~ "1 septic + cesspool to better septic and cesspool",
      unique_story == "1|5"    ~ "1 septic with a new aerobic installed",
      unique_story == "8|9"    ~ "1 cesspool + one other type installed",

      # 2. General Standard Logic
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1-7]$") ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$") ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$") ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1-7]$") ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$") ~ "multiple osds with new cesspool application",

      # 3. Errors and Lateral Modifications
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$") ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$") ~ "lateral change (clean to clean)",

      TRUE ~ "other scenario"
    )
  )

table_3_new_updated <- update_logic_refined_cl %>% 
  select(tmk, newpre_2009, starts_with("newclass"))

class_flag_forTable1 <- update_logic_refined_cl %>%
  select(tmk, class_match_flag = scenario_match_flag, unique_story, upgrade_scenario)

```

#### Complete Tables 1 - 3a by combining New, Unchanged and Updated

```{r complete-tables}

# read in table 1 new and unchanged versions, and add gapfill flags

table_1_new <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table1_tmk_registry_all.csv")) %>% 
  rename(in_dwelldat = is_dwelling) %>% 
  mutate(status_since_otp = "New Permit",
         n_addresses = as.character(n_addresses),
         tmk = as.character(tmk)) %>% 
  #select(-full_taxid) %>% 
  distinct()

table_1_unchanged <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step3/oahu_table1_tmk_registry_UNCHANGED.csv")) %>% 
  mutate(scenario = "No new permits") %>% 
  mutate(tmk = as.character(tmk)) %>% 
  distinct()

# are there duplicates? no

table_1_unchangeddups <- table_1_unchanged %>% filter(tmk %in% table_1_new$tmk)

table_1_complete <- bind_rows(table_1_new, table_1_unchanged)

# ------------ Table 2 --------------------- #

# 1. Unchanged parcels (historical baseline)
table_2_unchanged <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table2_beds_UNCHANGED.csv")) %>% 
  mutate(tmk = as.character(tmk))

# 2. Updated parcels (rename newbeds -> bedrooms_)
table_2_updatedforbind <- table_2_updated %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds"))

# 3. New parcels ONLY (MUST ALSO RENAME newbeds -> bedrooms_)
table_2_new_only <- table_2_new %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% table_2_updatedforbind$tmk) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds"))

table_2_complete <- bind_rows(table_2_unchanged, table_2_updatedforbind, table_2_new_only) %>% 
  distinct() #2000 ish commerical?

year_built <- oahu_dwell_alltmks %>% 
  filter(tmk %in% table_2_complete$tmk) %>% 
  select(tmk, year_built) %>%
  filter(!is.na(year_built)) %>% 
  distinct() %>% 
  group_by(tmk) %>%
  summarize(
    # str_flatten to join them with a comma
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    #count how many years were found
    count_buildyears = n()
  ) %>%
  ungroup() %>% 
  mutate(tmk = as.character(tmk))


table_2_flags <- table_2_complete %>% 
  mutate(
    bed_trend = case_when(
      bedrooms_2010 < bedrooms_2020  ~ "Expansion since OTP",
      bedrooms_2010 > bedrooms_2020  ~ "Shrink - double check",
      bedrooms_2010 == bedrooms_2020 ~ "Perfect Match",
      TRUE                          ~ NA_character_
    ),
  ) %>% 
  left_join(year_built, by = "tmk") %>% 
  select(tmk, bed_trend, year_built_combined, count_buildyears)

table_1_complete_bedflags <- table_1_complete %>% 
  mutate(gapfill_flag = "not gapfilled - no earlier info in legacy data") %>% 
  left_join(table_2_flags, by = "tmk")

# ------------- Table 2 Integeration 2

table_2_unchanged <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table2_beds_UNCHANGED.csv")) %>% 
  mutate(tmk = as.character(tmk))

table_2_updatedforbind <- table_2_updated %>% 
  rename_with(~ str_replace(.x, "newbeds", "bedrooms_"), starts_with("newbeds"))

table_2_new_only <- table_2_new %>% 
  filter(tmk %!in% table_2_updatedforbind$tmk) %>% 
  rename_with(~ str_replace(.x, "newbeds", "bedrooms_"), starts_with("newbeds"))

table_2_complete <- bind_rows(table_2_unchanged, table_2_updatedforbind, table_2_new_only) %>% 
  distinct(tmk, .keep_all = TRUE)

# -------------- Table 3 --------------------- #
  
table_3_unchanged <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table3_classes_UNCHANGED.csv"),
                              col_types = cols(tmk = col_character())) %>% 
  distinct() %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename(newclass2010 = osds_class_2010)

table_3_updatedforbind <- table_3_new_updated %>% 
  mutate(tmk = as.character(tmk))

table_3_newonly <- table_3_new %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  distinct()

table_3_complete <- bind_rows(table_3_updatedforbind, table_3_unchanged, table_3_newonly) %>%   rename_with(~ str_replace(.x, "newclass", "osds_class_"), 
              starts_with("newclass")) 

table_1_withClassflags <- table_1_complete_bedflags %>% 
  left_join(class_flag_forTable1, by = "tmk")


# ------------ Table 4 --------------------- # Left off here

# 1. Unchanged commercial locations
table_4_unchanged <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/oahu_table4_commercial_UNCHANGED_draft.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  select(tmk, any_of(c("x", "y")))

# 2. Updated commercial locations from current run
table_4_withUpdates_forbind <- table4_withUpdates %>% 
  mutate(tmk = as.character(tmk)) %>% 
  select(tmk, any_of(c("x", "y")))

# 3. New commercial locations from registry (isolate spatial columns)
table_4_new <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/oahu_table1_tmk_registry_commercial.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk) %>% 
  select(tmk, any_of(c("x", "y")))

# 4. Clean Bind
table_4_complete <- bind_rows(
  table_4_unchanged,
  table_4_new,
  table_4_withUpdates_forbind
) %>% 
  distinct(tmk, .keep_all = TRUE)



# ----------- Clean up Table 1 ---------------- #

table_1_final <- table_1_withClassflags %>% 
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete$tmk, TRUE,FALSE)
  ) %>% 
  mutate(bed_trend = "not able to discern bed trend") %>% 
  select(
    tmk,
    starts_with("n_"),
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag = gapfill_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4 = use_table4
  )

# ----------- Tmk checks ----------------- #

# tmk checks
# All TMKs in table 1 accounted for table 3? 

all(table_1_final$tmk %in% table_3_complete$tmk)

# All TMKS in table 2 in table 1?
all(table_2_complete$tmk %in% table_1_final$tmk)

# Authoritative list all in table 1?

all(master_tmk_list %in% table_1_final$tmk) # true
shouldbe_intable1 <- setdiff(master_tmk_list, table_1_final$tmk)

# Difference in table 2 and table 1 == table 4?

shouldbein_table_4 <- setdiff(table_1_final$tmk,table_2_complete$tmk) #2710

shouldbe_intable2 <- setdiff(shouldbein_table_4,table_4_complete$tmk) # 0

head(shouldbe_intable2)

all(shouldbe_intable2 %in% table_2_complete$tmk) #TRUE

table_4_complete_cl <- table_4_complete %>% 
  filter(tmk %!in% shouldbe_intable2)


# update use table 4 flag
table_1_final_final <- table_1_final %>%
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete_cl$tmk, TRUE,FALSE)
  )

# Save all! 


write_csv(table_1_final_final,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table1_tmk_registry_ALL.csv"))

write_csv(table_2_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table2_bedrooms_ALL.csv"))

write_csv(table_3_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table3a_permitsApplied_ALL.csv"))

write_csv(table_4_complete_cl,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table4_percommercial_draft.csv"))


```

## Step 4: Create table 3b

The additive encoding is what is known as a binary bit mask. Each class
represents a power of 2, so every unique sum corresponds to exactly one
combination of classes.

We can think of table 3a as an "event" table, since it records the
permit application event. However, we want to create a "status" table,
which records the actual OSDS present on the property in each year,
BASED on the events we know happened. This will be table 3b

We can make multiple versions of this, using different math. In the
first, I'll assume that: 1. If there was only one OSDS on the property,
the new application is replacing it 2. If there were multiple OSDS on
the property, the new application is replacing the grossest, ie highest
class. This correspond to the largest power of 2 within the encoded
value

This translates to the following math: 1. Identify the largest power of
2 inside the current value. 2. Subtract that power of 2. 3. Add the new
application value.

```{r table-3b}

# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% 
  rename(osds_cl_2008 = newpre_2009) %>% # note, this is a stand in for "pre 2009"
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(Applied,0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table3b_allyears_draft.csv"))

```

## Final Naming Schema From Oahu

| Table | FileName | Key Column Names |
|----|----|----|
| Table 1 | {island}\_table1_tmk_registry_ALL.csv | tmk, n_addresses, in_dwelldat, status_since_otp, n_buildings, bed_trend, year_built_combined, bed_gapfill_flag, permit_story, upgrade_scenario, use_table4 |
| Table 2 | {island}\_table2_bedrooms_ALL.csv | tmk, bedrooms_2010, bedrooms_2011, ..., bedrooms_2020" |
| Table 3 | {island}\_table3a_permitsApplied_ALL.csv | tmk, osds_class_pre2009, osds_class_2009, osds_class_2010, ..., osds_class_2020 |
| Table 4 | {island}\_table4_commercial_draft.csv | tmk, x, y |
