---
title: "Maui OSDS Timeseries"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))


```

### Step 0: SET UP and Create authoritative tmk lists

To start, we'll read in the iws data and the old otp data and create 4 lists:
- unchanged since OTP
- updated and present in both data sets
- new permits since OTP
- ALL tmks in both dataset - authortitative datasets


* we'll also clean up the datasets by standardizing the tmk column 

```{r}

# new data: iws permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk,"2"))

iws_maui_permits_tmk <- iws_maui_permits %>% pull(tmk)


# old data: otp 

# read in maui and molokai. check lanai?
otp_bob_maui <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Maui.csv")) %>% 
  clean_names()

otp_bob_molo <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Molokai.csv")) %>% 
  clean_names()

otp_maui <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk))

otp_maui_tmk <- otp_maui %>% 
  pull(tmk)

# new ones:

maui_unchanged <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% iws_maui_permits_tmk) # 12857

unchanged_tmks <- maui_unchanged %>% pull(tmk) %>% unique() #12857

maui_updated <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_maui_permits_tmk) #1574

updated_tmks <- maui_updated %>% pull(tmk) %>% unique() # 1574

maui_new <- iws_maui_permits %>% 
  filter(tmk %!in% otp_maui_tmk) # 837

new_tmks <- maui_new %>% pull(tmk) %>% unique() #659
# create authoritative tmk list - add ups!

maui_master_tmk_list <- c(iws_maui_permits_tmk,otp_maui_tmk)

maui_master_tmk_list <- unique(maui_master_tmk_list) # 15090


```


### Step 1: Clean Up Dwell Data

The Maui data comes to us pre-parsed, which is great. Here we will just generate the same summary tables as we did for Oahu. We will have to read in each excel file here and do some cleaning. 

Maui specific note: Iʻm not sure the condo type and stype codes are the same as for Oahu. But if they are, the following is what condo style means:

 1 = SINGLE FAMILY
 2 = DUPLEX
 3 = TOWNHOUSE
 4 = LOWRISE
 5 = HIGHRISE

Resources: 

https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/

```{r}

library(readxl)

# list files in the dropbox folder

maui_dwell_fp <- list.files(path = file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Maui_DwellRaw_12112025"),
                            full.names =  T,
                            pattern = ".XLSX")


file_types <- sapply(maui_dwell_fp, excel_format)

check <- data.frame(
  file = basename(maui_dwell_fp),
  format = file_types
)


maui_dwell_dfs <- lapply(maui_dwell_fp, read_excel,
                         skip =1,
                         col_names = TRUE)

# just look at one example:

df_2009 <- maui_dwell_dfs[[1]]


# 
collapse_duplicates <- function(df){
  
  #df <- maui_dwell_dfs[[1]]
  df_collapse <- df %>%  janitor::clean_names() %>% select(-number, -percent_complete) %>% distinct() # this will only remove exact duplicates
  return(df_collapse)
}


maui_collapsed_dfs <- lapply(maui_dwell_dfs, collapse_duplicates)

maui_dwell_allentries <- do.call(rbind,maui_collapsed_dfs) %>% 
    mutate(
      parid = str_trim(as.character(parid)),
      tax_id = substr(parid,1,8),
      tmk = paste0("2",tax_id)
      )

# create a version with just the tax ids that we can use to filter the permit data into "dwells" and "non dwells"
maui_dwelltmks_forjoin_noTXyear <- maui_dwell_allentries %>%
  select(-tax_year) %>%
  distinct()

# create a version with just the TMKS we need from our tmk master list 

maui_dwell_alltmks <- maui_dwell_allentries %>% 
  filter(tmk %in% maui_master_tmk_list) 

# write_csv(maui_dwell_alltmks,
#           file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/maui_dwell_clean_alltmks.csv"))

# check how many we should have
#maui_dwell_alltmks <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/maui_dwell_clean_alltmks.csv"))

length(unique(maui_dwell_alltmks$tmk)) #14402

```


### Step 2: Clean up in IWS data

```{r}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

# filter out just the permits in maui, and create permit ID tracker

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"2")) %>% 
  rename(tmk = tmk_char) %>%  
  rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) #2790 rows, 2233 unique tmks

tmks_permits <- unique(iws_maui_permits$tmk)

# execute join to see which tmks have dwell data. then, do the same using a filter and make sure the number of unique tmks match

iws_maui_dwell_join <- inner_join(iws_maui_permits, maui_dwelltmks_forjoin_noTXyear, by = "tmk") %>% 
  distinct() #8553, 2094 tmks

iws_maui_with_dwell <- iws_maui_permits %>% 
  filter(tmk %in% maui_dwelltmks_forjoin_noTXyear$tmk) #2094 tmks 

# now filter out the ones that do not have dwelling data and save

iws_maui_NOdwell <- iws_maui_permits %>% 
  filter(tmk %!in% iws_maui_dwell_join$tmk) # 191 rows, #139 tmks

write_csv(iws_maui_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**:
- There were 2790 permits issued for Maui Nui over the past 10 years
- 2599 have dwelling information, and this represents 2094 tmks
- 139 tmks do not have dwelling information

### Step 2b: Clean and Pivot Permit Data
Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 

##### Wastewater Class Encoding Reference Table

We only have classes 1 - 4 in this data. From the metadata, this
represents:

-   CLASS_I: Class I OSDS are those systems where the effluent is
    subjected to soil treatment
-   CLASS_II: Class II OSDS are septic systems discharging to a seepage
    pit.\
-   CLASS_III: Class III OSDS are aerobic systems discharging to a
    seepage pit.
-   CLASS_IV: Class IV OSDS are cesspools.\
-   EFFLUENT: The total effluent flux in million gallons per day
    discharged within the TMK parcel.
    

| Detected Classes | Encoded Sum | Description |
| :--- | :---: | :--- |
| **Class 1** | 1 | Cleanest Septic Type |
| **Class 2** | 2 | Secondary Septic Type |
| **Class 1 + 2** | 3 | **Combination of Class 1 + Class 2** |
| **Class 3** | 4 | Aerobic System |
| **Class 1 + 3** | 5 | Class 1 Septic + Aerobic |
| **Class 2 + 3** | 6 | Class 2 Septic + Aerobic |
| **Class 1 + 2 + 3** | 7 | Classes 1, 2, and 3 |
| **Class 4** | 8 | Cesspool |
| **Class 1 + 4** | 9 | Class 1 + Class 4 |
| **Class 2 + 4** | 10 | Class 2 + Class 4 |
| **Class 1 + 2 + 4** | 11 | Classes 1, 2, and 4 |
| **Class 3 + 4** | 12 | Aerobic + Class 4 |
| **Class 1 + 3 + 4** | 13 | Classes 1, 3, and 4 |
| **Class 2 + 3 + 4** | 14 | Classes 2, 3, and 4 |
| **All (1, 2, 3, 4)** | 15 | Full suite of permits |

```{r new-tables}

library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by separating out a year column, adding class, and standarizing the address

master_clean <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$tmk,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

# N rows = 2597 after distinct, 2790 before. Lost 193
# N tmks = 2233. Matches original permit list after distinct

master_clean2 <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) 

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #

### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(tmk) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(tmk %in% maui_dwelltmks_forjoin_noTXyear$tmk),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building", # change this to cards
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- maui_dwell_allentries %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    total_beds = sum(bed_rooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE)

table2b_bedrooms <- maui_dwell_allentries %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bed_rooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE)


#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(tmk, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(tmk, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)


# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table2a_newbedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table2b_newbedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table3_newclass_all.csv"))



```


### Step 3: Compare and Integrate Past Datasets

In this step we have to add in the old data, but differentiate between data that has been updated, unchanged. We also have to gapfill the bedroom and class data, because the old data is just a snapshot in time in time and in some cases we don't have the year information for that OSDS entry. 

#### Records Unchanged Since OTP

These have no new permit data. But, it would be nice to get a better understanding of their bedroom # over time. We have to make one assumption in order to merge our dwell data with this older data. 

For Maui, we also use pittcode here to help us filter non residential TMKS out. 


```{r unchanged}

library(tidyverse)

'%!in%' <- Negate('%in%')

maui_unchanged <- maui_unchanged %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  )

maui_dwell_alltmks <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/maui_dwell_clean_alltmks.csv"),
  col_types = cols(tmk = col_character())
) %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  )

# 2. Encode Classes
maui_unchanged_encoded <- maui_unchanged %>% 
  pivot_longer(
    cols = starts_with("class"),
    names_to = "class",
    values_to = "quantity"
  ) %>% 
  filter(!is.na(quantity) & quantity != 0 & quantity != -9999) %>% 
  rename(ttl_osds_2010 = any_of(c("osds_qty", "ttl_osds"))) %>%
  mutate(osds_class_otp = case_when(
    class == "class_iv"  ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii"  ~ 2,
    class == "class_i"   ~ 1,
    TRUE                 ~ NA_real_
  )) %>%
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010", any_of(c("pittcode", "pitt_code"))), 
    names_from = "class", 
    values_from = "osds_class_otp"
  ) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

# 3. Join to Dwelling Records
maui_unchgd_encoded_dwell <- left_join(maui_unchanged_encoded, maui_dwell_alltmks, by = "tmk") %>% 
  distinct()

# --- Table 1: TMK Registry ---
maui_card_dwell_summary <- maui_unchgd_encoded_dwell %>% 
  group_by(tmk) %>% 
  summarize(
    n_buildings = if ("card" %in% names(.)) n_distinct(card, na.rm = TRUE) else 1,
    in_dwelldat = any(!is.na(tax_year)),
    .groups = "drop"
  )

maui_otp_unchanged_table1 <- maui_unchgd_encoded_dwell %>%
  mutate(
    n_permits = case_when(
      encoded_sum %in% c(1, 2, 4, 8)               ~ 1,
      encoded_sum %in% c(3, 5, 6, 9, 10, 12)       ~ 2,
      encoded_sum %in% c(7, 11, 13, 14)            ~ 3,
      encoded_sum == 15                            ~ 4,
      TRUE                                         ~ 99999
    ),
    n_years = 1,
    min_year = 2010,
    max_year = 2010,
    n_addresses = "unknown",
    status_since_otp = "unchanged since otp",
    scenario = case_when(
      n_permits == 1 & n_years == 1 ~ "Simple",
      n_addresses == "unknown"      ~ "Unknown",
      n_years > 1                   ~ "Potential Upgrade",
      TRUE                          ~ "Multiple Permits Same Year"
    )
  ) %>% 
  left_join(maui_card_dwell_summary, by = "tmk") %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    scenario,
    n_buildings
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

# --- Table 2: Bedrooms ---
maui_otp_unchanged_table2 <- maui_unchgd_encoded_dwell %>%
  filter(!is.na(tax_year)) %>% 
  mutate(bed_rooms = as.numeric(bed_rooms)) %>% 
  group_by(tmk, tax_year) %>% 
  summarize(total_beds = sum(bed_rooms, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(
    id_cols = tmk,
    names_from = tax_year, 
    values_from = total_beds, 
    names_sort = TRUE
  ) %>% 
  select(tmk, matches("^[0-9]{4}")) %>% 
  rename_with(~ paste0("bedrooms_", .x), matches("^[0-9]{4}"))

# --- Table 3: OSDS Class Baseline ---
maui_otp_unchanged_table3 <- maui_unchgd_encoded_dwell %>% 
  mutate(year = 2010) %>% 
  select(tmk, year, encoded_sum) %>% 
  distinct() %>% 
  pivot_wider(
    id_cols = tmk,
    names_from = year, 
    values_from = encoded_sum,
    names_prefix = "osds_class_" # Standardized prefix matching Oahu
  )

# --- Table 4: Commercial Classification ---
maui_otp_unchanged_table4 <- maui_unchanged %>% 
  select(tmk, any_of(c("commercial", "pittcode", "pitt_code", "x", "y"))) %>% 
  rename(pitt_code = any_of(c("pittcode", "pitt_code"))) %>% 
  filter(!is.na(commercial) | !is.na(pitt_code)) %>%
  mutate(comm_assignment = case_when(
    pitt_code == 0                                ~ "Timeshare (Maui only)",
    pitt_code == 999                              ~ "Multiple PITT Codes",
    pitt_code >= 1000                             ~ "Commercialized Residential (Maui only)",
    pitt_code >= 100 & pitt_code < 200            ~ "Improved Residential",
    pitt_code >= 200 & pitt_code < 300            ~ "Apartment",
    pitt_code >= 300 & pitt_code < 400            ~ "Commercial",
    pitt_code >= 400 & pitt_code < 500            ~ "Industrial",
    pitt_code >= 500 & pitt_code < 600            ~ "Agricultural and Rural",
    pitt_code >= 600 & pitt_code < 700            ~ "Conservation",
    pitt_code >= 700 & pitt_code < 800            ~ "Hotel and Resort",
    pitt_code >= 800 & pitt_code < 900            ~ "Unimproved Residential",
    pitt_code >= 900 & pitt_code < 999            ~ "Homeowner",
    TRUE                                          ~ "Unknown"
  )) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# EXPORT INTERMEDIATE UNCHANGED FILES
# ==============================================================================

write_csv(
  maui_otp_unchanged_table1,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table1_tmk_registry_UNCHANGED.csv")
)

write_csv(
  maui_otp_unchanged_table2,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table2_bedrooms_UNCHANGED.csv")
)

write_csv(
  maui_otp_unchanged_table3,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table3_classes_UNCHANGED.csv")
)

write_csv(
  maui_otp_unchanged_table4,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table4_commercial_UNCHANGED_draft.csv")
)

```

#### Updates

These tmks are ones are present in OTP with OSDS recoreded (ie had an osds permit befor 2010) but also in the new data, meaning they have applied for a new permit since 2010 as well.  

```{r updated}

library(tidyverse)

# ==============================================================================
# 1. BEDROOM GAP-FILL LOGIC (MAUI)
# ==============================================================================

# Read step 2 new bedrooms sum and standardize TMK
table_2_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table2a_newbedrooms_sum.csv"),
  col_types = cols(tmk = col_character())
) %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  ) %>% 
  rename_with(~ paste0("newbeds", .x), !tmk)

maui_updated_forjoin <- maui_updated %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  ) %>% 
  select(tmk, any_of(c("pittcode", "pitt_code")), commercial, ttl_bdrms) %>% 
  rename(
    otp_ttl_bdrms = ttl_bdrms,
    pitt_code = any_of(c("pittcode", "pitt_code"))
  ) %>% 
  mutate(comm_assignment = case_when(
    pitt_code == 0                                ~ "Timeshare (Maui only)",
    pitt_code == 999                              ~ "Multiple PITT Codes",
    pitt_code >= 1000                             ~ "Commercialized Residential (Maui only)",
    pitt_code >= 100 & pitt_code < 200            ~ "Improved Residential",
    pitt_code >= 200 & pitt_code < 300            ~ "Apartment",
    pitt_code >= 300 & pitt_code < 400            ~ "Commercial",
    pitt_code >= 400 & pitt_code < 500            ~ "Industrial",
    pitt_code >= 500 & pitt_code < 600            ~ "Agricultural and Rural",
    pitt_code >= 600 & pitt_code < 700            ~ "Conservation",
    pitt_code >= 700 & pitt_code < 800            ~ "Hotel and Resort",
    pitt_code >= 800 & pitt_code < 900            ~ "Unimproved Residential",
    pitt_code >= 900 & pitt_code < 999            ~ "Homeowner",
    TRUE                                          ~ "Unknown"
  ))

table_2_updated <- table_2_new %>% 
  filter(tmk %in% maui_updated_forjoin$tmk)

updates_notInTable2 <- setdiff(maui_updated_forjoin$tmk, table_2_updated$tmk)

table4_withUpdates <- maui_updated_forjoin %>% 
  filter(tmk %in% updates_notInTable2) 

maui_otp_updated_withNewbeds <- maui_updated_forjoin %>% 
  left_join(table_2_updated, by = "tmk") %>% 
  filter(otp_ttl_bdrms != 0) 

year_cols <- maui_otp_updated_withNewbeds %>% select(starts_with("newbeds")) %>% names()
latest_bed_col <- if (length(year_cols) > 0) tail(year_cols, 1) else NULL

maui_otp_updaed_withBedFlags <- maui_otp_updated_withNewbeds %>% 
  mutate(
    match_index = {
      match_matrix <- pick(all_of(year_cols)) == otp_ttl_bdrms
      match_matrix[is.na(match_matrix)] <- FALSE 
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    match_name = year_cols[match_index],
    
    # Safely evaluate dynamic latest year (replaces hardcoded newbeds2023)
    latest_val = if (!is.null(latest_bed_col)) .data[[latest_bed_col]] else NA_real_,
    bed_trend = case_when(
      is.na(latest_val)             ~ NA_character_,
      otp_ttl_bdrms < latest_val    ~ "Expansion since OTP",
      otp_ttl_bdrms > latest_val    ~ "Shrink - double check",
      otp_ttl_bdrms == latest_val   ~ "Perfect Match",
      TRUE                          ~ NA_character_
    ),

    data_presence = if_else(
      if_all(starts_with("newbeds"), ~ .x == 0 | is.na(.x)),
      "Empty (All 0 or NA)",
      "Bedrooms Recorded"
    )
  )

# Get combined build years
year_built <- maui_dwell_alltmks %>% 
  filter(tmk %in% maui_otp_updaed_withBedFlags$tmk) %>% 
  select(tmk, year_built) %>%
  filter(!is.na(year_built)) %>% 
  distinct() %>% 
  group_by(tmk) %>%
  summarize(
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    count_buildyears = n(),
    .groups = "drop"
  ) %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  )

maui_updated_with_BEDflags_gapfilled <- maui_otp_updaed_withBedFlags %>% 
  left_join(year_built, by = "tmk") %>%
  mutate(
    gapfill_flag = if_else(
      data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0, 
      "Gapfilled with OTP value", 
      "Used dwell data"
    )
  ) %>%
  mutate(across(starts_with("newbeds"), ~ {
    fill_condition <- data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0
    if_else(fill_condition, as.numeric(otp_ttl_bdrms), as.numeric(.x))
  })) %>%
  mutate(across(starts_with("newbeds"), ~ {
    col_year <- as.numeric(str_extract(cur_column(), "\\d{4}"))
    min_build_yr <- as.numeric(str_extract(year_built_combined, "^\\d{4}"))
    
    case_when(
      gapfill_flag == "Used dwell data" ~ as.numeric(.x),
      is.na(min_build_yr) | min_build_yr <= 2010 ~ as.numeric(.x),
      min_build_yr > col_year ~ NA_real_,
      TRUE ~ as.numeric(.x)
    )
  }))

table_2_updatedGapfilled <- maui_updated_with_BEDflags_gapfilled %>% 
  select(tmk, starts_with("newbeds"))

bedFlags_forTable1 <- maui_updated_with_BEDflags_gapfilled %>% 
  select(
    tmk,
    bed_trend,
    data_presence,
    year_built_combined,
    count_buildyears,
    gapfill_flag
  )


write_csv(bedFlags_forTable1,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_bedflags_fortable1.csv"))


# ==============================================================================
# 2. OSDS CLASS GAP-FILL & UPGRADE LOGIC (MAUI)
# ==============================================================================

table_3_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/maui_table3_newclass_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(
    tmk = if ("full_taxid" %in% names(.)) full_taxid else tmk,
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  ) %>% 
  select(-any_of("full_taxid")) %>% 
  rename_with(~ paste0("newclass", .x), !tmk)

maui_updated_withClass_enc <- maui_updated %>% 
  mutate(
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  ) %>% 
  select(tmk, osds_qty, starts_with("class")) %>%
  select(-any_of(c("class"))) %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(!is.na(quantity) & quantity != 0 & quantity != -9999) %>% 
  rename(ttl_osds_2010 = osds_qty) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv"  ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii"  ~ 2,
    class == "class_i"   ~ 1,
    TRUE                 ~ NA_real_
  )) %>%
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "class", 
    values_from = "osds_class_otp"
  ) %>%
  mutate(otp_encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

maui_updated_withClassEnc_New <- maui_updated_withClass_enc %>%
  left_join(table_3_new, by = "tmk")

class_cols <- maui_updated_withClassEnc_New %>% select(starts_with("newclass")) %>% names()

table3_updated_classmatch <- maui_updated_withClassEnc_New %>% 
  mutate(
    match_index = {
      match_matrix <- pick(all_of(class_cols)) == otp_encoded_sum
      match_matrix[is.na(match_matrix)] <- FALSE 
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    match_name = class_cols[match_index],
    data_presence = if_else(
      if_all(starts_with("newclass"), ~ .x == "0" | is.na(.x) | .x == 0),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )
  )

table_1_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(
    tmk = if ("full_taxid" %in% names(.)) full_taxid else tmk,
    tmk = sprintf("%.0f", as.numeric(tmk)),
    tmk = str_pad(tmk, width = 9, side = "left", pad = "0")
  )

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  select(
    tmk, otp_encoded_sum, 
    any_of(c("min_year", "max_year", "n_addresses")), 
    starts_with("newclass")
  ) %>% 
  mutate(
    legacy_val = as.numeric(otp_encoded_sum),
    v2009 = if ("newclass2009" %in% names(.)) as.numeric(newclass2009) else NA_real_,
    v2010 = if ("newclass2010" %in% names(.)) as.numeric(newclass2010) else NA_real_,
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010)                 ~ "no data to compare",
      TRUE                                       ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = as.character(case_when(
    scenario_match_flag == "matched 2009 or 2010"               ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE                                                        ~ NA_real_
  ))) %>% 
  select(
    tmk, newpre_2009, 
    starts_with("newclass200"), starts_with("newclass201"), starts_with("newclass202"), 
    everything()
  ) %>% 
  rowwise() %>% 
  mutate(unique_story = {
    # Fix: select strictly pre2009 and permit class columns to build story string
    vals <- c_across(c(newpre_2009, starts_with("newclass")))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>% 
  ungroup() %>% 
  mutate(
    upgrade_scenario = case_when(
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1"  ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2"    ~ "septic and aerobic - replacement of aerobic",
      unique_story == "3|4"    ~ "septic combo to aerobic",
      unique_story == "1|4"    ~ "single septic to aerobic",
      unique_story == "7|4"    ~ "complex septic system to aerobic",
      unique_story == "8|5"    ~ "cesspool to septic / aerobic hybrid",
      
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1234]$")           ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$")             ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$")             ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1234]$")  ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$")       ~ "multiple osds with new cesspool application",
      
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$")  ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$")        ~ "lateral change (clean to clean)",
      
      TRUE ~ "other scenario"
    )
  )

table_3_updated_new <- update_logic_refined_cl %>% 
  select(tmk, newpre_2009, starts_with("newclass"))

class_flag_forTable1 <- update_logic_refined_cl %>% 
  select(tmk, class_match_flag = scenario_match_flag, unique_story, upgrade_scenario)

write_csv(class_flag_forTable1,
          file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_classflags_fortable1.csv"))

```

### Step 4: Assemble Final Tables by Island

To run this, need to run STEP 0. Otherwise all dependencies should be loaded here.

#### Complete Tables 1 - 3a, 4

```{r complete-tables}

'%!in%' <- Negate('%in%')

# helper function to guarantee 9-digit Maui TMKs starting with '2'
clean_maui_tmk <- function(df, col_name = "tmk") {
  df %>% 
    mutate(
      !!sym(col_name) := sprintf("%.0f", as.numeric(!!sym(col_name))),
      !!sym(col_name) := str_pad(!!sym(col_name), width = 9, side = "left", pad = "0")
    )
}


# ---- Table 1 --------------
table_1_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(tmk = if ("full_taxid" %in% names(.)) full_taxid else tmk) %>% 
  clean_maui_tmk("tmk") %>% 
  rename(in_dwelldat = any_of(c("is_dwelling", "in_dwelldat"))) %>% 
  mutate(
    status_since_otp = "New Permit",
    n_addresses = as.character(n_addresses)
  ) %>% 
  select(-any_of("full_taxid")) %>% 
  distinct(tmk, .keep_all = TRUE)

table_1_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table1_tmk_registry_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_maui_tmk("tmk") %>% 
  mutate(scenario = "No new permits") %>% 
  distinct(tmk, .keep_all = TRUE)


bedFlags_forTable1 <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_bedflags_fortable1.csv")) %>% 
    mutate(tmk = as.character(tmk)) %>%
  select(tmk,
        bed_gapfill_flag = gapfill_flag)

classFlags_forTable1 <- read_csv(file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_classflags_fortable1.csv")) %>% 
  mutate(tmk = as.character(tmk))

# Combine registry sources
table_1_complete <- bind_rows(table_1_new, table_1_unchanged) %>% 
  distinct(tmk, .keep_all = TRUE)

# Attach bedroom and class gap-fill flags
table_1_complete_flags <- table_1_complete %>% 
  #left_join(bedFlags_forTable1, by = "tmk") %>% 
  left_join(classFlags_forTable1, by = "tmk") %>% 
  select(-upgrade_scenario.x) %>% 
  rename(upgrade_scenario = upgrade_scenario.y) %>% 
  distinct(tmk, .keep_all = TRUE)

# ------------ Table 2 --------------------- #

table_2_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table2_bedrooms_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_maui_tmk("tmk") %>% 
  mutate(across(-tmk, as.numeric))

table_2_updated_forbind <- table_2_updatedGapfilled %>% 
  clean_maui_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds"))

table_2_newonly <- table_2_new %>% 
  clean_maui_tmk("tmk") %>% 
  filter(tmk %!in% table_2_updatedGapfilled$tmk) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds"))

table_2_complete <- bind_rows(
  table_2_unchanged, 
  table_2_updated_forbind, 
  table_2_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)


# ------------ Table 3a --------------------- #


table_3_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table3_classes_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_maui_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^osds_cl_", "osds_class_"), starts_with("osds_cl_"))

table_3_updatedforbind <- table_3_updated_new %>% 
  clean_maui_tmk("tmk") %>% 
  rename(osds_class_pre2009 = newpre_2009) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass"))

table_3_newonly <- table_3_new %>% 
  clean_maui_tmk("tmk") %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass"))

table_3_complete <- bind_rows(
  table_3_updatedforbind, 
  table_3_unchanged, 
  table_3_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)


# ------------ Table 4a --------------------- #

table_4_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/maui_table4_commercial_UNCHANGED_draft.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_maui_tmk("tmk") %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "pittcode", "x", "y"))) %>% 
  rename(pitt_code = any_of(c("pittcode", "pitt_code"))) %>% 
  mutate(across(everything(), as.character))

table_4_withUpdates_forbind <- table4_withUpdates %>% 
  clean_maui_tmk("tmk") %>% 
  filter(!is.na(commercial) | !is.na(pitt_code)) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_commercial.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(tmk = if ("full_taxid" %in% names(.)) full_taxid else tmk) %>% 
  clean_maui_tmk("tmk") %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk) %>% 
  mutate(
    commercial = "unknown",
    pitt_code = "-9999"
  ) %>% 
  mutate(pitt_code = as.character(pitt_code)) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_complete <- bind_rows(
  table_4_unchanged,
  table_4_new,
  table_4_withUpdates_forbind
) %>% 
  distinct(tmk, .keep_all = TRUE)

table_4_complete_cl <- table_4_complete %>% 
  mutate(
    pitt_code = if_else(is.na(pitt_code), "-9999", pitt_code),
    commercial = if_else(is.na(commercial) | commercial == "unknown", "check google places", commercial)
  )


# ------------------------------------------------------------------------------
# 5. FINAL RECONCILIATION & TABLE 1 OVERRIDES
# ------------------------------------------------------------------------------

# Identify commercial parcels that actually have residential bedroom data
shouldbein_table_4 <- setdiff(table_1_complete_flags$tmk, table_2_complete$tmk)
shouldbe_intable2 <- setdiff(table_4_complete_cl$tmk, shouldbein_table_4)

# Exclude residential properties from Table 4
table_4_complete_cl <- table_4_complete_cl %>% 
  filter(tmk %!in% shouldbe_intable2)

# Build Final Table 1 Master
table_1_final_final <- table_1_complete_flags %>% 
  mutate(use_table_4 = tmk %in% table_4_complete_cl$tmk) %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4 = use_table4
  ) %>% 
  distinct(tmk, .keep_all = TRUE)


# ------------------------------------------------------------------------------
# 6. AUTOMATED INTEGRITY CHECKS
# ------------------------------------------------------------------------------

cat("--- MAUI PIPELINE INTEGRITY CHECKS ---\n")
cat("1. All Table 1 TMKs present in Table 3a:", all(table_1_final_final$tmk %in% table_3_complete$tmk), "\n")
cat("2. All Table 2 TMKs present in Table 1:", all(table_2_complete$tmk %in% table_1_final_final$tmk), "\n")
if (exists("maui_master_tmk_list")) {
  cat("3. All Master TMKs present in Table 1:", all(maui_master_tmk_list %in% table_1_final_final$tmk), "\n")
}


# ------------------------------------------------------------------------------
# 7. EXPORT STATEWIDE-COMPATIBLE CSVs
# ------------------------------------------------------------------------------

write_csv(
  table_1_final_final,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step4/maui_table1_tmk_registry_ALL.csv")
)

write_csv(
  table_2_complete,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step4/maui_table2_bedrooms_ALL.csv")
)

write_csv(
  table_3_complete,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step4/maui_table3a_permitsApplied_ALL.csv")
)

write_csv(
  table_4_complete_cl,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step4/maui_table4_commercial_draft.csv")
)



```


#### Create table 3b

The additive encoding is what is known as a binary bit mask. Each class represents a power of 2, so every unique sum corresponds to exactly one combination of classes. 

We can think of table 3a as an "event" table, since it records the permit application event. However, we want to create a "status" table, which records the actual OSDS present on the property in each year, BASED on the events we know happened. This will be table 3b

We can make multiple versions of this, using different math. In the first, I'll assume that:
1. If there was only one OSDS on the property, the new application is replacing it
2. If there were multiple OSDS on the property, the new application is replacing the grossest, ie highest class. This correspond to the largest power of 2 within the encoded value

This translates to the following math: 
1. Identify the largest power of 2 inside the current value.
2. Subtract that power of 2.
3. Add the new application value.

```{r}

# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% 
  rename(osds_class_2008 = osds_class_pre2009) %>% # note, this is a stand in for "pre 2009"
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(as.numeric(Applied),0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/maui_table3b_activeOSDS.csv"))



```


