---
title: "Maui OSDS Timeseries"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))


```

### Step 0: Create authoritative tmk list

To start, we'll read in the iws data and the old otp data and create 4 lists:
- unchanged since OTP
- updated and present in both data sets
- new permits since OTP
- ALL tmks in both dataset - authortitative datasets


* we'll also clean up the datasets by standardizing the tmk column 

```{r}

# new data: iws permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk,"2"))

iws_maui_permits_tmk <- iws_maui_permits %>% pull(tmk)


# old data: otp 

# read in maui and molokai. check lanai?
otp_bob_maui <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Maui.csv")) %>% 
  clean_names()

otp_bob_molo <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Molokai.csv")) %>% 
  clean_names()

otp_maui <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk))

otp_maui_tmk <- otp_maui %>% 
  pull(tmk)

# new ones:

maui_unchanged <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% iws_maui_permits_tmk) # 12857

unchanged_tmks <- maui_unchanged %>% pull(tmk) %>% unique() #12857

maui_updated <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_maui_permits_tmk) #1574

updated_tmks <- maui_updated %>% pull(tmk) %>% unique() # 1574

maui_new <- iws_maui_permits %>% 
  filter(tmk %!in% otp_maui_tmk) # 837

new_tmks <- maui_new %>% pull(tmk) %>% unique() #659
# create authoritative tmk list - add ups!

master_tmk_list <- c(iws_maui_permits_tmk,otp_maui_tmk)
master_tmk_list <- unique(master_tmk_list) # 15090


```


### Step 1: Clean Up Dwell Data

The Maui data comes to us pre-parsed, which is great. Here we will just generate the same summary tables as we did for Oahu. We will have to read in each excel file here and do some cleaning. 

Maui specific note: Iʻm not sure the condo type and stype codes are the same as for Oahu. But if they are, the following is what condo style means:

 1 = SINGLE FAMILY
 2 = DUPLEX
 3 = TOWNHOUSE
 4 = LOWRISE
 5 = HIGHRISE

Resources: 

https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/

```{r}

library(readxl)

# list files in the dropbox folder

maui_dwell_fp <- list.files(path = file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Maui_DwellRaw_12112025"),
                            full.names =  T,
                            pattern = ".XLSX")


file_types <- sapply(maui_dwell_fp, excel_format)

check <- data.frame(
  file = basename(maui_dwell_fp),
  format = file_types
)


maui_dwell_dfs <- lapply(maui_dwell_fp, read_excel,
                         skip =1,
                         col_names = TRUE)

# just look at one example:

df_2009 <- maui_dwell_dfs[[1]]


# 
collapse_duplicates <- function(df){
  
  #df <- maui_dwell_dfs[[1]]
  df_collapse <- df %>%  janitor::clean_names() %>% select(-number, -percent_complete) %>% distinct() # this will only remove exact duplicates
  return(df_collapse)
}


maui_collapsed_dfs <- lapply(maui_dwell_dfs, collapse_duplicates)

maui_dwell_allentries <- do.call(rbind,maui_collapsed_dfs) %>% 
    mutate(
      parid = str_trim(as.character(parid)),
      tax_id = substr(parid,1,8),
      full_taxid = paste0("2",tax_id)
      )

# create a version with just the tax ids that we can use to filter the permit data into "dwells" and "non dwells"
maui_dwelltmks_forjoin_noTXyear <- maui_dwell_allentries %>%
  select(-tax_year) %>%
  mutate(
    parid = str_trim(as.character(parid)),
    tax_id = substr(parid,1,8),
    full_taxid = paste0("2",tax_id)) %>%
  distinct()


```


### Step 2: Clean up in IWS data

```{r}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

# filter out just the permits in maui, and create permit ID tracker

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"2")) %>% 
  rename(full_taxid = tmk_char) %>%  
  rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) #2790 rows, 2233 unique tmks

tmks_permits <- unique(iws_maui_permits$full_taxid)

# execute join to see which tmks have dwell data. then, do the same using a filter and make sure the number of unique tmks match

iws_maui_dwell_join <- inner_join(iws_maui_permits, maui_dwelltmks_forjoin_noTXyear, by = "full_taxid") %>% 
  distinct() #8553, 2094 tmks

iws_maui_with_dwell <- iws_maui_permits %>% 
  filter(full_taxid %in% maui_dwelltmks_forjoin_noTXyear$full_taxid) #2094 tmks 

# now filter out the ones that do not have dwelling data and save

iws_maui_NOdwell <- iws_maui_permits %>% 
  filter(full_taxid %!in% iws_maui_dwell_join$full_taxid) # 191 rows, #139 tmks

write_csv(iws_maui_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/maui_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**:
- There were 2790 permits issued for HI island over the past 10 years
- 2599 have dwelling information, and this represents 2094 tmks
- 139 tmks do not have dwelling information

### Step 2b: Clean and Pivot Permit Data
Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 

```{r}

library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by separating out a year column, adding class, and standarizing the address

master_clean <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

# N rows = 2597 after distinct, 2790 before. Lost 193
# N tmks = 2233. Matches original permit list after distinct

master_clean2 <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) 

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #

### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(full_taxid) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(full_taxid %in% maui_dwelltmks_forjoin_noTXyear$full_taxid),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- maui_dwell_allentries %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    total_beds = sum(bed_rooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)

table2b_bedrooms <- maui_dwell_allentries %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bed_rooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)


#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(full_taxid, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(full_taxid, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)


# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2a_bedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2b_bedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3_class_allcsv"))



```


### Step 3: Compare and Integrate Past Datasets

In this step we have to add in the old data, but differentiate between data that has been updated, unchanged. We also have to gapfill the bedroom and class data, because the old data is just a shot in time. 

#### Unchanged

These have no new permit data. But, it would be nice to get a better understanding of their bedroom # over time. We have to make one assumption in order to merge our dwell data with this older data. 

Here, we'll join the bedroom data to the unchanged otp tmks. Then, sort into our 3 tables as before.
We'll save these ALONE just as the unchanged tmks

```{r}




```

#### Updates

These ones are present in OTP but also in the new data. 

## Step 4: Create table 3b
