---
title: "Maui OSDS Timeseries"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))


```

### Step 0: SET UP and Create authoritative tmk lists

To start, we'll read in the iws data and the old otp data and create 4 lists:
- unchanged since OTP
- updated and present in both data sets
- new permits since OTP
- ALL tmks in both dataset - authortitative datasets


* we'll also clean up the datasets by standardizing the tmk column 

```{r}

# new data: iws permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk,"2"))

iws_maui_permits_tmk <- iws_maui_permits %>% pull(tmk)


# old data: otp 

# read in maui and molokai. check lanai?
otp_bob_maui <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Maui.csv")) %>% 
  clean_names()

otp_bob_molo <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Molokai.csv")) %>% 
  clean_names()

otp_maui <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk))

otp_maui_tmk <- otp_maui %>% 
  pull(tmk)

# new ones:

maui_unchanged <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% iws_maui_permits_tmk) # 12857

unchanged_tmks <- maui_unchanged %>% pull(tmk) %>% unique() #12857

maui_updated <- rbind(otp_bob_maui, otp_bob_molo) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_maui_permits_tmk) #1574

updated_tmks <- maui_updated %>% pull(tmk) %>% unique() # 1574

maui_new <- iws_maui_permits %>% 
  filter(tmk %!in% otp_maui_tmk) # 837

new_tmks <- maui_new %>% pull(tmk) %>% unique() #659
# create authoritative tmk list - add ups!

master_tmk_list <- c(iws_maui_permits_tmk,otp_maui_tmk)
master_tmk_list <- unique(master_tmk_list) # 15090


```


### Step 1: Clean Up Dwell Data

The Maui data comes to us pre-parsed, which is great. Here we will just generate the same summary tables as we did for Oahu. We will have to read in each excel file here and do some cleaning. 

Maui specific note: Iʻm not sure the condo type and stype codes are the same as for Oahu. But if they are, the following is what condo style means:

 1 = SINGLE FAMILY
 2 = DUPLEX
 3 = TOWNHOUSE
 4 = LOWRISE
 5 = HIGHRISE

Resources: 

https://thepapakeacollection.com/2023/03/18/understanding-the-tax-map-key-system-in-hawaii-a-comprehensive-guide/

```{r}

library(readxl)

# list files in the dropbox folder

maui_dwell_fp <- list.files(path = file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Maui_DwellRaw_12112025"),
                            full.names =  T,
                            pattern = ".XLSX")


file_types <- sapply(maui_dwell_fp, excel_format)

check <- data.frame(
  file = basename(maui_dwell_fp),
  format = file_types
)


maui_dwell_dfs <- lapply(maui_dwell_fp, read_excel,
                         skip =1,
                         col_names = TRUE)

# just look at one example:

df_2009 <- maui_dwell_dfs[[1]]


# 
collapse_duplicates <- function(df){
  
  #df <- maui_dwell_dfs[[1]]
  df_collapse <- df %>%  janitor::clean_names() %>% select(-number, -percent_complete) %>% distinct() # this will only remove exact duplicates
  return(df_collapse)
}


maui_collapsed_dfs <- lapply(maui_dwell_dfs, collapse_duplicates)

maui_dwell_allentries <- do.call(rbind,maui_collapsed_dfs) %>% 
    mutate(
      parid = str_trim(as.character(parid)),
      tax_id = substr(parid,1,8),
      full_taxid = paste0("2",tax_id)
      )

# create a version with just the tax ids that we can use to filter the permit data into "dwells" and "non dwells"
maui_dwelltmks_forjoin_noTXyear <- maui_dwell_allentries %>%
  select(-tax_year) %>%
  mutate(
    parid = str_trim(as.character(parid)),
    tax_id = substr(parid,1,8),
    full_taxid = paste0("2",tax_id)) %>%
  distinct()

# create a version with just the TMKS we need from our tmk master list 

maui_dwell_alltmks <- maui_dwell_allentries %>% 
  filter(full_taxid %in% master_tmk_list)

write_csv(maui_dwell_alltmks,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/maui_dwell_clean_alltmks.csv"))
```


### Step 2: Clean up in IWS data

```{r}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

# filter out just the permits in maui, and create permit ID tracker

iws_maui_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"2")) %>% 
  rename(full_taxid = tmk_char) %>%  
  rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) #2790 rows, 2233 unique tmks

tmks_permits <- unique(iws_maui_permits$full_taxid)

# execute join to see which tmks have dwell data. then, do the same using a filter and make sure the number of unique tmks match

iws_maui_dwell_join <- inner_join(iws_maui_permits, maui_dwelltmks_forjoin_noTXyear, by = "full_taxid") %>% 
  distinct() #8553, 2094 tmks

iws_maui_with_dwell <- iws_maui_permits %>% 
  filter(full_taxid %in% maui_dwelltmks_forjoin_noTXyear$full_taxid) #2094 tmks 

# now filter out the ones that do not have dwelling data and save

iws_maui_NOdwell <- iws_maui_permits %>% 
  filter(full_taxid %!in% iws_maui_dwell_join$full_taxid) # 191 rows, #139 tmks

write_csv(iws_maui_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/maui_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**:
- There were 2790 permits issued for HI island over the past 10 years
- 2599 have dwelling information, and this represents 2094 tmks
- 139 tmks do not have dwelling information

### Step 2b: Clean and Pivot Permit Data
Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 

##### Wastewater Class Encoding Reference Table

We only have classes 1 - 4 in this data. From the metadata, this
represents:

-   CLASS_I: Class I OSDS are those systems where the effluent is
    subjected to soil treatment
-   CLASS_II: Class II OSDS are septic systems discharging to a seepage
    pit.\
-   CLASS_III: Class III OSDS are aerobic systems discharging to a
    seepage pit.
-   CLASS_IV: Class IV OSDS are cesspools.\
-   EFFLUENT: The total effluent flux in million gallons per day
    discharged within the TMK parcel.
    

| Detected Classes | Encoded Sum | Description |
| :--- | :---: | :--- |
| **Class 1** | 1 | Cleanest Septic Type |
| **Class 2** | 2 | Secondary Septic Type |
| **Class 1 + 2** | 3 | **Combination of Class 1 + Class 2** |
| **Class 3** | 4 | Aerobic System |
| **Class 1 + 3** | 5 | Class 1 Septic + Aerobic |
| **Class 2 + 3** | 6 | Class 2 Septic + Aerobic |
| **Class 1 + 2 + 3** | 7 | Classes 1, 2, and 3 |
| **Class 4** | 8 | Cesspool |
| **Class 1 + 4** | 9 | Class 1 + Class 4 |
| **Class 2 + 4** | 10 | Class 2 + Class 4 |
| **Class 1 + 2 + 4** | 11 | Classes 1, 2, and 4 |
| **Class 3 + 4** | 12 | Aerobic + Class 4 |
| **Class 1 + 3 + 4** | 13 | Classes 1, 3, and 4 |
| **Class 2 + 3 + 4** | 14 | Classes 2, 3, and 4 |
| **All (1, 2, 3, 4)** | 15 | Full suite of permits |

```{r}

library(janitor)
# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}


# Define disposal types 
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function 
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1" # Catch-all default based on your manual fix
  )
}

## clean up permits by separating out a year column, adding class, and standarizing the address

master_clean <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) %>% 
  distinct() # this removes any EXACT duplicates. 

# N rows = 2597 after distinct, 2790 before. Lost 193
# N tmks = 2233. Matches original permit list after distinct

master_clean2 <- iws_maui_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = ifelse(tmk %in% iws_maui_NOdwell$TMK,"no","yes")
  ) %>%
  select(-permit_id) 

duplicates_to_check <- master_clean2 %>%
  group_by(across(everything())) %>% 
  filter(n() > 1) %>% 
  arrange(address_full) 

# for now, I'm going to assume exact duplicates are duplicate data. but we could also include them as additional osds.

# now create the three tables - bedroom will not be relevant for the non commercial #

### TABLE 1 #########

table1_tmk_registry <- master_clean %>%
  group_by(full_taxid) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year),
    max_year = max(year),
    is_dwelling = any(full_taxid %in% maui_dwelltmks_forjoin_noTXyear$full_taxid),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building", # change this to cards
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

#### TABLE 2 A and 2 B ########

table2a_bedrooms <- maui_dwell_allentries %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    total_beds = sum(bed_rooms, na.rm = TRUE),
    # mode_beds = calc_mode(rmbed, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)

table2b_bedrooms <- maui_dwell_allentries %>%
  filter(full_taxid %in% master_clean$full_taxid) %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>%
  group_by(full_taxid, tax_year) %>%
  summarize(
    #total_beds = sum(bedrooms, na.rm = TRUE),
    mode_beds = calc_mode(bed_rooms, na.rm = TRUE), # Uncomment if you want mode instead
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE) %>% 
  rename(tmk = full_taxid)


#### TABLE 3 ####

table3_classes <- master_clean %>%
  group_by(full_taxid, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(encoded_sum = 
    (str_detect(all_classes_text, "1") * 1) +
    (str_detect(all_classes_text, "2") * 2) +
    (str_detect(all_classes_text, "3") * 4) +
    (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(full_taxid, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)


# Filter out ommercial, ie dwell = no

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

### Save

write_csv(table1_tmk_registry,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"))

write_csv(table1_tmk_registry_commercial,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_commercial.csv"))

write_csv(table2a_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2a_bedrooms_sum.csv"))

write_csv(table2b_bedrooms,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2b_bedrooms_mode.csv"))

write_csv(table3_classes,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3_class_allcsv"))



```


### Step 3: Compare and Integrate Past Datasets

In this step we have to add in the old data, but differentiate between data that has been updated, unchanged. We also have to gapfill the bedroom and class data, because the old data is just a shot in time. 

#### Unchanged

These have no new permit data. But, it would be nice to get a better understanding of their bedroom # over time. We have to make one assumption in order to merge our dwell data with this older data. 

Here, 
1. apply class encoding 
2. we'll join the bedroom data to the unchanged otp tmks. 
3. Then, sort into our 3 tables as before.
We'll save these ALONE just as the unchanged tmks

We also use pittcode here to help us filter stuff out

```{r unchanged}

# first read in the dwelling data for all maui

 maui_dwell_alltmks <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/dwellings/maui_dwell_clean_alltmks.csv")) %>% 
  mutate(tmk = as.character(full_taxid))

# encode classes

maui_unchnaged_encoded <- maui_unchanged %>% 
  pivot_longer(cols = starts_with("class"),
               names_to = "class",
               values_to = "quantity") %>% 
  filter(quantity != 0) %>% 
  filter(quantity != -9999) %>% 
  rename(ttl_osds_2010 = osds_qty) %>%
  mutate(osds_class_otp = case_when(
    class == "class_iv" ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii" ~ 2,
    class == "class_i" ~ 1
  )) %>%
  pivot_wider(id_cols = c("tmk","ttl_osds_2010","pittcode"), names_from = "class", values_from = c("osds_class_otp")) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

# 2. join to dwells

maui_unchgd_encoded_dwell <- left_join(maui_unchnaged_encoded, maui_dwell_alltmks, by = "tmk") %>% 
  distinct()

# 3. Create the three tables 

maui_card_dwell_summary <-  maui_unchgd_encoded_dwell %>% 
  group_by(tmk) %>% 
  summarize(
    n_buildings = n_distinct(card),
    in_dwelldat = any(!is.na(tax_year))
  ) %>% 
  ungroup() 

maui_otp_unchanged_table1 <- maui_unchgd_encoded_dwell %>%
  mutate(n_permits = case_when(
    encoded_sum %in% c(1,2,4,8) ~ 1,
    encoded_sum %in% c(3,5,6,9,10,12) ~ 2,
    encoded_sum %in% c(7,11,13,14) ~ 3,
    encoded_sum == 15 ~ 4,
    TRUE ~ 99999),
    n_years = 1,
    min_year = 2010,
    max_year = 2010,
    n_addresses = "unknown",
    status_since_otp = "unchanged since otp"
  ) %>%
  mutate(
      scenario = case_when(
      n_permits == 1 & n_years == 1 ~ "Simple",
      n_addresses == "unknown" ~ "Unknown",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    )) %>% 
  left_join(maui_card_dwell_summary) %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp
  )
  
# ----------- Table 2 ------------------- # 

# no commercial here - removed at the end of this code chunk

maui_otp_unchanged_table2 <- maui_unchgd_encoded_dwell %>%
  mutate(bed_rooms = as.numeric(bed_rooms)) %>% 
  group_by(tmk, tax_year) %>% 
  summarize(total_beds = sum(bed_rooms, na.rm = TRUE),
            .groups = "drop") %>% 
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE) %>% 
  select(-`NA`) %>% 
  filter(!if_all(2:last_col(), is.na)) %>% 
  rename_with(~ paste0("bedrooms_", .), matches("^[0-9]"))
  


# --------- Table 3 ---------------- #

maui_otp_unchanged_table3 <- maui_unchgd_encoded_dwell %>% 
  mutate(year = 2010) %>% 
  select(tmk,year, encoded_sum) %>% 
  distinct() %>% 
  pivot_wider(id_cols = tmk,names_from = year, values_from = encoded_sum) %>% 
  rename_with(~ paste0("osds_cl_", .), matches("^[0-9]"))
  
  
# ---------- Table 4 -------------- #

# these should include all marked as commercial in the data + any we dont have dwell data for. ideally these lists are the same thing. 
maui_otp_unchanged_table4 <- maui_unchanged %>% 
  select(tmk, commercial, pitt_code = pittcode) %>% 
  filter(!is.na(commercial)) %>%
  mutate(comm_assignment = case_when(
    # Exact Matches (High Priority)
    pitt_code == 0    ~ "Timeshare (Maui only)",
    pitt_code == 999  ~ "Multiple PITT Codes",
    
    # kauai Specifics (Assumes you have a 'county' column)
    # county == "Kauai" & pitt_code == 100 ~ "Improved and Unimproved Residential",
    # county == "Kauai" & pitt_code == 800 ~ "Homestead Properties",

    # Range Matches (The "Hundreds")
    pitt_code >= 1000 ~ "Commercialized Residential (Maui only)",
    pitt_code >= 100 & pitt_code < 200 ~ "Improved Residential",
    pitt_code >= 200 & pitt_code < 300 ~ "Apartment",
    pitt_code >= 300 & pitt_code < 400 ~ "Commercial",
    pitt_code >= 400 & pitt_code < 500 ~ "Industrial",
    pitt_code >= 500 & pitt_code < 600 ~ "Agricultural and Rural",
    pitt_code >= 600 & pitt_code < 700 ~ "Conservation",
    pitt_code >= 700 & pitt_code < 800 ~ "Hotel and Resort",
    pitt_code >= 800 & pitt_code < 900 ~ "Unimproved Residential",
    pitt_code >= 900 & pitt_code < 999 ~ "Homeowner",
    
    # Catch-all for anything else
    TRUE ~ "Unknown"
  ))

#NOTE this hasnʻt been run yet! needs to be to finalize pittcode for known properties
maui_otp_unchanged_table4_all <- maui_otp_unchanged_table1 %>% 
  filter(in_dwelldat == FALSE)

# which are in the first, but not the second? Meaning, they have a commercial assignment but also bedrooms
difference <- setdiff(maui_otp_unchanged_table4$tmk,maui_otp_unchanged_table4_all$tmk)

difference_withno_info <- setdiff(maui_otp_unchanged_table4_all$tmk,
                                  maui_otp_unchanged_table4$tmk)

maui_tmks_forgoogleplaces <- maui_otp_unchanged_table1 %>% 
  filter(tmk %in% difference_withno_info)

# KEY POINT: for these, we'll keep them in tble 2 for now. Return to issue when I deal with Google places


  
# ------------- Final checks and save ----------------- #

write_csv(maui_otp_unchanged_table1,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_UNCHANGED.csv"))

write_csv(maui_otp_unchanged_table2,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2_classes_UNCHANGED.csv"))

write_csv(maui_otp_unchanged_table3,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3_classes_UNCHANGED.csv"))

write_csv(maui_otp_unchanged_table4,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_commercial_UNCHANGED_draft.csv"))

write_csv(maui_tmks_forgoogleplaces,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_googleplaces_UNCHANGED_draft.csv"))



```

#### Updates

These ones are present in OTP but also in the new data. 

```{r updated}


# ---------- gap fill bed rooms --------------- #
# read in the og table 2 with new info

table_2_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2a_bedrooms_sum.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename_with(., ~paste0("newbeds",.x), .cols= !tmk)

maui_updated_forjoin <- maui_updated %>% 
  select(tmk, pittcode, commercial,ttl_bdrms) %>% 
  rename(otp_ttl_bdrms = ttl_bdrms) %>%
  rename(pitt_code = pittcode) %>% 
  mutate(comm_assignment = case_when(
    # Exact Matches (High Priority)
    pitt_code == 0    ~ "Timeshare (Maui only)",
    pitt_code == 999  ~ "Multiple PITT Codes",
    
    # kauai Specifics (Assumes you have a 'county' column)
    # county == "Kauai" & pitt_code == 100 ~ "Improved and Unimproved Residential",
    # county == "Kauai" & pitt_code == 800 ~ "Homestead Properties",

    # Range Matches (The "Hundreds")
    pitt_code >= 1000 ~ "Commercialized Residential (Maui only)",
    pitt_code >= 100 & pitt_code < 200 ~ "Improved Residential",
    pitt_code >= 200 & pitt_code < 300 ~ "Apartment",
    pitt_code >= 300 & pitt_code < 400 ~ "Commercial",
    pitt_code >= 400 & pitt_code < 500 ~ "Industrial",
    pitt_code >= 500 & pitt_code < 600 ~ "Agricultural and Rural",
    pitt_code >= 600 & pitt_code < 700 ~ "Conservation",
    pitt_code >= 700 & pitt_code < 800 ~ "Hotel and Resort",
    pitt_code >= 800 & pitt_code < 900 ~ "Unimproved Residential",
    pitt_code >= 900 & pitt_code < 999 ~ "Homeowner",
    
    # Catch-all for anything else
    TRUE ~ "Unknown"
  )) # backfill with pittcode

# are all the updated ones in table 2? if they arenʻt are all those commercial?

table_2_updated <- table_2_new %>% 
  filter(tmk %in% maui_updated_forjoin$tmk) # missing 84 - check to see if those are commercial

updates_notInTable2 <- setdiff(maui_updated_forjoin$tmk, table_2_updated$tmk)

table4_withUpdates <- maui_updated_forjoin %>% 
  filter(tmk %in% updates_notInTable2) 

# like big island, some of these do have bedrooms. weʻll use that in our table

addons_fortable2 <- table4_withUpdates %>% 
  filter(is.na(commercial) & otp_ttl_bdrms != 0) %>% 
  select(tmk, pitt_assigment = comm_assignment, otp_ttl_bdrms)

# now join everything together and apply matching logic

maui_otp_updated_withNewbeds <- maui_updated_forjoin %>% 
  left_join(table_2_updated, by = "tmk") %>% 
  filter(otp_ttl_bdrms != 0) 

year_cols <- maui_otp_updated_withNewbeds %>% select(starts_with("new")) %>% names()
  
maui_otp_updaed_withBedFlags <- maui_otp_updated_withNewbeds %>% 
  mutate(
     # CREATE MATCH INDEX: 
    # This finds the index (1, 2, 3...) of the first column that matches OTP bedrooms
    # We use matrix comparison (df[cols] == df$otp) to keep it vectorized
    match_index = {
      match_matrix <- pick(all_of(year_cols)) == otp_ttl_bdrms
      # replace NAs with FALSE so max.col doesn't trip
      match_matrix[is.na(match_matrix)] <- FALSE 
      
      # If a row has at least one match, get the index. Otherwise, NA.
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    
    # CREATE MATCH NAME: Convert that index (e.g., 1) into the column name (e.g., "new_2010")
    match_name = year_cols[match_index],
    # COLUMN 1: The Match Year Flag
    # Priority: If it matches, show which one. Otherwise, it's unmatched.
    # match_status = case_when(
    #   !is.na(match_index) ~ paste0("Matches with ", match_name),
    #   TRUE                ~ "Unmatched"
    #),

    # COLUMN 2: The Direction of Change
    # Only applies to matches. We compare the OTP legacy to the matched year's value.
    bed_trend = case_when(
      otp_ttl_bdrms < newbeds2023  ~ "Expansion since OTP",
      otp_ttl_bdrms > newbeds2023  ~ "Shrink - double check",
      otp_ttl_bdrms == newbeds2023 ~ "Perfect Match",
      TRUE                          ~ NA_character_
    ),

    # COLUMN 3: Data Presence (The "All Zeros/NAs" check)
    # Checks if ALL columns starting with 'NEW' are either 0 or NA
    data_presence = if_else(
      if_all(starts_with("newbeds"), ~ .x == 0 | is.na(.x)),
      "Empty (All 0 or NA)",
      "Bedrooms Recorded"
    )
  )

# get year built, and add on to the flags

year_built <- maui_dwell_alltmks %>% 
  filter(tmk %in% maui_otp_updaed_withBedFlags$tmk) %>% 
  select(tmk, year_built) %>%
  filter(!is.na(year_built)) %>% 
  distinct() %>% 
  group_by(tmk) %>%
  summarize(
    # str_flatten to join them with a comma
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    #count how many years were found
    count_buildyears = n()
  ) %>%
  ungroup() %>% 
  mutate(tmk = as.character(tmk)) # missing 20 

maui_updated_with_BEDflags_gapfilled <- maui_otp_updaed_withBedFlags %>% 
  left_join(., year_built, by = "tmk") %>%
  # 1. Create the flags
  mutate(
    gapfill_flag = if_else(data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0, 
                           "Gapfilled with OTP value", "Used dwell data")
  ) %>%
  # 2. Apply the Gap-fill
  mutate(across(starts_with("newbeds"), ~ {
    fill_condition <- data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0
    if_else(fill_condition, as.numeric(otp_ttl_bdrms), as.numeric(.x))
  })) %>%
  # 3. Apply the CONSERVATIVE Year Built Correction
  mutate(across(starts_with("newbeds"), ~ {
    col_year <- as.numeric(str_extract(cur_column(), "\\d{4}"))
    
    case_when(
      # CONDITION A: If we DID NOT gap-fill this row, leave it exactly as is
      gapfill_flag == "Used dwell data" ~ as.numeric(.x),
      
      # CONDITION B: If we DID gap-fill, but house was built prior to 2010 or unknown, keep it
      is.na(year_built_combined) | year_built_combined <= 2010 ~ as.numeric(.x),
      
      # CONDITION C: If we DID gap-fill AND the house didn't exist yet, set to NA
      year_built_combined > col_year ~ NA_real_,
      
      # Fallback: Keep current value
      TRUE ~ as.numeric(.x)
    )
  })) # this will all be used for table 1 and table 2

# create table 2 updated and gapfilled

table_2_updatedGapfilled <- maui_updated_with_BEDflags_gapfilled %>% 
  select(tmk, starts_with("new"))

bedFlags_forTable1 <- maui_updated_with_BEDflags_gapfilled %>% 
  select(tmk,
         bed_trend,
         data_presence,
         year_built_combined,
         count_buildyears,
         gapfill_flag)


# --------- gap fill class ------------ # 

# read in table 3 class 

table_3_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3_class_allcsv")) %>% 
  mutate(tmk = as.character(full_taxid)) %>% 
  select(-full_taxid) %>% 
  rename_with(., ~paste0("newclass",.x), .cols= !tmk)

# join to maui updated, with specific columsn accepted. also do encoding

maui_updated_withClass_enc <- maui_updated %>% 
  select(tmk,osds_qty,starts_with("class")) %>%
  select(-class) %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity != 0) %>% 
  rename(ttl_osds_2010 = osds_qty) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv" ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii" ~ 2,
    class == "class_i" ~ 1
  )) %>%
  pivot_wider(id_cols = c("tmk","ttl_osds_2010"), names_from = "class", values_from = c("osds_class_otp")) %>%
  mutate(otp_encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

maui_updated_withClassEnc_New <- maui_updated_withClass_enc %>%
  left_join(table_3_new, by = "tmk")

# get cols to run match on
class_cols <- maui_updated_withClassEnc_New %>% select(starts_with("new")) %>% names()

table3_updated_classmatch <-maui_updated_withClassEnc_New %>% 
  mutate(
   # CREATE MATCH INDEX: 
    # This finds the index (1, 2, 3...) of the first column that matches OTP bedrooms
    # We use matrix comparison (df[cols] == df$otp) to keep it vectorized
    match_index = {
      match_matrix <- pick(all_of(class_cols)) == otp_encoded_sum
      # replace NAs with FALSE so max.col doesn't trip
      match_matrix[is.na(match_matrix)] <- FALSE 
      
      # If a row has at least one match, get the index. Otherwise, NA.
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    
    # CREATE MATCH NAME: Convert that index (e.g., 1) into the column name (e.g., "new_2010")
    match_name = class_cols[match_index],
    # COLUMN 1: The Match Year Flag
    # Priority: If it matches, show which one. Otherwise, it's unmatched.
    # match_status = case_when(
    #   !is.na(match_index) ~ paste0("Matches with ", match_name),
    #   TRUE                ~ "Unmatched"
    #),

    # COLUMN 3: Data Presence (The "All Zeros/NAs" check)
    # Checks if ALL columns starting with 'NEW' are either 0 or NA
    data_presence = if_else(
      if_all(starts_with("new"), ~ .x == 0 | is.na(.x)),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )) %>% 
  unite("class_history", starts_with("new"), sep = "|", remove = FALSE)


# separate df to create a flag. you need the iws table 1 registry for this one

table_1_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv")) %>% 
  mutate(tmk = as.character(full_taxid))

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  dplyr::select(tmk, otp_encoded_sum, min_year, max_year, n_addresses, class_history,
                starts_with("new")) %>% 
  # change this logic: first define siutation in "new", ie one or two upgrades etc. then compare that to OSDS. 
  # 
  mutate(
    legacy_val = as.numeric(otp_encoded_sum),
    v2009 = as.numeric(newclass2009),
    v2010 = as.numeric(newclass2010),
    
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010) ~ "no data to compare",
      TRUE ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = case_when(
    scenario_match_flag == "matched 2009 or 2010" ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE ~ NA_real_
  )) %>% 
  select(tmk, newpre_2009, newclass2009, newclass2010,
         starts_with("newclass201"),starts_with("newclass202"), everything()) %>% 
  # Use rowwise to build the unique story without the split/paste overhead
  rowwise() %>%
  mutate(unique_story = {
    vals <- c_across(starts_with("new"))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>%
  ungroup() %>%
  mutate(
    upgrade_scenario = case_when(
      # High-priority special overrides
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1" ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2" ~ "septic and aerobic - replacement of aerobic",
      
      unique_story == "3|4" ~ "septic combo to aerobic",
      
      unique_story == "1|4" ~ "single septic to aerobic",
      
      unique_story == "7|4" ~ "complex septic system to aerobic",
      
      unique_story == "8|5" ~ "cesspool to septic / aerobic hybrid",
      
      # Standard Logic
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1234]$") ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$") ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$") ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1234]$") ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$") ~ "multiple osds with new cesspool application",
      
      # Errors and Laterals
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$") ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$") ~ "lateral change (clean to clean)",
      
      TRUE ~ "other scenario"
    )
  )

# split into data for table 3 and a flag for table 1
table_3_new_updated <- update_logic_refined_cl %>% 
  select(tmk, starts_with("new"))

class_flag_forTable1 <- update_logic_refined_cl %>%
    select(tmk,class_match_flag = scenario_match_flag,unique_story,upgrade_scenario)

```

#### Complete Tables 1 - 3a, 4

``` {r complete-tables }
# --------------- Complete tables 1,2 and 3a for Maui --------------- #

# read in table 1 new and unchanged versions, and add gapfill flags

table_1_new <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv")) %>% 
  rename(in_dwelldat = is_dwelling) %>% 
  mutate(status_since_otp = "New Permit",
         n_addresses = as.character(n_addresses),
         tmk = as.character(full_taxid)) %>% 
  select(-full_taxid) %>% 
  distinct()

table_1_unchanged <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_UNCHANGED.csv")) %>% 
  mutate(scenario = "No new permits") %>% 
  mutate(tmk = as.character(tmk)) %>% 
  distinct()

# are there duplicates? no

table_1_unchangeddups <- table_1_unchanged %>% filter(tmk %in% table_1_new$tmk)

table_1_complete <- bind_rows(table_1_new, table_1_unchanged)

# ---------------- Table 2 ----------------- #
# this one should have all the tmks - commercial tmks 
table_2_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2_classes_UNCHANGED.csv")) %>% 
  mutate(tmk = as.character(tmk))

table_2_updated_forbind <- table_2_updatedGapfilled %>% 
  rename_with(~ str_replace(.x, "newbeds", "bedrooms_"), 
              starts_with("newbeds"))

table_2_newonly <- table_2_new %>% 
  filter(tmk %!in% table_2_updatedGapfilled$tmk)

table_2_complete <- bind_rows(table_2_unchanged, table_2_updated_forbind, table_2_newonly) #900 commerical?

table_1_complete_bedflags <- left_join(table_1_complete, bedFlags_forTable1, by = "tmk") %>% 
  distinct()
  
# -------------- Table 3 --------------------- #
  
table_3_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3_classes_UNCHANGED.csv")) %>% 
  distinct() %>% 
  mutate(tmk = as.character(tmk)) %>% 
  rename(newclass2010 = osds_cl_2010)

table_3_updatedforbind <- table_3_new_updated %>% 
  mutate(tmk = as.character(tmk))

table_3_newonly <- table_3_new %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  distinct()

table_3_complete <- bind_rows(table_3_new_updated, table_3_unchanged, table_3_newonly) %>%   rename_with(~ str_replace(.x, "newclass", "osds_class_"), 
              starts_with("newclass")) 

table_1_withClassflags <- table_1_complete_bedflags %>% 
  left_join(class_flag_forTable1, by = "tmk")

# ------------ Table 4 --------------------- # 

table_4_unchanged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_commercial_UNCHANGED_draft.csv")) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  select(tmk, commercial, pitt_code)

table_4_googleplaces <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_googleplaces_UNCHANGED_draft.csv")) %>% 
  mutate(
    tmk = as.character(tmk),
    commercial = "unknown",
    pitt_code = -9999) %>% 
  select(tmk, commercial,pitt_code)
  
table_4_withUpdates_forbind <- table4_withUpdates %>% 
  filter(!is.na(commercial)) %>% 
  select(tmk,commercial, pitt_code)
  
table_4_new <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_commercial.csv")) %>% 
  mutate(tmk = as.character(full_taxid)) %>% 
  select(-full_taxid) %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk) %>% 
  mutate(commercial = "unknown",
         pitt_code = -9999) %>% 
  select(tmk, commercial, pitt_code)
  

table_4_complete <- bind_rows(table_4_unchanged,
                              table_4_googleplaces,
                              table_4_new,
                              table_4_withUpdates_forbind)

# add pittcode to unknown. add flag for google places.


table_4_complete_cl <- table_4_complete %>% 
  mutate(
    pitt_code = case_when(
      is.na(pitt_code) ~ -9999,
      TRUE ~ pitt_code
    ),
    commercial = ifelse(is.na(commercial) | commercial == "unknown","check google places",commercial)
  )

# ----------- Clean up Table 1 ---------------- #

table_1_final <- table_1_withClassflags %>% 
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete_cl$tmk, TRUE,FALSE)
  ) %>% 
  select(
    tmk,
    starts_with("n_"),
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag = gapfill_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4 = use_table4
  )


# tmk checks
# All TMKs in table 1 accounted for table 3? 

all(table_1_final$tmk %in% table_3_complete$tmk)

# All TMKS in table 2 in table 1?
all(table_2_complete$tmk %in% table_1_final$tmk)

# Authoritative list all in table 1?

all(master_tmk_list %in% table_1_final$tmk)

# Difference in table 2 and table 1 == table 4?

shouldbein_table_4 <- setdiff(table_1_final$tmk,table_2_complete$tmk)

shouldbe_intable2 <- setdiff(table_4_complete_cl$tmk, shouldbein_table_4)

head(shouldbe_intable2)

all(shouldbe_intable2 %in% table_2_complete$tmk) # TRUE, yay

table_4_complete_cl <- table_4_complete_cl %>% 
  filter(tmk %!in% shouldbe_intable2)

# fix ones that were incorrectly classified as commecial 
table_1_final_final <- table_1_final %>%
  mutate(
    use_table4 = ifelse(tmk %in% table_4_complete_cl$tmk, TRUE,FALSE)
  )

# save them all!

write_csv(table_1_final_final,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_ALL.csv"))

write_csv(table_2_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2_bedrooms_ALL.csv"))

write_csv(table_3_complete,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3a_permitsApplied_ALL.csv"))

write_csv(table_4_complete_cl,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_percommercial_draft.csv"))


```


## Step 4: Create table 3b

The additive encoding is what is known as a binary bit mask. Each class represents a power of 2, so every unique sum corresponds to exactly one combination of classes. 

We can think of table 3a as an "event" table, since it records the permit application event. However, we want to create a "status" table, which records the actual OSDS present on the property in each year, BASED on the events we know happened. This will be table 3b

We can make multiple versions of this, using different math. In the first, I'll assume that:
1. If there was only one OSDS on the property, the new application is replacing it
2. If there were multiple OSDS on the property, the new application is replacing the grossest, ie highest class. This correspond to the largest power of 2 within the encoded value

This translates to the following math: 
1. Identify the largest power of 2 inside the current value.
2. Subtract that power of 2.
3. Add the new application value.

```{r}

# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% 
  rename(osds_cl_2008 = newpre_2009) %>% # note, this is a stand in for "pre 2009"
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(Applied,0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3b_allyears_draft.csv"))



```


## Step 5: Fill in Table 4 - happens in a different script
