---
title: "Hawaii OSDS Timeseries"
author: "Madeline Berger"
date: "2025-12-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))


# custom functions
calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

```


## Step 0: Create Authoritative TMK lists

```{r}

library(tidyverse)
library(janitor)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 0. HAWAII COUNTY (BIG ISLAND) TMK CLEANER FUNCTION
# ==============================================================================
clean_hawaii_tmk <- function(df, target_col = "tmk") {
  col_names <- names(df)
  actual_col <- if (target_col %in% col_names) {
    target_col
  } else if ("full_taxid" %in% col_names) {
    "full_taxid"
  } else if ("TMK" %in% col_names) {
    "TMK"
  } else {
    target_col
  }
  
  df %>% 
    mutate(
      raw_val = as.character(.data[[actual_col]]),
      raw_val = if_else(str_detect(raw_val, "e\\+"), sprintf("%.0f", as.numeric(raw_val)), raw_val),
      digits = str_replace_all(raw_val, "[^0-9]", ""),
      digits = if_else(nchar(digits) >= 9, substr(digits, 1, 9), digits),
      digits = case_when(
        nchar(digits) == 8 ~ paste0("3", digits),
        nchar(digits) == 9 & str_starts(digits, "0") ~ paste0("3", substr(digits, 2, 9)),
        TRUE ~ str_pad(digits, width = 9, side = "left", pad = "0")
      )
    ) %>% 
    mutate(tmk = digits) %>% 
    select(-any_of(c("raw_val", "digits"))) %>% 
    filter(!is.na(tmk) & tmk != "" & tmk != "000000000")
}

# ==============================================================================
# 1. READ PERMIT DATA (IWS 2009–PRESENT)
# ==============================================================================

iws_permit_data <- read_csv(
  file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_names()

iws_hawaii_permits <- iws_permit_data %>%
  clean_hawaii_tmk("tmk") %>% 
  filter(str_starts(tmk, "3"))

iws_hawaii_permits_tmk <- iws_hawaii_permits %>% pull(tmk) %>% unique() #7347

# ==============================================================================
# 2. READ BASELINE DATA (HIGIS 2010 OTP - HAWAII COUNTY)
# ==============================================================================

otp_hawaii_raw <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Hawaii_Island.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_names() %>% 
  clean_hawaii_tmk("tmk")

otp_hawaii_tmk <- otp_hawaii_raw %>% pull(tmk) %>% unique()

# ==============================================================================
# 3. PARTITION BASELINE & NEW PERMITS
# ==============================================================================

# Unchanged baseline TMKs (in HIGIS baseline, NO new permits)
hawaii_unchanged <- otp_hawaii_raw %>% 
  filter(tmk %!in% iws_hawaii_permits_tmk)

unchanged_tmks <- hawaii_unchanged %>% pull(tmk) %>% unique()

# Updated baseline TMKs (in HIGIS baseline AND has new permits)
hawaii_updated <- otp_hawaii_raw %>% 
  filter(tmk %in% iws_hawaii_permits_tmk)

updated_tmks <- hawaii_updated %>% pull(tmk) %>% unique()

# Brand new permit TMKs (in IWS permits, NOT in HIGIS baseline)
hawaii_new <- iws_hawaii_permits %>% 
  filter(tmk %!in% otp_hawaii_tmk)

new_tmks <- hawaii_new %>% pull(tmk) %>% unique()

# ==============================================================================
# 4. CREATE AUTHORITATIVE MASTER TMK LIST
# ==============================================================================

hawaii_master_tmk_list <- unique(c(iws_hawaii_permits_tmk, otp_hawaii_tmk))

cat("--- HAWAII COUNTY (BIG ISLAND) STEP 0 SUMMARY ---\n")
cat("Total Baseline 2010 OTP Records:", nrow(otp_hawaii_raw), "(Unique TMKs:", length(otp_hawaii_tmk), ")\n")
cat("Total IWS Permit Records:", nrow(iws_hawaii_permits), "(Unique TMKs:", length(iws_hawaii_permits_tmk), ")\n")
cat("--------------------------------------------------\n")
cat("Unchanged Baseline TMKs:", length(unchanged_tmks), "\n")
cat("Updated Baseline TMKs:", length(updated_tmks), "\n")
cat("Brand-New Permit TMKs:", length(new_tmks), "\n")
cat("Authoritative Master TMK Count:", length(hawaii_master_tmk_list), "\n")


```


## Step 1: Clean Up Dwell Data

The Hawaii data comes to us raw in text files, similar to Oahu. Here we will just generate the same summary tables as we did for Oahu. We will have to read in each excel file here and do some cleaning. 

This report should tell us:

- total number of TMKS assessed each year for Oahu
- how many TMKS are single owner versus multiple owner per year of data
- how many TMKS have more than one card (i.e. more than one building assessed on the parcel) per year
- Summary of bedrooms and bathrooms per year?


This might get long - the Hawaii dwell data is quite messy. A few notes from the data sent by the county: 

2010 - 2012 data did not come with extract formula sheets with them


#### First Define Col Specs for Parsing

I realized that specification docs were included for all years after 2012. I looked at them all side by side and there appear to be two distinct versions. The first (A) applies only to 2013, and 2019 - 2023. The second one (B) applies to 2014 - 2018.

They are actually identical until the year built column, which makes sense because when I was applying the ones above to different groups of years, the column that clued me into if it was working or not was year built. 

In A, yrbuilt is 434 to 437
In B, yrblt is 425 to 429


```{r col-specs}

library(readr)

# try estimated col speces
col_specs_hi <- fwf_positions(
  start = c(
    1, 15, 20, 28, 45, 49, 67, 83, 99, 115, 131, 152, 158, 164, 168, 172,
    178, 185, 199, 215, 230, 245, 260, 326, 334, 354, 371, 379, 387
  ),
  end = c(
    13, 18, 27, 44, 45, 53, 67, 83, 99, 115, 133, 153, 159, 166, 174, 180,
    185, 185, 201, 215, 230, 245, 260, 329, 338, 358, 376, 384, 392
  )
)

# these are from Oahu clean - they do better for 2016 - 2018
col_specs_old <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92, 112, 132, 152, 154,
    179, 186, 193, 199, 201, 202, 206, 211, 216, 218, 238, 239,
    259, 279, 299, 319, 343, 355, 357, 361, 364, 384, 404, 434,
    438, 443, 447, 448, 470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 18, 23, 43, 44, 47, 71, 91, 111, 131, 151, 153, 173,
    180, 187, 194, 200, 201, 201, 207, 212, 217, 237, 238, 258,
    278, 298, 318, 342, 354, 356, 360, 363, 383, 403, 433, 437,
    442, 446, 448, 469, 478, 489, 494, 502, 509, 515 
  )
)

# this one works best for anything before 2015
col_specs_2015 <- fwf_positions(
  start = c(1, 2, 15, 20, 24, 44, 49, 52, 72, 92,
            112, 132, 152, 154, 179, 186, 193, 199, 201, 202,
            206, 211, 217, 218, 238, 239, 259, 279, 299, 319,
            343, 349, 356, 358, 365, 373, 375, 379, 391, 411,
            413, 452, 456, 457, 461, 465, 487, 497, 508, 513,
            521, 528),
  end   = c(1, 14, 19, 23, 43, 48, 51, 71, 91, 111,
            131, 151, 153, 178, 185, 192, 198, 200, 201, 205,
            210, 216, 217, 237, 238, 258, 278, 298, 318, 342,
            348, 355, 357, 364, 372, 374, 378, 390, 410, 412,
            451, 455, 456, 460, 464, 486, 496, 507, 512, 520,
            527, 534)  
)

# try this one for later years?

col_specs_2022 <- fwf_positions(
  start = c(1, 2, 15, 20, 24, 44, 49, 52, 72, 92, 112, 132, 152, 154, 179, 186, 193, 199, 201, 202,
            206, 211, 217, 218, 238, 239, 259, 279, 299, 319,
            343, 349, 356, 358, 365, 373, 375, 379, 391, 411,
            413, 452, 457, 461, 462, 466, 488, 497, 508, 513,
            521, 528),
  end = c(1, 14, 19, 23, 43, 48, 51, 71, 91, 111,131, 151, 153, 178, 185, 192, 198, 200, 201, 205,
          210, 216, 217, 237, 238, 258, 278, 298, 318, 342,
          348, 355, 357, 364, 372, 374, 378, 390, 410, 412,
          451, 456, 460, 461, 465, 487, 496, 507, 512, 520,
          527,534))


colspec_A <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92,
    112, 132, 152, 154,
    179, 186, 193, 199, 201, 202,
    206, 211, 216, 218,
    238, 239, 259,
    279, 299, 319,
    343, 355, 357, 361, 364,
    384, 404,
    434, 438, 443, 444, 448,
    470, 479, 490, 495, 503, 510
  ),
  end = c(
    1, 14, 19, 23, 43, 47, 51, 71, 91, 111,
    131, 151, 153, 178,
    185, 192, 198, 200, 201, 205,
    210, 215, 217, 237,
    238, 258, 278,
    298, 318, 342,
    354, 356, 360, 363, 383,
    403, 433,
    437, 442, 443, 447, 469,
    478, 489, 494, 502, 509, 999
  ),
  col_names = c(
    "JUR","PARID","TAXYR","CARD","USER20","VALMETH","STORIES",
    "USER6","USER7","USER8","USER4","USER5","STYLE","USER9",
    "FIXBATH","FIXHALF","FIXADDL","FIXTOT","HEAT","ATTIC",
    "RMTOT","RMBED","RMFAM","USER10",
    "BSMT","USER11","USER12",
    "USER13","USER14","USER15",
    "RECROMAREA","CONDOLVL","CONDOTYP","CONDOVW","USER1",
    "USER2","USER3",
    "YRBLT","EFFYR","COND","GRADE","USER16",
    "USERFACT","ECNDEP","PCTCOMPLETE","FLR1AREA","ADDNAREA","SFLA"
  )
)



colspec_B <- fwf_positions(
  start = c(
    1, 2, 15, 20, 24, 44, 48, 52, 72, 92,
    112, 132, 152, 154,
    179, 186, 193, 199, 201, 202,
    206, 211, 216, 218,
    238, 239, 259,
    279, 299, 319,
    343, 355, 357, 361, 364,
    384, 404,
    425, 430, 434, 435, 439,
    461, 470, 481, 486, 494, 501
  ),
  end = c(
    1, 14, 19, 23, 43, 47, 51, 71, 91, 111,
    131, 151, 153, 178,
    185, 192, 198, 200, 201, 205,
    210, 215, 217, 237,
    238, 258, 278,
    298, 318, 342,
    354, 356, 360, 363, 383,
    403, 424,
    429, 433, 434, 438, 460,
    469, 480, 485, 493, 500, 999
  ),
  col_names = c(
    "JUR","PARID","TAXYR","CARD","USER20","VALMETH","STORIES",
    "USER6","USER7","USER8","USER4","USER5","STYLE","USER9",
    "FIXBATH","FIXHALF","FIXADDL","FIXTOT","HEAT","ATTIC",
    "RMTOT","RMBED","RMFAM","USER10",
    "BSMT","USER11","USER12",
    "USER13","USER14","USER15",
    "RECROMAREA","CONDOLVL","CONDOTYP","CONDOVW","USER1",
    "USER20_ALT","USER3",
    "YRBLT","EFFYR","COND","GRADE","USER16",
    "USERFACT","ECNDEP","PCTCOMPLETE","FLR1AREA","ADDNAREA","SFLA"
  )
)


# try on one year

hi_dwell_2013 <- read_fwf(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Hawaii_DwellRaw_12182025/2013/DWELDAT.TXT"),
                          col_positions = colspec_A,
                          na = c("", "NA"),
                          col_types = cols(.default = col_character())) %>%
  mutate(across(everything(), ~trimws(.)))

#YAY


```


##### Apply to data


For Hawaii, Iʻm going to try a slightly different approach where I filter by the TMKS with permits before worrying about the column parsing. Just to make things lighter 


```{r}

library(readr)
library(dplyr)
library(purrr)
library(janitor)
library(stringr)
library(tibble)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 1. READ PERMITS & EXTRACT TMK LIST
# ==============================================================================

iws_permit_data <- read_csv(
  file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"),
  col_types = cols(.default = "c")
)

# filter out just the permits in Hawaii, and create permit ID tracker

iws_hi_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"3")) %>% 
  rename(full_taxid = tmk_char) %>%  
  rowid_to_column("permit_id") # 7347

#tmks_permits <- iws_hi_permits$full_taxid

tmks_permits <- hawaii_master_tmk_list

# ------------ function to go get and read in each text file, parse, and filter -------------
read_and_filter_hi_dwell <- function(year, driver_dir, col_specs, tmks_permits) {
  file_path <- file.path(driver_dir,
                         "4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Hawaii_DwellRaw_12182025",
                         as.character(year),
                         "DWELDAT.TXT")
  
  read_fwf(file_path,
           col_positions = col_specs,
           na = c("", "NA"),
           col_types = cols(.default = col_character())) %>%
    mutate(across(everything(), ~trimws(.))) %>%
    # Rename first two columns
    rename(jursi_id = 1, tmk = 2) %>%
    # Create tmk_13 and tmk_9
    mutate(
      tmk_13 = paste0(jursi_id, tmk),
      tmk_9  = substr(tmk_13, 1, nchar(tmk_13) - 4)
    ) %>%
    # Filter by the TMKs we care about
    filter(tmk_9 %in% tmks_permits)
}

# ==============================================================================
# 3. PROCESS BY SCHEMA ERA
# ==============================================================================

# Era 1: 2010–2012 (Spec A - Requires Column Shift Correction)
years_2010_2012 <- 2010:2012
dwell_2010_2012 <- map_dfr(years_2010_2012, ~read_and_filter_hi_dwell(.x, driver_dir, colspec_A, tmks_permits)) %>% 
  mutate(
    year_built_fixed = paste0(YRBLT, substr(EFFYR, 1, 1)),
    effective_year_fixed = paste0(substr(EFFYR, 2, nchar(EFFYR)), substr(COND, 1, 1))
  ) %>% 
  clean_names() %>% 
  select(
    jursi_id, tmk, tmk_9, taxyr, card,
    fixbath, fixhalf, rmtot, rmbed,
    yrblt = year_built_fixed,
    effyr = effective_year_fixed
  )

# Era 2: 2013–2018 (Spec B)
years_2013_2018 <- 2013:2018
dwell_2013_2018 <- map_dfr(years_2013_2018, ~read_and_filter_hi_dwell(.x, driver_dir, colspec_B, tmks_permits)) %>% 
  clean_names() %>% 
  select(jursi_id, tmk, tmk_9, taxyr, card, fixbath, fixhalf, rmtot, rmbed, yrblt, effyr)

# Era 3: 2019–2023 (Spec A)
years_2019_2023 <- 2019:2023
dwell_2019_2023 <- map_dfr(years_2019_2023, ~read_and_filter_hi_dwell(.x, driver_dir, colspec_A, tmks_permits)) %>% 
  clean_names() %>% 
  mutate(
    # Re-stitch last digit of yrblt from the start of effyr if taxyr == 2023 and nchar(yrblt) == 3
    yrblt_fixed = if_else(
      taxyr == "2023" & nchar(yrblt) == 3,
      paste0(yrblt, substr(effyr, 1, 1)),
      yrblt
    ),
    effyr_fixed = if_else(
      taxyr == "2023" & nchar(yrblt) == 3,
      substr(effyr, 2, nchar(effyr)),
      effyr
    )
  ) %>% 
  select(
    jursi_id, tmk, tmk_9, taxyr, card, 
    fixbath, fixhalf, rmtot, rmbed, 
    yrblt = yrblt_fixed, 
    effyr = effyr_fixed
  )

# ==============================================================================
# 4. CONSOLIDATE AND EXPORT
# ==============================================================================

allyears_hi_dwell <- bind_rows(dwell_2010_2012, dwell_2013_2018, dwell_2019_2023) %>% 
  rename(full_taxid = tmk_9)

out_dwell_path <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/hi_dwell_osds_tmks.csv")
dir.create(dirname(out_dwell_path), recursive = TRUE, showWarnings = FALSE)
write_csv(allyears_hi_dwell, out_dwell_path)



```

Ok so it looks like col Spec B has worked really well for all years. Who knew! Iʻm going to quickly check for any mistakes below, then we can move on.

The yearbuilt column is important for gapfilling, so its worth making sure it looks right.

```{r}


# create new column with length of yearbuilt, and print unique instances

allyears_hi_dwell$yearblt_length <- nchar(allyears_hi_dwell$yrblt)

unique(allyears_hi_dwell$yearblt_length) # 4 and 3

# look at the 3

allyears_dwell_yrblt_3 <- allyears_hi_dwell %>% filter(yearblt_length == 3)

# which years does this apply to?

unique(allyears_dwell_yrblt_3$taxyr) # since 2023


# lets look at 2023s and see if we can figure out where the missing digit wrny

 # allyears_hi_dwell %>% 
 #  filter(taxyr == "2023", nchar(yrblt) == 3) %>% 
 #  select(tmk, taxyr, rmtot, rmbed, yrblt, effyr) %>% 
 #  head(10)
 # 
# after running this, I added code in the chunk above to manually fix 2023 values. 

# last check: are the year builts making sense? 

# Diagnostic: Find non-numeric, pre-1800, or future year_built values
invalid_yrblt <- allyears_hi_dwell %>% 
  mutate(yrblt_num = suppressWarnings(as.numeric(yrblt))) %>% 
  filter(!is.na(yrblt) & (is.na(yrblt_num) | yrblt_num < 1850 | yrblt_num > 2023))

cat("Total records with invalid/outlier year built:", nrow(invalid_yrblt), "\n")
```


## Step 2: Bring in IWS data

Here we join dwell data to the permit information for the ones that correspond to dwellings, and save the list of non-dwell TMKS.


```{r iws-dwell-match}


library(readr)
library(dplyr)
library(stringr)
library(tibble)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 1. READ CLEAN DWELLING DATA & STANDARDIZE TMK
# ==============================================================================

allyears_hi_dwell <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/hi_dwell_osds_tmks.csv"),
  col_types = cols(.default = "c")
) %>% 
  select(-tmk) %>% 
  rename(tmk = any_of("full_taxid")) %>% 
  mutate(tmk = as.character(tmk))

# ==============================================================================
# 2. READ PERMIT DATA & STANDARDIZE TMK
# ==============================================================================

iws_permit_data <- read_csv(
  file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"),
  col_types = cols(.default = "c")
)

iws_hi_permits <- iws_permit_data %>% 
  mutate(tmk = as.character(TMK)) %>% 
  filter(str_starts(tmk, "3")) %>% 
  rowid_to_column("permit_id")

tmks_permits <- iws_hi_permits$tmk

# ==============================================================================
# 3. JOIN PERMITS WITH DWELLING DATA BY TMK
# ==============================================================================

# Match permits to dwelling records
iws_hi_dwell_join <- inner_join(iws_hi_permits, allyears_hi_dwell, by = "tmk") %>% 
  distinct()

# Permits that matched at least one dwelling record
iws_hi_dwell_join_norepeats <- iws_hi_permits %>% 
  filter(tmk %in% allyears_hi_dwell$tmk)

# Permits with NO dwelling records
iws_hi_NOdwell <- iws_hi_permits %>% 
  filter(tmk %!in% iws_hi_dwell_join$tmk)

# ==============================================================================
# 4. EXPORT QA FILE
# ==============================================================================

out_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data")


write_csv(
  iws_hi_NOdwell,
  file.path(out_dir, "hi_iwspermits_nodwellinginfo.csv")
)

cat("Step 2 Part 1 Complete:\n")
cat("- Total Big Island Permits:", nrow(iws_hi_permits), "\n")
cat("- Permits matched to Dwelling Data:", nrow(iws_hi_dwell_join_norepeats), "\n")
cat("- Permits missing Dwelling Data saved to QA:", nrow(iws_hi_NOdwell), "\n")
```

**Results so Far**:
- There were 7347 permits issued for HI island over the past 10 years
- 6242 have dwelling information, and this represents 6011 tmks
- 1105 do not have dwelling information

### Step 2b: Clean and Pivot Permit Data

Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 


#### Wastewater Class Encoding Reference Table

| Detected Classes | Encoded Sum | Description |
| :--- | :---: | :--- |
| **Class 1** | 1 | Cleanest Septic Type |
| **Class 2** | 2 | Secondary Septic Type |
| **Class 1 + 2** | 3 | **Combination of Class 1 + Class 2** |
| **Class 3** | 4 | Aerobic System |
| **Class 1 + 3** | 5 | Class 1 Septic + Aerobic |
| **Class 2 + 3** | 6 | Class 2 Septic + Aerobic |
| **Class 1 + 2 + 3** | 7 | Classes 1, 2, and 3 |
| **Class 4** | 8 | Cesspool |
| **Class 1 + 4** | 9 | Class 1 + Class 4 |
| **Class 2 + 4** | 10 | Class 2 + Class 4 |
| **Class 1 + 2 + 4** | 11 | Classes 1, 2, and 4 |
| **Class 3 + 4** | 12 | Aerobic + Class 4 |
| **Class 1 + 3 + 4** | 13 | Classes 1, 3, and 4 |
| **Class 2 + 3 + 4** | 14 | Classes 2, 3, and 4 |
| **All (1, 2, 3, 4)** | 15 | Full suite of permits |


```{r new-tables}

library(tidyverse)
library(janitor)
library(lubridate)

# Helper: Mode calculation for bedrooms
calc_mode <- function(x, na.rm = TRUE) {
  if (na.rm) x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Helper: OSDS Class Assignment
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1"
  )
}

# ==============================================================================
# 1. BUILD MASTER CLEAN PERMIT DATASET
# ==============================================================================

master_clean <- iws_hi_permits %>%
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(
      is.na(project_street_address2), 
      project_street_address, 
      paste(project_street_address, "-", project_street_address2)
    ),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = if_else(tmk %in% iws_hi_NOdwell$tmk, "no", "yes")
  ) %>%
  select(-any_of("permit_id")) %>% 
  distinct()

# ==============================================================================
# 2. GENERATE TABLES 1, 2A, 2B, AND 3
# ==============================================================================

# --- TABLE 1: TMK Registry ---
table1_tmk_registry <- master_clean %>%
  group_by(tmk) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year, na.rm = TRUE),
    n_addresses = n_distinct(address_lower, na.rm = TRUE),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    is_dwelling = any(tmk %in% allyears_hi_dwell$tmk),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)

# --- TABLE 2A: Bedroom Counts (Sum) ---
table2a_bedrooms <- allyears_hi_dwell %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(rmbed = as.numeric(rmbed)) %>%
  group_by(tmk, taxyr) %>%
  summarize(
    total_beds = sum(rmbed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = taxyr, values_from = total_beds, names_sort = TRUE) %>% 
  rename_with(~ paste0("beds_", .), matches("^[0-9]"))

# --- TABLE 2B: Bedroom Counts (Mode) ---
table2b_bedrooms <- allyears_hi_dwell %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(rmbed = as.numeric(rmbed)) %>%
  group_by(tmk, taxyr) %>%
  summarize(
    mode_beds = calc_mode(rmbed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = taxyr, values_from = mode_beds, names_sort = TRUE) %>% 
  rename_with(~ paste0("beds_", .), matches("^[0-9]"))

# --- TABLE 3: Encoded OSDS Permit Classes by Year ---
table3_classes <- master_clean %>%
  filter(!is.na(year)) %>% 
  group_by(tmk, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(
    encoded_sum = 
      (str_detect(all_classes_text, "1") * 1) +
      (str_detect(all_classes_text, "2") * 2) +
      (str_detect(all_classes_text, "3") * 4) +
      (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(tmk, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE) %>% 
  rename_with(~ paste0("osds_cl_", .), matches("^[0-9]"))


# --- TABLE 4: Encoded OSDS Permit Classes by Year ---

# ==============================================================================
# 3. EXPORT TABLES
# ==============================================================================

out_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(table1_tmk_registry, file.path(out_dir, "hi_table1_tmk_registry_all.csv"))
write_csv(table2a_bedrooms,     file.path(out_dir, "hi_table2a_bedrooms_sum.csv"))
write_csv(table2b_bedrooms,     file.path(out_dir, "hi_table2b_bedrooms_mode.csv"))
write_csv(table3_classes,       file.path(out_dir, "hi_table3_class_all.csv"))
write_csv(table1_tmk_registry_commercial,    file.path(out_dir, "hi_table1_tmk_registry_commercial.csv"))

# ==============================================================================
# 4. READ BACK IN VERIFICATION
# ==============================================================================

hi_table1      <- read_csv(file.path(out_dir, "hi_table1_tmk_registry_all.csv"), col_types = cols(.default = "c"))
table2a_beds  <- read_csv(file.path(out_dir, "hi_table2a_bedrooms_sum.csv"), col_types = cols(.default = "c"))
table3_classes <- read_csv(file.path(out_dir, "hi_table3_class_all.csv"), col_types = cols(.default = "c"))

```

Ok so this works beautifully in that the bedroom table does not include any commercial tmks. Those will have their own table, which will specify "place", which dictates effluent, which 


## Step 3: Compare to Past Datasets to Complete the Time Series

```{r}

hi_otp <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Hawaii_Island.csv"))

unique(hi_otp$class) # weirdly this has no 4? but the class column iV is there

```

So the Hawaii database has a similar format to the others, but also has a column with "source", indicating that someone has already gone through the processes of bringing in additional information for TMKS not in the Dwelling Assessment. This means that we should use this for our "mismatches" in the new data. There are also notes for all the commercial properties which is great!

UNFORTUNATELY there is no information about how effluent was calculated for non dwells as far as I can tell, but maybe some snooping around later will fix that. And, is a good reason to apply a standardized approach that matches the other islands.

According to the page: Source =  State of Hawaii Department of Health, May, 2017.

We also have these! Which are helpful:

PITT Code Values:   0  Timeshare (Maui only)   100’s*  Improved Residential   200’s  Apartment   300’s  Commercial   400’s  Industrial   500’s  Agricultural and Rural   600’s   Conservation   700’s   Hotel and Resort   800’s**  Unimproved Residential   900’s  Homeowner    999   Multiple PITT Codes for a single TMK   1000’s  Commercialized Residential (Maui only)      *  Kauai:  100 = Improved and Unimproved Residential   **  Kauai:  800 = Homestead Properties 

Also: lots of -9999 in class? And no class 4 or 5. That makes no sense. 

Some of this data is messy and yuck but there is also a lot of helpful work that it looks like the department of health did in 2017 (or Joey did?) 

Another weird thing about this data is that they dropped the year column. 

We only have classes 1 - 3 in this data. From the metadata, this
represents:

-   CLASS_I: Class I OSDS are those systems where the effluent is
    subjected to soil treatment
-   CLASS_II: Class II OSDS are septic systems discharging to a seepage
    pit.\
-   CLASS_III: Class III OSDS are aerobic systems discharging to a
    seepage pit.
-   CLASS_IV: Class IV OSDS are cesspools.\
-   EFFLUENT: The total effluent flux in million gallons per day
    discharged within the TMK parcel.
    
    
The `osds_class` column also does not have any 4s, just 1-3, and -9999. I'll use the same additive coding here as I did with Maui Nui, Kauai (and eventually Oahu): 

### Unchanged OSDS since OTP

```{r unchanged}

library(tidyverse)
library(janitor)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 1. ENCODE BASELINE 2010 HIGIS OTP CLASSES
# ==============================================================================

hi_otp_clean <- otp_hawaii_raw %>% 
  clean_hawaii_tmk("tmk") %>% 
  select(tmk, starts_with("class"), osds_qty, commercial, pittcode, ttl_bdrms, no_struct)

hi_otp_class_enc <- hi_otp_clean %>% 
  pivot_longer(
    cols = starts_with("class"), 
    names_to = "class_name", 
    values_to = "quantity"
  ) %>% 
  filter(!is.na(quantity) & as.numeric(quantity) > 0) %>% 
  mutate(
    osds_class_val = case_when(
      class_name %in% c("class_iv", "class4") ~ 8, # Cesspool
      class_name %in% c("class_iii", "class3") ~ 4, # Aerobic / Class III
      class_name %in% c("class_ii", "class2") ~ 2, # Septic Class II
      class_name %in% c("class_i", "class1")  ~ 1, # Septic Class I
      TRUE ~ 0
    )
  ) %>% 
  group_by(tmk) %>% 
  summarize(
    encoded_sum = sum(osds_class_val, na.rm = TRUE),
    ttl_osds_2010 = first(osds_qty),
    commercial = first(commercial),
    pittcode = first(pittcode),
    ttl_bdrms = suppressWarnings(max(as.numeric(ttl_bdrms), na.rm = TRUE)),
    .groups = "drop"
  ) %>% 
  mutate(ttl_bdrms = if_else(is.infinite(ttl_bdrms), 0, ttl_bdrms))

# ==============================================================================
# 2. PARTITION UNCHANGED TMKs
# ==============================================================================

# Baseline TMKs NOT found in 2009-Present IWS permits. Run Step 0 for this
otp_unchanged <- hi_otp_class_enc %>% 
  filter(tmk %in% hawaii_unchanged$tmk)

# --- UNCHANGED TABLE 1 ---

hi_dwell_card_summary <- allyears_hi_dwell %>%
  group_by(tmk) %>%
  summarize(
    n_buildings = if ("card" %in% names(.)) n_distinct(card, na.rm = TRUE) else 1,
    in_dwelldat = any(!is.na(taxyr) & taxyr != ""),
    .groups = "drop"
  ) %>% 
  ungroup() 

hi_table1_unchanged <- otp_unchanged %>% 
  mutate(
    n_permits = case_when(
      encoded_sum %in% c(1, 2, 4, 8) ~ 1,
      encoded_sum %in% c(3, 5, 6, 9, 10, 12) ~ 2,
      encoded_sum %in% c(7, 11, 13, 14) ~ 3,
      encoded_sum == 15 ~ 4,
      TRUE ~ 1
    ),
    n_years = 1,
    n_addresses = "1",
    min_year = 2010,
    max_year = 2010,
    is_dwelling = if_else(is.na(commercial) | commercial == "Apartment/condo", TRUE, FALSE),
    status_since_otp = "unchanged since otp",
    scenario = "No new permits"
  ) %>% 
  left_join(hi_dwell_card_summary, by = "tmk") %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    scenario,
    n_buildings
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

# --- UNCHANGED TABLE 2 (Bedrooms with Baseline Gapfill) ---

hi_dwell_unchanged <- allyears_hi_dwell %>% 
  filter(tmk %in% otp_unchanged$tmk) %>% 
  mutate(rmbed = as.numeric(rmbed)) %>%
  group_by(tmk, taxyr) %>% 
  summarize(total_beds = sum(rmbed, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(names_from = taxyr, values_from = total_beds, names_sort = TRUE) %>% 
  rename_with(~ paste0("bedrooms_", .), matches("^[0-9]"))

hi_table2_unchanged <- otp_unchanged %>% 
  filter(is.na(commercial)) %>% 
  select(tmk, OTP_2010_beds = ttl_bdrms) %>% 
  left_join(hi_dwell_unchanged, by = "tmk") %>% 
  mutate(
    gapfill_flag = if_else(
      OTP_2010_beds > 0 & rowSums(pick(starts_with("bedrooms_")), na.rm = TRUE) == 0,
      "Gap-filled with OTP value",
      "Original dwell data"
    )
  ) %>% 
  mutate(across(starts_with("bedrooms_"), ~ if_else(gapfill_flag == "Gap-filled with OTP value", OTP_2010_beds, .x))) %>% 
  select(-OTP_2010_beds)

# --- UNCHANGED TABLE 3 (Baseline Classes) ---
hi_table3_unchanged <- otp_unchanged %>% 
  select(tmk, osds_class_pre2009 = encoded_sum)

# --- UNCHANGED TABLE 4 (Commercial Draft) ---
hi_table4_unchanged_draft <- otp_unchanged %>% 
  filter(!is.na(commercial)) %>% 
  select(tmk, commercial, pitt_code = pittcode) # 190. a little lpw

# ==============================================================================
# 3. EXPORT UNCHANGED CSVs
# ==============================================================================

step3_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3")
dir.create(step3_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(hi_table1_unchanged,         file.path(step3_dir, "hi_table1_tmk_registry_UNCHANGED.csv"))
write_csv(hi_table2_unchanged,         file.path(step3_dir, "hi_table2_bedrooms_UNCHANGED.csv"))
write_csv(hi_table3_unchanged,         file.path(step3_dir, "hi_table3_classes_UNCHANGED.csv"))
write_csv(hi_table4_unchanged_draft,   file.path(step3_dir, "hi_table4_commercial_UNCHANGED_draft.csv"))


```

Note for later: final Table 4 should have about 807 entries to match TMK registry. 386 will be unchanged, the rest will either new or updated. 

### Updated OTP

This one is a bit more complex. These tmks are already accounted for in our Tables 1 - 3 above, because they are updates. But, we want to know what those updates were. We also want to resolve any conflicts between new and old data. 

Like above, the person who updated this in 2017 may have had additional information that we do not get from the dwell data. 


```{r updated}

library(tidyverse)
library(janitor)

'%!in%' <- Negate('%in%')

# Set output directory for intermediate Step 3 files
step3_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3")
dir.create(step3_dir, recursive = TRUE, showWarnings = FALSE)

# ==============================================================================
# 0. IDENTIFY BASELINE UPDATED TMKs
# ==============================================================================

# TMKs in 2010 HIGIS OTP that ALSO have new IWS permits (2009–Present)
hi_updated_baseline <- hi_otp_class_enc %>% 
  filter(tmk %in% iws_hi_permits$tmk)

# ==============================================================================
# 1. BEDROOM GAP-FILL & TREND LOGIC (BIG ISLAND)
# ==============================================================================

# Read Step 2 permit bedroom data
table_2_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/hi_table2a_bedrooms_sum.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(tmk = as.character(tmk))

hi_updated_forjoin <- hi_updated_baseline %>% 
  select(tmk, any_of(c("pittcode", "pitt_code")), commercial, ttl_bdrms) %>% 
  rename(
    otp_ttl_bdrms = ttl_bdrms,
    pitt_code = any_of(c("pittcode", "pitt_code"))
  ) %>% 
  mutate(
    otp_ttl_bdrms = as.numeric(otp_ttl_bdrms),
    pitt_code_num = as.numeric(pitt_code),
    comm_assignment = case_when(
      pitt_code_num == 0                               ~ "Timeshare",
      pitt_code_num == 999                             ~ "Multiple PITT Codes",
      pitt_code_num >= 1000                            ~ "Commercialized Residential",
      pitt_code_num >= 100 & pitt_code_num < 200      ~ "Improved Residential",
      pitt_code_num >= 200 & pitt_code_num < 300      ~ "Apartment",
      pitt_code_num >= 300 & pitt_code_num < 400      ~ "Commercial",
      pitt_code_num >= 400 & pitt_code_num < 500      ~ "Industrial",
      pitt_code_num >= 500 & pitt_code_num < 600      ~ "Agricultural and Rural",
      pitt_code_num >= 600 & pitt_code_num < 700      ~ "Conservation",
      pitt_code_num >= 700 & pitt_code_num < 800      ~ "Hotel and Resort",
      pitt_code_num >= 800 & pitt_code_num < 900      ~ "Unimproved Residential",
      pitt_code_num >= 900 & pitt_code_num < 999      ~ "Homeowner",
      TRUE                                             ~ "Unknown"
    )
  ) %>% 
  select(-pitt_code_num)

# Filter permit bedroom table to updated TMKs
table_2_updated <- table_2_new %>% 
  filter(tmk %in% hi_updated_forjoin$tmk)

# Join baseline bedroom counts with new permit bedroom data
hi_updated_withNewbeds <- hi_updated_forjoin %>% 
  left_join(table_2_updated, by = "tmk") %>% 
  filter(!is.na(otp_ttl_bdrms) & otp_ttl_bdrms != 0)

bed_cols <- hi_updated_withNewbeds %>% select(starts_with("beds_")) %>% names()
latest_bed_col <- if (length(bed_cols) > 0) tail(bed_cols, 1) else NULL

hi_updated_withBedFlags <- hi_updated_withNewbeds %>% 
  mutate(
    match_index = {
      if (length(bed_cols) > 0) {
        match_matrix <- pick(all_of(bed_cols)) == otp_ttl_bdrms
        match_matrix[is.na(match_matrix)] <- FALSE 
        if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
      } else {
        NA_integer_
      }
    },
    match_name = if (!is.null(latest_bed_col)) bed_cols[match_index] else NA_character_,
    
    latest_val = if (!is.null(latest_bed_col)) as.numeric(.data[[latest_bed_col]]) else NA_real_,
    bed_trend = case_when(
      is.na(latest_val)             ~ NA_character_,
      otp_ttl_bdrms < latest_val   ~ "Expansion since OTP",
      otp_ttl_bdrms > latest_val   ~ "Shrink - double check",
      otp_ttl_bdrms == latest_val  ~ "Perfect Match",
      TRUE                          ~ NA_character_
    ),
    
    data_presence = if_else(
      if_all(starts_with("beds_"), ~ .x == "0" | is.na(.x) | .x == 0),
      "Empty (All 0 or NA)",
      "Bedrooms Recorded"
    )
  )

# Join Dwelling Build Years for Conservative Gap-Filling
year_built <- allyears_hi_dwell %>% 
  filter(tmk %in% hi_updated_withBedFlags$tmk) %>% 
  select(tmk, year_built = yrblt) %>% 
  filter(!is.na(year_built) & year_built != "") %>% 
  distinct() %>% 
  group_by(tmk) %>% 
  summarize(
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    count_buildyears = n(),
    .groups = "drop"
  )

hi_updated_with_BEDflags_gapfilled <- hi_updated_withBedFlags %>% 
  left_join(year_built, by = "tmk") %>% 
  mutate(
    gapfill_flag = if_else(
      data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0, 
      "Gapfilled with OTP value", 
      "Used dwell data"
    )
  ) %>% 
  mutate(across(starts_with("beds_"), ~ {
    fill_condition <- data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0
    if_else(fill_condition, as.numeric(otp_ttl_bdrms), as.numeric(.x))
  })) %>% 
  mutate(across(starts_with("beds_"), ~ {
    col_year <- as.numeric(str_extract(cur_column(), "\\d{4}"))
    min_build_yr <- as.numeric(str_extract(year_built_combined, "^\\d{4}"))
    
    case_when(
      gapfill_flag == "Used dwell data"                         ~ as.numeric(.x),
      is.na(min_build_yr) | min_build_yr <= 2010               ~ as.numeric(.x),
      min_build_yr > col_year                                   ~ NA_real_,
      TRUE                                                       ~ as.numeric(.x)
    )
  }))

table_2_updatedGapfilled <- hi_updated_with_BEDflags_gapfilled %>% 
  select(tmk, starts_with("beds_"))

bedFlags_forTable1 <- hi_updated_with_BEDflags_gapfilled %>% 
  select(
    tmk,
    bed_trend,
    data_presence,
    year_built_combined,
    count_buildyears,
    gapfill_flag
  )

write_csv(table_2_updatedGapfilled, file.path(step3_dir, "hi_table2_updated_gapfilled.csv"))
write_csv(bedFlags_forTable1,       file.path(step3_dir, "hi_bedflags_fortable1.csv"))

# ==============================================================================
# 2. OSDS CLASS GAP-FILL & UPGRADE LOGIC (BIG ISLAND)
# ==============================================================================

# Read Step 2 permit OSDS class table
table_3_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/hi_table3_class_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(tmk = as.character(tmk))

hi_updated_withClassEnc_New <- hi_updated_baseline %>% 
  select(tmk, otp_encoded_sum = encoded_sum) %>% 
  left_join(table_3_new, by = "tmk")

class_cols <- hi_updated_withClassEnc_New %>% select(starts_with("osds_cl_")) %>% names()

table3_updated_classmatch <- hi_updated_withClassEnc_New %>% 
  mutate(
    match_index = {
      if (length(class_cols) > 0) {
        match_matrix <- pick(all_of(class_cols)) == otp_encoded_sum
        match_matrix[is.na(match_matrix)] <- FALSE 
        if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
      } else {
        NA_integer_
      }
    },
    match_name = if (!is.null(class_cols)) class_cols[match_index] else NA_character_,
    data_presence = if_else(
      if_all(starts_with("osds_cl_"), ~ .x == "0" | is.na(.x) | .x == 0),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )
  )

# Join Table 1 Registry to examine baseline permit matches
table_1_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/hi_table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  mutate(tmk = as.character(tmk))

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  select(
    tmk, otp_encoded_sum, 
    any_of(c("min_year", "max_year", "n_addresses")), 
    starts_with("osds_cl_")
  ) %>% 
  mutate(
    legacy_val = as.numeric(otp_encoded_sum),
    v2009 = if ("osds_cl_2009" %in% names(.)) as.numeric(osds_cl_2009) else NA_real_,
    v2010 = if ("osds_cl_2010" %in% names(.)) as.numeric(osds_cl_2010) else NA_real_,
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010)                 ~ "no data to compare",
      TRUE                                       ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = as.character(case_when(
    scenario_match_flag == "matched 2009 or 2010"               ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE                                                        ~ NA_real_
  ))) %>% 
  select(
    tmk, newpre_2009, 
    starts_with("osds_cl_200"), starts_with("osds_cl_201"), starts_with("osds_cl_202"), 
    everything()
  ) %>% 
  rowwise() %>% 
  mutate(unique_story = {
    vals <- c_across(c(newpre_2009, starts_with("osds_cl_")))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>% 
  ungroup() %>% 
  mutate(
    upgrade_scenario = case_when(
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1"  ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2"    ~ "septic and aerobic - replacement of aerobic",
      unique_story == "3|4"    ~ "septic combo to aerobic",
      unique_story == "1|4"    ~ "single septic to aerobic",
      unique_story == "7|4"    ~ "complex septic system to aerobic",
      unique_story == "8|5"    ~ "cesspool to septic / aerobic hybrid",
      unique_story == "5|1"    ~ "septic and aerobic to septic",
      unique_story == "7|3"    ~ "2 septics and aerobic to just 2 septics - aerobic removal",
      unique_story == "7|1|4"  ~ "2 septics and aerobic + one septic repair + one aerobic repair",
      unique_story == "10|9"   ~ "1 septic + cesspool to better septic and cesspool",
      
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1234]$")           ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$")             ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$")             ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1234]$")  ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$")       ~ "multiple osds with new cesspool application",
      
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$") ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$")        ~ "lateral change (clean to clean)",
      
      TRUE ~ "other scenario"
    )
  )

# Build active time series carrying forward active permits bitwise
table3_updated_new <- update_logic_refined_cl %>% 
  select(tmk, osds_class_pre2009 = newpre_2009, starts_with("osds_cl_"))

class_flag_forTable1 <- update_logic_refined_cl %>% 
  select(
    tmk, 
    class_match_flag = scenario_match_flag, 
    unique_story, 
    upgrade_scenario
  )

# Export intermediate CSVs
write_csv(table3_updated_new,    file.path(step3_dir, "hi_table3_new_updated.csv"))
write_csv(class_flag_forTable1,   file.path(step3_dir, "hi_classflags_fortable1.csv"))

cat("Big Island Updated Parcels Processing Complete:\n")
cat("- Updated Bedroom Gap-Filling Done:", nrow(table_2_updatedGapfilled), "parcels\n")
cat("- Updated Class Scenarios Classified:", nrow(table3_updated_new), "parcels\n")


```


## Step 4 - Compile Final tables


```{r}

library(tidyverse)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 0. BIG ISLAND TMK CLEANER FUNCTION
# ==============================================================================
clean_hi_tmk <- function(df, target_col = "tmk") {
  col_names <- names(df)
  actual_col <- if (target_col %in% col_names) {
    target_col
  } else if ("full_taxid" %in% col_names) {
    "full_taxid"
  } else if ("TMK" %in% col_names) {
    "TMK"
  } else {
    target_col
  }
  
  df %>% 
    mutate(
      raw_val = as.character(.data[[actual_col]]),
      raw_val = if_else(str_detect(raw_val, "e\\+"), sprintf("%.0f", as.numeric(raw_val)), raw_val),
      digits = str_replace_all(raw_val, "[^0-9]", ""),
      digits = if_else(nchar(digits) >= 9, substr(digits, 1, 9), digits),
      digits = case_when(
        nchar(digits) == 8 ~ paste0("3", digits),
        nchar(digits) == 9 & str_starts(digits, "0") ~ paste0("3", substr(digits, 2, 9)),
        TRUE ~ str_pad(digits, width = 9, side = "left", pad = "0")
      )
    ) %>% 
    mutate(tmk = digits) %>% 
    select(-any_of(c("raw_val", "digits"))) %>% 
    filter(!is.na(tmk) & tmk != "" & tmk != "000000000")
}

# Directories
int_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate")
step2_dir <- file.path(int_dir,"step2-clean-permit-data")
step3_dir <- file.path(int_dir, "step3")
out_dir   <- file.path(int_dir, "step4")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ==============================================================================
# 1. TABLE 1: TMK MASTER REGISTRY
# ==============================================================================

table_1_new <- read_csv(
  file.path(step2_dir, "hi_table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  rename(in_dwelldat = any_of(c("is_dwelling", "in_dwelldat"))) %>% 
  mutate(
    status_since_otp = "New Permit",
    n_addresses = as.character(n_addresses)
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

table_1_unchanged <- read_csv(
  file.path(step3_dir, "hi_table1_tmk_registry_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  mutate(scenario = "No new permits") %>% 
  distinct(tmk, .keep_all = TRUE)

bedFlags_forTable1 <- read_csv(
  file.path(step3_dir, "hi_bedflags_fortable1.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk")

classFlags_forTable1 <- read_csv(
  file.path(step3_dir, "hi_classflags_fortable1.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk")

# Combine registries and attach gap-fill flags
table_1_complete <- bind_rows(table_1_new, table_1_unchanged) %>% 
  distinct(tmk, .keep_all = TRUE)

table_1_complete_flags <- table_1_complete %>% 
  left_join(bedFlags_forTable1, by = "tmk") %>% 
  left_join(classFlags_forTable1, by = "tmk") %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 2. TABLE 2: BEDROOM TIME-SERIES
# ==============================================================================

table_2_unchanged <- read_csv(
  file.path(step3_dir, "hi_table2_bedrooms_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  mutate(across(-tmk, as.numeric))

if (!exists("table_2_updatedGapfilled")) {
  table_2_updatedGapfilled <- read_csv(
    file.path(step3_dir, "hi_table2_updated_gapfilled.csv"),
    col_types = cols(.default = "c")
  )
}

if (!exists("table_2_new")) {
  table_2_new <- read_csv(
    file.path(step2_dir, "table2a_bedrooms_sum.csv"),
    col_types = cols(.default = "c")
  )
}

table_2_updated_forbind <- table_2_updatedGapfilled %>% 
  clean_hi_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^beds_", "bedrooms_"), starts_with("beds_")) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds")) %>% 
  mutate(across(-tmk, as.numeric))

table_2_newonly <- table_2_new %>% 
  clean_hi_tmk("tmk") %>% 
  filter(tmk %!in% table_2_updatedGapfilled$tmk) %>% 
  rename_with(~ str_replace(.x, "^beds_", "bedrooms_"), starts_with("beds_")) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds")) %>% 
  mutate(across(-tmk, as.numeric))

table_2_complete <- bind_rows(
  table_2_unchanged, 
  table_2_updated_forbind, 
  table_2_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 3. TABLE 3A: OSDS PERMIT & CLASS HISTORY
# ==============================================================================

table_3_unchanged <- read_csv(
  file.path(step3_dir, "hi_table3_classes_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^osds_cl_", "osds_class_"), starts_with("osds_cl_")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

if (!exists("table3_updated_new")) {
  table3_updated_new <- read_csv(
    file.path(step3_dir, "hi_table3_new_updated.csv"),
    col_types = cols(.default = "c")
  )
}

if (!exists("table_3_new")) {
  table_3_new <- read_csv(
    file.path(step2_dir, "table3_class_all.csv"),
    col_types = cols(.default = "c")
  )
}

table_3_updatedforbind <- table3_updated_new %>% 
  clean_hi_tmk("tmk") %>% 
  rename(osds_class_pre2009 = any_of(c("newpre_2009", "osds_class_pre2009"))) %>% 
  rename_with(~ str_replace(.x, "^osds_cl_", "osds_class_"), starts_with("osds_cl_")) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

table_3_newonly <- table_3_new %>% 
  clean_hi_tmk("tmk") %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  rename_with(~ str_replace(.x, "^osds_cl_", "osds_class_"), starts_with("osds_cl_")) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

table_3_complete <- bind_rows(
  table_3_updatedforbind, 
  table_3_unchanged, 
  table_3_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 4. TABLE 4: COMMERCIAL & NON-RESIDENTIAL PARCELS
# ==============================================================================

table_4_unchanged <- read_csv(
  file.path(step3_dir, "hi_table4_commercial_UNCHANGED_draft.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "pittcode", "x", "y"))) %>% 
  rename(pitt_code = any_of(c("pittcode", "pitt_code"))) %>% 
  mutate(across(everything(), as.character))

# table_4_googleplaces <- read_csv(
#   file.path(step3_dir, "hi_table4_googleplaces_UNCHANGED_draft.csv"),
#   col_types = cols(.default = "c")
# ) %>% 
#   clean_hi_tmk("tmk") %>% 
#   mutate(
#     commercial = "unknown",
#     pitt_code = "-9999"
#   ) %>% 
#   select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
#   mutate(across(everything(), as.character))

table_4_withUpdates_forbind <- if (exists("table4_withUpdates")) {
  table4_withUpdates %>% 
    clean_hi_tmk("tmk") %>% 
    filter(!is.na(commercial) | !is.na(pitt_code)) %>% 
    select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
    mutate(across(everything(), as.character))
} else {
  tibble(tmk = character(), commercial = character(), pitt_code = character(), x = character(), y = character())
}

table_4_new <- read_csv(
  file.path(step2_dir, "hi_table1_tmk_registry_commercial.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_hi_tmk("tmk") %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk) %>% 
  mutate(
    commercial = "unknown",
    pitt_code = "-9999"
  ) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_complete <- bind_rows(
  table_4_unchanged,
  #table_4_googleplaces,
  table_4_new,
  table_4_withUpdates_forbind
) %>% 
  distinct(tmk, .keep_all = TRUE)

table_4_complete_cl <- table_4_complete %>% 
  mutate(
    pitt_code = if_else(is.na(pitt_code), "-9999", pitt_code),
    commercial = if_else(is.na(commercial) | commercial == "unknown", "check google places", commercial)
  )

# ==============================================================================
# 5. FINAL RECONCILIATION & TABLE 1 OVERRIDES
# ==============================================================================

tmks_with_bedrooms <- table_2_complete %>% 
  filter(if_any(starts_with("bedrooms_"), ~ !is.na(.x) & .x > 0)) %>% 
  pull(tmk)

table_4_complete_cl <- table_4_complete_cl %>% 
  filter(tmk %!in% tmks_with_bedrooms) #yikes, this reduced by 200?

shouldbein_table_4 <- setdiff(table_1_complete_flags$tmk, table_2_complete$tmk) #1225

# Build Final Table 1 Master
table_1_final_final <- table_1_complete_flags %>% 
  mutate(use_table_4 = tmk %in% table_4_complete_cl$tmk) %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag = gapfill_flag,
    class_match_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 6. AUTOMATED INTEGRITY CHECKS
# ==============================================================================

cat("--- BIG ISLAND PIPELINE INTEGRITY CHECKS ---\n")
cat("1. All Table 1 TMKs present in Table 3a:", all(table_1_final_final$tmk %in% table_3_complete$tmk), "\n")
cat("2. All Table 2 TMKs present in Table 1:", all(table_2_complete$tmk %in% table_1_final_final$tmk), "\n")

if (exists("hawaii_master_tmk_list")) {
  clean_master_tmks <- tibble(tmk = as.character(hawaii_master_tmk_list)) %>% 
    clean_hi_tmk("tmk") %>% 
    filter(tmk != "000000000") %>% 
    pull(tmk) %>% 
    unique()
  
  missing_master <- setdiff(clean_master_tmks, table_1_final_final$tmk)
  
  cat("3. All Valid Master TMKs present in Table 1:", length(missing_master) == 0, "\n")
  if (length(missing_master) > 0) {
    cat("   -> Missing TMK count:", length(missing_master), "\n")
    cat("   -> Missing TMKs:", paste(head(missing_master, 10), collapse = ", "), "\n")
  }
}
# check our original dataset to see what these are
otp_missing <- hi_otp %>% filter(tmk %in% missing_master)

iws_hi_permits_missing <- iws_hi_permits %>%  filter(tmk %in% missing_master)

# ok cool so they are just empty. export.


# ==============================================================================
# 7. EXPORT STATEWIDE-COMPATIBLE CSVs
# ==============================================================================

write_csv(
  table_1_final_final,
  file.path(out_dir, "hi_table1_tmk_registry_ALL.csv")
)

write_csv(
  table_2_complete,
  file.path(out_dir, "hi_table2_bedrooms_ALL.csv")
)

write_csv(
  table_3_complete,
  file.path(out_dir, "hi_table3a_permitsApplied_ALL.csv")
)

write_csv(
  table_4_complete_cl,
  file.path(out_dir, "hi_table4_commercial_draft.csv")
)

cat("Big Island Step 4 Synthesis Complete! Files exported to:", out_dir, "\n")


```


Important: Logic for resolving between otp dataset and IWS permits.
We want to preserve records of changes. There are a few types of mistmatches:
1) years match, but permits don't
2) years match, but bedrooms don't
3) years match, but we are missing bedroom information that the old data has?

Some text on the thinking so far: 

Where OTP matched to ANY year in bedrooms, I just left as is for now.

For dwellings that didnt match, i looked at the shrinkage amount: 
**Housing Data Methodology**
To construct a consistent 10-year longitudinal profile of residential capacity, we integrated historical bedroom counts (Column A) with contemporary annual tax assessment records (2000–2024). A primary challenge identified during data reconciliation was "negative bedroom growth," where tax-assessed bedroom counts were significantly lower (up to 4 rooms) than historical baselines. Given that tax assessments strictly adhere to legal building codes (e.g., egress and closet requirements), they often underrepresent informal or non-conforming living spaces. To maintain a realistic proxy for human habitation and land-use intensity, we implemented a "high-water mark" logic: historical values were used as a functional floor. This prevents artificial "capacity drops" in the timeline and ensures that unpermitted but occupied bedrooms—frequently found in local housing contexts—are captured in the longitudinal analysis.

We found that for HI Island, there are 87 permit entries that have a smaller number of bedrooms in the Tax assessment data than in the OTP dataset. Most are different by 1 bedroom, but there are 2 that are different by 4. Spot checking these, they appear to be houses on homesteads. Therefore it is very possible that the tax assessment data is too conservative. 

Dropped by 4 (N = 2)
1) One is located at 32-068 Piha-Kahuku Rd. In the dwell data, we have year built at 2014 so it is possible that it was a 4 bed that was rebuilt to a 1 bed
2) The other is located in the Hawaiian Mountain View Estates. This one could certainly be in a lava field, however it was tax assessed every year since 2010 so I am not going to assume it was in the lava area until spatialization later. 


Dropped by 2 (N = 31)


Dropped by 1 (N = 52)
For these, none of the tax data shows up until 2013. It would be helpful to get the yr built for these. I am wondering if the "otp" data might be more recent than we realize. 
There are a also a number of cases where the permit is old (2009, 2010) but the house assessment doesnt show up until later. I am wondering if these are examples of old ones that were taken out in lava flows, and then were rebuilt. In this case, we'd want to account for it using the OTP beds until lava, and then pick up whatever the tax data has in the meantime. 


So there are a number here where there are more than one cesspool - we
may want to make an assumption that if there is one new permit for
something better than a cesspool (ie a septic tank), its replacing both
cesspools? OR, again another place for sensitivity analysis.

Also an important assumption: if there is a more recent application for
a place with cesspools, would that be for an ADDITIONAL cesspool or to
re-build one of them? not as important for effluent calculation, but
important for Chris and Aimee's work.

Last step! Tally up the changes and record as results below.

```{r}

# this is old, but keeping so I remember how I talled
# upgrade_tally <- last_permit_persentinboth %>% 
#   group_by(upgrade_yn) %>% 
#   summarize(
#     total_n = n()
#   )
# 
# cesspool_upgrade_tally <-  last_permit_persentinboth %>% 
#   group_by(cesspool_upgrade) %>% 
#   summarize(
#     total_n = n()
#   )

# using new table - do we get the same result?

update_summary <- table_1_complete %>% 
  group_by(upgrade_scenario) %>% 
  summarize(
    total_n = n()
  )

```


**Using old method**

WITHOUT COUNTING our more difficult scenarios: we have recorded 1514
upgrades   
351 possible additions or replacements of same kind   
114 where a change occurred on a tmk with many existing osds.

We also have 354 data overlaps, which I'll make sure aren't different,
and then delete duplicates.   
There are 4 where we record a "last class" that is a class lower than OTP. Need to double check these.

Of the upgrades (N = 1514), 
- 1195 are cesspool upgrades. 
- 101 are upgrades that occured on tks with many osds, of which one was a cesspool
- 133 are new cesspool permits since 2010

**New method**
- 1271 cases where one cesspool was likely upgraded to a better system (septic or aerobic)
- 119 are cases where we had multiple osds, followed by one improved application
- 3 are new cesspools on properties that already had multiple osds. that means there are about 130 new applications


#### Create Table 3b Apply bitmask logic to create "status" table from permit data

The additive encoding is what is known as a binary bit mask. Each class represents a power of 2, so every unique sum corresponds to exactly one combination of classes. 

We can think of table 3 as an "event" table, since it records the permit application event. However, we want to create a "status" table, which records the actual OSDS present on the property in each year, BASED on the events we know happened.

We can make multiple versions of this, using different math. In the first, I'll assume that:
1. If there was only one OSDS on the property, the new application is replacing it
2. If there were multiple OSDS on the property, the new application is replacing the grossest, ie highest class. This correspond to the largest power of 2 within the encoded value

This translates to the following math: 
1. Identify the largest power of 2 inside the current value.
2. Subtract that power of 2.
3. Add the new application value.

```{r}


# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% # note, this is a stand in for "pre 2009"
  rename(osds_class_2008 = osds_class_pre2009) %>% 
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(as.numeric(Applied),0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/hi_table3b_activeOSDS.csv"))


```




# OLD CODE 


#### Create Commercial Locations Prep

```{r}

# --------------------Create table 4: Commercial Locations -------------------------------- #
# these are dwellings we do not have any bedroom information for. should be about 807 of these guys

# read in the one from unchanged
table4_newunchaged <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table4_commercial_UNCHANGED_only.csv"))

# read in the one from updated
table4_updated <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table4_updated_draft.csv")) %>% 
  select(tmk, otp_commercial)

# what about new?
new_nodwell <- IWS_table1_tmk_registry %>% 
  filter(is_dwelling == FALSE) %>% #n = 807
  rename(tmk = full_taxid) %>% 
  full_join(table4_updated, by = "tmk") %>% 
  filter(tmk %!in% table2_complete$tmk) # this got rid of the extras

# list for google places api = whereever ^ this one is NA. need to fill all those.

tmks_for_fill <- new_nodwell %>% 
  filter(is.na(otp_commercial)) %>% 
  pull(tmk) # 677

# get addresses

address_for_googleplace <- iws_hi_permits %>% 
  filter(TMK %in% tmks_for_fill) %>% 
  select(-permit_id, -FinalApprovalDate) %>% 
  distinct()

write_csv(address_for_googleplace,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/addresses_for_googleplace.csv"))

```


B) NON DWELL scenarios

The old Hawaii osds data alread has categorized non dwells to some extent. Here I'll see how many of the new, non-dwell permits correspond to these old-places.

```{r}

# read in the non dwells - ie new permits for which we did not have any dwelling information

hi_permits_nodwellinfo <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/hi_iwspermits_nodwellinginfo.csv")) %>% clean_names() #1105


# hi otp commercial
hi_otp_commercial <- hi_otp %>% 
  filter(!is.na(commercial)) %>% 
  select(tmk, commercial, osds_qty, starts_with("class"), effluent,ttl_bdrms,source) %>%
  rename_with(~ paste0(.x, "_from_otp"), .cols = -1)
  

# left join, with hi
hi_newpermits_plusotpinfo <- left_join(hi_permits_nodwellinfo, hi_otp_commercial, by = "tmk")

# how many of the missing permits did we snag?

got_info <- hi_newpermits_plusotpinfo %>% filter(!is.na(commercial_from_otp)) # 289 only

# try another way- filter

hi_newpermits_noDwellfiltered <- hi_permits_nodwellinfo %>% 
  filter(tmk %in% hi_otp_commercial$tmk) # 289. same. 


# what about the rest of these? are the in the full otp, not filtered, dataset?

not_got_info <- hi_newpermits_plusotpinfo %>% filter(is.na(commercial_from_otp)) #816

hi_otp_filtered_byNGI <- hi_otp %>% 
  filter(tmk %in% not_got_info$tmk) %>% 
  mutate(dwell_flag = "tmk not in dwelling data, but someone added bedroom #")# 132

# any others we did not account for? no

######### save the hi_otp by NGI so we can recombine with other dwelling data ################

write_csv(hi_otp_filtered_byNGI,
          file.path())

```

Ok! So, the Non-dwelling data is somewhat solved for Big Island. Here is the breakdown:
- 289 are known commercial locations
- 816 are not known commercial locations, but whoever updated this last has marked all of them with source = something other than Dwell Data. But the good news is that we now have all the info needed to calculate old effluent + if these were upgraded!

```{r}

# create the two tables for the commercial locations

########### table 1: tmk + key identifying info

table_1 <- got_info %>% 
  select(permit_id,tmk, island, starts_with("project"), final_approval_date, commercial_from_otp)

############# table 2: class by year. we need to assign class to the new data, and reshape the old data. 

# disposal types by treatment

minimal_treatment <- c("Seepage Pit",
                       "Cesspool-New")

primary_treatment <- c("Bed",
                       "Trench",
                       "Infiltrators / Chambers",
                       "ET Bed-Irrigation System",
                       "Presby")

unknown_treatment <- c("Other","Unknown")

combo_treatment <- c("Combination of two types")

hi_commercial_odsd_YearClasstable <- got_info %>% 
  select(permit_id, tmk, project_street_address, project_street_address2, final_approval_date, permit_type, disposal_type, ends_with("from_otp")) %>% 
  select(-class_from_otp) %>% 
  mutate(new_osds_class = case_when(
    permit_type == "Septic Tank" & disposal_type %in% primary_treatment ~"1",
    permit_type == "Septic Tank" & disposal_type %in% minimal_treatment ~"2",
    permit_type == "Septic Tank" & disposal_type %in% unknown_treatment ~ "2", 
    permit_type == "Septic Tank" & disposal_type %in% combo_treatment ~ "1", 
    permit_type %in% c("Compost Toilet","Gray Water System","Compost Toilet/Gray Water System") & disposal_type %in% c("Other","Unknown")  ~ "3",
    permit_type %in% c("Compost Toilet","Compost Toilet/Gray Water System") & disposal_type == "Bed" ~ "1",
    permit_type == "Gray Water System" & disposal_type %in% c("Bed","Trench") ~ "3",
    permit_type == "Aerobic Unit" & disposal_type %in% minimal_treatment ~ "4",
    permit_type == "Cesspool-New" ~ "4",
    permit_type == "Cesspool-Existing" ~ "4",
    permit_type == "Aerobic Unit" & disposal_type %in% primary_treatment ~ "1",
    permit_type == "Aerobic Unit" & disposal_type %in% unknown_treatment ~ "3",
    permit_type == "Aerobic Unit" & disposal_type %in% combo_treatment ~ "1",
    permit_type == "Presby (passive, NSF 40)" ~ "1",
    permit_type == "Holding Tank (Government Only)" ~ "1",
    permit_type == "Incinerating Toilet" ~ "1",
    permit_type == "Other" & disposal_type %in% primary_treatment ~ "1", #because it getting soil treatment?
    TRUE ~ "Can't assign MB" # the only one I couldnt assign just had some extra notes on the permit type
  )) %>% 
  mutate(approval_date = mdy(final_approval_date),
         year = year(approval_date))

hi_commercial_odsd_new_clean <- hi_commercial_odsd_YearClasstable %>% 
  select(permit_id, tmk,starts_with("project"),final_approval_date,year,new_osds_class) %>% 
  pivot_wider(id_cols = c("permit_id","tmk","project_street_address","project_street_address2","final_approval_date"), names_from = year, values_from = new_osds_class)

hi_commercial_osds_olds <- got_info %>% 
  select(permit_id,tmk,project_street_address, project_street_address2,final_approval_date, ends_with("from_otp")) %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity != -9999) %>% 
  mutate(year = 2010) %>% 
  rename(ttl_osds_2010 = osds_qty_from_otp) %>% 
  mutate(osds_class_otp = case_when(
    str_detect(class, "class_iv") ~ 8, # cesspool 
    str_detect(class, "class_iii") ~ 4,
    str_detect(class, "class_ii") ~ 2,
    str_detect(class, "class_i") ~ 1
  )) %>%
  pivot_wider(id_cols = c("permit_id","tmk","project_street_address","final_approval_date","ttl_osds_2010"), names_from = c("class","year"), values_from = c("osds_class_otp")) %>%
  select(-class_from_otp_2010) %>% 
  mutate(combo_code = rowSums(across(starts_with("class")), na.rm = TRUE))

# join together

hi_commercial_olds_cl <- hi_commercial_osds_olds %>% 
  select(-starts_with("class")) %>% 
  rename(prior_class = combo_code)


table_2_final <- left_join(hi_commercial_olds_cl,hi_commercial_odsd_new_clean, by = c(
  "permit_id",
  "tmk",
  "project_street_address",
  "final_approval_date"
))

# use anti join to check that all of info in table 1 are in table 2

missing_rows <- anti_join(table_1, table_2_final, by = c(
  "permit_id",
  "tmk",
  "project_street_address",
  "final_approval_date"
)) # 0, yay!!



# save to drop box

write_csv(table1,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/commercial_finaltables/hi_comm_table1.csv"))

write_csv(table_2_final,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/commercial_finaltables/hi_comm_table2.csv"))

```

