---
title: "Kauai OSDS Timeseries"
author: "Madeline Berger"
date: "2026-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(here)

source(here("workflow/setup.R"))



```

## New Data: clean up and join 

### Step 1: Clean Up Dwell Data

The Kauai data comes to us pre-parsed, which is great. Here we will just generate the same summary tables as we did for Oahu. We will have to read in each excel file here and do some cleaning. 

The data was sent as one giant CSV.

- first I'll rename the columns
- select the ones we need for the "report"
- run the function


```{r dwell clean}

# use data table because this csv is large

library(data.table)
library(janitor)

kauai_raw_dwell <- fread(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Kauai_DwellRaw_1122025/KAUAI_CO_DWELDAT_UHDataRequest_010826.csv"))

kauai_clean_names <- kauai_raw_dwell %>% clean_names()

```



Create summary report - same function adjusted for column names

```{r}


create_summary_report = function(df){
  
  ### for testing ###
  
  #df <- maui_dwell_dfs[[1]]
  
  ######################
  
  dwells_clean_names <- janitor::clean_names(df) %>% mutate(card_char = as.character(card))
  
  #1 tmks per year, how many have > 1 card
  tmk_multi_by_year <- dwells_clean_names %>%
    group_by(tax_year, parid_full_parcel_id) %>%
    summarize(n_cards = n_distinct(card_char), .groups = "drop") %>%
    group_by(tax_year) %>%
    summarize(
      total_TMKs = n(), # total tmks
      TMKs_with_multiple_cards = sum(n_cards > 1), # how many with multiple cards
      pct_multi = 100 * TMKs_with_multiple_cards / total_TMKs) # percent with more than 1 card

  #2 beds and baths per year
  bedsbath_per_year <- dwells_clean_names %>% 
  mutate(bedrooms = ifelse(is.na(bedrooms), 0, as.numeric(bedrooms)),
         full_baths = ifelse(is.na(fullbath), 0, as.numeric(fullbath))) %>% # what if its NA?
  group_by(tax_year) %>% 
  summarize(
    n_beds = sum(bedrooms, na.rm = TRUE),
    n_baths = sum(full_baths, na.rm = TRUE)
  )
  
  #3  percent condo in each year. only relevent for Maui
  
  # condo_summary <- dwells_clean_names %>% 
  #   mutate(condo_type_char = str_to_lower(
  #     case_when(
  #     condo_type == 1 ~ "SINGLE FAMILY",
  #     condo_type == 2 ~ "DUPLEX",
  #     condo_type == 3 ~ "TOWNHOUSE",
  #     condo_type == 4 ~ "LOWRISE",
  #     condo_type == 5 ~ "HIGHRISE",
  #     TRUE ~ "no condo"
  #   ))) %>% 
  #   group_by(condo_type_char, tax_year) %>% 
  #   summarize(
  #     number_per_type = n()
  #   ) %>% 
  #   pivot_wider(id_cols = tax_year, names_from = condo_type_char, values_from = number_per_type)
  
  # bind all toghter
  
  per_year_info <- left_join(tmk_multi_by_year,
                             bedsbath_per_year,
                             by = "tax_year")
  
  # per_year_info_2 <- left_join(per_year_info,
  #                            condo_summary,
  #                            by = "tax_year")
  # 
  
  return(per_year_info)

  
}


# apply to kauai


kauai_summary_report <- create_summary_report(kauai_raw_dwell)


write_csv(kauai_summary_report,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/kauai_dwell_summary_report.csv"))

# collapse to get list of unique tmks. initially we have N = 534754

kaui_clean_names_coll <- kauai_clean_names %>%
  select(-tax_year)%>%
  mutate(full_taxid = as.character(tmk_includes_island_4)) %>% 
  distinct() #N = 53304

```


### Step 2: Bring in IWS Permit data 

- This is the "heart" of the database
- Execute Join
- Develop QA QC report 2 on which drop out and therefore likely represent commercial parcels. 


```{r}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))



# filter out Kauai using tmk digits, and then add an id by row 


# filter out just the permits in maui, and create permit ID tracker

iws_kauai_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"4")) %>% 
  rename(full_taxid = tmk_char) %>%  
  rowid_to_column("permit_id") #2406 permits since 2010

# join with collapsed tmks, and filter with tmks.the row number for both should be the same, and < 2406

iws_kauai_dwell_join <- inner_join(iws_kauai_permits, kaui_clean_names_coll, by = "full_taxid") %>% 
  distinct() # 12252 rows, 1741 tmks

iws_kauai_dwell_join_norepeats <- iws_kauai_permits %>% 
  filter(full_taxid %in% kaui_clean_names_coll$full_taxid) # 2115 rows, 1741 tmks. cool

# filter out permits without dwell data


iws_kauai_NOdwell <- iws_kauai_permits %>% 
  filter(full_taxid %!in% iws_kauai_dwell_join$full_taxid) # 291

write_csv(iws_kauai_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/kauai_iwspermits_nodwellinginfo.csv"))


```


Results so far:
- There are 2406 permits issued for Kauai in this dataset
- There are 2115 with dwell information, representing 1741 TMKS 
- There are 291 without Dwell information 
- Condos, particularly low rise and high rise have increased a lot in Maui over the past 10 years

### Step 2b: Identifying Upgrade Scenarios

Here we will work to identify easy, tricky and hard scenarios. The steps are as follows:
1. Create a flag that flags if a TMK is repeated in the same year or different year
2. Count up scenarios and save


```{r}

# working with just the permit data, analyze duplicates for 1 permit issued to same place, and which years

iws_kauai_permits$is_duplicate <- duplicated(iws_kauai_permits$TMK) | duplicated(iws_kauai_permits$TMK, fromLast = TRUE)
  
iws_kauai_permits_dups <- filter(iws_kauai_permits, is_duplicate == TRUE) # 805

iws_kauai_permits_cl <- iws_kauai_permits %>% 
  clean_names() %>%
  mutate(approval_date = mdy(final_approval_date),
         year = year(approval_date))

# create flag indicating if the duplicate is for the same year, or a different year

iws_kauai_permits_flag <- iws_kauai_permits_cl %>% 
  group_by(full_taxid) %>% 
  mutate(
    n_rows = n(),
    n_year = n_distinct(year)
  ) %>% 
  ungroup() %>% 
  mutate(
    osds_flag = case_when(
      n_rows > 1 & n_year > 1 ~ "possbile update or later addition",
      n_rows > 1 & n_year == 1 ~ "possible multiple osds permits",
      TRUE ~ NA_character_
    )
  )

check_dups_again <- filter(iws_kauai_permits_flag, is_duplicate == TRUE)

# count up the different scenarios

iws_duplicate_scenarios_kauai <- iws_kauai_permits_flag %>% 
  group_by(osds_flag) %>% 
  summarize(
    total_tmks = length(unique(full_taxid))
  )

# save to qaqc

write_csv(iws_duplicate_scenarios_kauai,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/kauai_iwspermits_scenario_summary.csv"))




```


Results: 
- 805 of the PERMITS issued during this time frame were issued to the same TMK, which represents 252 tnks
  - 82 indicate multiple permits for the same tmk for the same year
  - 170 indicate multiple permits for different years
  - The rest are for unique buildings or addresses on the same tmk
- 1601 are one building, one permit, one year

### Step 3: Assign Class to Permits

The assignment of class is based on a previously determined scheme by Whittier and El Kadi (2014). The scheme is as follows:

Class I = any system receiving soil treatment, with disposal types as bed, trench, and infiltration chambers
Class II = Septic system discharging into seepage pit
Class III = Effluent received primary treatment only 
Class IV = Aerobic units discharging into a seepage pit. The effluent receives primary and secondary treatment. 

Class V = multiple with at least one cesspool. So note this will only apply to those that weʻve flagged above that may have multiple

```{r}

library(janitor)

# disposal types by treatment

minimal_treatment <- c("Seepage Pit",
                       "Cesspool-New")

primary_treatment <- c("Bed",
                       "Trench",
                       "Infiltrators / Chambers",
                       "ET Bed-Irrigation System",
                       "Presby")

unknown_treatment <- c("Other","Unknown")

combo_treatment <- c("Combination of two types")


# apply to kauai data

iws_kauai_permits_withclass <- iws_kauai_permits_flag %>%
  clean_names() %>% 
  #filter(is.na(osds_flag)) %>% 
  mutate(osds_class = case_when(
    permit_type == "Septic Tank" & disposal_type %in% primary_treatment ~"1",
    permit_type == "Septic Tank" & disposal_type %in% minimal_treatment ~"2",
    permit_type == "Septic Tank" & disposal_type %in% unknown_treatment ~ "2", 
    permit_type == "Septic Tank" & disposal_type %in% combo_treatment ~ "1", 
    permit_type %in% c("Compost Toilet","Gray Water System","Compost Toilet/Gray Water System") & disposal_type %in% c("Other","Unknown")  ~ "3",
    permit_type %in% c("Compost Toilet","Compost Toilet/Gray Water System") & disposal_type == "Bed" ~ "1",
    permit_type == "Gray Water System" & disposal_type %in% c("Bed","Trench") ~ "3",
    permit_type == "Aerobic Unit" & disposal_type %in% minimal_treatment ~ "4",
    permit_type == "Cesspool-New" ~ "4",
    permit_type == "Aerobic Unit" & disposal_type %in% primary_treatment ~ "1",
    permit_type == "Aerobic Unit" & disposal_type %in% unknown_treatment ~ "3",
    permit_type == "Aerobic Unit" & disposal_type %in% combo_treatment ~ "1",
    permit_type == "Presby (passive, NSF 40)" ~ "1",
    permit_type == "Holding Tank (Government Only)" ~ "1",
    permit_type == "Incinerating Toilet" ~ "1",
    TRUE ~ "Can't assign MB" # the only one I couldnt assign just had some extra notes on the permit type
  )) %>% 
  mutate(osds_class = ifelse(osds_class == "Canʻt assign MB","1",osds_class)) # fix this one by hand



```

### Step 4: Count addresses and permits by TMK & find min and max years

Parallel to the class assignment, here I examine the address column to try and figure out if we can ID multiple building scenarios just from the permit data. 

```{r}

# how many addresses per TMK can we identify?
address_per_tmk <- iws_kauai_permits_cl %>%
  group_by(tmk) %>% 
  mutate(address_full = ifelse(is.na(project_street_address2),project_street_address,paste(project_street_address,"-",project_street_address2))) %>% 
  summarize(
    n_addresses = length(unique(address_full)), # how many buildings on this TMK
    n_permits = length(permit_id), # how many permits have they applied for in this time period
    min_year = min(year),
    last_year = max(year) # when these permits were issued
  ) # write this one out

# how many TMKS have more than one building?
nrow(address_per_tmk[address_per_tmk$n_addresses > 1, ]) #205

# add flag at the TMK level 
address_per_tmk_flag <- address_per_tmk %>% 
  mutate(building_permit_scenario = case_when(
    n_addresses == 1 & n_permits == 1  ~ "One building, one permit",
    n_addresses == 1 & n_permits > 1 ~ "One buidling, multiple permits",
    n_addresses > 1 & n_permits == 1 ~ "Multiple buildings, one permit",
    n_addresses > 1 & n_permits > 1 ~ "Multiple buildings, Mulitple permits"
  ),
  year_scenario = case_when(
    last_year == min_year ~ "All the same year",
    TRUE ~ "Mulitple Years"
  ))


scenario_summary <- address_per_tmk_flag %>% 
  group_by(building_permit_scenario,year_scenario) %>% 
  summarize(
    tmk_per_scenario = n()
  )

# write scenario summary  
write_csv(address_per_tmk,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/kauai_permits_and_buildings_pertmk.csv"),
         append = FALSE)


write_csv(scenario_summary,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/qaqc/kauai_building_scenario.csv"),
         append = FALSE)


```

**Results** from this tell us that:
- 205 of tmks that applied for permits in this time period have multiple buildings. 
- 1601 tmks are a one building, one permit scenario. Easy!
- 20 are one building, multiple permits, multiple years. This suggests a replacement or upgrade. Also easyish
- 27 are one building, multiple permits, all one year. 
- 55 are multiple permits for multiple buildings and one year. Easy if 1) all permits the same class and 2) building # = permit number. 
- 150 are multiple buildings, multiple permits, and multiple years. 


```{r}


easys <- address_per_tmk_flag %>% 
  filter(building_permit_scenario == "One building, one permit")

iws_kauai_permits_easyscenario <- iws_kauai_permits_withclass %>% 
  filter(tmk %in% easys$tmk) %>% 
  mutate(dwell_flag = ifelse(full_taxid %in% iws_kauai_dwell_join$full_taxid,"Dwelling","Other")) %>% 
  select(-n_rows,-n_year, -osds_flag)

write_csv(iws_kauai_permits_easyscenario,
         file.path(driver_dir,"4_OSDS_N_Runoff/output-data/iws_kauai_permits_easyscenario.csv"),
         append = FALSE)

# medium scenario is one building, multiple permits. assumption here is that these are multiple osds per building IF the year is the same. if not, its an upgrade. 

tricky_tmks <- address_per_tmk_flag %>% 
  filter(building_permit_scenario %in% c("One buidling, multiple permits"))

iws_kauai_permits_mediumscenario <- iws_kauai_permits_withclass %>% 
  filter(tmk %in% tricky_tmks$tmk) %>% 
  mutate(dwell_flag = ifelse(full_taxid %in% iws_kauai_dwell_join$full_taxid,"Dwelling","Other")) %>% 
  select(-n_rows,-n_year, -osds_flag)


# how many of these tmks are also in our dwell dataset? how many have multiple cards? 

hard_tmks <- address_per_tmk_flag %>% 
  filter(building_permit_scenario %in% c("Multiple buildings, Mulitple permits"))


iws_kauai_permits_hardscenario <- iws_kauai_permits_withclass %>% 
  filter(tmk %in% hard_tmks$tmk) %>% 
  mutate(dwell_flag = ifelse(full_taxid %in% iws_kauai_dwell_join$full_taxid,"Dwelling","Other")) %>% 
  select(-n_rows,-n_year, -osds_flag)





```

How many of the hard tmks are non dwellings?
```{r}


dwell_hard_tmks <- iws_kauai_dwell_join %>% filter(full_taxid %in% hard_tmks$tmk)

length(unique(dwell_hard_tmks$full_taxid)) # 173

length(unique(hard_tmks$tmk)) #205

# so indicates that 32 of the hard scenarios are non-dwellings




```

## Compare to Past Datasets to Complete the Time Series 

We have to bring in the past OSDS dataset to have a record of what each tmk was bedroom wise and class wise prior to 2009 (which is when our permit data starts). Without it, we can only track changes and upgrades within the 2009 - 2020. Weʻll add this as a column. Once we do this, we can generate cesspool upgrade list. 

There are two main categories of comparison from Step 2;
- Easy situations where we have one building, one permit per tmk
- More complex ones.


### A) Compare to OTP (2010 timestamp)

Let's start with the easy list. We'll read in the old OSDS dataset and assume that what ever is recorded for that tmk in there is what it was before 2010. If our new dataset has something different, we know its changed. 
Important Assumption from Julia's work: if a location goes from septic, or better, to CESSPOOL is probably wrong? 


First read in most recently available data for Kauai and Explore: 

```{r}

kauai_otp <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Kauai.csv"))

```

Like Maui, for Kauai the data I downloaded from HI GIS is a bit different than OAHU because it is in wide format, with a column for each class, and the value in the column indicating the quantity of that class. there is also a total quantity.

We only have classes 1 - 4 in this data.


From the metadata, this represents: 


We only have classes 1 - 4 in this data. From the metadata, this represents: 

- CLASS_I: Class I OSDS are those systems where the effluent is subjected to soil treatment 
- CLASS_II: Class II OSDS are septic systems discharging to a seepage pit.  
- CLASS_III: Class III OSDS are aerobic systems discharging to a seepage pit.
- CLASS_IV: Class IV OSDS are c  cesspools.  
- EFFLUENT: The total effluent flux in million gallons per day discharged within the TMK parcel. 


This makes it tricky to match it up to our "one building one permit" data. So what Iʻm going to do, which isnt a perfect solution but will help us sort out changes for now, is use additive encoding, that creates unique codes for each class AND each combination of class. The rule is for category n, use 2^(n-1) for the code. Therefore the class column becomes: 

| Combination                           | Sum |
| ------------------------------------- | --- |
| Class 1                               | 1   |
| Class 2                               | 2   |
| Class 1 + Class 2                     | 3   |
| Class 3                               | 4   |
| Class 1 + Class 3                     | 5   |
| Class 2 + Class 3                     | 6   |
| Class 1 + Class 2 + Class 3           | 7   |
| Class 4                               | 8   |
| Class 1 + Class 4                     | 9   |
| Class 2 + Class 4                     | 10  |
| Class 1 + Class 2 + Class 4           | 11  |
| Class 3 + Class 4                     | 12  |
| Class 1 + Class 3 + Class 4           | 13  |
| Class 2 + Class 3 + Class 4           | 14  |
| Class 1 + Class 2 + Class 3 + Class 4 | 15  |


```{r}

kauai_otp_clean <- kauai_otp %>% 
  clean_names() %>% 
  select(tmk, starts_with("class"), "osds_qty") # n = 13351 !

# read in the "easy scenario" new data. check for any duplciates

onebuilding_onepermit <- read.csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/iws_kauai_permits_easyscenario.csv")) #1601

onebuilding_onepermit_distinct <- onebuilding_onepermit %>% 
  select(-permit_id) %>% 
  distinct() # none, cool!

# -----------  Q1:  do all these new ones exist in the old data? if not, then they are NEW osds builds ------------- #

otp_kauai_filtered_byOBOP <- kauai_otp_clean %>% 
  mutate(tmk = as.integer(tmk)) %>% 
  filter(tmk %in% onebuilding_onepermit$tmk) %>% 
  distinct() %>% # KEY: remove EXACT duplicates. We are treating these as errors
  rowid_to_column("newbuild_permit_id") %>% 
  mutate(year_issued = 2010) %>% # note this isn't right, just a place hodler for "pre 2010" to maintain consistent data structure in this column 
  mutate(change_flag = "new permit detected since otp") # 1144


otp_kauai_NOTinOBOP <- kauai_otp_clean %>%
  mutate(tmk = as.integer(tmk)) %>% 
  filter(tmk %!in% onebuilding_onepermit$tmk) %>% 
  rowid_to_column("old_permit_id") %>% 
  mutate(change_flag = "no new permit application since otp") 
  

write_csv(otp_kauai_NOTinOBOP,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/otp_osds_wo_permitupgrade_kauai.csv")) 


# -----------  Q2:  Which of the new builds are new OSDS, rather than builds on tmks with existing OSDS? ------------- #


onebuilding_onepermit_NEW_since2010 <- onebuilding_onepermit %>% 
  filter(tmk %!in% otp_kauai_filtered_byOBOP$tmk) # 457 

length(unique(onebuilding_onepermit_NEW_since2010$tmk)) #457 yay

# use set diff to see if its the same
tmks_not_inbob <- setdiff(onebuilding_onepermit$tmk, otp_kauai_filtered_byOBOP$tmk) #457 YAY

onebuilding_onepermit_NEW_since2010_otherway <- onebuilding_onepermit %>% 
  filter(tmk %in% tmks_not_inbob) 


write_csv(onebuilding_onepermit_NEW_since2010_otherway,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/kauai_osds_new_since_otp.csv"))



# --------------- Old Data: Data Reshape ------------------------------------- #


# Here I take the tmks present in OTP AND the new permit data, which represent upgrades. I reassign class to match the additive encoding schema, and then pivot wider. 



otp_kauai_filtered_byOBOP_wide <- otp_kauai_filtered_byOBOP %>% 
  select(-class) %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity != 0) %>% 
  rename(ttl_osds_2010 = osds_qty) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv" ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii" ~ 2,
    class == "class_i" ~ 1
  )) %>%
  pivot_wider(id_cols = c("newbuild_permit_id","tmk","ttl_osds_2010"), 
              names_from = c("year_issued","class"), values_from = c("quantity"))



dups_check <- otp_kauai_filtered_byOBOP_wide %>% mutate(duplicated = duplicated(tmk)) # there was one duplicate the first time i ran this. Data is exactly the same, so I'm going to delete it. added that code at the beginning of the previous chunk with a distinct

otp_kauai_morethan1osds <- otp_kauai_filtered_byOBOP_wide %>% 
  filter(ttl_osds_2010 > 1) #493 ~ most of them!

# now, add the encoding 

otp_kauai_filtered_byOBOP_wideENC <- otp_kauai_filtered_byOBOP %>% 
  select(-class) %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(quantity != 0) %>% 
  rename(ttl_osds_2010 = osds_qty) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv" ~ 8, # cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii" ~ 2,
    class == "class_i" ~ 1
  )) %>%
  pivot_wider(id_cols = c("newbuild_permit_id","tmk","ttl_osds_2010"), names_from = c("year_issued","class"), values_from = c("osds_class_otp")) %>%
  mutate(combo_code = rowSums(across(starts_with("2010")), na.rm = TRUE))

otp_kauai_filtered_byOBOP_longENC <- otp_bob_filtered_byOBOP_wideENC %>% # collapse the wide columns
  select(-starts_with("2010"), -newbuild_permit_id)

# --------------- New Data: Data Reshape ------------------------------------- #


# now pivot the NEW data, for now just the "easy", that is also in OTP, one building one permit category, wider to match

onebuild_onepermit_inbob <- onebuilding_onepermit%>% 
  filter(tmk %in% otp_kauai_filtered_byOBOP$tmk)

onebuild_onepermit_kauai_wide <- onebuild_onepermit_inbob %>% 
  pivot_wider(id_cols = c("permit_id","tmk"), names_from = "year", values_from = c("permit_type","osds_class"))

dups_obop_kauai <- onebuild_onepermit_kauai_wide %>% mutate(duplicated = duplicated(tmk)) # no duplicates. great. 


# -------------- Combine:  Join our two wide tables together ----------------------------------- #

# IMPORTANT note: since 493 of the bob data have more than one OSDS, but the new data here are examples of just one building, one new permit, if any of the new data match to the old ones with more than one osds, weʻll have to make an assumption about which one was updated in this new data. 

all_osds_present_inboth <- left_join(dups_obop_kauai, otp_kauai_filtered_byOBOP_longENC, by = c("tmk")) # none dropped in the join. nice

# so the next analysis will determine the most recent year a tmk was upgraded. We can add a flag for if there was an upgrade and there are multiple OSDS on the property. hoping that most of these did not have an upgrade. 


all_permit_present_inboth_long <- all_osds_present_inboth %>% 
  rename(otp_class_2010 = combo_code) %>% # this is what it was in 2010
  select(-starts_with("permit_type")) %>% # we dont need this right now
  pivot_longer(cols = starts_with("osds"),
               names_to = "year",
               values_to = "class",
               values_drop_na = TRUE) %>% 
  mutate(
    year_num = as.integer(str_remove(year, "osds_class_"))  # extract numeric year rather than colname with year in it
  )


# here we'll use some logic to create a yes/no flag if upgraded, and a specific text flag for what kind of upgrade it was 


last_permit_persentinboth <- all_permit_present_inboth_long %>% 
  group_by(tmk) %>% 
  slice_max(year_num, n = 1, with_ties = FALSE) %>% # get the latest year we have records for a given tmk
  mutate(class_last = as.numeric(class),
         class_last_year = year_num,
         class_otp = as.numeric(otp_class_2010)) %>% 
  mutate(upgrade_yn = case_when(
    class_last < class_otp & class_otp %in% c(1,2,4,8) ~ "upgrade occured on a single osds",
    class_last < class_otp & class_otp %!in% c(1,2,4,8) ~ "change occured on one of many existing osds",
    class_last == class_otp & class_otp %in% c(1,2,4,8) & class_last_year > 2010 ~ "addition",
    class_last == class_otp & class_otp %in% c(1,2,4,8) & class_last_year <= 2010 ~ "data overlap",
    class_last > class_otp  ~ "check class assignment"
  )) %>% 
  mutate(cesspool_upgrade = case_when(
    class_otp == 8 & class_last < 4 ~ "cesspool upgraded",
    class_otp %in% c(9,10,11,12,13,14,15) & class_last < 4 ~ "maybe cesspool upgraded - multiple existing osds",
    class_last == 4 & class_last_year > 2010 ~ "new cesspool permit",
    class_otp == 4 & class_last == 4 & class_last_year %in% c(2009,2010) ~ "duplicate, not upgraded",
    class_otp %!in% c(9,10,11,12,13,14,15) ~ "na"
  ))


write_csv(last_permit_persentinboth,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/kauai_all_possible_OSDSupgrades_withflag.csv"))

```


So there are a number here where there are more than one cesspool - we may want to make an assumption that if there is one new permit for something better than a cesspool (ie a septic tank), its replacing both cesspools? OR, again another place for sensitivity analysis. 

Also an important assumption: if there is a more recent application for a place with cesspools, would that be for an ADDITIONAL cesspool or to re-build one of them? not as important for effluent calculation, but important for Chris and Aimee's work. 

Last step! Tally up the changes and record as results below. 

```{r}

upgrade_tally <- last_permit_persentinboth %>% 
  group_by(upgrade_yn) %>% 
  summarize(
    total_n = n()
  )

cesspool_upgrade_tally <-  last_permit_persentinboth %>% 
  group_by(cesspool_upgrade) %>% 
  summarize(
    total_n = n()
  )


```


RESULTS:

WITHOUT COUNTING our more difficult scenarios: 
we have recorded 636 upgrades
123 possible additions or replacements of same time
280 where a change occurred on a tmk with many existing osds. 

We also have 97 data overlaps, which I'll make sure aren't different, and then delete duplicates.
There are 8 where we record a "last class" that is a class lower than OTP. Given the number of multiunits in Kauai, I'm not super surprised but we should also double check these. 

Of the upgrades (N = 636): 
- 618 are cesspool upgrades
- an additional 265 are upgrades that occurred on tmks with many osds, one of which is a cesspool
- 12 are new cesspool permits since OTP

## Combine All Together

The key here is that once we combine old and new data, I create an authoritative list. 
