---
title: "Kauai OSDS Timeseries Streamlined"
author: "Madeline Berger"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))


```

### Step 0: Create authoritative tmk lists

```{r}

# new data: iws permits

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv")) %>% 
  clean_names()

iws_kauai_permits <- iws_permit_data %>%
  mutate(tmk = as.character(tmk)) %>% 
  filter(str_starts(tmk,"4"))

iws_kauai_permits_tmk <- iws_kauai_permits %>% pull(tmk)


# old data: otp 

# read in old kauai data

otp_kauai_raw <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Kauai.csv")) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk))

otp_kauai_tmk <- otp_kauai_raw %>%
  pull(tmk)

# new ones:

kauai_unchanged <- otp_kauai_raw %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %!in% iws_kauai_permits_tmk) # 12022

unchanged_tmks <- kauai_unchanged %>% pull(tmk) %>% unique() #11870

kauai_updated <- otp_kauai_raw %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% iws_kauai_permits_tmk) #1329

updated_tmks <- kauai_updated %>% pull(tmk) %>% unique() # 1328

kauai_new <- iws_kauai_permits %>% 
  filter(tmk %!in% otp_kauai_tmk) # 764

new_tmks <- kauai_new %>% pull(tmk) %>% unique() #525
# create authoritative tmk list - add ups!

master_tmk_list <- c(iws_kauai_permits_tmk,otp_kauai_tmk)
master_tmk_list <- unique(master_tmk_list) # 13351

kauai_master_tmk_list <- master_tmk_list


```



### Step 1: Clean Up Dwell Data

The Kauai data comes to us pre-parsed, which is great. Here we will just
generate the same summary tables as we did for Oahu. We will have to
read in each excel file here and do some cleaning.

The data was sent as one giant CSV.

-   first I'll rename the columns
-   select the ones we need for the "report"
-   run the function

```{r dwelldata}

library(data.table)
library(janitor)

kauai_raw_dwell <- fread(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Dwelling_Data_Raw/Kauai_DwellRaw_1122025/KAUAI_CO_DWELDAT_UHDataRequest_010826.csv"))

kauai_dwell_allentries <- kauai_raw_dwell %>% 
  clean_names() %>%
  mutate(
      parid = str_trim(as.character(parid_full_parcel_id)),
      tax_id = substr(parid,1,8),
      tmk = paste0("4",tax_id)
      )

length(unique(kauai_dwell_allentries$tmk)) # tmk = 18966

# collapsed for join

kauai_dwell_tmks_forjoin <- kauai_dwell_allentries %>%
  select(-tax_year)%>%
  distinct() 

length(unique(kauai_dwell_tmks_forjoin$tmk)) # tmk = 18966, cool

write_csv(kauai_dwell_allentries,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/kauai_dwell_clean_alltmks.csv"))

kauai_dwell_allentries <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/kauai_dwell_clean_alltmks.csv"))

kauai_dwell_osds <- kauai_dwell_allentries %>% filter(tmk %in% master_tmk_list) %>% 
  pull(tmk) %>% 
  unique() #13385

kauai_dwell_osds_df <- data.frame(
  dwells_w_osds_tmk = kauai_dwell_osds
)

write_csv(kauai_dwell_osds_df,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/kauai_dwell_osds_tmks.csv"))


```



### Step 2: Bring in Permit Data (IWS, Dept. of Health) and Build First Tables with New Data

Next we'll bring in the new IWS permit data and create our database for any new entries since 2010. Step 3 will focus on combining these new data with legacy data from OTP or other past efforts. 3

```{r permits}

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_kauai_permits <- iws_permit_data %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(str_starts(tmk,"4")) %>% 
  rowid_to_column("permit_id") %>%
  mutate(approval_date = mdy(final_approval_date),
         year = year(approval_date)) #2406 rows, 1853 permits

tmks_permits <- unique(iws_kauai_permits$tmk) #1853 tmks with new permits since 2010

# execute join to see which tmks have dwell data

iws_kauai_dwell_join <- inner_join(iws_kauai_permits, kauai_dwell_tmks_forjoin, by = "tmk") %>% 
  distinct() #12253 rows and 1741 tmks

iws_kauai_dwell_join_norepeats <- iws_kauai_permits %>% 
  filter(tmk %in% kauai_dwell_tmks_forjoin$tmk) #1741 tmks

# now filter out ones that have no data and save

iws_kauai_NOdwell <- iws_kauai_permits %>% 
  filter(tmk %!in% iws_kauai_dwell_join$tmk) # 291 rows, 112 tmks. perfect

write_csv(iws_kauai_NOdwell,
         file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_iwspermits_nodwellinginfo.csv"))

```

**Results so Far**:
- There were 2406 permits issued for HI island over the past 10 years, pertaining to 1853
- 1741 have dwelling information, and this represents 2094 tmks
- 112 tmks do not have dwelling information


#### Step 2b: Clean and Pivot New Permits

Instead of doing a bunch of stuff that identifies easy, tricky and hard scenarios, I'm condensing everything here into one big clean. 
are as follows: 
1. Assign class, using a custom function
2. Create master list (table 1)
3. Create tables 2 and 3 based on master list. 

##### Wastewater Class Encoding Reference Table

We only have classes 1 - 4 in this data. From the metadata, this
represents:

-   CLASS_I: Class I OSDS are those systems where the effluent is
    subjected to soil treatment
-   CLASS_II: Class II OSDS are septic systems discharging to a seepage
    pit.\
-   CLASS_III: Class III OSDS are aerobic systems discharging to a
    seepage pit.
-   CLASS_IV: Class IV OSDS are cesspools.\
-   EFFLUENT: The total effluent flux in million gallons per day
    discharged within the TMK parcel.
    

| Detected Classes | Encoded Sum | Description |
| :--- | :---: | :--- |
| **Class 1** | 1 | Cleanest Septic Type |
| **Class 2** | 2 | Secondary Septic Type |
| **Class 1 + 2** | 3 | **Combination of Class 1 + Class 2** |
| **Class 3** | 4 | Aerobic System |
| **Class 1 + 3** | 5 | Class 1 Septic + Aerobic |
| **Class 2 + 3** | 6 | Class 2 Septic + Aerobic |
| **Class 1 + 2 + 3** | 7 | Classes 1, 2, and 3 |
| **Class 4** | 8 | Cesspool |
| **Class 1 + 4** | 9 | Class 1 + Class 4 |
| **Class 2 + 4** | 10 | Class 2 + Class 4 |
| **Class 1 + 2 + 4** | 11 | Classes 1, 2, and 4 |
| **Class 3 + 4** | 12 | Aerobic + Class 4 |
| **Class 1 + 3 + 4** | 13 | Classes 1, 3, and 4 |
| **Class 2 + 3 + 4** | 14 | Classes 2, 3, and 4 |
| **All (1, 2, 3, 4)** | 15 | Full suite of permits |


```{r new-tables}

library(tidyverse)
library(janitor)

# ==============================================================================
# HELPER FUNCTIONS & DISPOSAL CLASSIFICATION
# ==============================================================================

calc_mode <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Kauai TMK Standardizer (Ensures 9-digit format starting with '4')
clean_kauai_tmk <- function(df, target_col = "tmk") {
  col_names <- names(df)
  actual_col <- if (target_col %in% col_names) {
    target_col
  } else if ("full_taxid" %in% col_names) {
    "full_taxid"
  } else if ("TMK" %in% col_names) {
    "TMK"
  } else {
    target_col
  }
  
  df %>% 
    mutate(
      raw_val = as.character(.data[[actual_col]]),
      # Convert scientific notation if present (e.g. "4.11e+08")
      raw_val = if_else(str_detect(raw_val, "e\\+"), sprintf("%.0f", as.numeric(raw_val)), raw_val),
      # Retain numeric digits only
      digits = str_replace_all(raw_val, "[^0-9]", ""),
      # Truncate CPR suffixes (12+ digits) to 9-digit parcel level
      digits = if_else(nchar(digits) >= 9, substr(digits, 1, 9), digits),
      # Prepend missing Kauai island prefix '4' if 8 digits
      digits = case_when(
        nchar(digits) == 8 ~ paste0("4", digits),
        nchar(digits) == 9 & str_starts(digits, "0") ~ paste0("4", substr(digits, 2, 9)),
        TRUE ~ str_pad(digits, width = 9, side = "left", pad = "0")
      )
    ) %>% 
    mutate(tmk = digits) %>% 
    select(-any_of(c("raw_val", "digits"))) %>% 
    filter(!is.na(tmk) & tmk != "" & tmk != "000000000")
}

# Disposal category vectors
primary_disposal <- c("Bed", "Trench", "Infiltrators / Chambers", "ET Bed-Irrigation System", "Presby")
minimal_disposal <- c("Seepage Pit", "Cesspool-New")
unknown_disposal <- c("Other", "Unknown")

# Class assignment function
assign_osds_class <- function(p_type, d_type) {
  case_when(
    p_type == "Septic Tank" & d_type %in% primary_disposal ~ "1",
    p_type == "Septic Tank" & d_type %in% c(minimal_disposal, unknown_disposal) ~ "2",
    p_type == "Septic Tank" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Compost Toilet", "Gray Water System", "Compost Toilet/Gray Water System") & 
      d_type %in% unknown_disposal ~ "3",
    p_type %in% c("Compost Toilet", "Compost Toilet/Gray Water System") & d_type == "Bed" ~ "1",
    p_type == "Gray Water System" & d_type %in% c("Bed", "Trench") ~ "3",
    
    p_type == "Aerobic Unit" & d_type %in% primary_disposal ~ "1",
    p_type == "Aerobic Unit" & d_type %in% minimal_disposal ~ "4",
    p_type == "Aerobic Unit" & d_type %in% unknown_disposal ~ "3",
    p_type == "Aerobic Unit" & d_type == "Combination of two types" ~ "1",
    
    p_type %in% c("Cesspool-New", "Cesspool-Existing") ~ "4",
    p_type %in% c("Presby (passive, NSF 40)", "Holding Tank (Government Only)", "Incinerating Toilet") ~ "1",
    
    p_type == "Other" & d_type %in% primary_disposal ~ "1",
    TRUE ~ "1"
  )
}

# ==============================================================================
# 1. CLEAN RAW PERMIT DATA & PREPARE MASTER CLEAN
# ==============================================================================

iws_kauai_NOdwell_clean <- clean_kauai_tmk(iws_kauai_NOdwell, "tmk")
kauai_dwell_tmks_forjoin_clean <- clean_kauai_tmk(kauai_dwell_tmks_forjoin, "tmk")
kauai_dwell_allentries_clean <- clean_kauai_tmk(kauai_dwell_allentries, "tmk")

master_clean <- iws_kauai_permits %>%
  clean_kauai_tmk("tmk") %>% 
  clean_names() %>%
  mutate(
    approval_date = mdy(final_approval_date),
    year = year(approval_date),
    address_full = if_else(
      is.na(project_street_address2), 
      project_street_address, 
      paste(project_street_address, "-", project_street_address2)
    ),
    address_lower = str_trim(str_to_lower(address_full)),
    osds_class = assign_osds_class(permit_type, disposal_type),
    dwelling_yn = if_else(tmk %in% iws_kauai_NOdwell_clean$tmk, "no", "yes")
  ) %>%
  select(-any_of("permit_id")) %>% 
  distinct()


# ==============================================================================
# 2. GENERATE TABLES (MATCHING MAUI STRUCTURE)
# ==============================================================================

# --- TABLE 1: TMK Registry ---
table1_tmk_registry <- master_clean %>%
  group_by(tmk) %>%
  summarize(
    n_permits = n(),
    n_years = n_distinct(year),
    n_addresses = n_distinct(address_lower),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    is_dwelling = any(tmk %in% kauai_dwell_tmks_forjoin_clean$tmk),
    scenario = case_when(
      n_addresses == 1 & n_permits == 1 ~ "Simple",
      n_addresses > 1 ~ "Multi-Building",
      n_years > 1 ~ "Potential Upgrade",
      TRUE ~ "Multiple Permits Same Year"
    ),
    .groups = "drop"
  )

# Commercial Subset
table1_tmk_registry_commercial <- table1_tmk_registry %>% 
  filter(is_dwelling == FALSE)


# --- TABLE 2A & 2B: Bedroom Aggregations ---
table2a_bedrooms <- kauai_dwell_allentries_clean %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bed_rooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    total_beds = sum(bed_rooms, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = total_beds, names_sort = TRUE)

table2b_bedrooms <- kauai_dwell_allentries_clean %>%
  filter(tmk %in% master_clean$tmk) %>%
  mutate(bed_rooms = as.numeric(bedrooms)) %>%
  group_by(tmk, tax_year) %>%
  summarize(
    mode_beds = calc_mode(bed_rooms, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = tax_year, values_from = mode_beds, names_sort = TRUE)


# --- TABLE 3: OSDS Class Encoding ---
table3_classes <- master_clean %>%
  group_by(tmk, year) %>%
  summarize(
    all_classes_text = paste(sort(unique(osds_class)), collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(
    encoded_sum = 
      (str_detect(all_classes_text, "1") * 1) +
      (str_detect(all_classes_text, "2") * 2) +
      (str_detect(all_classes_text, "3") * 4) +
      (str_detect(all_classes_text, "4") * 8)
  ) %>%
  select(tmk, year, encoded_sum) %>%
  pivot_wider(names_from = year, values_from = encoded_sum, names_sort = TRUE)


# ==============================================================================
# 3. SAVE STANDARDIZED OUTPUTS
# ==============================================================================

output_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data")
#dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(table1_tmk_registry, file.path(output_dir, "kauai_table1_tmk_registry_all.csv"))
write_csv(table1_tmk_registry_commercial, file.path(output_dir, "kauai_table1_tmk_registry_commercial.csv"))
write_csv(table2a_bedrooms, file.path(output_dir, "kauai_table2a_newbedrooms_sum.csv"))
write_csv(table2b_bedrooms, file.path(output_dir, "kauai_table2b_newbedrooms_mode.csv"))
write_csv(table3_classes, file.path(output_dir, "kauai_table3_newclass_all.csv"))


```


### Step 3: Compare and Integrate Past Datasets

In this step we have to add in the old data, but differentiate between data that has been updated, unchanged. We also have to gapfill the bedroom and class data, because the old data is just a snapshot in time in time and in some cases we don't have the year information for that OSDS entry. 

#### Records Unchanged Since OTP

These have no new permit data. But, it would be nice to get a better understanding of their bedroom # over time. We have to make one assumption in order to merge our dwell data with this older data. 

```{r unchanged}

library(tidyverse)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 0. KAUAI TMK CLEANER FUNCTION
# ==============================================================================
clean_kauai_tmk <- function(df, target_col = "tmk") {
  col_names <- names(df)
  actual_col <- if (target_col %in% col_names) {
    target_col
  } else if ("full_taxid" %in% col_names) {
    "full_taxid"
  } else if ("TMK" %in% col_names) {
    "TMK"
  } else {
    target_col
  }
  
  df %>% 
    mutate(
      raw_val = as.character(.data[[actual_col]]),
      raw_val = if_else(str_detect(raw_val, "e\\+"), sprintf("%.0f", as.numeric(raw_val)), raw_val),
      digits = str_replace_all(raw_val, "[^0-9]", ""),
      digits = if_else(nchar(digits) >= 9, substr(digits, 1, 9), digits),
      digits = case_when(
        nchar(digits) == 8 ~ paste0("4", digits),
        nchar(digits) == 9 & str_starts(digits, "0") ~ paste0("4", substr(digits, 2, 9)),
        TRUE ~ str_pad(digits, width = 9, side = "left", pad = "0")
      )
    ) %>% 
    mutate(tmk = digits) %>% 
    select(-any_of(c("raw_val", "digits"))) %>% 
    filter(!is.na(tmk) & tmk != "" & tmk != "000000000")
}

# ==============================================================================
# 1. READ BASELINE & DWELLING DATA
# ==============================================================================

kauai_unchanged <- kauai_unchanged %>% 
  clean_kauai_tmk("tmk")

kauai_dwell_alltmks <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step1-clean-dwelling-records/kauai_dwell_clean_alltmks.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk")

# Encode baseline 2010 OTP OSDS classes
kauai_unchanged_encoded <- kauai_unchanged %>% 
  select(-any_of(c("objectid", "x", "y"))) %>% 
  distinct() %>% 
  pivot_longer(
    cols = starts_with("class"),
    names_to = "class",
    values_to = "quantity"
  ) %>% 
  filter(!is.na(quantity) & quantity != 0 & quantity != -9999) %>% 
  rename(ttl_osds_2010 = any_of(c("osds_qty", "ttl_osds"))) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv"  ~ 8, # Cesspool
    class == "class_iii" ~ 4,
    class == "class_ii"  ~ 2,
    class == "class_i"   ~ 1,
    TRUE                 ~ NA_real_
  )) %>% 
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010", any_of(c("pittcode", "pitt_code"))), 
    names_from = "class", 
    values_from = "osds_class_otp",
    values_fn = max
  ) %>% 
  mutate(encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

# Join with Dwelling Records
kauai_unchgd_encoded_dwell <- left_join(kauai_unchanged_encoded, kauai_dwell_alltmks, by = "tmk") %>% 
  distinct()

# ==============================================================================
# 2. GENERATE UNCHANGED BASELINE TABLES (1–4)
# ==============================================================================

# --- Table 1: TMK Registry ---
kauai_card_dwell_summary <- kauai_unchgd_encoded_dwell %>% 
  group_by(tmk) %>% 
  summarize(
    n_buildings = if ("card" %in% names(.)) n_distinct(card, na.rm = TRUE) else 1,
    in_dwelldat = any(!is.na(tax_year) & tax_year != ""),
    .groups = "drop"
  )

kauai_otp_unchanged_table1 <- kauai_unchgd_encoded_dwell %>% 
  mutate(
    n_permits = case_when(
      encoded_sum %in% c(1, 2, 4, 8)       ~ 1,
      encoded_sum %in% c(3, 5, 6, 9, 10, 12) ~ 2,
      encoded_sum %in% c(7, 11, 13, 14)     ~ 3,
      encoded_sum == 15                     ~ 4,
      TRUE                                  ~ 99999
    ),
    n_years = 1,
    min_year = 2010,
    max_year = 2010,
    n_addresses = "unknown",
    status_since_otp = "unchanged since otp",
    scenario = case_when(
      n_permits == 1 & n_years == 1 ~ "Simple",
      n_addresses == "unknown"      ~ "Unknown",
      n_years > 1                   ~ "Potential Upgrade",
      TRUE                          ~ "Multiple Permits Same Year"
    )
  ) %>% 
  left_join(kauai_card_dwell_summary, by = "tmk") %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    scenario,
    n_buildings
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

# --- Table 2: Bedrooms ---
kauai_otp_unchanged_table2 <- kauai_unchgd_encoded_dwell %>% 
  filter(!is.na(tax_year) & tax_year != "") %>% 
  mutate(bed_rooms = as.numeric(bedrooms)) %>% 
  group_by(tmk, tax_year) %>% 
  summarize(total_beds = sum(bed_rooms, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(
    id_cols = tmk,
    names_from = tax_year, 
    values_from = total_beds, 
    names_sort = TRUE
  ) %>% 
  select(tmk, matches("^[0-9]{4}")) %>% 
  rename_with(~ paste0("bedrooms_", .x), matches("^[0-9]{4}"))

# --- Table 3: OSDS Class Baseline ---
kauai_otp_unchanged_table3 <- kauai_unchgd_encoded_dwell %>% 
  mutate(year = 2010) %>% 
  select(tmk, year, encoded_sum) %>% 
  distinct() %>% 
  pivot_wider(
    id_cols = tmk,
    names_from = year, 
    values_from = encoded_sum,
    names_prefix = "osds_class_" # Matches Maui & Oahu schema
  )

# --- Table 4: Commercial Classification ---
kauai_otp_unchanged_table4 <- kauai_unchanged %>% 
  select(tmk, any_of(c("commercial", "pittcode", "pitt_code", "x", "y"))) %>% 
  rename(pitt_code = any_of(c("pittcode", "pitt_code"))) %>% 
  filter(!is.na(commercial)) %>% 
  mutate(
    pitt_code_num = as.numeric(pitt_code),
    comm_assignment = case_when(
      pitt_code_num == 0                            ~ "Timeshare (Maui/Kauai)",
      pitt_code_num == 999                          ~ "Multiple PITT Codes",
      pitt_code_num >= 1000                         ~ "Commercialized Residential",
      pitt_code_num >= 100 & pitt_code_num < 200   ~ "Improved Residential",
      pitt_code_num >= 200 & pitt_code_num < 300   ~ "Apartment",
      pitt_code_num >= 300 & pitt_code_num < 400   ~ "Commercial",
      pitt_code_num >= 400 & pitt_code_num < 500   ~ "Industrial",
      pitt_code_num >= 500 & pitt_code_num < 600   ~ "Agricultural and Rural",
      pitt_code_num >= 600 & pitt_code_num < 700   ~ "Conservation",
      pitt_code_num >= 700 & pitt_code_num < 800   ~ "Hotel and Resort",
      pitt_code_num >= 800 & pitt_code_num < 900   ~ "Unimproved Residential",
      pitt_code_num >= 900 & pitt_code_num < 999   ~ "Homeowner",
      TRUE                                          ~ "Unknown"
    )
  ) %>% 
  select(-pitt_code_num) %>% 
  distinct(tmk, .keep_all = TRUE)

# Isolate non-residential TMKs missing commercial metadata for downstream Google Places lookup
kauai_otp_unchanged_table4_all <- kauai_otp_unchanged_table1 %>% 
  filter(in_dwelldat == FALSE)

difference_withno_info <- setdiff(kauai_otp_unchanged_table4_all$tmk, kauai_otp_unchanged_table4$tmk)

kauai_tmks_forgoogleplaces <- kauai_otp_unchanged_table1 %>% # these we really know nothing about
  filter(tmk %in% difference_withno_info) %>% 
  select(tmk)

# ==============================================================================
# 3. EXPORT INTERMEDIATE UNCHANGED FILES
# ==============================================================================

out_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  kauai_otp_unchanged_table1,
  file.path(out_dir, "kauai_table1_tmk_registry_UNCHANGED.csv")
)

write_csv(
  kauai_otp_unchanged_table2,
  file.path(out_dir, "kauai_table2_bedrooms_UNCHANGED.csv")
)

write_csv(
  kauai_otp_unchanged_table3,
  file.path(out_dir, "kauai_table3_classes_UNCHANGED.csv")
)

write_csv(
  kauai_otp_unchanged_table4,
  file.path(out_dir, "kauai_table4_commercial_UNCHANGED_draft.csv")
)

write_csv(
  kauai_tmks_forgoogleplaces,
  file.path(out_dir,"/kauai_table4_googleplaces_UNCHANGED_draft.csv")
)

```


#### Updated

These tmks are ones are present in OTP with OSDS recoreded (ie had an osds permit befor 2010) but also in the new data, meaning they have applied for a new permit since 2010 as well.  

```{r updated}

library(tidyverse)

# Note: clean_kauai_tmk() is assumed to be loaded in the R session from Part 1

# ==============================================================================
# 1. BEDROOM GAP-FILL LOGIC (KAUAI)
# ==============================================================================

# Read Step 2 permit bedroom data
table_2_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_table2a_newbedrooms_sum.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  rename_with(~ paste0("newbeds", .x), !tmk)

kauai_updated_forjoin <- kauai_updated %>% 
  clean_kauai_tmk("tmk") %>% 
  select(tmk, any_of(c("pittcode", "pitt_code")), commercial, ttl_bdrms) %>% 
  rename(
    otp_ttl_bdrms = ttl_bdrms,
    pitt_code = any_of(c("pittcode", "pitt_code"))
  ) %>% 
  mutate(
    pitt_code_num = as.numeric(pitt_code),
    comm_assignment = case_when(
      pitt_code_num == 0                            ~ "Timeshare",
      pitt_code_num == 999                          ~ "Multiple PITT Codes",
      pitt_code_num >= 1000                         ~ "Commercialized Residential",
      pitt_code_num >= 100 & pitt_code_num < 200   ~ "Improved Residential",
      pitt_code_num >= 200 & pitt_code_num < 300   ~ "Apartment",
      pitt_code_num >= 300 & pitt_code_num < 400   ~ "Commercial",
      pitt_code_num >= 400 & pitt_code_num < 500   ~ "Industrial",
      pitt_code_num >= 500 & pitt_code_num < 600   ~ "Agricultural and Rural",
      pitt_code_num >= 600 & pitt_code_num < 700   ~ "Conservation",
      pitt_code_num >= 700 & pitt_code_num < 800   ~ "Hotel and Resort",
      pitt_code_num >= 800 & pitt_code_num < 900   ~ "Unimproved Residential",
      pitt_code_num >= 900 & pitt_code_num < 999   ~ "Homeowner",
      TRUE                                          ~ "Unknown"
    )
  ) %>% 
  select(-pitt_code_num)

# Filter permit bedroom table to updated TMKs
table_2_updated <- table_2_new %>% 
  filter(tmk %in% kauai_updated_forjoin$tmk)

updates_notInTable2 <- setdiff(kauai_updated_forjoin$tmk, table_2_updated$tmk) # 33

table4_withUpdates <- kauai_updated_forjoin %>% 
  filter(tmk %in% updates_notInTable2)

# Join baseline bedroom counts with new permit bedroom data
kauai_updated_withNewbeds <- kauai_updated_forjoin %>% 
  left_join(table_2_updated, by = "tmk") %>% 
  filter(otp_ttl_bdrms != 0)

year_cols <- kauai_updated_withNewbeds %>% select(starts_with("newbeds")) %>% names()
latest_bed_col <- if (length(year_cols) > 0) tail(year_cols, 1) else NULL

kauai_updated_withBedFlags <- kauai_updated_withNewbeds %>% 
  mutate(
    # Find matching index across permit bedroom years
    match_index = {
      match_matrix <- pick(all_of(year_cols)) == otp_ttl_bdrms
      match_matrix[is.na(match_matrix)] <- FALSE 
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    match_name = year_cols[match_index],
    
    # Dynamic latest year trend calculation
    latest_val = if (!is.null(latest_bed_col)) as.numeric(.data[[latest_bed_col]]) else NA_real_,
    bed_trend = case_when(
      is.na(latest_val)              ~ NA_character_,
      otp_ttl_bdrms < latest_val     ~ "Expansion since OTP",
      otp_ttl_bdrms > latest_val     ~ "Shrink - double check",
      otp_ttl_bdrms == latest_val    ~ "Perfect Match",
      TRUE                           ~ NA_character_
    ),
    
    data_presence = if_else(
      if_all(starts_with("newbeds"), ~ .x == "0" | is.na(.x) | .x == 0),
      "Empty (All 0 or NA)",
      "Bedrooms Recorded"
    )
  )

# Join Dwelling Build Years for Conservative Gap-Filling
year_built <- kauai_dwell_alltmks %>% 
  filter(tmk %in% kauai_updated_withBedFlags$tmk) %>% 
  select(tmk, year_built = any_of(c("yrblt", "year_built"))) %>% 
  filter(!is.na(year_built) & year_built != "") %>% 
  distinct() %>% 
  group_by(tmk) %>% 
  summarize(
    year_built_combined = str_flatten(as.character(year_built), collapse = ", "),
    count_buildyears = n(),
    .groups = "drop"
  ) %>% 
  clean_kauai_tmk("tmk")

kauai_updated_with_BEDflags_gapfilled <- kauai_updated_withBedFlags %>% 
  left_join(year_built, by = "tmk") %>% 
  mutate(
    gapfill_flag = if_else(
      data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0, 
      "Gapfilled with OTP value", 
      "Used dwell data"
    )
  ) %>% 
  mutate(across(starts_with("newbeds"), ~ {
    fill_condition <- data_presence == "Empty (All 0 or NA)" & otp_ttl_bdrms > 0
    if_else(fill_condition, as.numeric(otp_ttl_bdrms), as.numeric(.x))
  })) %>% 
  mutate(across(starts_with("newbeds"), ~ {
    col_year <- as.numeric(str_extract(cur_column(), "\\d{4}"))
    min_build_yr <- as.numeric(str_extract(year_built_combined, "^\\d{4}"))
    
    case_when(
      gapfill_flag == "Used dwell data"                         ~ as.numeric(.x),
      is.na(min_build_yr) | min_build_yr <= 2010               ~ as.numeric(.x),
      min_build_yr > col_year                                   ~ NA_real_,
      TRUE                                                      ~ as.numeric(.x)
    )
  }))

table_2_updatedGapfilled <- kauai_updated_with_BEDflags_gapfilled %>% 
  select(tmk, starts_with("newbeds"))

bedFlags_forTable1 <- kauai_updated_with_BEDflags_gapfilled %>% 
  select(
    tmk,
    bed_trend,
    data_presence,
    year_built_combined,
    count_buildyears,
    gapfill_flag
  )

write_csv(
  bedFlags_forTable1,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_bedflags_fortable1.csv")
)


# ==============================================================================
# 2. OSDS CLASS GAP-FILL & UPGRADE LOGIC (KAUAI)
# ==============================================================================

# Read Step 2 permit OSDS class table
table_3_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_table3_newclass_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  rename_with(~ paste0("newclass", .x), !tmk)

kauai_updated_withClass_enc <- kauai_updated %>% 
  clean_kauai_tmk("tmk") %>% 
  select(tmk, osds_qty, starts_with("class")) %>% 
  select(-any_of(c("class"))) %>% 
  distinct() %>% 
  pivot_longer(cols = starts_with("class"), names_to = "class", values_to = "quantity") %>% 
  filter(!is.na(quantity) & quantity != 0 & quantity != -9999) %>% 
  rename(ttl_osds_2010 = osds_qty) %>% 
  mutate(osds_class_otp = case_when(
    class == "class_iv"  ~ 8, # Cesspool 
    class == "class_iii" ~ 4,
    class == "class_ii"  ~ 2,
    class == "class_i"   ~ 1,
    TRUE                 ~ NA_real_
  )) %>% 
  pivot_wider(
    id_cols = c("tmk", "ttl_osds_2010"), 
    names_from = "class", 
    values_from = "osds_class_otp"
  ) %>% 
  mutate(otp_encoded_sum = rowSums(across(starts_with("class")), na.rm = TRUE))

kauai_updated_withClassEnc_New <- kauai_updated_withClass_enc %>% 
  left_join(table_3_new, by = "tmk")

class_cols <- kauai_updated_withClassEnc_New %>% select(starts_with("newclass")) %>% names()

table3_updated_classmatch <- kauai_updated_withClassEnc_New %>% 
  mutate(
    match_index = {
      match_matrix <- pick(all_of(class_cols)) == otp_encoded_sum
      match_matrix[is.na(match_matrix)] <- FALSE 
      if_else(rowSums(match_matrix) > 0, max.col(match_matrix, ties.method = "first"), NA_integer_)
    },
    match_name = class_cols[match_index],
    data_presence = if_else(
      if_all(starts_with("newclass"), ~ .x == "0" | is.na(.x) | .x == 0),
      "Empty (All 0 or NA)",
      "OSDS Permit Recorded"
    )
  )

# Join TMK Registry to cross-examine 2009/2010 baseline matches
table_1_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk")

update_logic_df <- table3_updated_classmatch %>% 
  left_join(table_1_new, by = "tmk") %>% 
  select(
    tmk, otp_encoded_sum, 
    any_of(c("min_year", "max_year", "n_addresses")), 
    starts_with("newclass")
  ) %>% 
  mutate(
    legacy_val = as.numeric(otp_encoded_sum),
    v2009 = if ("newclass2009" %in% names(.)) as.numeric(newclass2009) else NA_real_,
    v2010 = if ("newclass2010" %in% names(.)) as.numeric(newclass2010) else NA_real_,
    scenario_match_flag = case_when(
      legacy_val == v2009 | legacy_val == v2010 ~ "matched 2009 or 2010",
      is.na(v2009) & is.na(v2010)                 ~ "no data to compare",
      TRUE                                       ~ "unmatched"
    )
  )

update_logic_refined_cl <- update_logic_df %>% 
  mutate(newpre_2009 = as.character(case_when(
    scenario_match_flag == "matched 2009 or 2010"               ~ NA_real_,
    scenario_match_flag %in% c("unmatched", "no data to compare") ~ legacy_val,
    TRUE                                                        ~ NA_real_
  ))) %>% 
  select(
    tmk, newpre_2009, 
    starts_with("newclass200"), starts_with("newclass201"), starts_with("newclass202"), 
    everything()
  ) %>% 
  rowwise() %>% 
  mutate(unique_story = {
    # Strictly select pre-2009 legacy and permit class columns
    vals <- c_across(c(newpre_2009, starts_with("newclass")))
    vals <- vals[!is.na(vals) & vals != "" & vals != "NA"]
    paste(unique(vals), collapse = "|")
  }) %>% 
  ungroup() %>% 
  mutate(
    upgrade_scenario = case_when(
      # High-priority overrides
      unique_story == "9|1|8"  ~ "cesspool and septic - both repaired, no change",
      unique_story == "10|2|9" ~ "cesspool and septic 2 - septic repaired twice, cesspool once",
      unique_story == "7|9|1"  ~ "potential data error: cesspool added, then possibly updated",
      unique_story == "4|2"    ~ "septic and aerobic - replacement of aerobic",
      unique_story == "3|4"    ~ "septic combo to aerobic",
      unique_story == "1|4"    ~ "single septic to aerobic",
      unique_story == "7|4"    ~ "complex septic system to aerobic",
      unique_story == "8|5"    ~ "cesspool to septic / aerobic hybrid",
      unique_story == "5|1"    ~ "septic and aerobic to septic",
      unique_story == "7|3"    ~ "2 septics and aerobic to just 2 septics - aerobic removal",
      unique_story == "7|1|4"  ~ "2 septics and aerobic + one septic repair + one aerobic repair",
      unique_story == "10|9"   ~ "1 septic + cesspool to better septic and cesspool",
      
      # Generalized Regex Scenarios
      unique_story == "" | !str_detect(unique_story, "\\|") ~ "no upgrade detected",
      str_detect(unique_story, "^8\\|.*[1234]$")           ~ "single cesspool upgraded to septic or aerobic",
      str_detect(unique_story, "^3\\|.*[12]$")             ~ "multiple septic with new permit: possible repair",
      str_detect(unique_story, "^[24]\\|.*1$")             ~ "septic or aerobic system refined to cleanest type (Class 1)",
      str_detect(unique_story, "^(9|1[0-5])\\|.*[1234]$")  ~ "multiple osds with cesspool upgraded",
      str_detect(unique_story, "^(9|1[0-5])\\|.*8$")       ~ "multiple osds with new cesspool application",
      
      # Errors and Laterals
      str_detect(unique_story, "^[1-7]\\|.*(8|9|1[0-5])$") ~ "potential data error: backslide to cesspool",
      str_detect(unique_story, "^[123]\\|.*[123]$")        ~ "lateral change (clean to clean)",
      
      TRUE ~ "other scenario"
    )
  )

# Split into updated class matrix and Table 1 flags
table_3_updated_new <- update_logic_refined_cl %>% 
  select(tmk, newpre_2009, starts_with("newclass"))

class_flag_forTable1 <- update_logic_refined_cl %>% 
  select(
    tmk, 
    class_match_flag = scenario_match_flag, 
    unique_story, 
    upgrade_scenario
  )

write_csv(
  class_flag_forTable1,
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_classflags_fortable1.csv")
)

```


### Step 4: Assemble Final Tables by Island

To run this, need to run STEP 0. Otherwise all dependencies should be loaded here.

#### Complete Tables 1 - 3a, 4

```{r complete-tables}

library(tidyverse)

'%!in%' <- Negate('%in%')

# ==============================================================================
# 0. KAUAI TMK CLEANER FUNCTION
# ==============================================================================
clean_kauai_tmk <- function(df, target_col = "tmk") {
  col_names <- names(df)
  actual_col <- if (target_col %in% col_names) {
    target_col
  } else if ("full_taxid" %in% col_names) {
    "full_taxid"
  } else if ("TMK" %in% col_names) {
    "TMK"
  } else {
    target_col
  }
  
  df %>% 
    mutate(
      raw_val = as.character(.data[[actual_col]]),
      raw_val = if_else(str_detect(raw_val, "e\\+"), sprintf("%.0f", as.numeric(raw_val)), raw_val),
      digits = str_replace_all(raw_val, "[^0-9]", ""),
      digits = if_else(nchar(digits) >= 9, substr(digits, 1, 9), digits),
      digits = case_when(
        nchar(digits) == 8 ~ paste0("4", digits),
        nchar(digits) == 9 & str_starts(digits, "0") ~ paste0("4", substr(digits, 2, 9)),
        TRUE ~ str_pad(digits, width = 9, side = "left", pad = "0")
      )
    ) %>% 
    mutate(tmk = digits) %>% 
    select(-any_of(c("raw_val", "digits"))) %>% 
    filter(!is.na(tmk) & tmk != "" & tmk != "000000000")
}

# ==============================================================================
# 1. TABLE 1: TMK MASTER REGISTRY
# ==============================================================================

table_1_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_table1_tmk_registry_all.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  rename(in_dwelldat = any_of(c("is_dwelling", "in_dwelldat"))) %>% 
  mutate(
    status_since_otp = "New Permit",
    n_addresses = as.character(n_addresses)
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

table_1_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_table1_tmk_registry_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  mutate(scenario = "No new permits") %>% 
  distinct(tmk, .keep_all = TRUE)

bedFlags_forTable1 <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_bedflags_fortable1.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk")

classFlags_forTable1 <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_classflags_fortable1.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk")

# Combine registries and attach gap-fill flags
table_1_complete <- bind_rows(table_1_new, table_1_unchanged) %>% 
  distinct(tmk, .keep_all = TRUE)

table_1_complete_flags <- table_1_complete %>% 
  left_join(bedFlags_forTable1, by = "tmk") %>% 
  left_join(classFlags_forTable1, by = "tmk") %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 2. TABLE 2: BEDROOM TIME-SERIES
# ==============================================================================

table_2_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_table2_bedrooms_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  mutate(across(-tmk, as.numeric))

table_2_updated_forbind <- table_2_updatedGapfilled %>% 
  clean_kauai_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds")) %>% 
  mutate(across(-tmk, as.numeric))

table_2_newonly <- table_2_new %>% 
  clean_kauai_tmk("tmk") %>% 
  filter(tmk %!in% table_2_updatedGapfilled$tmk) %>% 
  rename_with(~ str_replace(.x, "^newbeds", "bedrooms_"), starts_with("newbeds")) %>% 
  mutate(across(-tmk, as.numeric))

table_2_complete <- bind_rows(
  table_2_unchanged, 
  table_2_updated_forbind, 
  table_2_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 3. TABLE 3A: OSDS PERMIT & CLASS HISTORY
# ==============================================================================

table_3_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_table3_classes_UNCHANGED.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  rename_with(~ str_replace(.x, "^osds_cl_", "osds_class_"), starts_with("osds_cl_")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

table_3_updatedforbind <- table_3_updated_new %>% 
  clean_kauai_tmk("tmk") %>% 
  rename(osds_class_pre2009 = newpre_2009) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

table_3_newonly <- table_3_new %>% 
  clean_kauai_tmk("tmk") %>% 
  filter(tmk %!in% table_3_updatedforbind$tmk) %>% 
  rename_with(~ str_replace(.x, "^newclass", "osds_class_"), starts_with("newclass")) %>% 
  mutate(across(starts_with("bedrooms_"), as.numeric))

table_3_complete <- bind_rows(
  table_3_updatedforbind, 
  table_3_unchanged, 
  table_3_newonly
) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 4. TABLE 4: COMMERCIAL & NON-RESIDENTIAL PARCELS
# ==============================================================================


table_4_unchanged <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_table4_commercial_UNCHANGED_draft.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "pittcode", "x", "y"))) %>% 
  rename(pitt_code = any_of(c("pittcode", "pitt_code"))) %>% 
  mutate(across(everything(), as.character))

table_4_googleplaces <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step3/kauai_table4_googleplaces_UNCHANGED_draft.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  mutate(
    commercial = "unknown",
    pitt_code = "-9999"
  ) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_withUpdates_forbind <- table4_withUpdates %>% 
  clean_kauai_tmk("tmk") %>% 
  filter(!is.na(commercial) | !is.na(pitt_code)) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_new <- read_csv(
  file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step2-clean-permit-data/kauai_table1_tmk_registry_commercial.csv"),
  col_types = cols(.default = "c")
) %>% 
  clean_kauai_tmk("tmk") %>% 
  filter(tmk %!in% table_4_withUpdates_forbind$tmk) %>% 
  mutate(
    commercial = "unknown",
    pitt_code = "-9999"
  ) %>% 
  select(tmk, any_of(c("commercial", "pitt_code", "x", "y"))) %>% 
  mutate(across(everything(), as.character))

table_4_complete <- bind_rows(
  table_4_unchanged,
  table_4_googleplaces,
  table_4_new,
  table_4_withUpdates_forbind
) %>% 
  distinct(tmk, .keep_all = TRUE)

table_4_complete_cl <- table_4_complete %>% 
  mutate(
    pitt_code = if_else(is.na(pitt_code), "-9999", pitt_code),
    commercial = if_else(is.na(commercial) | commercial == "unknown", "check google places", commercial)
  )

# ==============================================================================
# 5. FINAL RECONCILIATION & TABLE 1 OVERRIDES
# ==============================================================================

tmks_with_bedrooms <- table_2_complete %>% 
  filter(if_any(starts_with("bedrooms_"), ~ !is.na(.x) & .x > 0)) %>% 
  pull(tmk)

table_4_complete_cl <- table_4_complete_cl %>% 
  filter(tmk %!in% tmks_with_bedrooms)

# Identify commercial parcels versus those with residential bedroom data. should match the number in table_4_complete_cl
shouldbein_table_4 <- setdiff(table_1_complete_flags$tmk, table_2_complete$tmk) # 309, nice


# Build Final Table 1 Master
table_1_final_final <- table_1_complete_flags %>% 
  mutate(use_table_4 = tmk %in% table_4_complete_cl$tmk) %>% 
  select(
    tmk,
    starts_with("n_"),
    min_year,
    max_year,
    in_dwelldat,
    status_since_otp,
    n_buildings,
    bed_trend,
    year_built_combined,
    bed_gapfill_flag = gapfill_flag,
    class_match_flag,
    permit_story = unique_story, 
    upgrade_scenario,
    use_table_4
  ) %>% 
  distinct(tmk, .keep_all = TRUE)

# ==============================================================================
# 6. AUTOMATED INTEGRITY CHECKS
# ==============================================================================

cat("--- KAUAI PIPELINE INTEGRITY CHECKS ---\n")
cat("1. All Table 1 TMKs present in Table 3a:", all(table_1_final_final$tmk %in% table_3_complete$tmk), "\n")
cat("2. All Table 2 TMKs present in Table 1:", all(table_2_complete$tmk %in% table_1_final_final$tmk), "\n")
if (exists("hi_master_tmk_list")) {
  cat("3. All Master TMKs present in Table 1:", all(kauai_master_tmk_list %in% table_1_final_final$tmk), "\n")
}

# extra check

if (exists("kauai_master_tmk_list")) {
  # Clean master list TMKs to ensure matching schema
  clean_master_tmks <- tibble(tmk = as.character(kauai_master_tmk_list)) %>% 
    clean_kauai_tmk("tmk") %>% 
    filter(tmk != "000000000") %>% 
    pull(tmk) %>% 
    unique()
  
  missing_master <- setdiff(clean_master_tmks, table_1_final_final$tmk)
  
  cat("3. All Valid Master TMKs present in Table 1:", length(missing_master) == 0, "\n")
  if (length(missing_master) > 0) {
    cat("   -> Missing TMK count:", length(missing_master), "\n")
    cat("   -> Missing TMKs:", paste(head(missing_master, 10), collapse = ", "), "\n")
  }
}

# if the above fails: 

missing_from_t1 <- setdiff(kauai_master_tmk_list, table_1_final_final$tmk)
cat("Number of missing TMKs:", length(missing_from_t1), "\n")
print(missing_from_t1)

# ==============================================================================
# 7. EXPORT STATEWIDE-COMPATIBLE CSVs
# ==============================================================================

out_dir <- file.path(driver_dir, "4_OSDS_N_Runoff/intermediate/step4")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  table_1_final_final,
  file.path(out_dir, "kauai_table1_tmk_registry_ALL.csv")
)

write_csv(
  table_2_complete,
  file.path(out_dir, "kauai_table2_bedrooms_ALL.csv")
)

write_csv(
  table_3_complete,
  file.path(out_dir, "kauai_table3a_permitsApplied_ALL.csv")
)

write_csv(
  table_4_complete_cl,
  file.path(out_dir, "kauai_table4_commercial_draft.csv")
)

```

#### Create table 3b

The additive encoding is what is known as a binary bit mask. Each class represents a power of 2, so every unique sum corresponds to exactly one combination of classes. 

We can think of table 3a as an "event" table, since it records the permit application event. However, we want to create a "status" table, which records the actual OSDS present on the property in each year, BASED on the events we know happened. This will be table 3b.

We can make multiple versions of this, using different math. In the first, I'll assume that:
1. If there was only one OSDS on the property, the new application is replacing it
2. If there were multiple OSDS on the property, the new application is replacing the grossest, ie highest class. This correspond to the largest power of 2 within the encoded value

This translates to the following math: 
1. Identify the largest power of 2 inside the current value.
2. Subtract that power of 2.
3. Add the new application value.

```{r}

# ------------------- Helper functions ------------------- #

# 1. Create a function for the logic
calculate_active_permits <- function(apps) {
  active_status <- numeric(length(apps))
  
  # Helper to find highest power of 2
  get_highest <- function(x) if (x == 0) 0 else 2^floor(log2(x))
  
  running_total <- apps[1]
  active_status[1] <- running_total
  
  for (i in 2:length(apps)) {
    # only trigger replacement if there is an application (>0)
    # If a new application value appears (it's an "event" year)
    #if (apps[i] != apps[i-1] & apps[i] != 0) 
    if(apps[i] >0){
      highest <- get_highest(running_total)
      # Remove highest bit, then Bitwise OR the new application
      running_total <- bitwOr(bitwAnd(running_total, bitwNot(highest)), apps[i])
    }
    active_status[i] <- running_total
  }
  return(active_status)
}

table_3a_long <- table_3_complete %>% # note, this is a stand in for "pre 2009"
  rename(osds_class_2008 = osds_class_pre2009) %>% 
  pivot_longer(cols = -tmk, names_to = "Year", values_to = "Applied") %>% 
  mutate(Applied = replace_na(as.numeric(Applied),0))
  
table_3b_long <- table_3a_long %>% 
  arrange(tmk,Year) %>% 
  group_by(tmk) %>% 
  mutate(active_status = calculate_active_permits(Applied)) %>% 
  ungroup()
  
table_3b_wide <- table_3b_long %>% 
  pivot_wider(id_cols = tmk, names_from = Year,values_from = active_status)

write_csv(table_3b_wide,
          file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/kauai_table3b_activeOSDS.csv"))



```
