---
title: "Spatialize TMKs"
author: "Madeline Berger"
date: "2026-04-30"
output: html_document
---

# Spatialize all tmks


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))

```


This script will read in our tmk registry from every island, combine them all, and then add spatial information including:
- TMK polygon
- TMK centroid
- Flow line - closes to TMK centroid 
- Flow lines(s) - all that intersect with tmk polygon


We'll also pull the old spatial information, for the points that had them, and include that to compare to cetroids. 


## Data Prep
### Read in and combine Table 1 from all Islands

```{r combine-table1}

in_dir <- file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4")

kauai <- read_csv(file.path(in_dir,"kauai_table1_tmk_registry_ALL.csv")) %>% 
  select(-class_match_flag) %>% 
  mutate(county = "Kauai")

maui_nui <- read_csv(file.path(in_dir,"maui_table1_tmk_registry_ALL.csv")) %>% mutate(county = "Maui Nui")
oahu <- read_csv(file.path(in_dir,"oahu_table1_tmk_registry_ALL.csv")) %>% 
  mutate(county = "Honolulu")

hawaii <- read_csv(file.path(in_dir,"hi_table1_tmk_registry_ALL.csv")) %>%
  select(-class_match_flag) %>% 
  mutate(county = "Hawaii")


# complete tmk registry 

all_osds_tmks <- rbind(kauai,
                       maui_nui,
                       oahu,
                       hawaii) %>% 
  mutate(tmk = as.character(tmk))


# save draft
write_csv(all_osds_tmks,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_ALL.csv"))

```


### Read in NEW spatial information 

```{r, message = FALSE}
# tmk - this will be the first join
tmk_parcels <- st_read(file.path(data_dir,"osds/raw-data-downloaded/tmk_state_parcels_shp/tmk_state.shp"))

crs(tmk_parcels)


# read in watersheds - spatial join 

watersheds_shp <- st_read(file.path(data_dir,"Coastlines_Watersheds/WBD_HU12_n83utm4_2023fixed.shp")) %>% 
  select(HUC12,Name)

crs(watersheds_shp)


# flow lines 
# flowlines <- st_read(file.path(driver_dir,"0_Input_Data_and_Models/Common_2024/Finalized/Coastlines_Watersheds/NHD_Flowlines/NHD_Flowlines.shp"))

```



### Read in OLD files for each island, isolate spatial info if available
These would all come from old otp data available. 

```{r}

library(janitor)

# oahu - already spatial
otp_oahu_raw <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/osds_oah.shp")) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk_9))

# all others, not spatial already 
molo_raw <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Molokai.csv"))

maui_raw <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Maui.csv"))

kauai_raw <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Kauai.csv"))

hawaii_raw <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/On-site_Sewage_Disposal_Systems_-_Hawaii_Island.csv"))

# spatialize these

library(purrr)

# 1. dataframes into a named list
df_list <- list(
  molo_raw, 
  maui_raw,
  kauai_raw,
  hawaii_raw
)

# 2. use map to convert all of them at once

sf_list <- df_list %>%
  map(~{
    .x %>%
      # Standardize names and ensure they are numeric
      mutate(X = as.numeric(X), Y = as.numeric(Y)) %>%
      select(X, Y, tmk) %>% 
      # Using CRS 32604 for UTM Zone 4N (Hawaii meters)
      st_as_sf(coords = c("X", "Y"), crs = 32604, remove = FALSE)
  })

mauinui_hi_kau_spatial <- do.call(rbind,sf_list)

otp_oahu_spatial <- st_transform(otp_oahu_raw, crs = 32604)

crs(mauinui_hi_kau_spatial) == crs(otp_oahu_spatial)


```


## Convert TMK Parcels to Points - point on surface

```{r}

sf_use_s2(FALSE)

### Step 1: Get TMKS centroid and join to table 1

tmk_osds <- tmk_parcels %>% 
  filter(TMK_txt %in% all_osds_tmks$tmk) %>% 
  mutate(is_valid = st_is_valid(geometry)) %>% 
  st_cast("MULTILINESTRING") %>%   # fix invalid non closures
  st_cast("LINESTRING") %>%        
  st_polygonize()

tmk_centroids <- st_centroid(tmk_osds)
tmk_pos <- st_point_on_surface(tmk_osds)

# Look at difference for which one to choose

kauai_centroids <- tmk_centroids %>% filter(county == "Kauai")
kauai_pos <- tmk_pos %>% filter(county == "Kauai")

mapview(kauai_centroids, col.regions = "blue")+
  mapview(kauai_pos, col.regions = "lightblue")


tmk_unique_points <- tmk_pos %>%
  # Select the ID and ensure we only have one row per ID
  distinct(TMK_txt, .keep_all = TRUE) %>%
  select(tmk = TMK_txt)

tmk_unique_centroids <- tmk_centroids %>%
  # Select the ID and ensure we only have one row per ID
  distinct(TMK_txt, .keep_all = TRUE) %>%
  select(tmk = TMK_txt)


# I guess I will start with pos, and can always switch to 
tmk_pos_join <- all_osds_tmks %>%
  left_join(tmk_unique_points, by = "tmk")


missing_tmks <- all_osds_tmks %>%
  anti_join(tmk_unique_points, by = "tmk")

nrow(missing_tmks)
head(missing_tmks)

osds_with_tmkpos_cl <- tmk_pos_join %>% 
  filter(tmk %!in% missing_tmks$tmk) %>% 
  st_as_sf() %>% 
  st_transform(crs = crs(watersheds_shp)) # 895 dont have a tmk point

# Save tmk_registry without spatial info

```


### Fill in missing spatial info using OTP coordinates

```{r}

# first read in full table 1
missing_spatial_info <- all_osds_tmks %>% 
  filter(tmk %in% missing_tmks$tmk)

# --------- filter out the tmks with missing spatial info from 

# --------- oahu
otp_oahu_spatial <- filter(otp_oahu_raw, tmk %in% missing_spatial_info$tmk) %>% 
  select(tmk,x,y) # N = 5

otp_oahu_spatial <- st_transform(otp_oahu_spatial, crs = 32604)


# -------- other islands
crs(mauinui_hi_kau_spatial) == crs(otp_oahu_spatial)


rest_islands_spatial <- filter(mauinui_hi_kau_spatial, tmk %in% missing_spatial_info$tmk) %>% 
  clean_names() # N = 7

all_for_gapfill <- rbind(otp_oahu_spatial,
                         rest_islands_spatial) %>% 
  mutate(tmk = as.character(tmk)) %>% 
  distinct(tmk, .keep_all = TRUE)


missing_info_gapfilled <- missing_spatial_info %>% 
  left_join(., all_for_gapfill, by = "tmk") %>% 
  st_as_sf() %>% 
  st_transform(crs = crs(osds_with_tmkpos_cl)) %>% 
  select(-x,-y) %>% 
  filter(!st_is_empty(geometry))


# add back to the rest

osds_spatial_gapfilled <- rbind(osds_with_tmkpos_cl,
                               missing_info_gapfilled) %>% 
  filter(!st_is_empty(geometry))


still_missing <- setdiff(all_osds_tmks$tmk,osds_spatial_gapfilled$tmk)

still_missing_df <- all_osds_tmks %>% filter(tmk %in% still_missing)

```


## Add Watersheds 

``` {r}

### Step 2: Spatial join to watersheds

osds_with_watersheds <- st_join(osds_spatial_gapfilled, watersheds_shp, join = st_intersects)


outside_count <- sum(is.na(osds_with_watersheds$Name)) # Name is watershed name
print(paste("Points with no watershed match:", outside_count))

insiders <- osds_with_watersheds %>% filter(!is.na(Name))
outsiders <- osds_with_watersheds %>% filter(is.na(Name))

mapview(outsiders)

# outsiders are mainly coastal. Add to nearest watershed. 
nearest_indices <- st_nearest_feature(outsiders, watersheds_shp)

dist <- st_distance(outsiders, watersheds_shp[nearest_indices, ], by_element = TRUE)


outsiders_fixed <- outsiders %>%
  select(-colnames(st_drop_geometry(watersheds_shp))) %>% 
  bind_cols(st_drop_geometry(watersheds_shp[nearest_indices, ]))


mapview(outsiders_fixed)

osds_final_watersheds <- bind_rows(insiders, outsiders_fixed) # 

outside_count2 <- sum(is.na(osds_final_watersheds$Name))

outsiders2 <- osds_final_watersheds %>% filter(is.na(Name))
mapview(outsiders2) #0 left

still_missing_watershed <- osds_final_watersheds %>% 
  filter(is.na(Name))

st_write(osds_final_watersheds,
         dsn = file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"),
         delete_dsn = T)

st_write(still_missing_df,
         dsn = file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withOUTSpatial.csv"),
         delete_dsn = T)

```

## Add Distance from Coastline

```{r}




```


### Potential addition: Flow Lines. We can think about how to categorize these later

```{r flowlines}



## read these back in, join but with spatial missing flag

osds_with_spatial <-  st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"))

osds_without_spatial <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withOUTSpatial.csv"))

mapview(osds_with_spatial, zcol = "HUC12")

osds_with_spatial_simple <- osds_with_spatial %>% 
  mutate(spatial_info = ifelse(st_is_empty(geometry),FALSE, TRUE)) %>% 
  select(tmk, status_since_otp = 7, spatial_info)

osds_without_spatial_simple <- osds_without_spatial %>% 
  mutate(spatial_info = FALSE) %>% 
  select(tmk, status_since_otp, spatial_info) %>%
  mutate(geometry = st_sfc(lapply(1:n(), function(x) st_geometrycollection()), 
                           crs = st_crs(osds_with_spatial))) %>% 
  st_as_sf() %>% 
  filter(tmk %!in% osds_with_spatial_simple$tmk)


table_1_all_raw <- rbind(osds_with_spatial_simple, osds_without_spatial_simple)

table_1_forChris <- table_1_all_raw %>% 
  arrange(tmk, desc(spatial_info)) %>% 
  distinct(tmk, .keep_all = TRUE)

any(duplicated(table_1_forChris$tmk))

# duplicated_tmks <- table_1_all_forChris %>%
#   group_by(tmk) %>%
#   filter(n() > 1) %>%
#   arrange(tmk)

st_write(table_1_forChris,
         dsn = file.path(here("data/osds/shared-data/to_ChrisShuler/table1_simple_07_15_2026.shp")))
```
