---
title: "Spatialize TMKs"
author: "Madeline Berger"
date: "2026-04-30"
output: html_document
---

# Spatialize all tmks


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

source(here("workflow/setup.R"))

```


This script will read in our tmk registry from every island, combine them all, and then add spatial information including:
- TMK polygon
- TMK centroid
- Flow line - closes to TMK centroid 
- Flow lines(s) - all that intersect with tmk polygon


We'll also pull the old spatial information, for the points that had them, and include that to compare to cetroids. 


## Data Prep
### Read in and combine Table 1 from all Islands

```{r}

kauai <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table1_tmk_registry_ALL.csv")) %>% 
  select(-use_table4) %>% 
  mutate(county = "Kauai")
maui_nui <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_ALL.csv")) %>% 
  select(-use_table4) %>% 
  mutate(county = "Maui Nui")
oahu <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_ALL.csv")) %>% 
  select(-use_table4) %>% 
  mutate(county = "Honolulu")
hawaii <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table1_tmk_registry_ALL_draft.csv")) %>% 
  rename(permit_story= unique_story ,
         bed_gapfill_flag = bedroom_data_source) %>% 
  select(-min_year,-max_year,-class_match_flag,-scenario,-pitt_assignment) %>% 
  mutate(n_buildings = "temp until fill",
         year_built_combined = "temp until fill",
         use_table_4 = "temp until fill",
         county = "Hawaii")

setdiff(colnames(maui_nui), colnames(hawaii))

# complete tmk registry 

all_osds_tmks <- rbind(kauai,
                       maui_nui,
                       oahu,
                       hawaii) %>% 
  mutate(tmk = as.character(tmk))


# save draft
write_csv(all_osds_tmks,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_ALL.csv"))

```


### Read in NEW spatial information 

```{r}
# tmk - this will be the first join
tmk_parcels <- st_read(file.path(data_dir,"osds/raw-data-downloaded/tmk_state_parcels_shp/tmk_state.shp"))

crs(tmk_parcels)


# read in watersheds - spatial join 

watersheds_shp <- st_read(file.path(data_dir,"Coastlines_Watersheds/WBD_HU12_n83utm4_2023fixed.shp")) %>% 
  select(HUC12,Name)

crs(watersheds_shp)


# flow lines 
flowlines <- st_read(file.path(driver_dir,"0_Input_Data_and_Models/Common_2024/Finalized/Coastlines_Watersheds/NHD_Flowlines/NHD_Flowlines.shp"))

```



### Read in OLD files for each island, isolate spatial info if available
These would all come from old otp data available. 

```{r}
# oahu 
otp_oahu_raw <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/OSDS_HIGIS_2010/osds_oah.shp")) %>% 
  clean_names() %>% 
  mutate(tmk = as.character(tmk_9))


```



## Analysis 

```{r}

sf_use_s2(FALSE)

### Step 1: Get TMKS centroid and join to table 1

tmk_osds <- tmk_parcels %>% 
  filter(TMK_txt %in% all_osds_tmks$tmk) %>% 
  mutate(is_valid = st_is_valid(geometry)) %>% 
  st_cast("MULTILINESTRING") %>%   # fix invalid non closures
  st_cast("LINESTRING") %>%        
  st_polygonize()

tmk_centroids <- st_centroid(tmk_osds)
tmk_pos <- st_point_on_surface(tmk_osds)

# Look at difference for which one to choose

kauai_centroids <- tmk_centroids %>% filter(county == "Kauai")
kauai_pos <- tmk_pos %>% filter(county == "Kauai")

mapview(kauai_centroids, col.regions = "blue")+
  mapview(kauai_pos, col.regions = "lightblue")


tmk_unique_points <- tmk_pos %>%
  # Select the ID and ensure we only have one row per ID
  distinct(TMK_txt, .keep_all = TRUE) %>%
  select(tmk = TMK_txt)

tmk_unique_centroids <- tmk_centroids %>%
  # Select the ID and ensure we only have one row per ID
  distinct(TMK_txt, .keep_all = TRUE) %>%
  select(tmk = TMK_txt)

tmk_pos_join <- all_osds_tmks %>%
  left_join(tmk_unique_points, by = "tmk")


missing_tmks <- all_osds_tmks %>%
  anti_join(tmk_unique_points, by = "tmk")

nrow(missing_tmks)
head(missing_tmks)

osds_with_tmkpos_cl <- tmk_pos_join %>% 
  filter(tmk %!in% missing_tmks$tmk) %>% 
  st_as_sf() %>% 
  st_transform(crs = crs(watersheds_shp)) # 895 dont have a tmk point

### Step 2: Spatial join to watersheds

osds_with_watersheds <- st_join(osds_with_tmkpos_cl, watersheds_shp, join = st_intersects)


outside_count <- sum(is.na(osds_with_watersheds$Name)) # Change to your column name
print(paste("Points with no watershed match:", outside_count))

insiders <- osds_with_watersheds %>% filter(!is.na(Name))
outsiders <- osds_with_watersheds %>% filter(is.na(Name))

nearest_indices <- st_nearest_feature(outsiders, watersheds_shp)

dist <- st_distance(outsiders, watersheds_shp[nearest_indices, ], by_element = TRUE)


outsiders_fixed <- outsiders %>%
  select(-colnames(st_drop_geometry(watersheds_shp))) %>% 
  bind_cols(st_drop_geometry(watersheds_shp[nearest_indices, ]))


mapview(outsiders)

osds_final_watersheds <- bind_rows(insiders, outsiders_fixed) # 13 donʻt have a watershed

outside_count2 <- sum(is.na(osds_final_watersheds$Name))

outsiders2 <- osds_final_watersheds %>% filter(is.na(Name))
mapview(outsiders2)


st_write(osds_final_watersheds,
         dsn = file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.csv"))

write_csv(missing_tmks,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withOUTSpatial.csv"))


### Last Step: Flow Lines. These donʻt have years? Question for Kim

```



