---
title: "Gapfill OSDS with No Dwelling information"
author: "Madeline Berger"
date: "2026-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

# Read in necessary data

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  rename(tmk = full_taxid) %>% 
  clean_names()

iws_oahu_permits_tmk <- iws_oahu_permits %>% pull(tmk)

iws_kauai_maui_permits <- iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"2") | str_starts(tmk_char,"4")) %>% 
  rename(tmk = tmk_char)

```


# Gapfilling OSDS missing dwell data 

### How many are we missing information for, across islands?

```{r}

# ------------------------ Read in Table 4s for each Island - missing Dwell Data ------------------ #

# read in table 4 from oahu - this one has x y info

oahu_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_percommercial_draft.csv")) %>% 
  mutate(tmk = as.character(tmk))

coords_fortbl4 <- oahu_table4 %>% 
  select(tmk, x,y) %>% 
  filter(!is.na(x) & !is.na(y)) %>% 
  st_as_sf(coords = c("x","y"), crs = 32604)

coordinates_latlon <- st_transform(coords_fortbl4, crs = 4326) #2633

mapview(coordinates_latlon)

api_ready_coords_oah <- coordinates_latlon %>%
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1],
    google_coords = paste(lat, lon, sep = ",")
  ) %>%
  st_drop_geometry()


# get addresses - if there are coordinates, then use these rather than address because it will be more precise.

addresses_fortbl4 <- iws_oahu_permits %>% 
  filter(tmk %in% oahu_table4$tmk) %>% 
  filter(tmk %!in% api_ready_coords_oah$tmk) %>% 
  select(tmk, starts_with("project")) %>% 
  mutate(street_house = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
         # Create the full address with a conditional check for the zip code
    address_full = if_else(is.na(project_zip_code) | project_zip_code == "NA",
                           paste(street_house, project_city, sep = ", "),
                           paste(street_house, project_city, project_zip_code, sep = ", ")),
    
    # Final step: convert everything to lower case
    address_full = str_to_lower(address_full)
    )


api_ready_address_oah <- addresses_fortbl4 %>%
  select(tmk, address_full)


# --------------------- Kauai Maui Hawaii ------------------------------ #

# these do NOT have xy data. So weʻll use the TMK centroids below 

mauinui_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_percommercial_draft.csv")) %>% 
  filter(commercial == "check google places")

kauai_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table4_percommercial_draft.csv")) %>% filter(commercial == "check google places")

hi_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table4_updated_draft.csv")) %>% 
  filter(data_presence == "Empty (All 0 or NA)")


kauai_maui_table4 <- bind_rows(mauinui_table4, kauai_table4) %>% 
  mutate(tmk = as.character(tmk))# 471


km_addresses_table4 <- iws_kauai_maui_permits %>% 
  clean_names() %>% 
  filter(tmk %in% kauai_maui_table4$tmk) %>% 
  #filter(tmk %!in% api_ready_coords_oah$tmk) %>% 
  select(tmk, starts_with("project")) %>% 
  mutate(street_house = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
         # Create the full address with a conditional check for the zip code
    address_full = if_else(is.na(project_zip_code) | project_zip_code == "NA",
                           paste(street_house, project_city, sep = ", "),
                           paste(street_house, project_city, project_zip_code, sep = ", ")),
    
    # Final step: convert everything to lower case
    address_full = str_to_lower(address_full)
    )


# --------------------- Read in Spatialized Table 1  ------------------------- #

## Alternative information to search with: TMK parcel centroids

table1_withspatial <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"))

tmk_for_api_kauaimaui <- table1_withspatial %>% 
  st_transform(crs = 4326) %>% 
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1]
  ) %>%
  #st_drop_geometry() %>%
  select(tmk, lat, lon) %>% 
  filter(tmk %in% kauai_maui_table4$tmk) #459 (we lost 12)


coords_fortbl4 <- tmk_for_api_kauaimaui %>% 
  select(tmk, lat,lon) %>% 
  filter(!is.na(lat) & !is.na(lon)) %>% 
  st_as_sf(coords = c("lat","lon"), crs = 4326)

coordinates_latlon <- st_transform(coords_fortbl4, crs = 4326) #2633

mapview(coordinates_latlon)

api_ready_coords_km <- coordinates_latlon %>%
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1],
    google_coords = paste(lat, lon, sep = ",")
  ) %>%
  st_drop_geometry()


tmk_for_api_oah <- table1_withspatial %>% 
  st_transform(crs = 4326) %>% 
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1]
  ) %>%
  #st_drop_geometry() %>%
  select(tmk, lat, lon) %>% 
  filter(tmk %in% oahu_table4$tmk) # 2218 (note we lost about 400, so these did not have dwell or tmk info?)


```

### First pass: Additional GIS Data for Flags

For this, weʻll use the TMKS.

Key info to help estimate effluent: 

"Hawaii’s public school system (which operates as a single statewide district) has an average school size of about 560 students. Statewide, there are roughly 295 public schools educating a total of about 165,000 students"

"teacher student ratio = 14:1"

```{r}

# read in additional layers 
state_landuse <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/State_Land_Use_Districts.geojson"))

public_schools <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/Public_Schools.csv")) %>%
  st_as_sf(coords = c("X","Y"), crs = 32604) %>% 
  mutate(
    enrollment = 560,
    faculty = 40,
    tuition = 0
  ) %>% 
  select(
    school_name = esis_name,
    address,
    city,
    zip,
    district,
    sch_type,
    complex_area,
    enrollment,
    faculty,
    tuition
  )

private_schools <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/Private_Schools.csv")) %>% 
  st_as_sf(coords = c("X","Y"), crs = 32604) %>% 
  mutate(
    sch_type = "private",
    complex_area = "unknown",
    district = "to add"
  ) %>% 
  select(
    school_name = school,
    address,
    city,
    zip = zipcode,
    district,
    sch_type,
    complex_area,
    enrollment = enrollment21,
    faculty = faculty21,
    tuition = tuition21
  )

all_schools <- rbind(public_schools, private_schools)

all_schools_nosf <- st_drop_geometry(all_schools)


# Pre-screening: Landuse code

landuse_km <- st_join(tmk_for_api_kauaimaui, state_landuse, join = st_intersects) %>% 
  select(tmk, lat, lon, ludcode,island) %>% 
  st_drop_geometry()

landuse_oah <- st_join(tmk_for_api_oah, state_landuse, join = st_intersects)

landuse_km_missing <- landuse_km %>% 
  filter(is.na(ludcode)) # these donʻt have spatial info, 17

landuse_oah_missing <- landuse_oah %>% 
  filter(is.na(ludcode)) # 40 

# which are NA?

mapview(landuse_oah_missing) # in the ocean

# Schools 

mapview(public_schools)+
  mapview(private_schools)+
  mapview(tmk_for_api_kauaimaui, col.regions = "red")


# Manual Inspection Dataframe

# here I actually looked at the map and confirmed ones that are schools, and a few other non-residences. We can use
# this to compare how well the spatial join works, as well 

manual_school_df <- data.frame(
  FID = c(
    427,
    428,
          285,
          35,
          36,
          29,
          17,
          4,
          87,
          54,
          53,
          52),
  School = c(
    "Molokai Middle School",
    "Molokai Middle School",
    "Seabury Hall",
    "Kauai High School",
    "Kauai Community College",
    "Koloa Elem School",
    "Kalaheo Elem School",
    "Kekaha Elem School",
    "Ke Kula Niihau O Kekaha-LPCS",
    "St. Catherine School",
    "St. Catherine School",
    "St. Catherine School"
  )
)

manual_other_df <- data.frame(
  FID = c(
    213,
    186,
    179,
    86,
    10,
    5,
    6
  ),
  Place = c(
    "Post office",
    "Post office",
    "Pot office",
    "Healthy Hut Natural Foods",
    "West Inn Hotel",
    "LDS Church",
    "Church"
  )
)

# join the school info to the osds points. then join with HI GIS school to get enrollment numbers

km_pts_with_fid <- tmk_for_api_kauaimaui %>% 
  mutate(FID = row_number()) %>% 
  left_join(manual_school_df, by = "FID") %>% 
  left_join(manual_other_df, by = "FID")

# check to make sure FID aligns
mapview(km_pts_with_fid)+
  mapview(tmk_for_api_kauaimaui, col.regions = "red")

# create matrix and test 

within_200m_matrix <- st_is_within_distance(km_pts_with_fid, all_schools, dist = 200, sparse = FALSE)

validation_set <- km_pts_with_fid %>%
  mutate(
    buffer_flagged_200m = rowSums(within_200m_matrix) > 0
  )

km_pts_with_allinfo <- km_pts_with_fid %>% 
  rename(school_name = School) %>% 
  mutate(school_name = ifelse(is.na(school_name),"not a school",school_name)) %>% 
  left_join(all_schools_nosf, by = "school_name") %>% 
  mutate(tuition = case_when(
    school_name == "Seabury Hall" ~ "22270",
    school_name == "St. Catherine School" ~ "7300",
    TRUE ~ tuition)) %>% 
  left_join(landuse_km, by = c("tmk","lat","lon"))


# join back to table 4 for kauai and maui 

kauai_maui_table4_withinfo <- kauai_maui_table4 %>% 
  left_join(km_pts_with_allinfo, by = "tmk") %>% 
  select(tmk, x,y,lat,lon, school_name,classification,enrollment, faculty, tuition)

kauai_maui_table4_schools <- kauai_maui_table4_withinfo %>% 
  filter(school_name != "not a school")

kauai_maui_table4_other<- kauai_maui_table4_withinfo %>% 
  filter(!is.na(Place))

kauai_maui_identified <- rbind(kauai_maui_table4_schools, kauai_maui_table4_other)


write_csv(kauai_maui_identified,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/table4_pass1_complete_kauaimaui.csv"))


# filter out list for api

kauai_maui_api <- api_ready_coords_km %>% 
  filter(tmk %!in% kauai_maui_identified)
  

```


``` {r}
# ok its pretty good! letʻs try it on Oahu

oahu_schools <- all_schools %>% 
  filter(district %in% c("Honolulu",
                         "Central",
                         "Leeward",
                         "Windward"))

oahu_private_schools <- private_schools %>% 
  filter(city %in% c("Honolulu",
                     "Aiea",
                     "Kailua",
                     "Ewa",
                     "Kane'ohe",
                     "Kapolei",
                     "Mililani",
                     "Mililani Town",
                     "Pearl City",
                     "Waianae",
                     "Wahiawa",
                     "Waialua",
                     "Waipahu"))


all_oahu_schools <- rbind(oahu_schools, oahu_private_schools)

all_oahu_schools_nosf <- st_drop_geometry(all_oahu_schools)

oah_pts_with_fid <- tmk_for_api_oah %>% 
  mutate(FID = row_number())


within_200m_matrix <- st_is_within_distance(oah_pts_with_fid, all_oahu_schools, dist = 200, sparse = FALSE)

validation_set <- oah_pts_with_fid %>%
  mutate(
    buffer_flagged_200m = rowSums(within_200m_matrix) > 0
  )

possible_schools <- validation_set %>% 
  filter(buffer_flagged_200m == TRUE)

# manually check

mapview(possible_schools, col.regions = "yellow", alpha = 0.5) + 
  mapview(all_oahu_schools, col.regions = "lightgreen")

```


``` {r}
# 1. Using the FID from "possible schools" 
schools_confirmed_df <- tibble(
  FID2 = list(
    22, 55, 34, c(30, 31), 210, 42, 89, c(40, 41), 37, 
    96, 106, 97, c(68, 74), 73, 121, c(82, 112, 113), 
    44, c(46, 47, 48), 43, 45
  ),
  School_Name = c(
    "Kahaluu Elem School",
    "Enchanted Lake Elem School",
    "Malama Honua - PCS",
    "Pope Elem School",
    "Mary Star of the Sea Early Learning Center", # Check this TMK
    "La Pietra - Hawaii School for Girls",
    "Liholiho Elem School",
    "Kaimuki High School",
    "Hokulani Elem School",
    "Mid-Pacific Institute",
    "Lunalilo Elem School",
    "Maryknoll School",
    "Hawaiian Mission Academy - 9-12 Campus",
    "Roosevelt High School",
    "Stevenson Middle School",
    "Lincoln Elem School",
    "Kawaiaha'o Church School",
    "St. Andrew's Schools - The Priory",
    "Central Middle School",
    "Royal Elem School"
  )
) %>% 
  unnest(FID2) # Separates combined arrays into individual rows automatically

# 2. Clean Other Land Uses Data Frame
other_landuse_df <- tibble(
  FID2 = list(
    21, 62, 57, 128, 129, 125, 39, 123, 122, 124, 95, 104,
    c(83, 84, 85, 86, 87), 88, c(77, 78), 76, c(35, 36)
  ),
  Classification = c(
    "residence",
    "powerplant",
    "HIDOE School Food Services Branch",
    "The Arc in Hawaii Diamond Head",
    "Student Services Support Branch",
    "ACE Hardware",
    "Hawaii Storage",
    "Palolo Public Housing",
    "Anuenue School",
    "Gussies House",
    "Manoa Marketplace",
    "Church with pottery guild",
    "Makiki Skate Park or St Clements School",
    "Residence or Hawaii Mission Academy Ka Lama Iki",
    "apartment building",
    "residence",
    "mixed commercial"
  )
) %>% 
  unnest(FID2)


#2. Confirmed with other FID from tmk_api_oahu

added_schools_confirmed_df <- tibble(
  FID = list(
    111, 1875, 590, 597, 2115, 2113, 972, 1446, 1448, 1852, 
    1794, 121, 308, c(304, 306), 294, c(70, 263), 217, c(53, 230, 231), 
    41, c(1131, 1132, 1134)
  ),
  School_Name = c(
    "Kawananakoa Middle School",
    "Hongwanji Mission School",
    "Waolani Judd Nazarene School",
    "Lanakila Elem School",
    "Hawaii Technology Academy-PCS",
    "Waipahu Elem School",
    "Iroquois Point Elem School",
    "Barbers Point Elem School",
    "Island Pacific Academy",
    "Nanaikapono Elem School",
    "Mililani Ike Elem School",
    "Waialua High & Inter School",
    "St. Michael School",
    "Waialua Elem School",
    "Haleiwa Elem School",
    "Sunset Beach Elem School",
    "Laie Elem School",
    "Hauula Elem School",
    "Kaaawa Elem School",
    "Waiahole Elem School"
  )
) %>% 
  unnest(FID)

# 2. Clean Other Land Uses Data Frame (Added FID Set)
added_other_landuse_df <- tibble(
  FID = list(
    1880, 1862, 591, 1951, 2119, 977, 986, 1854, 2008, 295, 
    c(262, 264), 214
  ),
  Classification = c(
    "Auwaiolimu Neighborhood Park",
    "Pauoa Substation",
    "City Memorial Park",
    "House across the street from the Children's House",
    "Hawaii Plantation Village",
    "Strip mall in Ewa Beach",
    "Reynolds Recycling / Mall in Ewa Beach",
    "Longs Drugs",
    "HDOT Waianae Baseyard",
    "House Across the street from Haleiwa Elementary School",
    "Ehukai Beach Park",
    "Kahuku Sugar Mill"
  )
) %>% 
  unnest(FID)

# join everything together
possible_schools_withname <- possible_schools %>% 
  mutate(FID2 = row_number()) %>% 
  left_join(schools_confirmed_df, by = "FID2") %>% 
  left_join(other_landuse_df, by = "FID2") %>% 
  left_join(added_schools_confirmed_df, by = "FID") %>% 
  left_join(added_other_landuse_df, by = "FID") %>% 
  mutate(school_name = coalesce(School_Name.x, School_Name.y)) %>% 
  mutate(Classification = coalesce(Classification.x, Classification.y)) %>% 
  select(-Classification.x, -Classification.y,-School_Name.x, -School_Name.y)

# join enrollment info
confirmed_schools_withinfo <- possible_schools_withname %>% 
  mutate(school_name = ifelse(is.na(school_name),"not a school",school_name)) %>% 
  left_join(all_oahu_schools_nosf, by = "school_name") %>% 
  filter(school_name != "not a school") %>%
  mutate(classification = "school")


confirmed_others <- possible_schools_withname %>% 
  filter(!is.na(Classification))
  

# add landuse, schoool or classification info back to original table 4s. 

oahu_table4_wSchools <- oahu_table4 %>% 
  left_join(possible_schools_withname, by = "tmk") %>%
  mutate(Classification = ifelse(!is.na(school_name),"school",Classification))
 


# save oahu table 4

write_csv(oahu_table4_wSchools,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass1_complete.csv"))

# read in pass1 if starting here

oahu_pass1_complete <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass1_complete.csv"))

# filter out missig for API
oahu_for_pass2 <- oahu_pass1_complete %>% 
  filter(is.na(Classification))


oahu_api_ready_coords <- api_ready_coords_oah %>% 
  filter(tmk %in% oahu_for_pass2$tmk)


```


### Second pass: Google Places, using Coordinations

Activate Google Places API

```{r}

# See if we have the key

# Load the key from the local file
source(here("config.txt"))

# Set it as an environment variable so packages can find it
Sys.setenv(GOOGLE_PLACES_KEY = GOOGLE_PLACES_KEY)

# Verify (should print your key)
print(Sys.getenv("GOOGLE_PLACES_KEY"))


# Verify it immediately
if(Sys.getenv("GOOGLE_PLACES_KEY") == "") {
  stop("API Key is missing!")
} else {
  print("Success! Key is loaded.")
}

key <- Sys.getenv("GOOGLE_PLACES_KEY")

# Load packages
library(googleway)
library(dplyr)

# ----------- Search by coordinates improved

library(tidyr)
library(purrr)

boring_types <- c("premise", "establishment", "point_of_interest", "political", 
                  "floor", "locality", "neighborhood", "geocode","route")

api_count <- 0
safety_limit <- 3500

safe_google_call <- possibly(
  function(lat, lon, key) { google_places(location = c(lat, lon), radius = 50, key = key) }, 
  otherwise = NULL
)

# Unified Function: One API Call, Smart Local Filtering
get_place_category <- function(lat, lon) {
  
  # Increment global counter
  api_count <<- api_count + 1 
  if (api_count > safety_limit) {
    stop(paste("SAFETY STOP: Limit of", safety_limit, "reached! Full pipeline killed."))
  }
  
  # Call once with a reasonable operational radius (e.g., 50 meters)
  res <- google_places(location = c(lat, lon), radius = 50, key = key)
  
  # Check if we got valid results back
  if (!is.null(res$results) && length(res$results) > 0 && nrow(res$results) > 0) {
    
    # Process the results completely inside R memory (Free!)
    # We unnest the types list column to look at all categories returned per place
    df_results <- res$results %>%
      select(name, types) %>%
      mutate(row_order = row_number()) %>% # Keep track of Google's prominence ranking
      unnest(types)
    
    # Strategy A: Try to find the highest-ranked "interesting" business type
    interesting_places <- df_results %>%
      filter(!types %in% boring_types)
    
    if (nrow(interesting_places) > 0) {
      # Grab the top-ranked non-boring place
      top_match <- interesting_places %>%
        filter(row_order == min(row_order)) %>%
        slice(1)
      return(list(name = top_match$name, type = top_match$types))
    }
    
    # Strategy B: Fallback if EVERYTHING returned was labeled boring
    # Just take Google's absolute #1 top result's first type
    fallback_place <- res$results[1, ]
    fallback_type <- if (!is.null(fallback_place$types[[1]])) fallback_place$types[[1]][1] else "unknown"
    return(list(name = fallback_place$name, type = fallback_type))
  }
  
  # Strategy C: Absolute fallback if Google found nothing at those coordinates
  return(list(name = "No result", type = "unknown"))
}

# --- EXECUTION OAHU --- #

# Safely wrap the function to prevent an API error from crashing your whole loop
safe_get_place <- possibly(get_place_category, otherwise = list(name = "Error", type = "error"))

processed_oah_data_all <- oahu_api_ready_coords %>%
  rowwise() %>%
  mutate(
    api_res = list(safe_get_place(lat, lon)),
    place_name = api_res$name,
    final_category = api_res$type
  ) %>%
  select(-api_res)

boring_types_updated <- boring_types[c(1,2,4:9)] 
  
oahu_pass2_successful <- processed_oah_data_all %>% 
  filter(final_category %!in% boring_types_updated) %>% 
  filter(final_category != "unknown") # N = 964

oahu_for_pass3 <- processed_oah_data_all %>% 
  filter(tmk %!in% oahu_pass2_successful$tmk) # N = 1628

# save both for future processing. 

write_csv(oahu_pass2_successful,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass2_complete.csv"))

write_csv(oahu_for_pass3,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass3_stillunknown.csv"))

# old 

processed_data2 <- oahu_api_ready_coords %>%
  filter(tmk %!in% c("149004001","149004001","149001007","148010007","156003038")) %>% 
  head(5) %>% # Test cluster
  rowwise() %>%
  mutate(
    api_res = list(safe_get_place(lat, lon)),
    place_name = api_res$name,
    final_category = api_res$type
  ) %>%
  select(-api_res) # Drop the temporary list column

processed_data3 <- oahu_api_ready_coords %>%
  filter(tmk %!in% c("149004001","149004001","149001007","148010007","156003038")) %>% 
  filter(tmk %!in% c("156002028","156002001","156005009","156006044","156002013")) %>% 
  #head(5) %>% # Test cluster
  rowwise() %>%
  mutate(
    api_res = list(safe_get_place(lat, lon)),
    place_name = api_res$name,
    final_category = api_res$type
  ) %>%
  select(-api_res)

# check out results? 


processed_data3_sf <- st_as_sf(processed_data3, coords = c("lon","lat"), crs = 4326) %>% 
  st_transform(crs = 4326) %>% 
  filter(final_category != "error")

mapview(processed_data3_sf, zcol = "final_category")


# --- EXECUTION KAUAI MAUI --- #

kauai_maui_1 <- kauai_maui_api %>% 
  rowwise() %>%
  mutate(
    api_res = list(safe_get_place(lat, lon)),
    place_name = api_res$name,
    final_category = api_res$type
  ) %>%
  select(-api_res)

unique(kauai_maui_1$final_category)

kauai_maui_pass2_sucess <- kauai_maui_1 %>% 
  filter(final_category %!in% boring_types) %>% 
  filter(final_category != "unknown") # N = 220

# join to table 4

kauai_maui_table4_pass2 <- kauai_maui_table4 %>% 
  left_join(kauai_maui_pass2_sucess, by = "tmk")

# still needed - add address

kauai_maui_pass3 <- kauai_maui_table4_pass2 %>% 
  filter(tmk %!in% kauai_maui_pass2_sucess$tmk) # N = 222

api_pass3 <- kauai_maui_api %>% 
  filter(tmk %in% kauai_maui_pass3$tmk)

# read in iws permits for kauai and maui to get the addresses

km_addresses_forpass3 <- km_addresses_table4 %>% 
  mutate(tmk = as.character(tmk)) %>% 
  filter(tmk %in% kauai_maui_pass3$tmk) %>% 
  left_join(api_pass3, by = "tmk")

km_addresses_forpass3_sf <- km_addresses_forpass3 %>% 
  filter(!is.na(lat)) %>% 
  st_as_sf(coords = c("lon","lat"), crs = 4326)

mapview(km_addresses_forpass3_sf)

```

### Third pass - Search by Address

```{r}


# --- SEARCH BY ADDRESS ---
# try one from address dataframe
addr_example <- "66-1009 kaukona hua road, waialua"

res_addr <- google_find_place(
  input = addr_example,
  key = key,
  inputtype = "textquery",
  fields = c("formatted_address", "name", "type")
)

addr_types <- res_addr$candidates$types[[1]]
print(paste("Address is a:", paste(addr_types, collapse = ", ")))


# function 
# 2. Logic for Address Search + Coordinate Extraction
get_detailed_address_info <- function(addr_string) {
  # Safety count
  # Access the counter from the global environment
  api_count <<- api_count + 1 
  
  # THE TOLL BOOTH: Check the limit BEFORE calling Google
  if (api_count > safety_limit) {
    stop(paste("Safety limit of", safety_limit, "reached. Stopping to save your bank account!"))
  }
  
  # We ask for 'geometry' so we get the lat/lon even if the type is boring
  res <- google_find_place(
    input = addr_string, 
    key = key,
    inputtype = "textquery", 
    fields = c("name", "type", "geometry")
  )
  
  if (length(res$candidates) > 0 && nrow(res$candidates) > 0) {
    
    # isolate top match row
    top_cand <- res$candidates[1, ]
    # extract type from top match
    main_type <- if (!is.null(top_cand$types[[1]])) top_cand$types[[1]][1] else "unknown"
    
    # get coordinates safely from the nested geometry data frame
    lat_val <- top_cand$geometry$location$lat[1]
    lon_val <- top_cand$geometry$location$lng[1] # Note the 'lng'
    
    # Flag if it lands on a boring type
    needs_check <- main_type %in% boring_types
    
    return(list(
      name = top_cand$name[1], 
      type = main_type, 
      lat = lat_val,
      lon = lon_val,
      check = needs_check
    ))
  }
  return(list(name = NA, type = NA, lat = NA, lon = NA, check = FALSE))
}


# 4. Execute on address for kauai and Maui

results_addresses <- km_addresses_forpass3 %>%
  rowwise() %>%
  mutate(
    # Step 1: Initial Address Lookup
    init = list(get_detailed_address_info(address_full)),
    found_name = init$name,
    found_type = init$type,
    temp_lat = init$lat,
    temp_lon = init$lon,
    check = init$check
  )






```

### Build Final Table 4s!

Combine all the findings from pass 1, 2 and 3. Everything else will have to be a manual classification. First - we need to map to what Bob found.

Report here: chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://health.hawaii.gov/wastewater/files/2015/09/OSDS_OAHU.pdf

```{r}

# recreate table 3 form the Oahu priortization report for type to effluent mapping 

table_3_Oahu <- data.frame(
  type = c("baseyard",
           "business",
           "cemetary",
           "church",
           "golf course",
           "non-profit organization",
           "non-profit organzation with showers",
           "park",
           "school - with gym",
           "school - without gym"),
  min_estEffluent_gpd = c(
    195,
    130,
    1200,
    540,
    540,
    240,
    1825,
    200,
    600,
    12625
  ),
    max_estEffluent_gpd = c(
    390,
    4590,
    2400,
    2600,
    1080,
    2200,
    2500,
    800,
    5640,
    42425
  )
) %>% 
  mutate(median_effluent_gpd = (max_estEffluent_gpd + min_estEffluent_gpd) / 2)

# other islands?

```


#### Kauai and Maui Nui
```{r}

# pass 1 # need to add table 3 mapping
kauai_maui_identified_clean <- kauai_maui_identified %>% 
  select(tmk,lat, lon, school_name,place_name = Place, enrollment, faculty,ludcode) %>% 
  mutate(place_name = ifelse(school_name != "not a school","school",place_name)) %>% 
  mutate(type_table3 = case_when(
    place_name == "school" ~ "school with showers",
    place_name == "Healthy Hut Natural Foods" ~ "business",
    place_name %in% c("Post office","Pot office") ~ "non-profit organization",
    place_name %in% c("Church","LDS Church") ~ "church",
    place_name %in% "West Inn Hotel" ~ "business",
    TRUE ~ place_name
  )) %>% 
  mutate(class_method = "manual ID")


# pass 2 # tmk, lat , lon, place_name, final_cat. need to add school details
kauai_maui_pass2_forjoin <- kauai_maui_pass2_sucess %>% 
  select(-google_coords) %>% 
  mutate(
    table_3_type = case_when(
      # --- CHURCH ---
      final_category == "church" ~ "church",
      
      # --- CEMETARY ---
      final_category %in% c("cemetery", "cemetary") ~ "cemetary",
      
      # --- PARK ---
      final_category %in% c("park", "campground") ~ "park",
      
      # --- SCHOOLS ---
      # Defaulting to 'without gym' since the API doesn't specify structural facilities
      final_category %in% c("school", "primary_school", "secondary_school") ~ "school - without gym",
      
      # --- NON PROFIT --
      # Defaulting without shower
      final_category %in% c("local_government_office", "post_office") ~ "non-profit organization",
      
      # --- BASEYARD ---
      final_category %in% c("general_contractor", "roofing_contractor", 
                            "storage", "car_repair", "hardware_store", "laundry") ~ "baseyard",
      
      # --- BUSINESS (Food, Retail, Offices, Medical) ---
      final_category %in% c(
        "restaurant", "food", "cafe", "meal_takeaway", "bakery", "bar",
        "store", "grocery_or_supermarket", "supermarket", "convenience_store", 
        "liquor_store", "clothing_store", "jewelry_store", "home_goods_store", "shopping_mall",
        "real_estate_agency", "accounting", "finance", "bank", "atm", "travel_agency",
        "health", "doctor", "dentist", "physiotherapist", "spa", "beauty_salon", "gym",
        "lodging", "gas_station", "tourist_attraction", "art_gallery"
      ) ~ "business",
      
      # --- CATCH-ALL FOR ERRORS / BLANKS ---
      final_category == "error" ~ NA_character_,
      is.na(final_category)     ~ NA_character_,
      
      # Safety line: anything missed gets flagged so your code doesn't crash
      TRUE ~ "unclassified_check_me"
    )
  ) %>% 
  mutate(class_method = "google places coordiate search") %>%
  rename(found_type = final_category)


# pass 3 
results_addresses_forjoin <- results_addresses %>% 
  select(tmk, lat,lon,found_name,found_type) %>% 
  distinct() %>% 
  rowwise() %>% # Ensure rowwise is active if processing row-by-row text
  mutate(
    # Step 1: Clean up the 'establishment' tags by looking directly at the names
    table_3_type = case_when(
      found_type != "establishment" ~ "unknown", # If Google gave a good tag, keep it!
      
      # --- Manual Name Overrides for 'establishment' rows ---
      
      # PARKS
      str_detect(found_name, "(?i)Park|Beach") ~ "park",
      
      # SCHOOLS / COLLEGES
      str_detect(found_name, "(?i)School|Academy|College|KIPA") ~ "school - with gym",
      
      # INDUSTRIAL / TRUCKING / MILLS (Baseyards)
      str_detect(found_name, "(?i)Concrete|Mill|Farming Operation|Biodiesel") ~ "baseyard",
      
      # INSTITUTIONAL / CIVIL (Non-Profit No Shower)
      str_detect(found_name, "(?i)Postal|Civic Center|Neighborhood Center|Foundation") ~ "non-profit organization",
      
      # RETAIL / FOOD / MARKETS / TOURISM (Business)
      str_detect(found_name, "(?i)Bar|Restaurant|Joe's|Duke's|Market|Center|Zipline|Tour|Weddings|Housing|Camp|Spa") ~ "business",
      
      # Fallback if an establishment name doesn't hit a keyword
      TRUE ~ "establishment"
    )
  ) %>%
  ungroup() %>% 
  mutate(class_method = "google places address search") %>% 
  rename(place_name = found_name)



# combine all together, compare to table 4, and mark missing

kauai_maui_identified_final <- kauai_maui_identified_clean %>% 
  select(tmk,lat,lon, place_name,table_3_type = type_table3,class_method) %>% 
  mutate(found_type = NA) %>% 
  rbind(kauai_maui_pass2_forjoin) %>% 
  rbind(results_addresses_forjoin)

kauai_maui_table4_final <- kauai_maui_table4 %>% 
  left_join(kauai_maui_identified_final, by = "tmk") # duplicates we need to get rid of

kauai_maui_table4_deduplicated <- kauai_maui_table4_final %>%
  # Step 1: Create ordering scores for logic
  mutate(
    # Method rank: lower number = better reliability
    method_rank = case_when(
      class_method == "manual ID" ~ 1,
      class_method == "google places coordiate search"    ~ 2,
      TRUE                            ~ 3
    ),
    
    # Category rank: lower number = more specific
    category_rank = case_when(
      # Specific Table 3 categories are the absolute best
      table_3_type %in% c("school - without gym", "school - with gym", 
                            "church", "cemetary", "park", "baseyard", "business", "non-profit organization") ~ 1,
      
      # Vague API results are second-best
      table_3_type == "establishment" ~ 2,
      
      # Premise/Route/NA are the least useful
      found_type %in% c("premise", "route", "street_address","error") ~ 3,
      is.na(found_type) ~ 4,
      
      TRUE ~ 5
    )
  ) %>% 
  # Step 2: Group by unique spatial keys
  group_by(tmk, lat, lon) %>%
  # Step 3: Sort the duplicates inside each group by ranks
  # R will sort by method_rank first, and break ties using category_rank
  arrange(method_rank, category_rank, .by_group = TRUE) %>%
  # Step 4: Grab the #1 best row for each TMK combination
  slice(1) %>%
  # Step 5: Clean up by unlocking the dataframe and dropping temporary rank columns
  ungroup() %>%
  select(-method_rank, -category_rank) # 247


kauai_maui_table4_still_missing <- kauai_maui_table4_deduplicated %>% 
  filter(table_3_type == "unknown" | is.na(table_3_type))  # N 224
  

# SAVE!!

write_csv(kauai_maui_table4_deduplicated,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/table4_kauaimaui_Final_allentries.csv"))


```

#### Oahu 

```{r}

oahu_pass1_id <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass1_complete.csv"))

oahu_pass2_id <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass2_complete.csv"))

oahu_still_missing <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_pass3_stillunknown.csv"))

# ------- map classes in pass 1 to table 3

oahu_p1_mapped <- oahu_pass1_id %>% 
  select(tmk,lat, lon, school_name,Classification) %>% 
  mutate(classification = str_to_lower(Classification)) %>% 
  mutate(name_detect = case_when(
    str_detect(classification,"church") ~ "church",
    str_detect(classification,"school") ~ "school",
    str_detect(classification,"park") ~ "park",
    str_detect(classification,"apartment") | str_detect(classification,"house") | str_detect(classification,"housing") | str_detect(classification, "residence") ~ "residential",
    str_detect(classification,"powerplant") | str_detect(classification,"substation") ~ "powerplant",
    str_detect(classification,"branch") | str_detect(classification,"hdot") | str_detect(classification,"arc") ~ "nonprofit or government",
    TRUE ~ "business"
  )) %>% 
  mutate(name_detect = ifelse(is.na(classification),NA,name_detect)) %>% 
  mutate(table_3_type = case_when(
    name_detect == "school" ~ "school with gym",
    name_detect == "business" ~ "business",
    name_detect == "park" ~ "park",
    name_detect == "church" ~ "church",
    name_detect == "nonprofit or government" ~ "nonprofit",
    name_detect == "powerplant" ~ "nonprofit",
    name_detect == "residential" ~ "residential",
    TRUE ~ NA
  )) %>% 
  mutate(class_method = "manual ID")

oahu_pass1_mapped_forjoin <-oahu_p1_mapped %>% 
  select(tmk, lat, lon, place_name = classification,table_3_type,class_method) %>% 
  mutate(found_type = NA)
  
  
# ------------- map classes from pass 2 to table 3

oahu_pass2_mapped <- oahu_pass2_id %>% 
  select(-google_coords) %>% 
  mutate(
    table_3_type = case_when(
      # --- CHURCH ---
      final_category %in% c("church","place_of_worship") ~ "church",
      
      # --- CEMETARY ---
      final_category %in% c("cemetery", "cemetary") ~ "cemetary",
      
      # --- PARK ---
      final_category %in% c("park", "campground") ~ "park",
      
      # --- SCHOOLS ---
      # Defaulting to 'without gym' since the API doesn't specify structural facilities
      final_category %in% c("school", "primary_school", "secondary_school") ~ "school - without gym",
      final_category %in% c("university") ~ "school - with gym",
      
      # --- NON PROFIT --
      # Defaulting without shower
      final_category %in% c("local_government_office", "post_office","bus_station","library","transit_station", "museum","police") ~ "non-profit organization",
      
      final_category %in% c("embassy","fire_station") ~ "non-profit organization with showers",
      
      # --- BASEYARD ---
      final_category %in% c("general_contractor", "roofing_contractor", 
                            "storage", "car_repair", "hardware_store", "laundry") ~ "baseyard",
      
      # --- BUSINESS (Food, Retail, Offices, Medical) ---
      final_category %in% c(
        "restaurant", "food", "cafe", "meal_takeaway", "bakery", "bar",
        "store", "grocery_or_supermarket", "supermarket", "convenience_store", 
        "liquor_store", "clothing_store", "jewelry_store", "home_goods_store", "shopping_mall",
        "real_estate_agency", "accounting", "finance", "bank", "atm", "travel_agency",
        "health", "doctor", "dentist", "physiotherapist", "spa", "beauty_salon", "gym",
        "lodging", "gas_station", "tourist_attraction", "art_gallery","airport","point_of_interest"
      ) ~ "business",
      
      # --- CATCH-ALL FOR ERRORS / BLANKS ---
      final_category == "error" ~ NA_character_,
      is.na(final_category)     ~ NA_character_,
      
      # Safety line: anything missed gets flagged so your code doesn't crash
      TRUE ~ "unclassified_check_me"
    )
  ) %>% 
  mutate(table_3_type = ifelse(table_3_type == "unclassified_check_me","business",table_3_type)) %>% 
  mutate(class_method = "google places coordiate search") %>%
  rename(found_type = final_category) %>% 
  mutate(table_3_type = case_when(
    place_name %in% c("Manoa Falls Trail","Barbers Point Lighthouse","Ulupō Heiau State Historic Site","Hāwea heiau complex","East-West Center Japanese Garden","Diamond Head Lighthouse","The Clark Hatch Bark Park at Diamond Head","Hula Kahiko Dancer Statue & Pond") ~ "park",
    place_name %in% c("Sunset Beach Christian Church","St Michael","Waialua Hongwanji Mission") ~ "church",
    TRUE ~ table_3_type
  ))

oahu_pass2_mapped %>% 
  filter(table_3_type == "unclassified_check_me") %>% 
  pull(found_type) %>% 
  unique()

# Join pass 1 and 2 together, apply logic to prioritize when both show up

oahu_identified_final <- oahu_pass2_mapped %>% 
  select(tmk,lat,lon, place_name,table_3_type,class_method, found_type) %>% 
  rbind(oahu_pass1_mapped_forjoin) %>% 
  filter(!is.na(table_3_type))

# tmk check 1 - are all the tmks in the OG table 4 present in this joined version?

oahu_identified_final %>% filter(tmk %!in% oahu_table4$tmk) # nice!


# de duplicate

oahu_table4_deduplicated <- oahu_identified_final %>%
  # Step 1: Create ordering scores for logic
  mutate(
    # Method rank: lower number = better reliability
    method_rank = case_when(
      class_method == "manual ID" ~ 1,
      class_method == "google places coordiate search"    ~ 2,
      TRUE                            ~ 3
    ),
    
    # Category rank: lower number = more specific
    category_rank = case_when(
      # Specific Table 3 categories are the absolute best
      table_3_type %in% c("school - without gym", "school - with gym", 
                            "church", "cemetary", "park", "baseyard", "business", "non-profit organization") ~ 1,
      
      # Vague API results are second-best
      table_3_type == "establishment" ~ 2,
      
      # Premise/Route/NA are the least useful
      found_type %in% c("premise", "route", "street_address","error") ~ 3,
      is.na(found_type) ~ 4,
      
      TRUE ~ 5
    )
  ) %>% 
  # Step 2: Group by unique spatial keys
  group_by(tmk, lat, lon) %>%
  # Step 3: Sort the duplicates inside each group by ranks
  # R will sort by method_rank first, and break ties using category_rank
  arrange(method_rank, category_rank, .by_group = TRUE) %>%
  # Step 4: Grab the #1 best row for each TMK combination
  slice(1) %>%
  # Step 5: Clean up by unlocking the dataframe and dropping temporary rank columns
  ungroup() %>%
  select(-method_rank, -category_rank)

# Filter out residences and still missing

oahu_still_missing <- oahu_table4 %>% 
  filter(tmk %!in% oahu_table4_deduplicated$tmk)

oahu_residences <- oahu_table4_deduplicated %>% 
  filter(table_3_type == "residential")

ohau_table4_final <- oahu_table4_deduplicated %>% 
  filter(tmk %!in% oahu_still_missing$tmk) %>% 
  filter(tmk %!in% oahu_residences$tmk)


write_csv(ohau_table4_final,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/table4_oahu_allentries.csv"))

write_csv(oahu_still_missing,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_still_unidentified.csv"))

write_csv(oahu_residences,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_residences_notInDwell.csv"))


```

#### Combine with Big Island + Effluent Ranges to create final table 4

```{r}


# read in our 3 tables 

oahu_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/table4_oahu_allentries.csv")) %>% 
  select(tmk, table_3_type)

kauai_mauitable_4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/table4_kauaimaui_Final_allentries.csv")) %>% 
  select(tmk, table_3_type)

hi_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table4_updated_draft.csv"))


# how many hi have beds?

hi_table4_withbeds <- hi_table4 %>% filter(data_presence == "Bedrooms Recorded") # about 1/4. We will keep them for now to be more consistent with the other states

hi_types <- hi_table4_withbeds %>% pull(otp_commercial)
  
# map hi categories to the table 3 categories

hi_table4_mapped <- hi_table4_withbeds %>% 
  mutate(table_3_type = case_when(
    otp_commercial %in% c("small business",
                          "resturant") ~ "business",
    str_detect(otp_commercial, "church") ~ "church",
    str_detect(otp_commercial, "gov") | str_detect(otp_commercial, "non-profit") ~ "non-profit organization",
    str_detect(otp_commercial, "park") ~ "park",
    otp_commercial %in% c("small school, no gym","small school/no gym") ~ "school - without gym",
    otp_commercial == "small school/w gym" ~ "school - with gym",
    TRUE ~ NA
  )) %>% 
  select(tmk, table_3_type)
  
  
  
# bind all together and join to effluent calculations
  
table_4_allislands <- rbind(hi_table4_mapped,
                            kauai_mauitable_4,
                            oahu_table4) %>% 
  rename(type = table_3_type) %>% 
  filter(type %!in% c("unknown","unclassified_check_me")) %>% 
  filter(!is.na(type)) %>% 
  left_join(table_3_Oahu, by = "type")
  

still_missing <- table_4_allislands <- rbind(hi_table4_mapped,
                            kauai_mauitable_4,
                            oahu_table4) %>% 
  rename(type = table_3_type) %>% 
  filter(type %in% c("unknown","unclassified_check_me")) # 69
# save final table

write_csv(table_4_allislands, file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))
  
```

### Check against New Tables

Since we don't want to re-run these (it cost about 20) I'll just compare what we did before to the new and potentially improved table 4s, and see which are still missing. 

```{r reconcile}

# table classified
classified_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands_preclean.csv"))

# new table 4s post script alignment and cleaning
int_path <- file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4")

fps <- list.files(path = int_path,
                  pattern = "table4",
                  full.names = TRUE)
new_table4s <- lapply(fps, read_csv)
new_table4s_combined <- do.call(rbind, new_table4s[1:3])

oahu_table4 <- read_csv(fps[4]) %>% 
  mutate(commercial = NA, 
         pitt_code = NA)


new_table4s_all <- rbind(
  new_table4s_combined,
  oahu_table4)

# join with the classified one 
# clean up the usable table 4s, add effluent calculation, and save!

all_table4s_combined <- new_table4s_all %>% 
  left_join(classified_table4, by = "tmk")


usable_table4s <- all_table4s_combined %>% 
  filter(!is.na(type) | (!is.na(commercial) & commercial != "check google places")) %>%  
  mutate(comm_class_clean = str_to_lower(coalesce(type, commercial))) %>% 
  select(-type,-commercial)

missing_effluent_types <- usable_table4s %>% 
  filter(is.na(min_estEffluent_gpd) | min_estEffluent_gpd == "") %>% 
  distinct(comm_class_clean) %>% 
  pull(comm_class_clean)

unique(missing_effluent_types)


usable_table4s_witheffluent <- usable_table4s %>%
  select(-ends_with("gpd")) %>% 
  mutate(
    type = case_when(
      # Baseyards
      str_detect(comm_class_clean, "baseyard") ~ "baseyard",
      
      # Cemetaries
      str_detect(comm_class_clean, "cemet") ~ "cemetary",
      
      # Churches
      str_detect(comm_class_clean, "church") ~ "church",
      
      # Golf Courses
      str_detect(comm_class_clean, "golf course") ~ "golf course",
      
      # Schools with Gym/Showers
      str_detect(comm_class_clean, "school.*w/gym|school.*gym|school with showers|school with gym") ~ "school - with gym",
      
      # Schools without Gym
      str_detect(comm_class_clean, "school") ~ "school - without gym",
      
      # Non-profits with Gyms/Showers
      str_detect(comm_class_clean, "non-profit.*w/gym|non-profit.*showers") ~ "non-profit organzation with showers",
      
      # General Non-profits
      str_detect(comm_class_clean, "non-profit|nonprofit") ~ "non-profit organization",
      
      # Parks
      str_detect(comm_class_clean, "park") ~ "park",
      
      # Businesses & General Commercial (includes gov't, resturants, hotels, condos, etc.)
      str_detect(comm_class_clean, "business|gov't|resturant|hotel|establishment|apartment|condo|gym") ~ "business",
      
      TRUE ~ NA_character_
    )
  ) %>% 
  left_join(table_3_Oahu, by = "type")


# filter out still missing 
still_missing_table4s <- all_table4s_combined %>% 
  filter(tmk %!in% usable_table4s_witheffluent$tmk) #2555

write_csv(usable_table4s_witheffluent,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

write_csv(still_missing_table4s,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_stillmissing_allislands.csv"))

```

