---
title: "Gapfill OSDS with No Dwelling information"
author: "Madeline Berger"
date: "2026-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

# Read in necessary data

iws_permit_data <- read_csv(file.path(data_dir, "osds/shared-data/2009_Present_IWSList_Permits.csv"))

iws_oahu_permits <- iws_permit_data %>% 
  mutate(tmk_char = as.character(TMK)) %>% 
  filter(str_starts(tmk_char,"1")) %>% 
  rename(full_taxid = tmk_char) %>% # this is where fulltax id is created
    rowid_to_column("permit_id") %>% 
  mutate(approval_date = mdy(FinalApprovalDate),
         year = year(approval_date)) %>% 
  clean_names()

iws_oahu_permits_tmk <- iws_oahu_permits %>% pull(tmk)

```


# Gapfilling OSDS missing dwell data 

### How many are we missing information for, across islands?

```{r}

# ------------------------ Read in Table 4s for each Island - missing Dwell Data ------------------ #

# read in table 4 from oahu - this one has x y info

oahu_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table4_percommercial_draft.csv"))

coords_fortbl4 <- oahu_table4 %>% 
  select(tmk, x,y) %>% 
  filter(!is.na(x) & !is.na(y)) %>% 
  st_as_sf(coords = c("x","y"), crs = 32604)

coordinates_latlon <- st_transform(coords_fortbl4, crs = 4326) #2633

mapview(coordinates_latlon)

api_ready_coords_oah <- coordinates_latlon %>%
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1],
    google_coords = paste(lat, lon, sep = ",")
  ) %>%
  st_drop_geometry()


# get addresses - if there are coordinates, then use these rather than address because it will be more precise.

addresses_fortbl4 <- iws_oahu_permits %>% 
  filter(tmk %in% oahu_table4$tmk) %>% 
  filter(tmk %!in% api_ready_coords_oah$tmk) %>% 
  select(tmk, starts_with("project")) %>% 
  mutate(street_house = if_else(is.na(project_street_address2), 
                           project_street_address, 
                           paste(project_street_address, "-", project_street_address2)),
         # Create the full address with a conditional check for the zip code
    address_full = if_else(is.na(project_zip_code) | project_zip_code == "NA",
                           paste(street_house, project_city, sep = ", "),
                           paste(street_house, project_city, project_zip_code, sep = ", ")),
    
    # Final step: convert everything to lower case
    address_full = str_to_lower(address_full)
    )


api_ready_address_oah <- addresses_fortbl4 %>%
  select(tmk, address_full)


# --------------------- Kauai Maui ------------------------------ #

# these do NOT have xy data. So weʻll use the TMK centroids below 

mauinui_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table4_percommercial_draft.csv")) %>% 
  filter(commercial == "check google places")

kauai_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table4_percommercial_draft.csv")) %>% filter(commercial == "check google places")

hi_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table4_updated_draft.csv")) %>% 
  filter(data_presence == "Empty (All 0 or NA)")


kauai_maui_table4 <- bind_rows(mauinui_table4, kauai_table4) # 471


# --------------------- Read in Spatialized Table 1  ------------------------- #

## Alternative information to search with: TMK parcel centroids

table1_withspatial <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"))

tmk_for_api_kauaimaui <- table1_withspatial %>% 
  st_transform(crs = 4326) %>% 
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1]
  ) %>%
  #st_drop_geometry() %>%
  select(tmk, lat, lon) %>% 
  filter(tmk %in% kauai_maui_table4$tmk) #459 (we lost 12)

tmk_for_api_oah <- table1_withspatial %>% 
  st_transform(crs = 4326) %>% 
  mutate(
    lat = st_coordinates(.)[,2],
    lon = st_coordinates(.)[,1]
  ) %>%
  #st_drop_geometry() %>%
  select(tmk, lat, lon) %>% 
  filter(tmk %in% oahu_table4$tmk) # 2218 (note we lost about 400, so these did not have dwell or tmk info?)

```

### First pass: Additional GIS Data for Flags

For this, weʻll use the TMKS.

Key info to help estimate effluent: 

"Hawaii’s public school system (which operates as a single statewide district) has an average school size of about 560 students. Statewide, there are roughly 295 public schools educating a total of about 165,000 students"

"teacher student ratio = 14:1"

```{r}

# read in additional layers 
state_landuse <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/State_Land_Use_Districts.geojson"))

public_schools <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/Public_Schools.csv")) %>%
  st_as_sf(coords = c("X","Y"), crs = 32604) %>% 
  mutate(
    enrollment = 560,
    faculty = 40,
    tuition = 0
  ) %>% 
  select(
    school_name = esis_name,
    address,
    city,
    zip,
    district,
    sch_type,
    complex_area,
    enrollment,
    faculty,
    tuition
  )

private_schools <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/large-input-data/Additional_Data_for_ID/Private_Schools.csv")) %>% 
  st_as_sf(coords = c("X","Y"), crs = 32604) %>% 
  mutate(
    sch_type = "private",
    complex_area = "unknown",
    district = "to add"
  ) %>% 
  select(
    school_name = school,
    address,
    city,
    zip = zipcode,
    district,
    sch_type,
    complex_area,
    enrollment = enrollment21,
    faculty = faculty21,
    tuition = tuition21
  )

all_schools <- rbind(public_schools, private_schools)


# Pre-screening: Landuse code

landuse_km <- st_join(tmk_for_api_kauaimaui, state_landuse, join = st_intersects)

landuse_oah <- st_join(tmk_for_api_oah, state_landuse, join = st_intersects)

landuse_km_missing <- landuse_km %>% 
  filter(is.na(ludcode)) # these donʻt have spatial info, 17

landuse_oah_missing <- landuse_oah %>% 
  filter(is.na(ludcode)) # 40 

# which are NA?

mapview(landuse_oah_missing) # in the ocean

# Schools 

mapview(public_schools)+
  mapview(private_schools)+
  mapview(tmk_for_api_kauaimaui, col.regions = "red")


# Manual Inspection Dataframe

# here I actually looked at the map and confirmed ones that are schools, and a few other non-residences. We can use
# this to compare how well the spatial join works, as well 

manual_school_df <- data.frame(
  FID = c(
    427,
    428,
          285,
          35,
          36,
          29,
          17,
          4,
          87,
          54,
          53,
          52),
  School = c(
    "Molokai Middle School",
    "Molokai Middle School",
    "Seabury Hall",
    "Kauai High School",
    "Kauai Community College",
    "Koloa Elem School",
    "Kalaheo Elem School",
    "Kekaha Elem School",
    "Ke Kula Niihau O Kekaha-LPCS",
    "St. Catherine School",
    "St. Catherine School",
    "St. Catherine School"
  )
)

manual_other_df <- data.frame(
  FID = c(
    213,
    186,
    179,
    86,
    10,
    5,
    6
  ),
  Place = c(
    "Post office",
    "Post office",
    "Pot office",
    "Healthy Hut Natural Foods",
    "West Inn Hotel",
    "LDS Church",
    "Church"
  )
)

# join the school info to the osds points. then join with HI GIS school to get enrollment numbers

km_pts_with_fid <- tmk_for_api_kauaimaui %>% 
  mutate(FID = row_number()) %>% 
  left_join(manual_school_df, by = "FID") %>% 
  left_join(manual_other_df, by = "FID")

# check to make sure FID aligns
mapview(km_pts_with_fid)+
  mapview(tmk_for_api_kauaimaui, col.regions = "red")

# create matrix

within_200m_matrix <- st_is_within_distance(km_pts_with_fid, all_schools, dist = 200, sparse = FALSE)
validation_set <- km_pts_with_fid %>%
  mutate(
    buffer_flagged_200m = rowSums(within_200m_matrix) > 0
  )



```



### Second pass: Google Places

Activate Google Places API

```{r}

# See if we have the key

# Load the key from the local file
source(here("config.txt"))

# Set it as an environment variable so packages can find it
Sys.setenv(GOOGLE_PLACES_KEY = GOOGLE_PLACES_KEY)

# Verify (should print your key)
print(Sys.getenv("GOOGLE_PLACES_KEY"))


# Verify it immediately
if(Sys.getenv("GOOGLE_PLACES_KEY") == "") {
  stop("API Key is missing!")
} else {
  print("Success! Key is loaded.")
}

key <- Sys.getenv("GOOGLE_PLACES_KEY")

# Load packages
library(googleway)
library(dplyr)


# --- PART A: SEARCH BY ADDRESS ---
# try one from address dataframe
addr_example <- "66-1009 kaukona hua road, waialua"

res_addr <- google_find_place(
  input = addr_example,
  key = key,
  inputtype = "textquery",
  fields = c("formatted_address", "name", "type")
)

addr_types <- res_addr$candidates$types[[1]]
print(paste("Address is a:", paste(addr_types, collapse = ", ")))

# set up saftey limit

# 1. Initialize the counter at the very top
api_count <- 0
safety_limit <- 200

# 2. Logic for Address Search + Coordinate Extraction
get_detailed_address_info <- function(addr_string) {
  # Safety count
  # Access the counter from the global environment
  api_count <<- api_count + 1 
  
  # THE TOLL BOOTH: Check the limit BEFORE calling Google
  if (api_count > safety_limit) {
    stop(paste("Safety limit of", safety_limit, "reached. Stopping to save your bank account!"))
  }
  
  # We ask for 'geometry' so we get the lat/lon even if the type is boring
  res <- google_find_place(
    input = addr_string, 
    key = key,
    inputtype = "textquery", 
    fields = c("name", "type", "geometry")
  )
  
  if (length(res$candidates) > 0) {
    top_cand <- res$candidates
    main_type <- top_cand$types[[1]][1]
    coords <- top_cand$geometry$location
    
    # Check if we need to dig deeper
    needs_sweep <- main_type %in% boring_types
    
    return(list(
      name = top_cand$name[1], 
      type = main_type, 
      lat = coords$lat[1], 
      lon = coords$lng[1], 
      sweep = needs_sweep
    ))
  }
  return(list(name = NA, type = NA, lat = NA, lon = NA, sweep = FALSE))
}

# 3. The Radius Sweep (Same as before, using the extracted coordinates)
radius_sweep <- function(lat, lon) {
  if (is.na(lat)) return(NA)
  
  # Saftey stop for bank account
  api_count <- api_count + 1
  if (api_count > safety_limit){
    stop("Saftey limit reached during radius sweep")
  }
  
  res <- google_places(location = c(lat, lon), radius = 100, key = key)
  
  if (nrow(res$results) > 0) {
    # Look for the first thing that ISN'T boring
    interesting_finds <- res$results %>%
      unnest(cols = c(types)) %>%
      filter(!types %in% boring_types)
    
    if (nrow(interesting_finds) > 0) {
      return(paste0("Part of: ", interesting_finds$name[1], " (", interesting_finds$types[1], ")"))
    }
  }
  return("No larger landmark found")
}

# 4. Execute on address df

results_addresses <- addresses_fortbl4 %>%
  head(5) %>% # Test batch
  rowwise() %>%
  mutate(
    # Step 1: Initial Address Lookup
    init = list(get_detailed_address_info(address_full)),
    found_name = init$name,
    found_type = init$type,
    temp_lat = init$lat,
    temp_lon = init$lon,
    needs_sweep = init$sweep
  )

results_sweep <- results_addresses %>% 
  mutate(
    # Step 2: If address was 'premise', look around that coordinate
    final_insight = if_else(needs_sweep, 
                            radius_sweep(temp_lat, temp_lon), 
                            found_type)
  ) %>%
  select(-init) # Clean up the list column


# --------------- PART B: Search by Coordinates ----------------------
# Workflow: 1) if it has a specific place, get it. if it doesn't flag for radius search

boring_types <- c("premise", "establishment", "point_of_interest", "political", "floor", "locality")

# 2. Logic for Step 1: Specific Check
get_initial_info_coords <- function(lat, lon) {
  # Safety count
  api_count <<- api_count + 1 
  
  # THE TOLL BOOTH: Check the limit BEFORE calling Google
  if (api_count > safety_limit) {
    stop(paste("Safety limit of", safety_limit, "reached. Stopping to save your bank account!"))
  }
  # Perform the search
  res <- google_places(location = c(lat, lon), radius = 10, key = key)
  
  # Check if 'results' exists and has rows safely
  if (!is.null(res$results) && length(res$results) > 0 && nrow(res$results) > 0) {
    top_place <- res$results[1, ]
    
    # Safely extract the first type
    main_type <- if (!is.null(top_place$types[[1]])) top_place$types[[1]][1] else "unknown"
    
    # Check if we need a radius sweep (boring_types defined earlier)
    needs_sweep <- main_type %in% boring_types
    
    return(list(name = top_place$name, type = main_type, sweep = needs_sweep))
  }
  
  # Fallback if nothing is found at all
  return(list(name = "No result", type = "unknown", sweep = TRUE))
}
# 3. Logic for radius sweep

radius_sweep_coords <- function(lat, lon) {
  # We look up to 100 meters away for a "landmark" type
  res <- google_places(location = c(lat, lon), radius = 100, key = key)
  
  if (nrow(res$results) > 0) {
    # Flatten all types found in the area
    all_nearby <- res$results %>%
      select(name, types) %>%
      tidyr::unnest(types) %>%
      filter(!types %in% boring_types) # Keep only the interesting ones
    
    if (nrow(all_nearby) > 0) {
      # Return the name of the most prominent interesting thing
      return(paste0("Part of: ", all_nearby$name[1], " (", all_nearby$types[1], ")"))
    }
  }
  return("No landmark found")
}

# --- 4. EXECUTION ---
# This assumes your df has 'lat' and 'lon'

safe_sweep <- possibly(radius_sweep_coords, otherwise = "Sweep Failed")

processed_data <- api_ready_coords %>%
  head(5) %>% # Test with 5 first
  rowwise() %>%
  mutate(
    init = list(get_initial_info_coords(lat, lon)),
    place_name = init$name,
    primary_type = init$type,
    needs_sweep = init$sweep
  ) %>%
  mutate(
    final_category = if_else(needs_sweep, 
                             safe_sweep(lat, lon), 
                             primary_type)
  )

```

### Third pass - Manual  


### Standardize and effluent 