---
title: "Compile Final Database and Query"
author: "Madeline Berger"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

```

# Calculate Effluent from Database

This markdown contains code to assemble and clean the final compiled database. It also contains the development code for the effluent query.

Progress note 08-11-2026: Tables should be aligned upstream. Can dive right into query next session.

## Assemble Tables Into Final Database and Final Gapfilling

Up until this point, I've worked on each dataset separately due to small differences in each islands' most recent dataset. 
Here I will combine them so we have one database with 5 tables:

-Table 1: Authoritative TMK List, Spatial Information, Analysis Flags
-Table 2: Bedrooms per year
-Table 3a: Permits applied per year
-Table 3b: Cummulative OSDS status per year
-Table 4: Non-residential OSDS


At this stage, we should also check to ensure there is data for at least one year for each TMK. Some buildings in the dwell dat, for example, did not have bedrooms listed (esp in Maui)
We will gapfill these with the following assumption:
- IF the tmk is in the tax record, with total_rooms, but no bedrooms, then we will assign the number of bedrooms as 50% of the total rooms
- This comes from the following statistics on housing inventory: median number of rooms for HI homes is 4.7, and 55.4 percent of homes have 3 or more bedrooms. Therefore, a typical house has 2 or 3 bedrooms out of 4.7 rooms, so roughly 50% of a building can be considered bedrooms. 

```{r}

#

# ------------------- Table 1 -----------------------------

oahu_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table1_tmk_registry_ALL.csv"))
maui_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/maui_table1_tmk_registry_ALL.csv"))


table_1_complete <- rbind(oahu_1,maui_1) # oahu is missing min year and max year. 


# ------------------ Table 2 ----------------- #


oahu_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table2_bedrooms_ALL.csv"))
maui_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/maui_table2_bedrooms_ALL.csv"))


# Create lists of tmks that are 0s for bedroms

# NOTE - we don't have most recent data for oahu
maui_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2_bedrooms_ALL.csv"))

kauai_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table2_bedrooms_ALL.csv"))

hi_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table2_bedrooms_ALL_draft.csv"))

final_table2_allislands <- rbind(oahu_2,
                                 maui_2,
                                 kauai_2,
                                 hi_2)

write_csv(final_table2_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands.csv"))

# table 3as

oahu_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_all.csv")) %>% 
  rename(tmk = full_taxid)
names(oahu_3a)[-1] <- paste0("osds_class_",names(oahu_3a)[-1])


maui_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

kauai_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

hi_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3_osdscl_ALL_draft.csv")) %>% 
  select(-osds_cl_pre_2009)
names(hi_3a)[-1] <- gsub("osds_cl_", "osds_class_", names(hi_3a)[-1])

# removed years not present in oahu - 2009 and pre2009 
final_table3a_allislands <- rbind(oahu_3a,
                                  maui_3a,
                                  kauai_3a,
                                  hi_3a)
write_csv(final_table3a_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))

# 3b - need to change
oahu_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3b_allyears_draft.csv")) 
names(oahu_3b)[-1] <- gsub("osds_class_", "osds_present_", names(oahu_3b)[-1])
names(oahu_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(oahu_3b)[-1])

maui_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3b_allyears_draft.csv"))
names(maui_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(maui_3b)[-1])
names(maui_3b)[-1] <- gsub("osds_class_", "osds_present_", names(maui_3b)[-1])

kauai_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3b_allyears_draft.csv"))
names(kauai_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(kauai_3b)[-1])
names(kauai_3b)[-1] <- gsub("osds_class_", "osds_present_", names(kauai_3b)[-1])

hi_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3b_allyears_draft.csv"))
names(hi_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(hi_3b)[-1])

final_table3b_allislands <- rbind(oahu_3b,
                                  maui_3b,
                                  kauai_3b,
                                  hi_3b)

write_csv(final_table3b_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))

# table 4s - this one has already been combined 

table_4_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

```

### Check to ensure Table 1 TMKS are accounted for. 

```{r}

# --------- Read in Final Tables --------- #

final_table2_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands.csv"))
final_table3a_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))
final_table3b_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))


table_1_tmks_all <- final_table1_allislands$tmk

# are all of table 2 tmks in table 1?

all(final_table2_allislands$tmk %in% table_1_tmks_all) # TRUE

# do table 2 and table 3b tmks match?

setequal(final_table2_allislands$tmk, final_table3b_allislands$tmk) # FALSE, because some of these are commercial

# Elements in Table 2 but missing from Table 3b - this one should be 0 
setdiff(final_table2_allislands$tmk, final_table3b_allislands$tmk)

# Elements in Table 3b but missing from Table 2 - this one will have table 4 classes
length(setdiff(final_table3b_allislands$tmk, final_table2_allislands$tmk)) #89488. Way too many

# are all of table 3b tmks in table 1?

all(final_table3b_allislands$tmk %in% table_1_tmks_all) # TRUE


# are all table 4 tms in table 1?

final_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

all(final_table4$tmk %in% table_1_tmks_all) # TRUE

# are the remaining unaccounted table 1s all accounted for in our "missing info" group?

# identify tmks in table one that so far have not shown up in table 4 or 3b

accounted_tmks <- unique(c(final_table4$tmk, final_table2_allislands$tmk))
unaccounted_table1_tmks <- setdiff(table_1_tmks_all, accounted_tmks) # 88229

# compare with dfs of missing info - these will either be missing spatial info OR missing variable info 

still_missing_spatial <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withOUTSpatial.csv"))

still_missing_identifiers 


```

From this, I can see that we need to:
1) review table 2
2) review table 3a
3) confirm and collect all missing data

### 