---
title: "Compile Final Database and Query"
author: "Madeline Berger"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

```

# Final Databse and Query

This markdown contains code to assemble and clean the final compiled database. It also contains the development code for the effluent query.

## Assemble Tables Into Final Dataabse

Up until this point, I've worked on each dataset separately due to small differences in each islands' most recent dataset. 
Here I will combine them so we have one database with 5 tables:

-Table 1: Authoritative TMK List, Spatial Information, Analysis Flags
-Table 2: Bedrooms per year
-Table 3a: Permits applied per year
-Table 3b: Cummulative OSDS status per year
-Table 4: Non-residential OSDS

```{r}


# table1s

oahu_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table1_tmk_registry_all.csv"))
maui_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table1_tmk_registry_all.csv"))
kauai_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table1_tmk_registry_all.csv"))

hi_1 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table1_tmk_registry_ALL_draft.csv")) %>% select(-class_match_flag, -pitt_assignment,-min_year,-max_year,-scenario) %>% 
  rename(permit_story = unique_story,bed_gapfill_flag = bedroom_data_source) %>% 
  mutate(year_built_combined = 9999)

# missing key column names for Hi

final_table1_kmo <- rbind(oahu_1,
                                 maui_1) %>% 
  rbind(kauai_1) %>% 
  select(-starts_with("use"), -n_buildings)
  
final_table1_allislands <-  rbind(hi_1,final_table1_kmo)

# table 2
oahu_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table2_bedrooms_ALL.csv")) %>% 
  mutate(`2021` = NA,
         `2022` = NA,
         `2023` = NA)

# NOTE - we don't have most recent data for oahu
maui_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table2_bedrooms_ALL.csv"))

kauai_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table2_bedrooms_ALL.csv"))

hi_2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table2_bedrooms_ALL_draft.csv"))

final_table2_allislands <- rbind(oahu_2,
                                 maui_2,
                                 kauai_2,
                                 hi_2)

write_csv(final_table2_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands.csv"))

# table 3as

oahu_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3_class_all.csv")) %>% 
  rename(tmk = full_taxid)
names(oahu_3a)[-1] <- paste0("osds_class_",names(oahu_3a)[-1])


maui_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

kauai_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3a_permitsApplied_ALL.csv")) %>% 
  select(-starts_with("new"))

hi_3a <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3_osdscl_ALL_draft.csv")) %>% 
  select(-osds_cl_pre_2009)
names(hi_3a)[-1] <- gsub("osds_cl_", "osds_class_", names(hi_3a)[-1])

# removed years not present in oahu - 2009 and pre2009 
final_table3a_allislands <- rbind(oahu_3a,
                                  maui_3a,
                                  kauai_3a,
                                  hi_3a)
write_csv(final_table3a_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))

# 3b - need to change
oahu_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/oahu_tables/table3b_allyears_draft.csv")) 
names(oahu_3b)[-1] <- gsub("osds_class_", "osds_present_", names(oahu_3b)[-1])
names(oahu_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(oahu_3b)[-1])

maui_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/maui_tables/table3b_allyears_draft.csv"))
names(maui_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(maui_3b)[-1])
names(maui_3b)[-1] <- gsub("osds_class_", "osds_present_", names(maui_3b)[-1])

kauai_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/kauai_tables/table3b_allyears_draft.csv"))
names(kauai_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(kauai_3b)[-1])
names(kauai_3b)[-1] <- gsub("osds_class_", "osds_present_", names(kauai_3b)[-1])

hi_3b <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/int/hi_tables/table3b_allyears_draft.csv"))
names(hi_3b)[-1] <- gsub("osds_cl_", "osds_present_", names(hi_3b)[-1])

final_table3b_allislands <- rbind(oahu_3b,
                                  maui_3b,
                                  kauai_3b,
                                  hi_3b)

write_csv(final_table3b_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))

# table 4s - this one has already been combined 

table_4_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

```

### Check to ensure Table 1 TMKS are accounted for. 

```{r}

# --------- Read in Final Tables --------- #

final_table2_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands.csv"))
final_table3a_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))
final_table3b_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))


table_1_tmks_all <- final_table1_allislands$tmk

# are all of table 2 tmks in table 1?

all(final_table2_allislands$tmk %in% table_1_tmks_all) # TRUE

# do table 2 and table 3b tmks match?

setequal(final_table2_allislands$tmk, final_table3b_allislands$tmk) # FALSE, because some of these are commercial

# Elements in Table 2 but missing from Table 3b - this one should be 0 
setdiff(final_table2_allislands$tmk, final_table3b_allislands$tmk)

# Elements in Table 3b but missing from Table 2 - this one will have table 4 classes
length(setdiff(final_table3b_allislands$tmk, final_table2_allislands$tmk)) #89488. Way too many

# are all of table 3b tmks in table 1?

all(final_table3b_allislands$tmk %in% table_1_tmks_all) # TRUE


# are all table 4 tms in table 1?

final_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

all(final_table4$tmk %in% table_1_tmks_all) # TRUE

# are the remaining unaccounted table 1s all accounted for in our "missing info" group?

# identify tmks in table one that so far have not shown up in table 4 or 3b

accounted_tmks <- unique(c(final_table4$tmk, final_table2_allislands$tmk))
unaccounted_table1_tmks <- setdiff(table_1_tmks_all, accounted_tmks) # 88229

# compare with dfs of missing info - these will either be missing spatial info OR missing variable info 

still_missing_spatial <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withOUTSpatial.csv"))

still_missing_identifiers 


```

From this, I can see that we need to:
1) review table 2
2) review table 3a
3) confirm and collect all missing data

### 