---
title: "Compile Final Database and Query"
author: "Madeline Berger"
date: "2026-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

```

# Calculate Effluent from Database

This markdown contains code to assemble and clean the final compiled database. It also contains the development code for the effluent query.

Progress note 08-11-2026: Tables should be aligned upstream. Can dive right into query next session.

## Assemble Tables Into Final Database and Assess Final Gapfilling

Up until this point, I've worked on each dataset separately due to small differences in each islands' most recent dataset. 
Here I will combine them so we have one database with 5 tables:

-Table 1: Authoritative TMK List, Spatial Information, Analysis Flags
-Table 2: Bedrooms per year
-Table 3a: Permits applied per year
-Table 3b: Cummulative OSDS status per year
-Table 4: Non-residential OSDS


At this stage, we should also check to ensure there is data for at least one year for each TMK. Some buildings in the dwell dat, for example, did not have bedrooms listed (esp in Maui)


We will gapfill these with the following assumption:
- IF the tmk is in the tax record, with total_rooms, but no bedrooms, then we will assign the number of bedrooms as 50% of the total rooms
- This comes from the following statistics on housing inventory: median number of rooms for HI homes is 4.7, and 55.4 percent of homes have 3 or more bedrooms. Therefore, a typical house has 2 or 3 bedrooms out of 4.7 rooms, so roughly 50% of a building can be considered bedrooms. 
- note this should happen upstream, but for now we'll do it here. Add this as a clean up step for scripts

```{r}

# File path for intermediate folder

int_dir <- file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/")

# ------------------- Table 1 -----------------------------

oahu_1 <- read_csv(file.path(int_dir,"step4/oahu_table1_tmk_registry_ALL.csv"))
maui_1 <- read_csv(file.path(int_dir,"step4/maui_table1_tmk_registry_ALL.csv"))
kauai_1 <- read_csv(file.path(int_dir,"step4/kauai_table1_tmk_registry_ALL.csv")) |> 
  select(-class_match_flag)# remove class match flag upsteam
hi_1 <- read_csv(file.path(int_dir,"step4/hi_table1_tmk_registry_ALL.csv")) |> 
  select(-class_match_flag)# remove class match flag upstream

table_1_complete <- rbind(oahu_1,maui_1, kauai_1, hi_1) # 

# check these - not sure whats going on
table1_napermitstory <- table_1_complete |> 
  filter(is.na(permit_story))

write_csv(table_1_complete ,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_allislands_allosds.csv"))

# double check in old scripts that permit story is added for new and non updated ones. 

# ------------------ Table 2 ----------------- #


oahu_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/oahu_table2_bedrooms_ALL.csv"))# oahu missing 4 years (2009, 2021, 2022, 2023)
maui_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/maui_table2_bedrooms_ALL.csv")) |>
  select(-ends_with(c("2008","2009","2021","2022", "2023")))
kauai_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/kauai_table2_bedrooms_ALL.csv")) |> 
  select(-ends_with(c("2009","2021","2022", "2023","2024", "2025"))) # goes up to 2025
hi_table2 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/intermediate/step4/hi_table2_bedrooms_ALL.csv")) |> 
select(-ends_with(c("2021","2022", "2023")), -gapfill_flag)# upstream - remove gapfill flag 


table2_all_islands_2010_2020 <- rbind(oahu_table2,
                                      maui_table2,
                                      kauai_table2,
 
                                                                           hi_table2) # this indicated that we should have about 4922 in table 4 

# ------------------ Gapfill Table 2  ----------------- #

empty_tmks <- table2_all_islands_2010_2020 %>%
  filter(if_all(-tmk, ~ is.na(.) | . == 0)) %>%
  pull(tmk)

table_2_needsgapfill <- table2_all_islands_2010_2020 |> 
  filter(tmk %in% empty_tmks)

# read in clean dwell data

oahu_dwell <- read_csv(file.path(int_dir,"step1-clean-dwelling-records/oahu_dwell_clean_allyears.csv")) |> 
  select(tmk, total_rooms, year_built)
maui_dwell <- read_csv(file.path(int_dir,"step1-clean-dwelling-records/maui_dwell_clean_alltmks.csv")) |> 
  select(tmk, total_rooms, year_built)
kauai_dwell <- read_csv(file.path(int_dir,"step1-clean-dwelling-records/kauai_dwell_clean_alltmks.csv")) |> 
  select(tmk, total_rooms = tot_rooms, year_built = yrblt)
hi_dwell <- read_csv(file.path(int_dir,"step1-clean-dwelling-records/hi_dwell_osds_tmks.csv")) |> 
  select(tmk, total_rooms = rmtot, year_built = yrblt)

lookup_allislands <- rbind(oahu_dwell,
                           maui_dwell,
                           kauai_dwell,
                           hi_dwell) |>
filter(tmk %in% empty_tmks) %>%
  filter(!is.na(total_rooms) & !is.na(year_built)) %>%
  # Sort by TMK, and put the oldest year_built first
  arrange(tmk, year_built) %>%
  # Keep only the first record for each TMK (which will be the oldest year_built)
  distinct(tmk, .keep_all = TRUE)


# next we'll build a loop that will look for total room and year built information in the dwell data. 

year_cols <- setdiff(names(table2_all_islands_2010_2020), "tmk")

# gapfill using vectorized operations
table2_gapfilled <- table_2_needsgapfill |> 

  left_join(lookup_allislands, by = "tmk") |> 
  
  # Step A: fill ALL year columns with total_rooms
  mutate(total_rooms = as.numeric(total_rooms)) |> 
  mutate(across(all_of(year_cols), ~ ifelse(total_rooms == 1, 1,round(total_rooms / 2)))) |> 
  
  # Step B: pivot to long format to easily zero-out years prior to year_built
  pivot_longer(
    cols = starts_with("bedrooms"),
    names_to = "year_col",
    values_to = "rooms"
  ) |> 
  mutate(
    year_num = as.numeric(str_extract(year_col, "\\d{4}")),
    # if the column year is older than year_built, set value to NA
    rooms = case_when(
      !is.na(year_built) & year_num < year_built ~ NA_real_,
      !is.na(year_built) & year_num >= year_built ~ rooms,
      is.na(year_built) ~ rooms # Keeps total_rooms if year_built is missing
    )
  ) |> 
  
  # Step C: Pivot back to wide format matching original table structure
  pivot_wider(
    id_cols = tmk,
    names_from = year_col,
    values_from = rooms
  )
  
  # Step D: Drop temporary lookup columns (optional)
  #select(-total_rooms, -year_built, -year_num)


still_empty_tmks <- table2_gapfilled |> 
  filter(if_all(-tmk, ~ is.na(.) | . == 0)) %>%
  pull(tmk) # yikes


# Join gapfilled data back to other data, minus the emptys. Save the emptys

table_2_withgapfill <- table2_all_islands_2010_2020 |> 
  filter(tmk %!in% empty_tmks) |> 
  rbind(table2_gapfilled) |> 
  filter(tmk %!in% still_empty_tmks) # still missing a lot


table2_incomplete <- table2_all_islands_2010_2020 |> 
  filter(tmk %in% still_empty_tmks)


write_csv(table_2_withgapfill,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands_complete_entries_2010_2020.csv"))

write_csv(table2_incomplete, file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands_INcomplete_entries_2010_2020.csv"))

# ------------------ Table 3as ----------------- #

oahu_3a <- read_csv(file.path(int_dir,"step4/oahu_table3a_permitsApplied_ALL.csv")) |> 
  rename(osds_class_pre2009 = newpre_2009) # fix name upstream
maui_3a <- read_csv(file.path(int_dir,"step4/maui_table3a_permitsApplied_ALL.csv"))
kauai_3a <- read_csv(file.path(int_dir,"step4/kauai_table3a_permitsApplied_ALL.csv"))
hi_3a <- read_csv(file.path(int_dir,"step4/hi_table3a_permitsApplied_ALL.csv"))

final_table3a_allislands <- rbind(oahu_3a,
                                  maui_3a,
                                  kauai_3a,
                                  hi_3a)

final_table3a_allislands_until2020 <- final_table3b_allislands |> 
  select(-ends_with(c("2008","2021","2022", "2023")))

write_csv(final_table3a_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands.csv"))
write_csv(final_table3a_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands_until2020.csv"))

# ------------------ Table 3bs ----------------- #

oahu_3b <- read_csv(file.path(int_dir,"step4/oahu_table3b_activeOSDS.csv")) |> 
  rename(osds_class_2008 = osds_cl_2008) # fix name upstream
maui_3b <- read_csv(file.path(int_dir,"step4/maui_table3b_activeOSDS.csv"))
kauai_3b <- read_csv(file.path(int_dir,"step4/kauai_table3b_activeOSDS.csv"))
hi_3b <- read_csv(file.path(int_dir,"step4/hi_table3b_activeOSDS.csv"))

final_table3b_allislands <- rbind(oahu_3b,
                                  maui_3b,
                                  kauai_3b,
                                  hi_3b)

final_table3b_allislands_until2020 <- final_table3b_allislands |> 
  select(-ends_with(c("2008","2021","2022", "2023")))

write_csv(final_table3b_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands.csv"))
write_csv(final_table3b_allislands,file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands_until2020.csv"))

# ------------------ Table 4 ----------------- #

table_4_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))


# quick check - are the missing table 2 entries in here?

check_still_missing <- table_4_allislands |> 
  filter(tmk %in% still_empty_tmks) # only one. 


still_empty_tmks <- still_empty_tmks[still_empty_tmks != 239002108]

table2_incomplete <- table2_all_islands_2010_2020 |> 
  filter(tmk %in% still_empty_tmks) |> 
  distinct()

write_csv(table2_incomplete, file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands_INcomplete_entries_2010_2020.csv"))
```

### Final check to ensure Table 1 TMKS are accounted for - this will go into the paper on which had not enough information!


```{r}

# --------- Read in Final Tables --------- #

final_table1_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_allislands_allosds.csv"))

final_table2_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands_complete_entries_2010_2020.csv"))

final_table3b_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands_until2020.csv"))

final_table4_allislands <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

# are all of table 2 tmks in table 1?

all(final_table2_allislands$tmk %in% final_table1_allislands$tmk) # TRUE

# are all of table 2 tmks in table 3b?

setequal(final_table2_allislands$tmk, final_table3b_allislands$tmk) # FALSE, because some of these are commercial


# are all of table 3b tmks in table 1?

all(final_table3b_allislands$tmk %in% final_table1_allislands$tmk) # TRUE


# are all table 4 tmk in table 1?

final_table4 <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))

all(final_table4$tmk %in% final_table1_allislands$tmk) # TRUE

# are the remaining unaccounted table 1s all accounted for in our "missing info" group?

# identify tmks in table one that so far have not shown up in table 4 or 3b

accounted_tmks <- unique(c(final_table4$tmk, final_table2_allislands$tmk))
unaccounted_table1_tmks <- setdiff(final_table1_allislands$tmk, accounted_tmks) # 4332 - not bad

# compare with dfs of missing info - these will either be missing spatial info OR missing variable info 
```

These numbers are looking a lot better - the last step will be to look at the 4000 we can't account for. 

## Query 2010 - 2020

Write a function that calculates the effluent for a given row based on the following question on the clean tables. 
Output should be table 1 with the average for each year you want. 

Effluent is calculated using the following equation: 


Effluent (g/day)  = 200 * Number of bedrooms in residence

N flux (kg / day) = Effluent ga/ per day * 3.78541 L/gal x [Ni] mg / L * 10-6

Where Ni represents the estimated N pollution from each class 1 through 4 of OSDS, and number of bedrooms in a residence serves as a proxy for intensity of use 

N loading by class assignment - note this is averaging the ranges in Whittier and El Kadi

| Class | N Loading (mg/L) Range | Average |
|:--------------|:--------------|:--------------|
| Class 1| 1 | 1 |
| Class 2 | 39 – 82 | 60.5 |
| Class 3 | 7 - 60 | 33.5 |
| Class 4 | 15-90 | 37.5 |



```{r query-effluent}

# read in and pivot table  

osds_main <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_allislands_allosds.csv")) |> 
  select(tmk)

bedrooms_wide <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table2_allislands_complete_entries_2010_2020.csv"))

class_wide <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands_until2020.csv"))


commercial_long <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table4_identified_allislands.csv"))


# class look up - include encoded! Key assumption in this table. Think about how it would work to split bedroom number.

ni_lookup <- data.frame(
  class = c(0,1,2,4,8),
  ni_mg_l = c(0,1,60.5,33.5,37.5)
)

# ------------ expand ni look up to include average classess -------- # 

library(tidyverse)

# 1. Define your base classes and their specific [Ni] concentrations (mg/L)
# Note: Bit flags MUST be powers of 2 (1, 2, 4, 8, 16, etc.)
base_classes <- tribble(
  ~bit_flag, ~class_name, ~ni_mg_l,
  0,         "No class",  0,
  1,         "Class_1",   1.0,
  2,         "Class_2",   60.5,
  4,         "Class_3",   33.5,
  8,         "Class_4",   37.5
)

# 2. Function to decode a bitwise sum into its constituent bit flags
decode_bitwise <- function(sum_val, valid_flags) {
  # Bitwise AND logic: checks which flags are set in sum_val
  valid_flags[bitwAnd(sum_val, valid_flags) > 0]
}

# 3. Generate all possible bitwise combinations dynamically
max_combination <- sum(base_classes$bit_flag)

ni_bitwise_lookup <- tibble(encoded_class = 1:max_combination) %>%
  # Extract active bit flags for each encoded sum
  mutate(
    active_flags = map(encoded_class, ~ decode_bitwise(.x, base_classes$bit_flag))
  ) %>%
  # Filter out invalid combinations if sum doesn't match any flag combination
  filter(map_int(active_flags, length) > 0) %>%
  
  # Calculate average [Ni] and list included classes
  mutate(
    # Average [Ni] of all active classes in the sum
    ni_mg_l = map_dbl(active_flags, function(flags) {
      base_classes %>%
        filter(bit_flag %in% flags) %>%
        summarize(mean_ni = mean(ni_mg_l)) %>%
        pull(mean_ni)
    }),
    
    # Human-readable string of included classes
    classes_included = map_chr(active_flags, function(flags) {
      base_classes %>%
        filter(bit_flag %in% flags) %>%
        pull(class_name) %>%
        paste(collapse = " + ")
    }),
    
    # Count of active classes in the sum
    num_classes = map_int(active_flags, length)
  ) %>%
  # Clean up table output
  select(encoded_class, classes_included, num_classes, ni_mg_l) |> 
  # add 0
  rbind(data.frame(
    encoded_class = 0,
    classes_included = "No classes",
    num_classes = 0,
    ni_mg_l = 0
  ))

# Inspect the result
print(ni_bitwise_lookup)

# ------------------- # ---------------------------- #

# pivot longer) 
bedrooms_long <- bedrooms_wide |> 
  pivot_longer(cols = -tmk,
               names_to = "year",
               values_to = "bedrooms") |> 
  mutate(year_num = as.numeric(substr(year,10,13))) |> 
  mutate(bedrooms = ifelse(is.na(bedrooms),0,bedrooms)) # crucial so that years with no record are 0 effluent in stead of NA


class_long <- class_wide |> 
  pivot_longer(cols = -tmk,
               names_to = "year",
               values_to = "encoded_class") |> 
  mutate(year_num = as.numeric(substr(year,12,15)))


#target_years <- 2010:2014  # use c(2015) for a single year

annual_effluent_nflux_res <- bedrooms_long %>%
  # Join class by both ID and Year
  inner_join(class_long, by = c("tmk", "year_num")) %>%
  
  # Join main OSDS table for static metadata (e.g., island, location)
  inner_join(osds_main, by = "tmk") %>%
  
  # Join Nitrogen concentration lookup by class
  left_join(ni_bitwise_lookup, by = "encoded_class") %>%
  
  # Apply Equations
  mutate(
    # Effluent (gal/day) = 200 * bedrooms
    effluent_gpd  = 200 * bedrooms,
    
    # N flux (kg/day) = Effluent * 3.78541 L/gal * [Ni] mg/L * 10^-6 kg/mg
    n_flux_kg_day = effluent_gpd * 3.78541 * ni_mg_l * 1e-6,
    n_flux_kg_year = n_flux_kg_day * 365
  )

# run for commercial. not sure this i s interesting because they don't change year over year. 

# idea: create a random matrix of year and min max combiations, which we can sample for different runs

year_seq = seq(from = 2010, to = 2020, by = 1)

tmk_year_df <- data.frame(
  tmk = rep(unique(commercial_long$tmk),11),
  year_num = rep(year_seq,2358)
) 

commercial_long <- distinct(commercial_long, tmk, .keep_all = T)


annual_effluent_nflux_comm <- commercial_long %>%
  # Add year - this will be arbitraty but we need to preserve the year in class
  full_join(tmk_year_df, by = "tmk") |> 
  # join class by both ID and Year
  inner_join(class_long, by = c("tmk","year_num")) |>  # these are changing year over year
  
  # Join main OSDS table for static metadata (e.g., island, location)
  #inner_join(osds_main, by = "tmk") %>%
  
  # Join Nitrogen concentration lookup by class
  left_join(ni_bitwise_lookup, by = "encoded_class") %>%
  
  # Apply Equations
  mutate(
    # Effluent (gal/day) = 200 * bedrooms
    effluent_gpd  = median_effluent_gpd,
    
    # N flux (kg/day) = Effluent * 3.78541 L/gal * [Ni] mg/L * 10^-6 kg/mg
    n_flux_kg_day = effluent_gpd * 3.78541 * ni_mg_l * 1e-6,
    
    n_flux_kg_year = n_flux_kg_day * 365
  ) 
  

# ------------ Validate,  Clean up and Save Huge Tables --------------- #

# get list of tmks from table 1 that should show changes over time (ie updated)

tmks_updates <- table_1_complete |> 
  filter(upgrade_scenario == "multiple osds with cesspool upgraded") |> 
  pull(tmk)

residential_effluent_everyear_clean <- annual_effluent_nflux_res |> 
  select(tmk,year_num, n_flux_kg_day,n_flux_kg_year) |> 
  mutate(year_num = as.numeric(year_num)) |> 
  arrange(tmk, year_num)

res_change_check <- residential_effluent_everyear_clean |> 
  filter(tmk %in% tmks_updates)

commercial_effluent_everyear_clean <- annual_effluent_nflux_comm |> 
  select(tmk,year_num, n_flux_kg_day,n_flux_kg_year) |> 
  mutate(year_num = as.numeric(year_num)) |> 
  arrange(tmk, year_num)

# save - this wont run without wifi but thats ok.

write_csv(residential_effluent_everyear_clean,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/residential_kg_pyear.csv"))

write_csv(commercial_effluent_everyear_clean,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/non_residential_kg_pyear.csv"))


```

Versioning Notes:
v0 will be using what I found in the tables but clearly needs to be updated.
What is the most up to date version>

## Join Calculated Effluent to Spatial Points - Save

Use table one with spatial information here! This step should probably be moved to the effluent calculation model. 

```{r}

# calculate effluent for 2010 - 2014 per point and 2015 - 2020 per point

alltypes_effluent_peryear <- rbind(residential_effluent_everyear_clean,
                                   commercial_effluent_everyear_clean)

effluent_multi_year_averages <- alltypes_effluent_peryear |> 
  group_by(tmk) |> 
  summarize(
    n_avg_2010_2014 = mean(n_flux_kg_year[year_num >= 2010 & year_num <= 2014]),
    n_avg_2015_2020 = mean(n_flux_kg_year[year_num >= 2015 & year_num <= 2020])
  ) |>
  mutate(tmk = as.character(tmk))

# cool!!
change_check <- effluent_multi_year_averages |> 
  filter(n_avg_2010_2014 != n_avg_2015_2020) 

# 0.1461011, or 14.6 percent have shown some change in effluent

# read in spatial points

table_1_spatial <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_tmk_registry_withSpatial.shp"))

length(unique(table_1_spatial$tmk))

# Join res effluent

table_1_spatial_effleunt <- table_1_spatial |> 
  left_join(effluent_multi_year_averages, by = "tmk") |> 
  filter(!is.na(n_avg_2010_2014) & !is.na(n_avg_2015_2020)) 

# 95.7 % remain. SICK, N = 4302 dropped out

st_write(table_1_spatial_effleunt,
          file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/alleffluent_5yraverages_spatial.gpkg"))

```

