---
title: "6_DatabasePaper_ResultsFigs"
author: "Madeline Berger"
date: "2026-08-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(janitor)
source(here("workflow/setup.R"))

# color schemes for manuscript 

# option 1 1 - 4
county_pal <- c(
  "Oahu"     = "#395B50", # Granite green
  "Maui Nui" = "#C3ACCE", # Lilac
  "Kauai"    = "#5A7684", # Blue slate
  "Hawaii"   = "#8EAA80"  # Sage
  #"Hawaii"   = "#CFEE9E"# Alternate, Lime Cream
)

# osds class pal: 1 - 4

class_pal <- c(
  "Class 1" = "#F2E3C6", # Cream / Low
  "Class 2" = "#E69F00", # Warm Amber
  "Class 3" = "#D55E00", # Vermilion / Medium-High
  "Class 4" = "#730000"  # Deep Crimson / High
)

class_pal_earthy <- c( "Class 1" = "#1C6E8C", #-cerulean blue:
              "Class 2" = "#89bd9eff",  #--muted-teal: 
              "Class 3" = "#f0c987ff", #--apricot-cream: 
               "Class 4" = "#a53f2bff") #--reddish-brown: 


island_pal <- c(
  "Oahu"     = "#0072B2", # Ocean Blue
  "Maui" = "#009E73", # Jade / Seafoam
  "Kauai"    = "#E6B800", # Sun Yellow
  "Hawaii"   = "#CC79A7", # Orchid Pink / Lava
  "Molokai"  = "#332288"  # Deep Indigo
)

```

# Summary

This script houses code for tallying results and creating figures for the OSDS Database publication


```{r}

# read in final databases 

residential_effluent_everyear_clean <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/residential_kg_pyear.csv"))

commercial_effluent_everyear_clean <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/non_residential_kg_pyear.csv"))

# full table 1

table_1_spatial_effleunt <- st_read(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/calculated_effluent/v0/alleffluent_5yraverages_spatial.gpkg"))

table1_nonspatial <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table1_allislands_allosds.csv"))


```

## Result One: Total Number of OSDS by Location and Island

```{r result-1}


# total total 

total_tmks <- length(unique(table1_nonspatial$tmk))

# totals with effluent
total_tmks_accounted <- length(unique(commercial_effluent_everyear_clean$tmk)) + length(unique(residential_effluent_everyear_clean$tmk))

# unaccounted for due to missing effluent variable

unaccounted <- setdiff(c(unique(commercial_effluent_everyear_clean$tmk),unique(residential_effluent_everyear_clean$tmk)),unique(residential_effluent_everyear_clean$tmk))


# use other lists to assign to residential, non residential or unknown

table1_assigned <- table1_nonspatial |> 
  mutate(
    county_code = substr(tmk, 1, 1),
    category = case_when(
      tmk %in% residential_effluent_everyear_clean$tmk ~ "res",
      tmk %in% commercial_effluent_everyear_clean$tmk  ~ "nonres",
      TRUE                                              ~ "unknown"
    )
  )


# by county

table1_nonspatial_county <- table1_assigned |> 
  mutate(county_code = substr(tmk,1,1)) |> 
  group_by(county_code,category) |> 
  summarize(
    tmk_n = length(unique(tmk)),
    permit_n = sum(n_permits, na.rm = T),
    .groups = "drop"
  ) |> 
  pivot_wider(id_cols = county_code, names_from = category, values_from = tmk_n, values_fill = 0) |> 
  # row totals (totals across categories for each county)
  mutate(
    tmk_n_total = rowSums(across(-starts_with("county")))
  ) |> 
  # bottom row for column totals (totals across all counties)
  adorn_totals("row", fill = "-", na.rm = TRUE)



```


## Result 2: Temporal Shifts in OSDS Over the Past Decade 

```{r results-2}

library(purrr)

# read in table 3 a and 3 b

table3a_wide <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3a_allislands_until2020.csv"))

table3b_wide <- read_csv(file.path(driver_dir,"4_OSDS_N_Runoff/output-data/final_tables/table3b_allislands_until2020.csv"))

# pivot longer for line plot and re explode bits 

class_bits <- c("Class 1" = 1, "Class 2" = 2, "Class 3" = 4, "Class 4" = 8)

county_codes <- c("1" = "Oahu", "2" = "Maui Nui", "3" = "Hawaii", "4" = "Kauai")

table_3a_long <- table3a_wide |>
  mutate(county = county_codes[substr(tmk, 1, 1)]) |>
  pivot_longer(cols = starts_with("osds"), names_to = "year_raw", values_to = "raw_val") |>
  mutate(year_num = parse_number(year_raw)) |>
  filter(!is.na(raw_val)) |>
  mutate(
    osds_class = map(raw_val, \(val) {
      names(class_bits)[bitwAnd(val, class_bits) > 0]
    })
  ) |>
  unnest(osds_class)


plot_data <- table_3a_long |>
  filter(year_num %!in% c(2009,2010,2021,2022,2023)) |>
  group_by(year_num, county, osds_class) |>
  summarize(count = n(), .groups = "drop") |>
  # fill in year x county x class combos with no permits so bars drop to zero
  # instead of getting skipped
  complete(year_num, county, osds_class, fill = list(count = 0))

plot_a <- ggplot(plot_data, aes(x = year_num, y = count, fill = factor(osds_class), group = factor(osds_class))) +
  geom_col(position = "stack", color = "white", linewidth = 0.3) +
  scale_fill_manual(values = class_pal) +
  scale_x_continuous(breaks = unique(plot_data$year_num), expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ county, scales = "free_y") +
  labs(
    title = "OSDS Permits Applied for Over Time",
    x = "Year",
    y = "Number of OSDS Permits",
    fill = "OSDS Class"
  ) +
  theme_minimal()

plot_a


# repeat for table 3b
table_3b_long <- table3b_wide |> 
  pivot_longer(cols = starts_with("osds"), names_to = "year", values_to = "osds_class")


table_3b_long <- table3b_wide |>
  mutate(county = county_codes[substr(tmk, 1, 1)]) |>
  pivot_longer(cols = starts_with("osds"), names_to = "year_raw", values_to = "raw_val") |>
  mutate(year_num = parse_number(year_raw)) |>
  filter(!is.na(raw_val)) |>
  mutate(
    osds_class = map(raw_val, \(val) {
      names(class_bits)[bitwAnd(val, class_bits) > 0]
    })
  ) |>
  unnest(osds_class)


plot_data_b <- table_3b_long |>
  filter(year_num %!in% c(2008,2009,2021,2022,2023)) |>
  group_by(year_num, county, osds_class) |>
  summarize(count = n(), .groups = "drop")

plot_b <- ggplot(plot_data_b, aes(x = year_num, y = count, fill = factor(osds_class), group = factor(osds_class))) +
  geom_area(alpha = 0.85, color = "white", linewidth = 0.3) +
  scale_fill_manual(values = class_pal) +
  scale_x_continuous(breaks = unique(plot_data_b$year_num), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ county, scales = "free_y") +
  labs(
    title = "Active OSDS Over Time",
    x = "Year",
    y = "Number of Active OSDSS",
    fill = "OSDS Class"
  ) +
  theme_minimal()

plot_b

# combine with patchwork, sharing one class legend

library(patchwork)

combined_plot <- (plot_a / plot_b) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 6),
    strip.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.35, "cm")
  )

combined_plot

ggsave(
  filename = "osds_permits_active_by_county_Bar.png",
  plot = combined_plot,
  path = here("OSDS-script/figs"),
  width = 6.5,
  height = 8,
  dpi = 300
)

# same two plots, but colored by county and faceted by class instead

plot_a_bycounty <- ggplot(plot_data, aes(x = year_num, y = count, fill = county, group = county)) +
  geom_col(position = "stack", color = "white", linewidth = 0.3) +
  scale_fill_manual(values = county_pal) +
  scale_x_continuous(breaks = unique(plot_data$year_num), expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ osds_class, scales = "free_y") +
  labs(
    title = "OSDS Permits Applied for Over Time by County",
    x = "Year",
    y = "Number of OSDS Permits",
    fill = "County"
  ) +
  theme_minimal()

plot_a_bycounty

plot_b_bycounty <- ggplot(plot_data_b, aes(x = year_num, y = count, fill = county, group = county)) +
  geom_area(alpha = 0.85, color = "white", linewidth = 0.3) +
  scale_fill_manual(values = county_pal) +
  scale_x_continuous(breaks = unique(plot_data_b$year_num), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ osds_class, scales = "free_y") +
  labs(
    title = "Active OSDS Over Time by County",
    x = "Year",
    y = "Number of Active OSDS",
    fill = "County"
  ) +
  theme_minimal()

plot_b_bycounty

combined_plot_bycounty <- (plot_a_bycounty / plot_b_bycounty) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 10),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 6),
    strip.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.35, "cm")
  )

combined_plot_bycounty

ggsave(
  plot = combined_plot_bycounty,
  filename = "osds_permits_active_by_class_Bar.png",
  path = here("OSDS-script/figs"),
  width = 6.5,
  height = 8,
  dpi = 300
)


## bar graph for effluent by county by year




```

### Active OSDS by county, 2010 vs 2020 (Class 1-3 vs Class 4)

```{r results-2-summary}

# lump classes 1-3 together, keep class 4 on its own

class_group_lookup <- c(
  "Class 1" = "Class 1-3",
  "Class 2" = "Class 1-3",
  "Class 3" = "Class 1-3",
  "Class 4" = "Class 4"
)

active_osds_summary <- table_3b_long |>
  filter(year_num %in% c(2010, 2020)) |>
  mutate(class_group = class_group_lookup[osds_class]) |>
  group_by(county, year_num, class_group) |>
  summarize(count = n(), .groups = "drop") |>
  arrange(county, year_num, class_group)

active_osds_summary

# wide format: one row per county, columns for each year x class group

active_osds_summary_wide <- active_osds_summary |>
  pivot_wider(names_from = c(year_num, class_group), values_from = count, values_fill = 0) |> 
  mutate(change_1_3 = `2010_Class 1-3` - `2020_Class 1-3`,
         change_4 = `2010_Class 4` - `2020_Class 4`)

active_osds_summary_wide

```



## Result 3: Spatial Shifts in OSDS over the past Decade by Moku

Here we will designate moku by increase or decrease over the 10 year span. 

Step 1: Add moku to spatialized effluent calcuations
- sum effluent by moku
- sum osds by moku?
